### 2022-12-17 submission.tex
first commit

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage[margin=2cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usepackage[export]{adjustbox}
\usepackage{indentfirst}
\usepackage{wrapfig}
\usepackage{listings}
\usepackage{color}
\usepackage{enumerate}
\usepackage{amssymb, bm}
\usepackage{csvsimple}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{lscape}
\newcommand\tab[1][1cm]{\hspace*{#1}}
\definecolor{mygreen}{rgb}{0,0.6,0}
\definecolor{mygray}{rgb}{0.5,0.5,0.5}
\definecolor{mymauve}{rgb}{0.58,0,0.82}
\lstset{ %
    backgroundcolor=\color{gray!10!white},
  basicstyle=\tiny, %footnotesize,        % the size of the fonts that are used for the code
  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
  breaklines=true,                 % sets automatic line breaking
  captionpos=b,                    % sets the caption-position to bottom
  commentstyle=\color{mygreen},    % comment style
  deletekeywords={...},            % if you want to delete keywords from the given language
  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
  frame=single,	                   % adds a frame around the code
  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
  keywordstyle=\color{blue},       % keyword style
  language=Python,                 % the language of the code
  morekeywords={*,...},           % if you want to add more keywords to the set
  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
  numbersep=10pt,                   % how far the line-numbers are from the code
  numberstyle=\tiny\color{mygray}, % the style that is used for the line-numbers
  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
  showstringspaces=false,          % underline spaces within strings only
  showtabs=false,                  % show tabs within strings adding particular underscores
  stepnumber=1,                    % the step between two line-numbers. If it's 1, each line will be numbered
  stringstyle=\color{mymauve},     % string literal style
  tabsize=1,	                   % sets default tabsize to 2 spaces
  title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
}
\graphicspath{ }
\usetikzlibrary{arrows}

\title{\textbf{COMP0085 Summative Assignment}}
%\author{Jian Shu (James) Wu \\ }
\date{Jan 4, 2023}

\begin{document}
\maketitle
\section*{Question 1}

\begin{enumerate}

\item[(a)]

\item[(b)]

\end{enumerate}

\newpage
\section*{Question 2}

\newpage
The Python code:
\lstinputlisting[language=Python]{src/solutions/q2.py}
\newpage
\section*{Question 3}

\newpage
\section*{Appendix 1: constants.py}
\lstinputlisting[language=Python]{src/constants.py}
\newpage
\section*{Appendix 2: main.py}
\lstinputlisting[language=Python]{main.py}
\end{document}

### 2022-12-17 demo_code/MStep.py

@@ -1,0 +1,41 @@
+import numpy as np
+
+
+def m_step(X, ES, ESS):
+    """
+    mu, sigma, pie = MStep(X,ES,ESS)
+
+    Inputs:
+    -----------------
+           X: shape (N, D) data matrix
+          ES: shape (N, K) E_q[s]
+         ESS: shape (K, K) sum over data points of E_q[ss'] (N, K, K)
+                           if E_q[ss'] is provided, the sum over N is done for you.
+
+    Outputs:
+    --------
+          mu: shape (D, K) matrix of means in p(y|{s_i},mu,sigma)
+       sigma: shape (,)    standard deviation in same
+         pie: shape (1, K) vector of parameters specifying generative distribution for s
+    """
+    N, D = X.shape
+    if ES.shape[0] != N:
+        raise TypeError("ES must have the same number of rows as X")
+    K = ES.shape[1]
+    if ESS.shape == (N, K, K):
+        ESS = np.sum(ESS, axis=0)
+    if ESS.shape != (K, K):
+        raise TypeError("ESS must be square and have the same number of columns as ES")
+
+    mu = np.dot(np.dot(np.linalg.inv(ESS), ES.T), X).T
+    sigma = np.sqrt(
+        (
+            np.trace(np.dot(X.T, X))
+            + np.trace(np.dot(np.dot(mu.T, mu), ESS))
+            - 2 * np.trace(np.dot(np.dot(ES.T, X), mu))
+        )
+        / (N * D)
+    )
+    pie = np.mean(ES, axis=0, keepdims=True)
+
+    return mu, sigma, pie

### 2022-12-17 demo_code/genimages.py

@@ -1,0 +1,54 @@
+import matplotlib.pyplot as plt
+import numpy as np
+
+N = 400  # number of data points - you can increase this if you want to
+# learn better features (but it will take longer).
+D = 16  # dimensionality of the data
+
+np.random.seed(0)
+
+# Define the basic shapes of the features
+
+m1 = [0, 0, 1, 0, 0, 1, 1, 1, 0, 0, 1, 0, 0, 0, 0, 0]
+
+m2 = [0, 1, 0, 0, 0, 1, 0, 0, 0, 1, 0, 0, 0, 1, 0, 0]
+
+m3 = [1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0]
+
+m4 = [1, 0, 0, 0, 0, 1, 0, 0, 0, 0, 1, 0, 0, 0, 0, 1]
+
+m5 = [0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 1, 1, 0, 0]
+
+m6 = [1, 1, 1, 1, 1, 0, 0, 1, 1, 0, 0, 1, 1, 1, 1, 1]
+
+m7 = [0, 0, 0, 0, 0, 1, 1, 0, 0, 1, 1, 0, 0, 0, 0, 0]
+
+m8 = [0, 0, 0, 1, 0, 0, 0, 1, 0, 0, 0, 1, 0, 0, 0, 1]
+
+nfeat = 8  # number of features
+rr = 0.5 + np.random.rand(nfeat, 1) * 0.5  # weight of each feature between 0.5 and 1
+mut = np.array(
+    [
+        rr[0] * m1,
+        rr[1] * m2,
+        rr[2] * m3,
+        rr[3] * m4,
+        rr[4] * m5,
+        rr[5] * m6,
+        rr[6] * m7,
+        rr[7] * m8,
+    ]
+)
+s = np.random.rand(N, nfeat) < 0.3  # each feature occurs with prob 0.3 independently
+
+# Generate Data - The Data is stored in Y
+
+Y = np.dot(s, mut) + np.random.randn(N, D) * 0.1  # some Gaussian noise is added
+
+nrows = 13
+for k in range(16):
+    plt.subplot(4, 4, k + 1)
+    plt.imshow(np.reshape(Y[k], (4, 4)), cmap=plt.gray(), interpolation="none")
+    plt.axis("off")
+
+plt.show()

### 2022-12-17 main.py

@@ -1,0 +1,13 @@
+import os
+
+from src.constants import CO2_FILE_PATH, IMAGES_FILE_PATH, OUTPUTS_FOLDER
+from src.solutions import q2, q3, q4, q5, q6
+
+if __name__ == "__main__":
+    if not os.path.exists(OUTPUTS_FOLDER):
+        os.makedirs(OUTPUTS_FOLDER)
+
+    # Question 2
+    Q2_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q2")
+    if not os.path.exists(Q2_OUTPUT_FOLDER):
+        os.makedirs(Q2_OUTPUT_FOLDER)

### 2022-12-17 src/constants.py

@@ -1,0 +1,10 @@
+import os
+
+DATA_FOLDER = "data"
+
+CO2_FILE_PATH = os.path.join(DATA_FOLDER, "co2.txt")
+IMAGES_FILE_PATH = os.path.join(DATA_FOLDER, "images.jpg")
+
+OUTPUTS_FOLDER = "outputs"
+
+DEFAULT_SEED = 0

### 2022-12-25 submission.tex
update so far

@@
 \usepackage{amssymb, bm}
+\usepackage{amsmath}
 \usepackage{csvsimple}
@@
 
-\begin{enumerate}
+\subsection*{(a)} The directed acyclic graph:
 
-\item[(a)]
+\begin{center}
+\begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}]
+    \node[main] (A) {$A$};
+    \node[main] (B) [below left of=A] {$B$};
+    \node[main] (C) [below right of=A] {$C$};
+    \node[main] (E) [below right of=C] {$E$};
+    \node[main] (D) [below left of=C] {$D$};
+    \node[main] (H) [below of=E] {$H$};
+    \node[main] (G) [right of=E] {$G$};
+    \node[main] (I) [below left of=H] {$I$};
+    \node[main] (F) [above left of=I] {$F$};
 
-\item[(b)]
+    \draw[-latex] (A) -- (B);
+    \draw[-latex] (A) -- (C);
+    \draw[-latex] (B) -- (F);
+    \draw[-latex] (C) -- (D);
+    \draw[-latex] (C) -- (E);
+    \draw[-latex] (G) -- (H);
+    \draw[-latex] (D) -- (F);
+    \draw[-latex] (E) -- (H);
+    \draw[-latex] (F) -- (I);
+    \draw[-latex] (H) -- (I);
+\end{tikzpicture}
+\end{center}
 
-\end{enumerate}
+\subsection*{(b)} The moralised graph:
+
+\begin{center}
+\begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}]
+    \node[main] (A) {$A$};
+    \node[main] (B) [below left of=A] {$B$};
+    \node[main] (C) [below right of=A] {$C$};
+    \node[main] (E) [below right of=C] {$E$};
+    \node[main] (D) [below left of=C] {$D$};
+    \node[main] (H) [below of=E] {$H$};
+    \node[main] (G) [right of=E] {$G$};
+    \node[main] (I) [below left of=H] {$I$};
+    \node[main] (F) [above left of=I] {$F$};
+
+    \draw (A) -- (B);
+    \draw (A) -- (C);
+    \draw (B) -- (F);
+    \draw (B) -- (D);
+    \draw (C) -- (D);
+    \draw (C) -- (E);
+    \draw (E) -- (G);
+    \draw (G) -- (H);
+    \draw (D) -- (F);
+    \draw (E) -- (H);
+    \draw (F) -- (H);
+    \draw (F) -- (I);
+    \draw (H) -- (I);
+\end{tikzpicture}
+\end{center}
+
+\newpage
+
+An effective triangulation:
+
+%\begin{center}
+%\begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}]
+%    \node[main] (A) {$A$};
+%    \node[main] (B) [below left of=A] {$B$};
+%    \node[main] (C) [below right of=A] {$C$};
+%    \node[main] (E) [below right of=C] {$E$};
+%    \node[main] (D) [below left of=C] {$D$};
+%    \node[main] (H) [below of=E] {$H$};
+%    \node[main] (G) [right of=E] {$G$};
+%    \node[main] (I) [below left of=H] {$I$};
+%    \node[main] (F) [above left of=I] {$F$};
+%
+%    \draw (A) -- (B);
+%    \draw (A) -- (C);
+%    \draw (B) -- (F);
+%    \draw (B) -- (D);
+%    \draw (C) -- (D);
+%    \draw (C) -- (E);
+%    \draw (E) -- (G);
+%    \draw (G) -- (H);
+%    \draw (D) -- (F);
+%    \draw (E) -- (H);
+%    \draw (F) -- (H);
+%    \draw (F) -- (I);
+%    \draw (H) -- (I);
+%
+%    \draw[dashed] (I) -- (E);
+%    \draw[dashed] (F) -- (E);
+%    \draw[dashed] (C) -- (F);
+%    \draw[dashed] (B) -- (D);
+%    \draw[dashed] (B) -- (C);
+%
+%\end{tikzpicture}
+%\end{center}
+
+\begin{center}
+\begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}]
+    \node[main] (A) {$A$};
+    \node[main] (B) [below left of=A] {$B$};
+    \node[main] (C) [below right of=A] {$C$};
+    \node[main] (E) [below right of=C] {$E$};
+    \node[main] (D) [below left of=C] {$D$};
+    \node[main] (H) [below of=E] {$H$};
+    \node[main] (G) [right of=E] {$G$};
+    \node[main] (I) [below left of=H] {$I$};
+    \node[main] (F) [above left of=I] {$F$};
+
+    \draw (A) -- (B);
+    \draw (A) -- (C);
+    \draw (B) -- (F);
+    \draw (B) -- (D);
+    \draw (C) -- (D);
+    \draw (C) -- (E);
+    \draw (E) -- (G);
+    \draw (G) -- (H);
+    \draw (D) -- (F);
+    \draw (E) -- (H);
+    \draw (F) -- (H);
+    \draw (F) -- (I);
+    \draw (H) -- (I);
+
+    \draw[dashed] (C) -- (H);
+    \draw[dashed] (C) -- (F);
+    \draw[dashed] (B) -- (D);
+    \draw[dashed] (B) -- (C);
+
+\end{tikzpicture}
+\end{center}
+
+where the dashed lines are edges added to triangulate the moralised graph.
+
+\newpage
+
+The resulting junction tree:
+
+%\begin{center}
+%\begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}, square/.style={regular polygon,regular polygon sides=4}]
+%    \node[main] (ABC) {$ABC$};
+%    \node[square] (BC) [square,draw] [below of=ABC] {$BC$};
+%    \node[main] (BCDF) [below of=BC] {$BCDF$};
+%    \node[square] (CF) [square,draw] [below of=BCDF] {$CF$};
+%    \node[main] (CEF) [below of=CF] {$CEF$};
+%    \node[square] (EF) [square,draw] [below of=CEF] {$EF$};
+%    \node[main] (EFHI) [below of=EF] {$EFHI$};
+%    \node[square] (EH) [square,draw] [below of=EFHI] {$EH$};
+%    \node[main] (EGH) [below of=EH] {$EGH$};
+%    \draw (ABC) -- (BC);
+%    \draw (BC) -- (BCDF);
+%    \draw (BCDF) -- (CF);
+%    \draw (CF) -- (CEF);
+%    \draw (CEF) -- (EF);
+%    \draw (EF) -- (EFHI);
+%    \draw (EFHI) -- (EH);
+%    \draw (EH) -- (EGH);
+%\end{tikzpicture}
+%\end{center}
+
+\begin{center}
+\begin{tikzpicture}[node distance={25mm}, main/.style = {draw, circle}, square/.style={regular polygon,regular polygon sides=4}]
+    \node[main] (ABC) {$ABC$};
+    \node[main] (BCDF) [below of=ABC] {$BCDF$};
+    \node[main] (CFH) [below of=BCDF] {$CFH$};
+    \node[main] (FHI) [below left of=CFH] {$FHI$};
+    \node[main] (CEH) [below right of=CFH] {$CEH$};
+    \node[main] (EGH) [below of=CEH] {$EGH$};
+    \draw (ABC) -- (BCDF);
+    \draw (BCDF) -- (CFH);
+    \draw (CFH) -- (FHI);
+    \draw (CFH)  -- (CEH);
+    \draw (EGH)  -- (CEH);
+\end{tikzpicture}
+\end{center}
+
+where the circular nodes are cliques.
+
+\newpage
+
+The junction tree redrawn as a factor graph:
+
+\begin{center}
+\begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}, square/.style={regular polygon,regular polygon sides=4}]
+    \node[main] (ABC) {$ABC$};
+    \node[square] (BC) [square,draw] [below of=ABC] {$BC$};
+    \node[main] (BCDF) [below of=BC] {$BCDF$};
+    \node[square] (CF) [square,draw] [below of=BCDF] {$CF$};
+    \node[main] (CFH) [below of=CF] {$CFH$};
+    \node[square] (FH) [square,draw] [below left of=CFH] {$FH$};
+    \node[main] (FHI) [below left of=FH] {$FHI$};
+    \node[square] (CH) [square,draw] [below right of=CFH] {$CH$};
+    \node[main] (CEH) [below right of=CH] {$CEH$};
+    \node[square] (EH) [square,draw] [below of=CEH] {$EH$};
+    \node[main] (EGH) [below of=EH] {$EGH$};
+    \draw (ABC) -- (BC);
+    \draw (BC) -- (BCDF);
+    \draw (BCDF) -- (CF);
+    \draw (CF) -- (CFH);
+    \draw (CFH) -- (FH);
+    \draw (FH) -- (FHI);
+    \draw (CFH) -- (CH);
+    \draw (CH) -- (CEH);
+    \draw (CEH) -- (EH);
+    \draw (EGH) -- (EH);
+\end{tikzpicture}
+\end{center}
+
+where the circular nodes are cliques and the square nodes are separators/factors.
+
+\newpage
+
+\subsection*{(c)}
+
+\begin{center}
+\begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}]
+    \node[main] (A) [fill=black!20] {$A$};
+    \node[main] (B) [below left of=A] {$B$};
+    \node[main] (C) [below right of=A] {$C$};
+    \node[main] (E) [below right of=C, fill=black!20] {$E$};
+    \node[main] (D) [below left of=C, fill=black!20] {$D$};
+    \node[main] (H) [below of=E, fill=black!20] {$H$};
+    \node[main] (G) [right of=E] {$G$};
+    \node[main] (I) [below left of=H] {$I$};
+    \node[main] (F) [above left of=I, fill=black!20] {$F$};
+
+    \draw[-latex] (A) -- (B);
+    \draw[-latex] (A) -- (C);
+    \draw[-latex] (B) -- (F);
+    \draw[-latex] (C) -- (D);
+    \draw[-latex] (C) -- (E);
+    \draw[-latex] (G) -- (H);
+    \draw[-latex] (D) -- (F);
+    \draw[-latex] (E) -- (H);
+    \draw[-latex] (F) -- (I);
+    \draw[-latex] (H) -- (I);
+\end{tikzpicture}
+\end{center}
+
+The set $\{A, D, E, F, H\}$ is a non-unique smallest set of molecules such that if the concentrations of the species within the set are known, the concentrations of the others $\{B, C, G, I\}$ would all be independent (conditioned on the measured ones).
+
+\subsection*{(d)}
+
+\newpage
+
+\subsection*{(e)}
 
@@
 
+\subsection*{(a)}
+
+We want the posterior mean and covariance over $a$ and $b$.
+Defining a weight vector $\textbf{w}$:
+
+\[\textbf{w} = \begin{bmatrix}
+                 a \\
+                 b
+         \end{bmatrix}\]
+
+Our distribution for $\textbf{w}$:
+
+\[P(\textbf{w}) = \mathcal{N} \left(
+\begin{bmatrix}
+                 \mu_a \\
+                 \mu_b
+         \end{bmatrix} ,
+\begin{bmatrix}
+                 \sigma_a^2 & 0 \\
+                 0 & \sigma_b^2
+         \end{bmatrix}
+\right)
+  = \mathcal{N}(\mu_{\textbf{w}}, \Sigma_{\textbf{w}})
+\]
+
+Moreover, for our data $\mathcal{D} = \{\textbf{X}, \textbf{Y}\}$:
+
+\[P(\mathcal{D} | \textbf{w}) = \mathcal{N} \left( \textbf{Y} - \textbf{w}^T \textbf{X}, \sigma^2 \textbf{I}
+\right)
+\]
+
+where $\textbf{X} =  \begin{bmatrix}
+                 t_1 & t_2 \cdots t_N \\
+                 1 & 1 \cdots 1
+         \end{bmatrix} \in  \mathbb{R}^{2 \times N}$ and $\textbf{Y} \in \mathbb{R}^{1 \times N}$.
+
+Knowing:
+
+\[P(\textbf{w} | \mathcal{D}) \propto P(\mathcal{D} | \textbf{w}) P(\textbf{w})\]
+
+we can substitute the above distributions:
+
+\[P(\textbf{w} | \mathcal{D}) \propto
+  \exp \left(\frac{-1}{2 \sigma^2} \left( \textbf{Y} - \textbf{w}^T \textbf{X}\right) \left( \textbf{Y} - \textbf{w}^T \textbf{X}\right)^T \right)
+\exp \left(\frac{-1}{2} \left( \textbf{w} -\mu_{\textbf{w}}\right)^T \Sigma_{\textbf{w}}^{-1} \left( \textbf{w} -\mu_{\textbf{w}}\right)\right)
+\]
+
+expanding:
+
+\[\log P(\textbf{w} | \mathcal{D}) \propto
+  \frac{-1}{2} \left( \frac{\textbf{Y} \textbf{Y}^T}{\sigma^2} - 2\textbf{w}^T \frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \textbf{w}^T \frac{\textbf{X} \textbf{X}^T}{\sigma^2} \textbf{w} +  \textbf{w}^T \Sigma_{\textbf{w}}^{-1}  \textbf{w} - 2\textbf{w}^T \Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} + \mu_{\textbf{w}}^T\Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} \right)
+\]
+
+collecting $\textbf{w}$ terms:
+
+\[\log P(\textbf{w} | \mathcal{D}) \propto
+  \frac{-1}{2} \left( \textbf{w}^T \left(\frac{\textbf{X} \textbf{X}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \right)  \textbf{w} - 2\textbf{w}^T \left(\frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right)  \right)
+\]
+
+Knowing that the posterior $P(\textbf{w} | \mathcal{D})$ will be Gaussian with mean $\bar{\mu}_w$ and covariance $\bar{\Sigma}_w$, we can see that expanding the exponent component would have the form:
+
+\[\left( \textbf{w} - \bar{\mu}_w \right)^T \bar{\Sigma}_w^{-1} \left( \textbf{w} - \bar{\mu}_w \right) = \textbf{w}^T \bar{\Sigma}_w^{-1} \textbf{w} -2 \textbf{w}^T \bar{\Sigma}_w^{-1} \bar{\mu}_w + \bar{\mu}_w^T \bar{\Sigma}_w^{-1} \bar{\mu}_w\]
+
+Thus we can identify the posterior covariance:
+
+\[\bar{\Sigma}_w = \left(\frac{\textbf{X} \textbf{X}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \right)^{-1}\]
+
+and the posterior mean:
+
+\[\bar{\mu}_w = \bar{\Sigma}_w \left(\frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right) \]
+
 \newpage
@@
 \lstinputlisting[language=Python]{src/solutions/q2.py}
+
 \newpage
 \section*{Question 3}
+
+\subsection*{(a)}
+
+The free energy is can be calculated as:
+\[\mathcal{F}(q, \theta) = \langle \log P(\textbf{x}, \textbf{s} | \theta)\rangle_{q(\textbf{s})} + H[Q(\textbf{s})]\]
+
+%Thus,
+%\[\mathcal{F}(Q, \theta) = \langle \log P(\textbf{x}, \textbf{s} | \theta)\rangle_{\prod_{i=1}^{K} q_{i} (s_i)} + H\left[ \prod_{i=1}^{K} q_{i} (s_i)\right] \]
+
+Knowing,
+\[\log P(\textbf{x}, \textbf{s} | \theta) = \log P(\textbf{x} | \textbf{s}, \theta) + \log P(\textbf{s} | \theta)\]
+
+we can write:
+
+\[\mathcal{F}(Q, \theta) = \langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} + \langle \log P(\textbf{s} | \theta)\rangle_{q(\textbf{s})} + H\left[ q(\textbf{s})\right] \]
+
+Moreover, our mean field approximation:
+\[q(\textbf{s}) = \prod_{i=1}^{K} q_{i} (s_i) \]
+
+where $q_{i} (s_i) = \lambda_i^{s_i} (1-\lambda_i)^{(1-s_i)}$.
+
+To compute the first term:
+
+\[ P(\textbf{x} | \textbf{s}, \theta)  = \mathcal{N}\left( \sum_{i=1}^{K} s_i \mu_i, \sigma^2 \textbf{I} \right)}\]
+
+substituting the appropriate terms:
+
+\[ P(\textbf{x} | \textbf{s}, \theta)  = 2\pi^{-\frac{d}{2}} |\sigma^2 \textbf{I}|^{-\frac{1}{2}} \exp \left( -\frac{1}{2} \left(\textbf{x} - \sum_{i=1}^{K} s_i \mu_i\right)^T \frac{1}{\sigma^2} \textbf{I}  \left(\textbf{x} - \sum_{i=1}^{K} s_i \mu_i\right) \right) \]
+
+with $d$ being the number of dimensions.
+
+Taking the logarithm:
+
+\[ \log P(\textbf{x} | \textbf{s}, \theta)  = -\frac{d}{2} \log (2 \pi) - \log(\sigma)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} s_i \mu_i   + \sum_{i=1}^{K} \sum_{i=1}^{K} s_i s_j \mu_i^T\mu_j \right) \]
+
+The expectation distributed to the relevant terms:
+
+\[
+\langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} \\
+= -\frac{d}{2} \log (2 \pi) - \log(\sigma)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} \langle s_i \rangle_{q_{i} (s_i)} \mu_i   + \sum_{i=1}^{K} \sum_{j=1}^{K} \langle s_i s_j \rangle_{q_{i} (s_i) q_{j} (s_j)} \mu_i^T\mu_j \right)\]
+
+Evaluating the expectations:
+
+\[
+\langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} \\
+= -\frac{d}{2} \log (2 \pi) - \log(\sigma)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right)\]
+
+where $\langle s_i s_i \rangle_{q_{i} (s_i)} = \langle s_i \rangle_{q_{i} (s_i)}$ because $s_i \in \{0, 1\}$.
+
+To compute the second term:
+\[ P(\textbf{s} | \theta) = \prod_{i=1}^{K}\pi_i^{s_i} (1-\pi_i)^{(1-s_i)}\]
+
+Taking the logarithm:
+
+\[ \log P(\textbf{s} | \theta) = \sum_{i=1}^{K}s_i \log \pi_i + (1-s_i) \log (1-\pi_i)}\]
+
+The expectation distributed to the relevant terms:
+
+\[ \langle \log P(\textbf{s} | \theta)\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \langle s_i \rangle_{q_{i} (s_i)} \log \pi_i + (1-\langle s_i\rangle_{q_{i} (s_i)}) \log (1-\pi_i)}\]
+
+Evaluating the expectations:
+
+\[ \langle \log P(\textbf{s} | \theta)\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \lambda_i \log \pi_i + (1-\lambda_i) \log (1-\pi_i)}\]
+
+To compute the third term, we use the mean field factorisation:
+
+\[H\left[ q(\textbf{s})\right] = \sum_{i=1}^K H\left[ q_{i} (s_i)\right] \]
+
+Thus,
+
+\[H\left[ q(\textbf{s})\right] = - \sum_{i=1}^K \sum_{s_i \in \{0, 1\}} q_{i} (s_i) \log q_{i} (s_i) \]
+
+Substituting the appropriate values:
+
+\[H\left[ q(\textbf{s})\right] = - \sum_{i=1}^K \lambda_i \log \lambda_i + (1-\lambda_i) \log (1-\lambda_i) \]
+
+Combining, we have our free energy expression:
+
+\[
+\begin{array}{l}
+\tab \mathcal{F}(q, \theta) = \\
+\tab \tab \frac{-d}{2} \log (2 \pi) - \log(\sigma)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right) \\
+\tab \tab + \sum_{i=1}^{K} \lambda_i \log \pi_i + (1-\lambda_i) \log (1-\pi_i)}\\
+\tab \tab - \sum_{i=1}^K \lambda_i \log \lambda_i + (1-\lambda_i) \log (1-\lambda_i)
+\end{array}
+\]
+
+To derive the partial update for $q_i(s_i)$ we take the variational derivative of the Lagrangian, enforcing the normalisation of $q_i$:
+
+\[\frac{\partial}{\partial q_i}\left( \mathcal{F}(q, \theta) + \lambda^{LG} \int q_i -1)\right) = \langle \log P(\textbf{x}, \textbf{s} | \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} - \log q_i(s_i) - 1 + \lambda^{LG}\]
+
+where $\lambda^{LG}$ is the Lagrange multiplier.
+
+Setting this to zero we can solve for the $\lambda_i$ that maximises the free energy:
+
+\[\log q_i(s_i) = \langle \log P(\textbf{x}, \textbf{s} | \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} - 1 + \lambda^{LG}\]
+
+Similar to our free energy derivation:
+
+\[\langle \log P(\textbf{x} |  \textbf{s}, \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} \propto  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} \langle s_k \rangle_{\prod_{j\neq i} q_j(s_j)} \mu_i   + \sum_{k=1}^{K} \sum_{j=1}^{K} \langle s_k s_j \rangle_{\prod_{j\neq i} q_j(s_j)} \right)\]
+
+and
+
+\[ \langle \log P(\textbf{s} | \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} = \sum_{k=1}^{K} \langle s_k \rangle_{\prod_{j\neq i} q_j(s_j)} \log \pi_k + (1-\langle s_k\rangle_{\prod_{j\neq i} q_j(s_j)}) \log (1-\pi_k)}\]
+
+We can write:
+
+\[\log q_i(s_i) \propto  \log P(\textbf{x} |  \textbf{s}, \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} + \langle \log P(\textbf{s} | \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} \]
+
+Substituting the relevant terms:
+
+\[\log q_i(s_i) \propto  -\frac{1}{2 \sigma^2} \left( - 2 s_i \textbf{x}^T \mu_i   + s_i s_i  \mu_i^T\mu_i +  2 \sum_{j=1, j\neq i}^{K}  s_i \lambda_j  \mu_i^T\mu_j  \right) + s_i \log \pi_i + (1- s_i ) \log (1-\pi_i)}
+\]
+
+Knowing $\log q_i(s_i) = s_i \log \lambda_i + (1-s_i) \log (1-\lambda_i)$:
+
+\[\log q_i(s_i) \propto s_i \log \frac{\lambda_i}{1-\lambda_i}\]
+
+Thus,
+
+\[s_i \log \frac{\lambda_i}{1-\lambda_i} \propto  -\frac{1}{2 \sigma^2} \left( - 2 s_i \textbf{x}^T \mu_i   + s_i s_i  \mu_i^T\mu_i +  2 \sum_{j=1, j\neq i}^{K}  s_i \lambda_j  \mu_i^T\mu_j  \right) + s_i \log \frac{\pi_i}{1-\pi_i}}\]
+
+Also, because $s_i \in \{0, 1\}}$ we know that $s_i^2 = s_i$:
+
+\[s_i \log \frac{\lambda_i}{1-\lambda_i} \propto  -\frac{1}{2 \sigma^2} \left( - 2 s_i \textbf{x}^T \mu_i   +  s_i  \mu_i^T\mu_i +  2 \sum_{j=1, j\neq i}^{K}  s_i \lambda_j  \mu_i^T\mu_j  \right) + s_i \log \frac{\pi_i}{1-\pi_i}}\]
+
+Because we have only kept terms with $s_i$, this is an equality:
+
+\[s_i \log \frac{\lambda_i}{1-\lambda_i} =  \frac{  s_i \mu_i^T}{2 \sigma^2} \left( 2\textbf{x} - \mu_i -  2 \sum_{j=1, j\neq i}^{K}   \lambda_j  \mu_j \right) + s_i \log \frac{\pi_i}{1-\pi_i}}\]
+
+Solving for $\lambda_i$:
+
+\[ \lambda_i =  \frac{1}{1+\exp\left[ - \left(\frac{  \mu_i^T}{\sigma^2} \left( \textbf{x} -  \frac{\mu_i}{2} -  \sum_{j=1, j\neq i}^{K}   \lambda_j  \mu_j \right) + \log \frac{\pi_i}{1-\pi_i}}\right) \right]}\]
+
+we have our partial update.
+
+\newpage
+The Python code:
+\lstinputlisting[language=Python]{src/solutions/q3.py}
+\newpage
+\section*{Question 4}
+
+\newpage
+\section*{Question 5}
+
+\subsection*{(a)}
+
+The log-joint probability for a single observation-source pair:
+
+\[\log p(\textbf{s}, \textbf{x}) = \log p(\textbf{s}) + \logp(\textbf{x}|\textbf{s})\]
+
+Knowing $p(\textbf{s}) = \prod_{i=1}^{K}p(s_i| \pi_i)$ and $p(\textbf{x}|\textbf{s}) = \mathcal{N}(\sum_{i=1}^{K} s_i \mu_i, \sigma^2 \textbf{I})$:
+
+\[\log p(\textbf{s}, \textbf{x})  \propto \frac{-1}{2}\left( \textbf{x} - \sum_{i=1}^{K}s_i \mu_i\right)^T \frac{1}{\sigma^2} \textbf{I} \left( \textbf{x} - \sum_{i=1}^{K} s_i \mu_i\right) + \sum_{i=1}^{K} \left(s_i \log\pi_i + (1-s_i)\log(1-\pi_i)\right)\]
+
+Expanding:
+
+\[\log p(\textbf{s}, \textbf{x})  \propto \frac{-1}{2\sigma^2} \left( \textbf{x}^T\textbf{x} - 2\textbf{x}^T\sum_{i=1}^{K}s_i \mu_i + \sum_{i=1}^{K}\sum_{j=1}^{K}s_i s_j \mu_i^T \mu_j\right) + \sum_{i=1}^{K} \left(s_i \log\pi_i + (1-s_i)\log(1-\pi_i)\right)\]
+
+Collecting terms pertaining to $s_i$:
+
+\[\log p(\textbf{s}, \textbf{x})  =    \sum_{i=1}^{K} \left(\left(\frac{\textbf{x}^T \mu_i}{\sigma^2} +\log\frac{\pi_i}{1-\pi_i} \right) s_i\right) + \sum_{i=1}^{K}\sum_{j=1}^{K}\left( \frac{ - \mu_i^T \mu_j}{2\sigma^2} s_i s_j \right) + C\]
+
+where $C$ are all other terms without $s_i$.
+
+Knowing that $s_i^2= s_i$:
+
+\[\log p(\textbf{s}, \textbf{x})  =    \sum_{i=1}^{K} \left(\left(\frac{\textbf{x}^T \mu_i}{\sigma^2} +\log\frac{\pi_i}{1-\pi_i} - \frac{\mu_i^T \mu_j}{2\sigma^2} \right) s_i\right) + \sum_{i=1}^{K}\sum_{j=1}^{i-1}\left( \frac{ - \mu_i^T \mu_j}{\sigma^2} s_i s_j \right) + C\]
+
+Thus:
+
+\[\log p(\textbf{s}, \textbf{x})  =    \sum_{i=1}^{K} \log f_i(s_i) + \sum_{i=1}^{K}\sum_{j=1}^{i-1}\log g_{ij}(s_i, s_j)\right) \]
+
+where the factors are defined:
+
+\[\log f_i(s_i) = \left(\frac{\textbf{x}^T \mu_i}{\sigma^2} +\log\frac{\pi_i}{1-\pi_i} - \frac{\mu_i^T \mu_j}{2\sigma^2} \right) s_i\]
+
+and
+
+\[\log g_{ij}(s_i, s_j) = \frac{- \mu_i^T \mu_j}{\sigma^2} s_i s_j\]
+
+as required. Note that $C$ can simply be absorbed into any one of these factors.
+
+The Boltzmann Machine can be defined:
+
+\[P(\textbf{s}| \textbf{W}, \textbf{b}) = \frac{1}{Z} \exp\left( \sum_{i=1}^{K}\sum_{j=1}^{i-1}W_{ij}s_i s_j - \sum_{i=1}^{K} b_i s_i\right)}\]
+
+where $s_i \in \{0, 1\}$, the same as our source variables.
+
+From our factorisation, we can see that $p(\textbf{s}, \textbf{x})$ is a Boltzmann Machine with:
+
+\[W_{ij} = \frac{- \mu_i^T \mu_j}{\sigma^2}\]
+
+and
+
+\[b_i = -\left(\frac{\textbf{x}^T \mu_i}{\sigma^2} +\log\frac{\pi_i}{1-\pi_i} - \frac{\mu_i^T \mu_j}{2\sigma^2}\right)\]
+
+and
+
+\[\log Z = -C \]
+
+\subsection*{(b)}
+
+For $f_i(s_i)$, we will choose a Bernoulli approximation:
+
+\[\tilde{f}_i(s_i) = \lambda_i^{s_i} + (1-\lambda_i)^{1-s_i}\]
+
+Thus,
+
+\[\log \tilde{f}_i(s_i) \propto \log \left(\frac{\lambda_i}{1-\lambda_i} \right)s_i\]
+
+For $g_{ij}(s_i, s_j)$, we will choose a product of Bernoulli's approximation:
+
+\[\tilde{g}_{ij}(s_i, s_j) = \left( \theta_i^{s_i} + (1-\theta_i)^{1-s_i}\right)\left( \theta_j^{s_j} + (1-\theta_j)^{1-s_j}\right)\]
+
+Thus,
+
+\[\log \tilde{g}_{ij}(s_i, s_j) \propto \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j\]
+
+To derive the a message passing scheme, we first define:
+
+\[q(\textbf{s}) = \left(\prod_{i=1}^{K} \tilde{f}_i(s_i) \right) \left(\prod_{i=1}^{K}  \prod_{j=1}^{i-1} \tilde{g}_{ij} (s_i, s_j) \right)\]
+
+Thus, we can derive cavity distributions:
+
+\[q_{\neg \tilde{f}_i(s_i)}(\textbf{s}) = \left(\prod_{i'=1, i'\neq i}^{K} \tilde{f}_{i'}(s_{i'}) \right) \left(\prod_{i=1}^{K}  \prod_{j=1}^{i-1} \tilde{g}_{ij} (s_i, s_j) \right)\]
+
+and
+
+\[q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) = \left(\prod_{i'=1}^{K} \tilde{f}_{i'}(s_{i'}) \right) \left(\prod_{i'=1}^{K}  \prod_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1} \tilde{g}_{i'j'} (s_{i'}, s_{j'}) \right)\]
+
+For $\tilde{f}_{i}(s_{i})$, we do not need to make an approximation step.
+This is because we are minimising:
+
+\[\tilde{f}_{i}(s_{i}) = \arg \min \textbf{KL} \left[ f_{i}(s_{i}) q_{\neg \tilde{f}_i(s_i)}(\textbf{s}) \| \tilde{f}_{i}(s_{i}) q_{\neg \tilde{f}_i(s_i)}(\textbf{s}) \right]\]
+
+We know that the factor $\log f_i(s_i)$ is a Bernoulli of the form $b_i s_i$. Because our approximation is also Bernoulli, we can simply solve for $\lambda_i$ in $\log \tilde{f}_{i}(s_{i})$:
+
+\[\log \tilde{f}_i(s_i) = \log f_{i}(s_{i})\]
+
+\[\log \left(\frac{\lambda_i}{1-\lambda_i} \right)s_i = b_i s_i\]
+
+\[\lambda_i = \frac{1}{1+\exp(-b_i)}\]
+
+On the other hand, for $\tilde{g}_{ij}(s_i, s_j)$, we will approximate with:
+
+\[\tilde{g}_{ij}(s_i, s_j) = \arg \min \textbf{KL} \left[ g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \| \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \right]\]
+
+Note that because $\tilde{g}_{ij}(s_i, s_j)$ is the product of two Bernoulli distributions, we only require the natural parameters:
+
+\[ \phi_{ij}(\theta) = \begin{bmatrix} \theta_i \\ \theta_j\right] \end{bmatrix}\]
+
+the mean with respect to $s_i$ and $s_j$ respectively.
+
+We can write:
+%\[\log \tilde{g}_{ij}q_{\neg \tilde{g}_{ij}} \propto \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j + \sum_{i=1}^{K}  \log \left(\frac{\lambda_i}{1-\lambda_i} \right) +  \sum_{i'=1}^{K}\sum_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1}  \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j\]
+
+\[
+\begin{array}{l}
+ \log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j\\
+\tab \tab \tab \tab \tab + \sum_{i'=1}^{K}  \log \left(\frac{\lambda_{i'}}{1-\lambda_{i'}} \right) s_{i'} \\
+\tab \tab \tab \tab \tab +  \sum_{i'=1}^{K}\sum_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1}  \log \left(\frac{\theta_{i'}}{1-\theta_{i'}} \right) s_{i'} + \log \left(\frac{\theta_{j'}}{1-\theta_{j'}} \right) s_{j'}
+\end{array}
+\]
+
+Only including terms with $s_i$ and $s_j$:
+
+%\[
+% \log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \left(K \log \left(\frac{\theta_i}{1-\theta_i} \right) + \log \left(\frac{\lambda_i}{1-\lambda_i} \right)\right) s_i +   \left(
+%K \log \left(\frac{\theta_j}{1-\theta_j} \right) + \log \left(\frac{\lambda_j}{1-\lambda_j} \right)\right) s_j
+%\]
+
+\[
+\begin{array}{l}
+ \log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \left((K-1) \log \left(\frac{\theta_i}{1-\theta_i} \right) + \log \left(\frac{\lambda_i}{1-\lambda_i} \right)\right) s_i\\
+\tab \tab \tab \tab \tab +\left(
+(K-1) \log \left(\frac{\theta_j}{1-\theta_j} \right) + \log \left(\frac{\lambda_j}{1-\lambda_j} \right)\right) s_j
+\end{array}
+\]
+
+Moreover:
+
+\[
+\begin{array}{l}
+ \log g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \frac{- \mu_i^T \mu_j}{\sigma^2} s_i s_j\\
+\tab \tab \tab \tab \tab + \sum_{i'=1}^{K}  \log \left(\frac{\lambda_{i'}}{1-\lambda_{i'}} \right) s_{i'} \\
+\tab \tab \tab \tab \tab +  \sum_{i'=1}^{K}\sum_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1}  \log \left(\frac{\theta_{i'}}{1-\theta_{i'}} \right) s_{i'} + \log \left(\frac{\theta_{j'}}{1-\theta_{j'}} \right) s_{j'}
+\end{array}
+\]
+
+Only including terms with $s_i$ and $s_j$:
+
+\[
+\begin{array}{l}
+ \log g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \frac{- \mu_i^T \mu_j}{\sigma^2} s_i s_j\\
+\tab \tab \tab \tab \tab + \left((K-2)\log \left(\frac{\theta_i}{1-\theta_i} \right) + \log \left(\frac{\lambda_i}{1-\lambda_i} \right)\right) s_i \\
+\tab \tab \tab \tab \tab +  \left((K-2)\log \left(\frac{\theta_j}{1-\theta_j} \right) + \log \left(\frac{\lambda_j}{1-\lambda_j} \right)\right) s_j
+\end{array}
+\]
 

### 2022-12-25 main.py

@@ -1,13 +1,70 @@
 import os
-
+import pandas as pd
+import numpy as np
 from src.constants import CO2_FILE_PATH, IMAGES_FILE_PATH, OUTPUTS_FOLDER
 from src.solutions import q2, q3, q4, q5, q6
+from src.solutions.q2 import LinearRegressionParameters
+from src.generate_images import generate_images
+
 
 if __name__ == "__main__":
     if not os.path.exists(OUTPUTS_FOLDER):
         os.makedirs(OUTPUTS_FOLDER)
 
+    with open(CO2_FILE_PATH) as file:
+        lines = [line.rstrip().split() for line in file]
+
     # Question 2
     Q2_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q2")
     if not os.path.exists(Q2_OUTPUT_FOLDER):
         os.makedirs(Q2_OUTPUT_FOLDER)
+    df_co2 = pd.DataFrame(
+        np.array([line for line in lines if line[0] != "#"]).astype(float)
+    )
+    column_names = lines[max([i for i, line in enumerate(lines) if line[0] == "#"])][1:]
+    df_co2.columns = column_names
+    t = df_co2.decimal.values[:] - np.min(df_co2.decimal.values[:])
+    y = df_co2.average.values[:].reshape(1, -1)
+
+    sigma = 1
+    mean = np.array([0, 360]).reshape(-1, 1)
+    covariance = np.array(
+        [
+            [10**2, 0],
+            [0, 100**2],
+        ]
+    )
+
+    prior_linear_regression_parameters = LinearRegressionParameters(
+        mean=mean,
+        covariance=covariance,
+    )
+    posterior_linear_regression_parameters = q2.a(
+        t,
+        y,
+        sigma,
+        prior_linear_regression_parameters,
+        save_path=os.path.join(Q2_OUTPUT_FOLDER, "a"),
+    )
+    q2.b(
+        t_year=df_co2.decimal.values[:],
+        t=t,
+        y=y,
+        linear_regression_parameters=posterior_linear_regression_parameters,
+        error_mean=0,
+        error_variance=1,
+        save_path=os.path.join(Q2_OUTPUT_FOLDER, "b"),
+    )
+
+    # # Question 3
+    # Q3_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q3")
[…]

### 2022-12-25 src/constants.py

@@ -8,3 +8,19 @@ IMAGES_FILE_PATH = os.path.join(DATA_FOLDER, "images.jpg")
 OUTPUTS_FOLDER = "outputs"
 
 DEFAULT_SEED = 0
+
+M1 = [0, 0, 1, 0, 0, 1, 1, 1, 0, 0, 1, 0, 0, 0, 0, 0]
+
+M2 = [0, 1, 0, 0, 0, 1, 0, 0, 0, 1, 0, 0, 0, 1, 0, 0]
+
+M3 = [1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0]
+
+M4 = [1, 0, 0, 0, 0, 1, 0, 0, 0, 0, 1, 0, 0, 0, 0, 1]
+
+M5 = [0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 1, 1, 0, 0]
+
+M6 = [1, 1, 1, 1, 1, 0, 0, 1, 1, 0, 0, 1, 1, 1, 1, 1]
+
+M7 = [0, 0, 0, 0, 0, 1, 1, 0, 0, 1, 1, 0, 0, 0, 0, 0]
+
+M8 = [0, 0, 0, 1, 0, 0, 0, 1, 0, 0, 0, 1, 0, 0, 0, 1]

### 2022-12-25 src/generate_images.py

@@ -1,0 +1,38 @@
+import numpy as np
+from src.constants import M1, M2, M3, M4, M5, M6, M7, M8, DEFAULT_SEED
+
+
+def generate_images(n: int = 400, seed: int = DEFAULT_SEED):
+    """
+
+    :param n: number of data points
+    :param seed: random seed
+    :return:
+    """
+    d = 16  # dimensionality of the data
+    np.random.seed(seed)
+
+    # Define the basic shapes of the features
+    number_of_features = 8  # number of features
+    rr = (
+        0.5 + np.random.rand(number_of_features, 1) * 0.5
+    )  # weight of each feature between 0.5 and 1
+    mut = np.array(
+        [
+            rr[0] * M1,
+            rr[1] * M2,
+            rr[2] * M3,
+            rr[3] * M4,
+            rr[4] * M5,
+            rr[5] * M6,
+            rr[6] * M7,
+            rr[7] * M8,
+        ]
+    )
+    s = (
+        np.random.rand(n, number_of_features) < 0.3
+    )  # each feature occurs with prob 0.3 independently
+
+    # Generate Data - The Data is stored in Y
+
+    return np.dot(s, mut) + np.random.randn(n, d) * 0.1  # some Gaussian noise is added

### 2022-12-25 src/solutions/q3.py

@@ -1,0 +1,318 @@
+from dataclasses import dataclass
+
+import numpy as np
+
+from demo_code.MStep import m_step
+
+
+@dataclass
+class MeanFieldApproximation:
+    """
+    lambda_matrix: parameters variational approximation (number_of_points, number_of_latent_variables)
+    """
+
+    lambda_matrix: np.ndarray
+
+    def lambda_matrix_exclude(self, exclude_latent_index: int) -> np.ndarray:
+        #  (number_of_points, number_of_latent_variables-1)
+        return np.concatenate(
+            (
+                self.lambda_matrix[:, :exclude_latent_index],
+                self.lambda_matrix[:, exclude_latent_index + 1 :],
+            ),
+            axis=1,
+        )
+
+    @property
+    def log_lambda_matrix(self):
+        return np.log(self.lambda_matrix)
+
+    @property
+    def log_one_minus_lambda_matrix(self):
+        return np.log(1 - self.lambda_matrix)
+
+    @property
+    def n(self):
+        return self.lambda_matrix.shape[0]
+
+    @property
+    def k(self):
+        return self.lambda_matrix.shape[1]
+
+
+def init_mean_field_approximation(k: int, n: int) -> MeanFieldApproximation:
+    return MeanFieldApproximation(
+        lambda_matrix=np.random.random(size=(n, k)),
+    )
+
+
+@dataclass
+class BinaryLatentFactorModel:
+    """
+    mu: matrix of means (number_of_dimensions, number_of_latent_variables)
+    sigma: gaussian noise parameter
+    pi: vector of priors (1, number_of_latent_variables)
+    """
+
+    mu: np.ndarray
+    sigma: float
+    pi: np.ndarray
+
+    def mu_exclude(self, exclude_latent_index: int) -> np.ndarray:
+        #  (number_of_dimensions, number_of_latent_variables-1)
+        return np.concatenate(
+            (self.mu[:, :exclude_latent_index], self.mu[:, exclude_latent_index + 1 :]),
+            axis=1,
+        )
+
+    @property
+    def log_pi(self):
+        return np.log(self.pi)
+
+    @property
+    def log_one_minus_pi(self):
+        return np.log(1 - self.pi)
+
+    @property
+    def variance(self):
+        return self.sigma**2
+
+    @property
+    def precision(self):
+        return 1 / self.variance
+
+    @property
+    def d(self):
+        return self.mu.shape[0]
+
+    @property
+    def k(self):
+        return self.mu.shape[1]
+
+
[…]

### 2022-12-25 src/solutions/q3.py
fix free energy

@@ -120,29 +120,33 @@ def _compute_expectation_log_p_x_s_given_theta(
         binary_latent_factor_model.mu.T @ binary_latent_factor_model.mu,
     )
 
-    expectation_log_p_x_given_s_theta = (
-        (-binary_latent_factor_model.d / 2) * np.log(2 * np.pi)
-        - np.log(binary_latent_factor_model.sigma)
-        - (0.5 * binary_latent_factor_model.precision)
-        * (
-            np.sum(np.multiply(x, x))
-            - 2 * np.sum(np.multiply(x, mu_lambda))
-            + np.sum(expectation_s_i_s_j_mu_i_mu_j)
-            - np.trace(
-                expectation_s_i_s_j_mu_i_mu_j
-            )  # remove incorrect E[s_i s_i] = lambda_i * lambda_i
-            + np.sum(  # add correct E[s_i s_i] = lambda_i
-                mean_field_approximation.lambda_matrix
-                @ np.multiply(
-                    binary_latent_factor_model.mu, binary_latent_factor_model.mu
-                ).T
-            )
+    expectation_log_p_x_given_s_theta = -(
+        mean_field_approximation.n * binary_latent_factor_model.d / 2
+    ) * np.log(2 * np.pi * binary_latent_factor_model.variance) - (
+        0.5 * binary_latent_factor_model.precision
+    ) * (
+        np.sum(np.multiply(x, x))
+        - 2 * np.sum(np.multiply(x, mu_lambda))
+        + np.sum(expectation_s_i_s_j_mu_i_mu_j)
+        - np.trace(
+            expectation_s_i_s_j_mu_i_mu_j
+        )  # remove incorrect E[s_i s_i] = lambda_i * lambda_i
+        + np.sum(  # add correct E[s_i s_i] = lambda_i
+            mean_field_approximation.lambda_matrix
+            @ np.multiply(
+                binary_latent_factor_model.mu, binary_latent_factor_model.mu
+            ).T
         )
     )
     expectation_log_p_s_given_theta = np.sum(
-        mean_field_approximation.lambda_matrix @ binary_latent_factor_model.log_pi.T
-        + (1 - mean_field_approximation.lambda_matrix)
-        @ binary_latent_factor_model.log_one_minus_pi.T
+        np.multiply(
+            mean_field_approximation.lambda_matrix,
+            binary_latent_factor_model.log_pi,
+        )
+        + np.multiply(
+            1 - mean_field_approximation.lambda_matrix,
[…]

### 2022-12-25 submission.tex
fix free energy
update to gp question

@@
 Defining a weight vector $\textbf{w}$:
-
 \[\textbf{w} = \begin{bmatrix}
@@
 Our distribution for $\textbf{w}$:
-
 \[P(\textbf{w}) = \mathcal{N} \left(
@@
 Moreover, for our data $\mathcal{D} = \{\textbf{X}, \textbf{Y}\}$:
-
 \[P(\mathcal{D} | \textbf{w}) = \mathcal{N} \left( \textbf{Y} - \textbf{w}^T \textbf{X}, \sigma^2 \textbf{I}
@@
 Knowing:
-
 \[P(\textbf{w} | \mathcal{D}) \propto P(\mathcal{D} | \textbf{w}) P(\textbf{w})\]
@@
 we can substitute the above distributions:
-
 \[P(\textbf{w} | \mathcal{D}) \propto
@@
 expanding:
-
 \[\log P(\textbf{w} | \mathcal{D}) \propto
@@
 collecting $\textbf{w}$ terms:
-
 \[\log P(\textbf{w} | \mathcal{D}) \propto
@@
 Knowing that the posterior $P(\textbf{w} | \mathcal{D})$ will be Gaussian with mean $\bar{\mu}_w$ and covariance $\bar{\Sigma}_w$, we can see that expanding the exponent component would have the form:
-
 \[\left( \textbf{w} - \bar{\mu}_w \right)^T \bar{\Sigma}_w^{-1} \left( \textbf{w} - \bar{\mu}_w \right) = \textbf{w}^T \bar{\Sigma}_w^{-1} \textbf{w} -2 \textbf{w}^T \bar{\Sigma}_w^{-1} \bar{\mu}_w + \bar{\mu}_w^T \bar{\Sigma}_w^{-1} \bar{\mu}_w\]
@@
 Thus we can identify the posterior covariance:
-
 \[\bar{\Sigma}_w = \left(\frac{\textbf{X} \textbf{X}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \right)^{-1}\]
@@
 and the posterior mean:
-
 \[\bar{\mu}_w = \bar{\Sigma}_w \left(\frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right) \]
@@
 \newpage
-The Python code:
+Computing the posterior mean and covariance over $a$ and $b$ given by the $CO_2$ data:
+
+\begin{figure}[h]
+\centering
+\begin{minipage}{.5\textwidth}
+  \centering
+\includegraphics[scale = 0.5]{outputs/q2/a-mean}
+\caption{The Posterior Mean}
+\label{fig:fig2-a-mean}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+\includegraphics[scale = 0.5]{outputs/q2/a-covariance}
+\caption{t=The Posterior Covariance}
+\label{fig:fig2-a-covariance}
+\end{minipage}
+\end{figure}
+
+\subsection*{(b)}
+
+Plotting the residuals:
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.5]{outputs/q2/b-residuals-timeseries}
+\caption{$g_{obs}(t)$}
+\label{fig:b-residuals-timeseries}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.5]{outputs/q2/b-residuals-density-estimation}
+\caption{Density Estimation of Residuals vs $e(t) \sim \mathcal{N}(0, 1)$}
+\label{fig:b-residuals-density-estimation}
+\end{figure}
+
+We can see that the residuals do not perfectly conform to our prior over $e(t) \sim \mathcal{N}(0, 1)$.
+The density estimation shows that a mean of zero is a reasonable prior belief however the data does not seem to exhibit unit variance.
+
+\newpage
+\subsection*{(c \& d)}
+
+We are considering the kernel:
+
+\[k(s, t) = \theta^2 \left( \exp\left( - \frac{2\sin^2(\pi(s-t)/\tau)}{\sigma^2}\right) + \phi^2 \exp\left( - \frac{(s-t)^2}{2\eta^2}\right)\right) + \zeta^2 \delta_{s=t}\]
+
+We can make qualitative observations this kernel by visualising the covariance (gram) matrix:
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.5]{outputs/q2/c-gram-matrix}
+\caption{Gram Matrix}
+\label{fig:c-gram-matrix}
+\end{figure}
+
+We can observe a striped pattern which indicate higher covariance at regular intervals.
+This can be attributed to the sinusoidal term in the kernel and encourages sinusoidal functions.
+Additionally, we can see that covariance values also decay as they are further away from the diagonal.
+This can be attributed to the exponential term in the kernel, encouraging points closer in time to be more correlated and vice versa.
+From our $CO_2$ data, we would want a class of functions which exhibit both of these behaviours as the data looks sinusoidal (seasonal with respect to each year) and correlations locally.
+
+We can also visualise some samples from a Gaussian Process with the same covariance matrix and zero mean.
+This verifies our observations about the covariance matrix.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.5]{outputs/q2/c-samples}
+\caption{Samples from a zero mean GP with the provided covariance kernel}
+\label{fig:c-samples}
+\end{figure}
+
+\newpage
+More specifically, we can see how changing each hyper-parameter will affect the characteristics of the function.
+
+\begin{figure}[h]
+\centering
+\begin{minipage}{.5\textwidth}
+  \centering
+\includegraphics[scale = 0.35]{outputs/q2/c-parameter-grams}
+\caption{Covariances for different parameters}
+\label{fig:fig2-c-parameter-grams}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+\includegraphics[scale = 0.35]{outputs/q2/c-parameter-samples}
+\caption{Samples for different parameters}
+\label{fig:fig2-c-parameter-samples}
+\end{minipage}
+\end{figure}
+
+\begin{itemize}
+\item [$\theta$:] As $\theta$ increases, we see more pronounced periodic behavior in the sample function.
+    The covariance matrix shows how increasing $\theta$ visually reveals the striped periodic component.
+        This is expected because it is the parameter that adjusts the weight of $\exp\left( - \frac{2\sin^2(\pi(s-t)/\tau)}{\sigma^2}\right)$.
+\item [$\sigma$:] As $\sigma$ increases, we see smoother periodic behaviour in the sample function.
+    The covariance matrix shows how increasing $\sigma$ will increase covariance values in the off-diagonals.
+    This is expected because it adjusts the lengthscale of the periodic portion of the kernel.
+\item [$\phi$:] As $\phi$ increases, we see less smooth behaviour in the sample function.
+    The covariance matrix shows how increasing $\sigma$ will increase covariance values in the off-diagonals.
+    This is expected because it adjusts the lengthscale of the periodic portion of the kernel.
+\item [$\eta$:]
+\item [$\tau$:]
+\item [$\zeta$:]
+\end{itemize}
+\newpage
+\subsection*{(e)}
+
+\subsection*{(f)}
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.5]{outputs/q2/f-extrapolation}
+\caption{Extrapolation of $CO_2$ concentration levels}
+\label{fig:f-extrapolation}
+\end{figure}
+
+\subsection*{(g)}
+
+\newpage
+The Python code for Bayesian Linear Regression:
+\lstinputlisting[language=Python]{src/models/bayesian_linear_regression.py}
+
+\newpage
+The Python code for kernels:
+\lstinputlisting[language=Python]{src/models/kernels.py}
+
+\newpage
+The Python code for Gaussian Process Regression:
+\lstinputlisting[language=Python]{src/models/gaussian_process_regression.py}
+
+\newpage
+The rest of the Python code for question 2:
 \lstinputlisting[language=Python]{src/solutions/q2.py}
@@
 
-\[ \log P(\textbf{x} | \textbf{s}, \theta)  = -\frac{d}{2} \log (2 \pi) - \log(\sigma)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} s_i \mu_i   + \sum_{i=1}^{K} \sum_{i=1}^{K} s_i s_j \mu_i^T\mu_j \right) \]
+\[ \log P(\textbf{x} | \textbf{s}, \theta)  = -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} s_i \mu_i   + \sum_{i=1}^{K} \sum_{i=1}^{K} s_i s_j \mu_i^T\mu_j \right) \]
 
@@
 \langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} \\
-= -\frac{d}{2} \log (2 \pi) - \log(\sigma)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} \langle s_i \rangle_{q_{i} (s_i)} \mu_i   + \sum_{i=1}^{K} \sum_{j=1}^{K} \langle s_i s_j \rangle_{q_{i} (s_i) q_{j} (s_j)} \mu_i^T\mu_j \right)\]
+= -\frac{d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} \langle s_i \rangle_{q_{i} (s_i)} \mu_i   + \sum_{i=1}^{K} \sum_{j=1}^{K} \langle s_i s_j \rangle_{q_{i} (s_i) q_{j} (s_j)} \mu_i^T\mu_j \right)\]
 
@@
 \langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} \\
-= -\frac{d}{2} \log (2 \pi) - \log(\sigma)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right)\]
+= -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right)\]
 
@@
 \tab \mathcal{F}(q, \theta) = \\
-\tab \tab \frac{-d}{2} \log (2 \pi) - \log(\sigma)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right) \\
+\tab \tab \frac{-d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right) \\
 \tab \tab + \sum_{i=1}^{K} \lambda_i \log \pi_i + (1-\lambda_i) \log (1-\pi_i)}\\

### 2022-12-25 main.py

@@ -1,23 +1,30 @@
 import os
-import pandas as pd
-import numpy as np
-from src.constants import CO2_FILE_PATH, IMAGES_FILE_PATH, OUTPUTS_FOLDER
-from src.solutions import q2, q3, q4, q5, q6
-from src.solutions.q2 import LinearRegressionParameters
-from src.generate_images import generate_images
 
+import jax
+import jax.numpy as jnp
+import numpy as np
+import pandas as pd
+
+from src.constants import CO2_FILE_PATH, IMAGES_FILE_PATH, OUTPUTS_FOLDER
+from src.generate_images import generate_images
+from src.models.bayesian_linear_regression import LinearRegressionParameters
+from src.models.kernels import CombinedKernel, CombinedKernelParameters
+from src.models.gaussian_process_regression import GaussianProcessParameters
+from src.solutions import q2, q3, q4, q5, q6
+from dataclasses import asdict
+jax.config.update("jax_enable_x64", True)
 
 if __name__ == "__main__":
     if not os.path.exists(OUTPUTS_FOLDER):
         os.makedirs(OUTPUTS_FOLDER)
 
-    with open(CO2_FILE_PATH) as file:
-        lines = [line.rstrip().split() for line in file]
-
     # Question 2
     Q2_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q2")
     if not os.path.exists(Q2_OUTPUT_FOLDER):
         os.makedirs(Q2_OUTPUT_FOLDER)
+    with open(CO2_FILE_PATH) as file:
+        lines = [line.rstrip().split() for line in file]
+
     df_co2 = pd.DataFrame(
         np.array([line for line in lines if line[0] != "#"]).astype(float)
     )
@@ -33,6 +40,15 @@ if __name__ == "__main__":
             [10**2, 0],
             [0, 100**2],
         ]
+    )
+    kernel = CombinedKernel()
+    kernel_parameters = CombinedKernelParameters(
+        log_theta=jnp.log(1),
+        log_sigma=jnp.log(1),
+        log_phi=jnp.log(1),
+        log_eta=jnp.log(1),
+        log_tau=jnp.log(1),
+        log_zeta=jnp.log(1e-1),
     )
 
     prior_linear_regression_parameters = LinearRegressionParameters(
[…]

### 2022-12-25 src/generate_images.py

@@ -1,5 +1,6 @@
 import numpy as np
-from src.constants import M1, M2, M3, M4, M5, M6, M7, M8, DEFAULT_SEED
+
+from src.constants import DEFAULT_SEED, M1, M2, M3, M4, M5, M6, M7, M8
 
 
 def generate_images(n: int = 400, seed: int = DEFAULT_SEED):

### 2022-12-25 src/models/gaussian_process_regression.py

@@ -1,0 +1,177 @@
+from dataclasses import dataclass
+from typing import Any, Dict, Tuple
+
+import jax
+import jax.numpy as jnp
+import optax
+from jax import grad
+from optax import GradientTransformation
+
+from src.models.kernels import Kernel
+
+
+@dataclass
+class GaussianProcessParameters:
+    """
+    Parameters for a Gaussian Process:
+        log_sigma: logarithm of the noise parameter
+        kernel: parameters for the chosen kernel
+    """
+
+    log_sigma: float
+    kernel: Dict[str, Any]
+
+    @property
+    def variance(self) -> float:
+        return self.sigma**2
+
+    @property
+    def sigma(self) -> float:
+        return jnp.exp(self.log_sigma)
+
+    @sigma.setter
+    def sigma(self, value: float) -> None:
+        self.log_sigma = jnp.log(value)
+
+
+class GaussianProcess:
+    """
+    A Gaussian measure defined with a kernel, better known as a Gaussian Process.
+    """
+
+    Parameters = GaussianProcessParameters
+
+    def __init__(self, kernel: Kernel, x: jnp.ndarray, y: jnp.ndarray) -> None:
+        """Initialising requires a kernel and data to condition the distribution.
+
+        Args:
+            kernel: kernel for the Gaussian Process
+            x: design matrix (number_of_features, number_of_dimensions)
+            y: response vector (number_of_features, )
+        """
+        self.number_of_train_points = x.shape[0]
+        self.x = x
+        self.y = y
+        self.kernel = kernel
+
+    def _compute_kxx_shifted_cholesky_decomposition(
+        self, parameters
+    ) -> Tuple[jnp.ndarray, bool]:
+        """
+        Cholesky decomposition of (kxx + (1/σ^2)*I)
+
+        Args:
+            parameters: parameters dataclass for the Gaussian Process
+
+        Returns:
+            cholesky_decomposition_kxx_shifted: the cholesky decomposition (number_of_features, number_of_features)
+            lower_flag: flag indicating whether the factor is in the lower or upper triangle
+        """
+        kxx = self.kernel(self.x, **parameters.kernel)
+        kxx_shifted = kxx + parameters.variance * jnp.eye(self.number_of_train_points)
+        kxx_shifted_cholesky_decomposition, lower_flag = jax.scipy.linalg.cho_factor(
+            a=kxx_shifted, lower=True
+        )
[…]

### 2022-12-25 src/models/kernels.py

@@ -1,0 +1,253 @@
+from abc import ABC, abstractmethod
+from dataclasses import dataclass
+
+import jax.numpy as jnp
+from jax import vmap
+
+
+@dataclass
+class KernelParameters(ABC):
+    """
+    An abstract dataclass containing the parameters for a kernel.
+    """
+
+
+class Kernel(ABC):
+    """
+    An abstract kernel.
+    """
+
+    Parameters: KernelParameters = None
+
+    @abstractmethod
+    def _kernel(
+        self, parameters: KernelParameters, x: jnp.ndarray, y: jnp.ndarray
+    ) -> jnp.ndarray:
+        """Kernel evaluation between a single feature x and a single feature y.
+
+        Args:
+            parameters: parameters dataclass for the kernel
+            x: ndarray of shape (number_of_dimensions,)
+            y: ndarray of shape (number_of_dimensions,)
+
+        Returns:
+            The kernel evaluation. (1, 1)
+        """
+        raise NotImplementedError
+
+    def kernel(
+        self, parameters: KernelParameters, x: jnp.ndarray, y: jnp.ndarray = None
+    ) -> jnp.ndarray:
+        """Kernel evaluation for an arbitrary number of x features and y features. Compute k(x, x) if y is None.
+        This method requires the parameters dataclass and is better suited for parameter optimisation.
+
+        Args:
+            parameters: parameters dataclass for the kernel
+            x: ndarray of shape (number_of_x_features, number_of_dimensions)
+            y: ndarray of shape (number_of_y_features, number_of_dimensions)
+
+        Returns:
+            A gram matrix k(x, y), if y is None then k(x,x). (number_of_x_features, number_of_y_features)
+        """
+        # compute k(x, x) if y is None
+        if y is None:
+            y = x
+
+        # add dimension when x is 1D, assume the vector is a single feature
+        x = jnp.atleast_2d(x)
+        y = jnp.atleast_2d(y)
+
+        assert (
+            x.shape[1] == y.shape[1]
+        ), f"Dimension Mismatch: {x.shape[1]=} != {y.shape[1]=}"
+
+        return vmap(
+            lambda x_i: vmap(
+                lambda y_i: self._kernel(parameters, x_i, y_i),
+            )(y),
+        )(x)
+
+    def __call__(
+        self, x: jnp.ndarray, y: jnp.ndarray = None, **parameter_args
+    ) -> jnp.ndarray:
[…]

### 2022-12-25 src/models/mean_field_learning.py

@@ -1,0 +1,324 @@
+from dataclasses import dataclass
+
+import numpy as np
+
+from demo_code.MStep import m_step
+
+
+@dataclass
+class MeanFieldApproximation:
+    """
+    lambda_matrix: parameters variational approximation (number_of_points, number_of_latent_variables)
+    """
+
+    lambda_matrix: np.ndarray
+
+    def lambda_matrix_exclude(self, exclude_latent_index: int) -> np.ndarray:
+        #  (number_of_points, number_of_latent_variables-1)
+        return np.concatenate(
+            (
+                self.lambda_matrix[:, :exclude_latent_index],
+                self.lambda_matrix[:, exclude_latent_index + 1 :],
+            ),
+            axis=1,
+        )
+
+    @property
+    def log_lambda_matrix(self):
+        return np.log(self.lambda_matrix)
+
+    @property
+    def log_one_minus_lambda_matrix(self):
+        return np.log(1 - self.lambda_matrix)
+
+    @property
+    def n(self):
+        return self.lambda_matrix.shape[0]
+
+    @property
+    def k(self):
+        return self.lambda_matrix.shape[1]
+
+
+def init_mean_field_approximation(k: int, n: int) -> MeanFieldApproximation:
+    return MeanFieldApproximation(
+        lambda_matrix=np.random.random(size=(n, k)),
+    )
+
+
+@dataclass
+class BinaryLatentFactorModel:
+    """
+    mu: matrix of means (number_of_dimensions, number_of_latent_variables)
+    sigma: gaussian noise parameter
+    pi: vector of priors (1, number_of_latent_variables)
+    """
+
+    mu: np.ndarray
+    sigma: float
+    pi: np.ndarray
+
+    def mu_exclude(self, exclude_latent_index: int) -> np.ndarray:
+        #  (number_of_dimensions, number_of_latent_variables-1)
+        return np.concatenate(
+            (self.mu[:, :exclude_latent_index], self.mu[:, exclude_latent_index + 1 :]),
+            axis=1,
+        )
+
+    @property
+    def log_pi(self):
+        return np.log(self.pi)
+
+    @property
+    def log_one_minus_pi(self):
+        return np.log(1 - self.pi)
+
+    @property
+    def variance(self):
+        return self.sigma**2
+
+    @property
+    def precision(self):
+        return 1 / self.variance
+
+    @property
+    def d(self):
+        return self.mu.shape[0]
+
+    @property
+    def k(self):
+        return self.mu.shape[1]
+
+
[…]

### 2022-12-25 src/solutions/q2.py

@@ -1,64 +1,25 @@
-import scipy
-import numpy as np
-from dataclasses import dataclass
-import pandas as pd
+from dataclasses import asdict, fields
+import optax
 import dataframe_image as dfi
+import jax
+import jax.numpy as jnp
 import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import scipy
 
+from src.models.bayesian_linear_regression import (
+    LinearRegressionParameters,
+    Theta,
+    compute_linear_regression_posterior,
+)
+from src.models.gaussian_process_regression import (
+    GaussianProcess,
+    GaussianProcessParameters,
+)
+from src.models.kernels import CombinedKernel, CombinedKernelParameters
 
-@dataclass
-class LinearRegressionParameters:
-    mean: np.ndarray
-    covariance: np.ndarray
-
-    @property
-    def precision(self):
-        return np.linalg.inv(self.covariance)
-
-    def predict(self, x: np.ndarray) -> np.ndarray:
-        return self.mean.T @ x
-
-
-@dataclass
-class Theta:
-    linear_regression_parameters: LinearRegressionParameters
-    sigma: float
-
-    @property
-    def variance(self):
-        return self.sigma**2
-
-    @property
-    def precision(self):
-        return 1 / self.variance
-
-
-def compute_posterior(
-    x: np.ndarray,
-    y: np.ndarray,
-    prior_linear_regression_parameters: LinearRegressionParameters,
-    residuals_precision: float,
-) -> LinearRegressionParameters:
-    """
-    Compute the parameters of the posterior distribution on the linear regression weights
-
-    :param x: design matrix (number of features, number of data points)
-    :param y: response matrix (1, number of data points)
-    :param prior_linear_regression_parameters: parameters for the prior distribution on the linear regression weights
-    :param residuals_precision: the precision of the residuals of the linear regression
-    :return: parameters for the posterior distribution on the linear regression weights
-    """
-    posterior_covariance = np.linalg.inv(
-        residuals_precision * x @ x.T + prior_linear_regression_parameters.precision
-    )
-    posterior_mean = posterior_covariance @ (
-        residuals_precision * x @ y.T
-        + prior_linear_regression_parameters.precision
-        @ prior_linear_regression_parameters.mean
-    )
[…]

### 2022-12-25 src/solutions/q3.py

@@ -1,322 +1,0 @@
-from dataclasses import dataclass
-
-import numpy as np
-
-from demo_code.MStep import m_step
-
-
-@dataclass
-class MeanFieldApproximation:
-    """
-    lambda_matrix: parameters variational approximation (number_of_points, number_of_latent_variables)
-    """
-
-    lambda_matrix: np.ndarray
-
-    def lambda_matrix_exclude(self, exclude_latent_index: int) -> np.ndarray:
-        #  (number_of_points, number_of_latent_variables-1)
-        return np.concatenate(
-            (
-                self.lambda_matrix[:, :exclude_latent_index],
-                self.lambda_matrix[:, exclude_latent_index + 1 :],
-            ),
-            axis=1,
-        )
-
-    @property
-    def log_lambda_matrix(self):
-        return np.log(self.lambda_matrix)
-
-    @property
-    def log_one_minus_lambda_matrix(self):
-        return np.log(1 - self.lambda_matrix)
-
-    @property
-    def n(self):
-        return self.lambda_matrix.shape[0]
-
-    @property
-    def k(self):
-        return self.lambda_matrix.shape[1]
-
-
-def init_mean_field_approximation(k: int, n: int) -> MeanFieldApproximation:
-    return MeanFieldApproximation(
-        lambda_matrix=np.random.random(size=(n, k)),
-    )
-
-
-@dataclass
-class BinaryLatentFactorModel:
-    """
-    mu: matrix of means (number_of_dimensions, number_of_latent_variables)
-    sigma: gaussian noise parameter
-    pi: vector of priors (1, number_of_latent_variables)
-    """
-
-    mu: np.ndarray
-    sigma: float
-    pi: np.ndarray
-
-    def mu_exclude(self, exclude_latent_index: int) -> np.ndarray:
-        #  (number_of_dimensions, number_of_latent_variables-1)
-        return np.concatenate(
-            (self.mu[:, :exclude_latent_index], self.mu[:, exclude_latent_index + 1 :]),
-            axis=1,
-        )
-
-    @property
-    def log_pi(self):
-        return np.log(self.pi)
-
-    @property
-    def log_one_minus_pi(self):
-        return np.log(1 - self.pi)
-
-    @property
-    def variance(self):
-        return self.sigma**2
-
-    @property
-    def precision(self):
-        return 1 / self.variance
-
-    @property
-    def d(self):
-        return self.mu.shape[0]
-
-    @property
-    def k(self):
-        return self.mu.shape[1]
-
-
[…]

### 2022-12-25 src/solutions/q2.py
add untrained and trained hyperparameter gp extrapolations

@@ -188,10 +188,36 @@ def f(
         residuals_precision=prior_theta.precision,
     )
 
-    # Train Gaussian Process Regression (Hyperparameter Tune)
     residuals = y_train - posterior_linear_regression_parameters.predict(x_train)
     gaussian_process = GaussianProcess(kernel, t_train.reshape(-1, 1), residuals.reshape(-1))
 
+    # Prediction
+    x_test = construct_design_matrix(t_test)
+    linear_prediction = posterior_linear_regression_parameters.predict(x_test).reshape(-1)
+    mean_prediction, covariance_prediction = gaussian_process.posterior_distribution(
+        t_test.reshape(-1, 1), **asdict(gaussian_process_parameters)
+    )
+
+    # Plot
+    plt.figure(figsize=(7, 7))
+    plt.scatter(t_train+min_year, y_train.reshape(-1), s=2, color='blue', label="historical data")
+    plt.plot(t_test+min_year, linear_prediction + mean_prediction, color="gray", label="prediction")
+    plt.fill_between(
+        t_test+min_year,
+        linear_prediction+mean_prediction-1*jnp.diagonal(covariance_prediction),
+        linear_prediction+mean_prediction+1*jnp.diagonal(covariance_prediction),
+        facecolor=(0.8, 0.8, 0.8),
+        label="error bound (one stdev)"
+    )
+    plt.xlabel("date (decimal year)")
+    plt.ylabel("parts per million")
+    plt.title("Global Mean CO_2 Concentration Prediction (Untrained Hyperparameters)")
+    plt.legend()
+    plt.tight_layout()
+    plt.savefig(save_path+"-extrapolation-untrained", bbox_inches='tight')
+    plt.close()
+
+    # Train Gaussian Process Regression (Hyperparameter Tune)
     optimizer = optax.adam(learning_rate)
     gaussian_process_parameters = gaussian_process.train(
         optimizer, number_of_iterations, **asdict(gaussian_process_parameters)
@@ -205,7 +231,7 @@ def f(
     )
 
     # Plot
-    plt.figure(figsize=(10, 10))
+    plt.figure(figsize=(7, 7))
     plt.scatter(t_train+min_year, y_train.reshape(-1), s=2, color='blue', label="historical data")
     plt.plot(t_test+min_year, linear_prediction + mean_prediction, color="gray", label="prediction")
     plt.fill_between(
@@ -217,6 +243,8 @@ def f(
     )
[…]

### 2022-12-27 submission.tex
add untrained and trained hyperparameter gp extrapolations
update EP and mean field questions

@@
 \includegraphics[scale = 0.5]{outputs/q2/a-covariance}
-\caption{t=The Posterior Covariance}
+\caption{The Posterior Covariance}
 \label{fig:fig2-a-covariance}
@@
 The density estimation shows that a mean of zero is a reasonable prior belief however the data does not seem to exhibit unit variance.
+Also we know it's not iid because timeseries.
 
@@
 \includegraphics[scale=0.5]{outputs/q2/c-gram-matrix}
-\caption{Gram Matrix}
+\caption{Covariance Matrix}
 \label{fig:c-gram-matrix}
@@
 \subsection*{(f)}
+
 \begin{figure}[h]
 \centering
-\includegraphics[scale=0.5]{outputs/q2/f-extrapolation}
-\caption{Extrapolation of $CO_2$ concentration levels}
-\label{fig:f-extrapolation}
+\begin{minipage}{.5\textwidth}
+  \centering
+\includegraphics[scale=0.5]{outputs/q2/f-extrapolation-untrained}
+\caption{Without hyperparameter tuning}
+\label{fig:f-extrapolation-untrained}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+\includegraphics[scale=0.5]{outputs/q2/f-extrapolation-trained}
+\caption{With hyperparameter tuning}
+\label{fig:f-extrapolation-trained}
+\end{minipage}
 \end{figure}
+%
+%\begin{figure}[h]
+%\centering
+%\includegraphics[scale=0.5]{outputs/q2/f-extrapolation-untrained}
+%\caption{Extrapolation of $CO_2$ concentration levels}
+%\label{fig:f-extrapolation-untrained}
+%\end{figure}
+%
+%\begin{figure}[h]
+%\centering
+%\includegraphics[scale=0.5]{outputs/q2/f-extrapolation-trained}
+%\caption{Extrapolation of $CO_2$ concentration levels}
+%\label{fig:f-extrapolation-trained}
+%\end{figure}
 
@@
 \newpage
-The Python code:
+
+\subsection*{(b)}
+
+The provided derivations for the M step of the mean parameter $\mu$:
+
+\[\mu = \left( \langle \textbf{s}\textbf{s}^T \rangle_{q(\textbf{s})} \right)^{-1}\langle \textbf{s}\rangle_{q(\textbf{s})} \textbf{x}\]
+
+where $\mu \in \mathbb{R}^{K \times D}$, \textbf{s} \in \mathbb{R}^{K \times N}$, and $\textbf{x} \in \mathbb{R}^{N \times D}$.
+
+This mimics the least squares solution:
+
+\[\hat{\beta} = (\textbf{X}\textbf{X}^T) \textbf{X} \textbf{Y}\]
+
+for the linear regression problem $\textbf{Y} = \textbf{X}^T \beta$ where $\beta$ corresponds to the mean parameters $\mu$, the design matrix $\textbf{X}$ corresponds to the input $\textbf{s}$ and the response $Y$ corresponds to the image pixels denoted $\textbf{x}$.
+This makes sense because our resulting images $\textbf{x}$ are modeled as linear combinations of features $\mu$, weighted by $\textbf{s}$.
+
+\subsection*{(c)}
+
+The computational complexity of the implemented M step function can be broken down for each parameter:
+
+\begin{itemize}
+\item [$\mu$:]\begin{itemize}
+\item The inversion $\text{ESS}^{-1}$ where $\text{ESS} \in \mathbb{R}^{K \times K}$ is $\mathcal{O}(K^3)$
+\item The dot product $\text{ESS}^{-1} \text{ES}^T$ where $\text{ESS}^{-1} \in \mathbb{R}^{K \times K}$ and $\text{ES} \in \mathbb{R}^{N \times K}$ is  $\mathcal{O}(K^2N)$
+\item The dot product $(\text{ESS}^{-1} \text{ES}^T)\textbf{x}$ where $(\text{ESS}^{-1} \text{ES}^T)\ \in \mathbb{R}^{K \times N}$ and $\textbf{x} \in \mathbb{R}^{N \times D}$ is  $\mathcal{O}(KND )$
+\end{itemize}
+
+\item [$\sigma$:] \begin{itemize}
+\item The dot product $(\textbf{x}^{T} \textbf{x})$ where $\textbf{x} \in \mathbb{R}^{N \times D}$ is  $\mathcal{O}(D^2N)$
+\item The dot product $\mu^{T} \mu$ where $\mu \in \mathbb{R}^{D \times K}$ is  $\mathcal{O}(K^2D)$
+\item The dot product $(\mu^{T} \mu)\text{ESS}$ where $\mu^{T} \mu \in \mathbb{R}^{K \times K}$ and $\text{ESS} \in \mathbb{R}^{K \times K}$ is $\mathcal{O}(K^3)$
+\end{itemize}
+\item [$\pi$:] \begin{itemize}
+    \item The mean operation for $\text{ES} \in \mathbb{R}^{N \times K}$ along the first dimension is $\mathcal{O}(NK)$
+    \end{itemize}
+\end{itemize}
+
+Thus, the computational complexity of the M step is $\mathcal{O}(K^3+K^2N + KND + D^2N + K^2D)$ where we do not assume that any of $N$, $K$, or $D$ is large compared to the others.
+\newpage
+\subsection*{(d)}
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.4]{data/images}
+\caption{Images generated by randomly combined features with noise}
+\label{fig:3d-images}
+\end{figure}
+
+\subsection*{(e)}
+
+We can plot the free energy to make sure it increases each iteration:
+
+\subsection*{(f)}
+
+\subsection*{(g)}
+
+\newpage
+The Python code for mean field learning:
+\lstinputlisting[language=Python]{src/models/mean_field_learning.py}
+
+\newpage
+The rest of the Python code for question 3:
 \lstinputlisting[language=Python]{src/solutions/q3.py}
@@
 
-\[q(\textbf{s}) = \left(\prod_{i=1}^{K} \tilde{f}_i(s_i) \right) \left(\prod_{i=1}^{K}  \prod_{j=1}^{i-1} \tilde{g}_{ij} (s_i, s_j) \right)\]
+\[q(\textbf{s}) = \prod_{i'=1}^{K} \left( \tilde{f}_{i'}(s_{i'})
+\prod_{j\in ne(i)}^{K} \mathcal{M}_{j \rightarrow i}
+\right)
+\]
 
-Thus, we can derive cavity distributions:
+where $ne(i)$ are indices of neighbouring nodes of node $i$ and $\mathcal{M}_{j \rightarrow i}$ is the message from node $j$ to $i$.
 
-\[q_{\neg \tilde{f}_i(s_i)}(\textbf{s}) = \left(\prod_{i'=1, i'\neq i}^{K} \tilde{f}_{i'}(s_{i'}) \right) \left(\prod_{i=1}^{K}  \prod_{j=1}^{i-1} \tilde{g}_{ij} (s_i, s_j) \right)\]
+Thus, the cavity distributions:
+
+\[q_{\neg \tilde{f}_i(s_i)}(\textbf{s}) =
+\prod_{j\in ne(i)}^{K} \mathcal{M}_{j \rightarrow i}
+\]
 
@@
 
-\[q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) = \left(\prod_{i'=1}^{K} \tilde{f}_{i'}(s_{i'}) \right) \left(\prod_{i'=1}^{K}  \prod_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1} \tilde{g}_{i'j'} (s_{i'}, s_{j'}) \right)\]
+\[q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) = \left( \tilde{f}_{i}(s_{i}) \prod_{k\in ne(i), k\neq j}^{K} \mathcal{M}_{k \rightarrow i}
+\right)\left( \tilde{f}_{j}(s_{j}) \prod_{k\in ne(j), k\neq j}^{K} \mathcal{M}_{k \rightarrow j}
+\right)
+\]
 
@@
 
-\[\tilde{f}_{i}(s_{i}) = \arg \min \textbf{KL} \left[ f_{i}(s_{i}) q_{\neg \tilde{f}_i(s_i)}(\textbf{s}) \| \tilde{f}_{i}(s_{i}) q_{\neg \tilde{f}_i(s_i)}(\textbf{s}) \right]\]
+\[\tilde{f}_{i}(s_{i}) = \arg \min_{\tilde{f}_{i}(s_{i})} \textbf{KL} \left[ f_{i}(s_{i}) q_{\neg \tilde{f}_i(s_i)}(\textbf{s}) \| \tilde{f}_{i}(s_{i}) q_{\neg \tilde{f}_i(s_i)}(\textbf{s}) \right]\]
 
-We know that the factor $\log f_i(s_i)$ is a Bernoulli of the form $b_i s_i$. Because our approximation is also Bernoulli, we can simply solve for $\lambda_i$ in $\log \tilde{f}_{i}(s_{i})$:
+We know that the factor $\log f_i(s_i)$ is a Bernoulli of the form $b_i s_i$.
+Because our approximation for this site is also Bernoulli, we can simply solve for $\lambda_i$ in $\log \tilde{f}_{i}(s_{i})$:
 
@@
 
-\[\tilde{g}_{ij}(s_i, s_j) = \arg \min \textbf{KL} \left[ g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \| \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \right]\]
+\[\tilde{g}_{ij}(s_i, s_j) = \arg \min_{\tilde{f}_{i}(s_{i})} \textbf{KL} \left[ g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \| \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \right]\]
 
-Note that because $\tilde{g}_{ij}(s_i, s_j)$ is the product of two Bernoulli distributions, we only require the natural parameters:
+Defining the natural parameters $\eta_{ij}$ and $\eta_{ji}$ for the messages:
+
+\[\mathcal{M}_{i \rightarrow j} = \exp(\eta_{ij} s_i) = \tilde{f}_{i}(s_{i}) \prod_{k\in ne(i), k\neq j}^{K} \mathcal{M}_{k \rightarrow i} \]
+
+\[\mathcal{M}_{j \rightarrow i}  = \exp(\eta_{ji} s_j) = \tilde{f}_{j}(s_{j}) \prod_{k\in ne(j), k\neq i}^{K} \mathcal{M}_{k \rightarrow j}\]
+
+Note that because $\tilde{g}_{ij}(s_i, s_j)$ is the product of two Bernoulli distributions, updates to this site approximation involves updating to the natural parameters:
 
@@
 
-the mean with respect to $s_i$ and $s_j$ respectively.
+the mean of $s_i$ and $s_j$ respectively.
 
@@
 
-\[
-\begin{array}{l}
- \log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j\\
-\tab \tab \tab \tab \tab + \sum_{i'=1}^{K}  \log \left(\frac{\lambda_{i'}}{1-\lambda_{i'}} \right) s_{i'} \\
-\tab \tab \tab \tab \tab +  \sum_{i'=1}^{K}\sum_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1}  \log \left(\frac{\theta_{i'}}{1-\theta_{i'}} \right) s_{i'} + \log \left(\frac{\theta_{j'}}{1-\theta_{j'}} \right) s_{j'}
-\end{array}
-\]
+\[\log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j + \eta_{ij} s_i + \eta_{ji} s_j\]
 
-Only including terms with $s_i$ and $s_j$:
+Simplifying:
+
+\[\log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \log \left(\frac{\theta_i}{1-\theta_i} + \eta_{ij} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} + \eta_{ji} \right) s_j \]
 
 %\[
-% \log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \left(K \log \left(\frac{\theta_i}{1-\theta_i} \right) + \log \left(\frac{\lambda_i}{1-\lambda_i} \right)\right) s_i +   \left(
-%K \log \left(\frac{\theta_j}{1-\theta_j} \right) + \log \left(\frac{\lambda_j}{1-\lambda_j} \right)\right) s_j
+%\begin{array}{l}
+% \log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j\\
+%\tab \tab \tab \tab \tab + \sum_{i'=1}^{K}  \log \left(\frac{\lambda_{i'}}{1-\lambda_{i'}} \right) s_{i'} \\
+%\tab \tab \tab \tab \tab +  \sum_{i'=1}^{K}\sum_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1}  \log \left(\frac{\theta_{i'}}{1-\theta_{i'}} \right) s_{i'} + \log \left(\frac{\theta_{j'}}{1-\theta_{j'}} \right) s_{j'}
+%\end{array}
 %\]
 
-\[
-\begin{array}{l}
- \log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \left((K-1) \log \left(\frac{\theta_i}{1-\theta_i} \right) + \log \left(\frac{\lambda_i}{1-\lambda_i} \right)\right) s_i\\
-\tab \tab \tab \tab \tab +\left(
-(K-1) \log \left(\frac{\theta_j}{1-\theta_j} \right) + \log \left(\frac{\lambda_j}{1-\lambda_j} \right)\right) s_j
-\end{array}
-\]
+%Only including terms with $s_i$ and $s_j$:
+%
+%\[
+%\begin{array}{l}
+% \log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \left((K-1) \log \left(\frac{\theta_i}{1-\theta_i} \right) + \log \left(\frac{\lambda_i}{1-\lambda_i} \right)\right) s_i\\
+%\tab \tab \tab \tab \tab +\left(
+%(K-1) \log \left(\frac{\theta_j}{1-\theta_j} \right) + \log \left(\frac{\lambda_j}{1-\lambda_j} \right)\right) s_j
+%\end{array}
+%\]
+
+Thus, the first moments:
+
+\[\mathbb{E}_{s_i}[\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \frac{1}{1+\exp{\left(-\left(\theta_i} + \eta_{ij}\right)}\right)}\]
+
+and
+
+\[\mathbb{E}_{s_j}[\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \frac{1}{1+\exp{\left(-\left(\theta_j + \eta_{ji}\right)}\right)}\]
 
@@
 
-\[
-\begin{array}{l}
- \log g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \frac{- \mu_i^T \mu_j}{\sigma^2} s_i s_j\\
-\tab \tab \tab \tab \tab + \sum_{i'=1}^{K}  \log \left(\frac{\lambda_{i'}}{1-\lambda_{i'}} \right) s_{i'} \\
-\tab \tab \tab \tab \tab +  \sum_{i'=1}^{K}\sum_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1}  \log \left(\frac{\theta_{i'}}{1-\theta_{i'}} \right) s_{i'} + \log \left(\frac{\theta_{j'}}{1-\theta_{j'}} \right) s_{j'}
-\end{array}
+\[\log g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto W_{ij} s_i s_j
+ + \eta_{ij} s_i + \eta_{ji} s_j\]
+
+%\[
+%\begin{array}{l}
+% \log g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \frac{- \mu_i^T \mu_j}{\sigma^2} s_i s_j\\
+%\tab \tab \tab \tab \tab + \sum_{i'=1}^{K}  \log \left(\frac{\lambda_{i'}}{1-\lambda_{i'}} \right) s_{i'} \\
+%\tab \tab \tab \tab \tab +  \sum_{i'=1}^{K}\sum_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1}  \log \left(\frac{\theta_{i'}}{1-\theta_{i'}} \right) s_{i'} + \log \left(\frac{\theta_{j'}}{1-\theta_{j'}} \right) s_{j'}
+%\end{array}
+%\]
+
+%Only including terms with $s_i$ and $s_j$:
+%
+%\[
+%\begin{array}{l}
+% \log g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \frac{- \mu_i^T \mu_j}{\sigma^2} s_i s_j\\
+%\tab \tab \tab \tab \tab + \left((K-2)\log \left(\frac{\theta_i}{1-\theta_i} \right) + \log \left(\frac{\lambda_i}{1-\lambda_i} \right)\right) s_i \\
+%\tab \tab \tab \tab \tab +  \left((K-2)\log \left(\frac{\theta_j}{1-\theta_j} \right) + \log \left(\frac{\lambda_j}{1-\lambda_j} \right)\right) s_j
+%\end{array}
+%\]
+%
+%We can write this as:
+%
+%\[
+%\log g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto W_{ij} s_i s_j+ A_{s_i} s_i + B_{s_j} s_j
+%\]
+%
+%where:
+%
+%\[A_{s_i} = (K-2)\log \left(\frac{\theta_i}{1-\theta_i} \right) + \log \left(\frac{\lambda_i}{1-\lambda_i} \right)\]
+%
+%and
+%
+%\[B_{s_j} = (K-2)\log \left(\frac{\theta_j}{1-\theta_j} \right) + \log \left(\frac{\lambda_j}{1-\lambda_j} \right)\]
+
+To derive the first moment for $ g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})$ with respect to $s_i$, we first marginalise out $s_j$:
+
+\[ \sum_{s_j \in \{0, 1\}} g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \exp\left( W_{ij} s_i + \eta_{ij} s_i + \eta_{ji}\right) +  \exp\left(\eta_{ij} s_i \right)
 \]
 
-Only including terms with $s_i$ and $s_j$:
+Thus, the first moment:
 
-\[
-\begin{array}{l}
- \log g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \frac{- \mu_i^T \mu_j}{\sigma^2} s_i s_j\\
-\tab \tab \tab \tab \tab + \left((K-2)\log \left(\frac{\theta_i}{1-\theta_i} \right) + \log \left(\frac{\lambda_i}{1-\lambda_i} \right)\right) s_i \\
-\tab \tab \tab \tab \tab +  \left((K-2)\log \left(\frac{\theta_j}{1-\theta_j} \right) + \log \left(\frac{\lambda_j}{1-\lambda_j} \right)\right) s_j
-\end{array}
-\]
+\[\mathbb{E}_{s_i}[g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \frac{\exp\left( W_{ij} + \eta_{ij} + \eta_{ji}\right) +  \exp\left(\eta_{ij} \right)}{[\exp\left( W_{ij} + \eta_{ij} + \eta_{ji}\right) +  \exp\left(\eta_{ij} \right)]+[\exp\left( \eta_{ji}\right) +  1]}\]
+
+Simplifying:
+
+\[\mathbb{E}_{s_i}[g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \frac{\left( \exp\left( W_{ij} + \eta_{ji}\right) +  1 \right)\exp\left(\eta_{ij} \right)}{[\left( \exp\left( W_{ij} + \eta_{ji}\right) +  1 \right)\exp\left(\eta_{ij} \right)]+[\exp\left( \eta_{ji}\right) +  1]}\]
+
+similarly:
+
+\[\mathbb{E}_{s_j}[g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \frac{\left( \exp\left( W_{ij} + \eta_{ij}\right) +  1 \right)\exp\left(\eta_{ji} \right)}{[\left( \exp\left( W_{ij} + \eta_{ij}\right) +  1 \right)\exp\left(\eta_{ji} \right)]+[\exp\left( \eta_{ij}\right) +  1]}\]
+
+By setting:
+
+\[\mathbb{E}_{s_i}[\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \mathbb{E}_{s_i}[g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})]\]
+
+and
+
+\[\mathbb{E}_{s_j}[\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \mathbb{E}_{s_j}[g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})]\]
+
+we can solve for the parameters of $\tilde{g}_{ij}(s_i, s_j)$ with moment matching:
+
+\[\frac{1}{1+\exp{\left(-\left(\theta_i} + \eta_{ij}\right)}\right)}} = \frac{\left( \exp\left( W_{ij} + \eta_{ji}\right) +  1 \right)\exp\left(\eta_{ij} \right)}{[\left( \exp\left( W_{ij} + \eta_{ji}\right) +  1 \right)\exp\left(\eta_{ij} \right)]+[\exp\left( \eta_{ji}\right) +  1]}\]
+
+\[\exp\left( \eta_{ji}\right) +  1 =\exp{\left(-\left(\theta_i + \eta_{ij}\right)}\right)}\left( \exp\left( W_{ij} + \eta_{ji}\right) +  1 \right)\exp\left(\eta_{ij} \right)\]
+
+\[\frac{\exp\left( \eta_{ji}\right) +  1}{ \exp\left( W_{ij} + \eta_{ji}\right) +  1 } =\exp{\left(-\theta_i\right)}\]
+
+\[\theta_i = \log \left( \frac{ 1+\exp\left( W_{ij} + \eta_{ji}\right) }{1+\exp\left( \eta_{ji}\right)} \right)\]
+
+Similarly:
+
+\[\theta_j = \log \left( \frac{ 1+ \exp\left( W_{ij} + \eta_{ij}\right)  }{1+\exp\left( \eta_{ij}\right)} \right)\]
+
+\newpage
+
+\subsection*{(c)}
+
+Our message passing approximations:
+
+\[\exp(\eta_{ij} s_i) = \tilde{f}_{i}(s_{i}) \prod_{k\in ne(i), k\neq j}^{K} \mathcal{M}_{k \rightarrow i} \]
+
+\[\exp(\eta_{ji} s_j) = \tilde{f}_{j}(s_{j}) \prod_{k\in ne(j), k\neq i}^{K} \mathcal{M}_{k \rightarrow j}\]
+
+where each message $\mathcal{M}_{j \rightarrow i}$ has a factored approximation:
+
+\[\mathcal{M}_{k \rightarrow i} = \exp(\eta_{ki} s_k) \]
+
+because each site $\tilde{g}_{jk}(s_j s_k)$ is approximated as a product of two messages $\mathcal{M}_{j \rightarrow k}\mathcal{M}_{k \rightarrow j}$, each a Bernoulli.
+
+Thus, the natural parameters of the messages are updated with:
+
+\[\eta_{ij} = b_i + \sum_{k\in ne(i), k\neq j}^{K} \eta_{ki} \]
+
+The summation of the natural parameters of the singleton factor for node $i$ with the natural parameters of messages from all the neighbouring nodes.
+
+This leads to a loopy BP algorithm because the nodes are fully connected (i.e.
+every node is the neighbour of all other nodes).
+Thus, we cannot simply move from one end of the graph to the other like BP for tree structured graphs.
+
+\subsection*{(d)}
 

### 2022-12-27 main.py

@@ -12,6 +12,7 @@ from src.models.kernels import CombinedKernel, CombinedKernelParameters
 from src.models.gaussian_process_regression import GaussianProcessParameters
 from src.solutions import q2, q3, q4, q5, q6
 from dataclasses import asdict
+
 jax.config.update("jax_enable_x64", True)
 
 if __name__ == "__main__":
@@ -86,7 +87,7 @@ if __name__ == "__main__":
         log_sigma=jnp.log(1),
     )
     years_to_predict = 15
-    t_new = t[-1]+np.linspace(0, years_to_predict, years_to_predict*12)
+    t_new = t[-1] + np.linspace(0, years_to_predict, years_to_predict * 12)
     t_test = np.concatenate((t, t_new))
     q2.f(
         t_train=t,

### 2022-12-27 src/solutions/q2.py

@@ -124,8 +124,12 @@ def c(
     plt.close()
 
     fig_samples, ax_samples = plt.subplots(
-        len(fields(kernel_parameters.__class__)), len(log_theta_range),
-        figsize=(len(log_theta_range) * 2, len(fields(kernel_parameters.__class__))*2),
+        len(fields(kernel_parameters.__class__)),
+        len(log_theta_range),
+        figsize=(
+            len(log_theta_range) * 2,
+            len(fields(kernel_parameters.__class__)) * 2,
+        ),
         frameon=False,
     )
     for i, field in enumerate(fields(kernel_parameters.__class__)):
@@ -138,15 +142,21 @@ def c(
                     jnp.zeros(gram.shape[0]), gram, size=1
                 ).reshape(-1),
             )
-            ax_samples[i][j].set_title(f"{field.name.strip('log_')}={np.round(np.exp(log_value), 2)}")
+            ax_samples[i][j].set_title(
+                f"{field.name.strip('log_')}={np.round(np.exp(log_value), 2)}"
+            )
         setattr(kernel_parameters, field.name, default_value)
     plt.tight_layout()
-    plt.savefig(save_path + f"-parameter-samples", bbox_inches='tight')
+    plt.savefig(save_path + f"-parameter-samples", bbox_inches="tight")
     plt.close(fig_samples)
 
     fig_gram, ax_gram = plt.subplots(
-        len(fields(kernel_parameters.__class__)), len(log_theta_range),
-        figsize=(len(log_theta_range) * 2, len(fields(kernel_parameters.__class__))*2),
+        len(fields(kernel_parameters.__class__)),
+        len(log_theta_range),
+        figsize=(
+            len(log_theta_range) * 2,
+            len(fields(kernel_parameters.__class__)) * 2,
+        ),
         frameon=False,
     )
     for i, field in enumerate(fields(kernel_parameters.__class__)):
@@ -155,25 +165,27 @@ def c(
             setattr(kernel_parameters, field.name, log_value)
             gram = kernel(t, **asdict(kernel_parameters))
             ax_gram[i][j].imshow(gram)
-            ax_gram[i][j].set_title(f"{field.name.strip('log_')}={np.round(np.exp(log_value), 2)}")
+            ax_gram[i][j].set_title(
+                f"{field.name.strip('log_')}={np.round(np.exp(log_value), 2)}"
+            )
         setattr(kernel_parameters, field.name, default_value)
     plt.tight_layout()
[…]

### 2022-12-28 submission.tex
update assignment

@@
 The free energy is can be calculated as:
-\[\mathcal{F}(q, \theta) = \langle \log P(\textbf{x}, \textbf{s} | \theta)\rangle_{q(\textbf{s})} + H[Q(\textbf{s})]\]
+\[\mathcal{F}(q, \theta) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{q(\textbf{s})} + H[Q(\textbf{s})]\]
 
 %Thus,
-%\[\mathcal{F}(Q, \theta) = \langle \log P(\textbf{x}, \textbf{s} | \theta)\rangle_{\prod_{i=1}^{K} q_{i} (s_i)} + H\left[ \prod_{i=1}^{K} q_{i} (s_i)\right] \]
+%\[\mathcal{F}(Q, \theta) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{i=1}^{K} q_{i} (s_i)} + H\left[ \prod_{i=1}^{K} q_{i} (s_i)\right] \]
 
@@
 
-\[\mathcal{F}(Q, \theta) = \langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} + \langle \log P(\textbf{s} | \theta)\rangle_{q(\textbf{s})} + H\left[ q(\textbf{s})\right] \]
+\[\mathcal{F}(Q, \theta) = \left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} + \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})} + H\left[ q(\textbf{s})\right] \]
 
@@
 \[
-\langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} \\
-= -\frac{d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} \langle s_i \rangle_{q_{i} (s_i)} \mu_i   + \sum_{i=1}^{K} \sum_{j=1}^{K} \langle s_i s_j \rangle_{q_{i} (s_i) q_{j} (s_j)} \mu_i^T\mu_j \right)\]
+\left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} \\
+= -\frac{d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} \left\langle s_i \right\rangle_{q_{i} (s_i)} \mu_i   + \sum_{i=1}^{K} \sum_{j=1}^{K} \left\langle s_i s_j \right\rangle_{q_{i} (s_i) q_{j} (s_j)} \mu_i^T\mu_j \right)\]
 
@@
 \[
-\langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} \\
+\left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} \\
 = -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right)\]
 
-where $\langle s_i s_i \rangle_{q_{i} (s_i)} = \langle s_i \rangle_{q_{i} (s_i)}$ because $s_i \in \{0, 1\}$.
+where $\left\langle s_i s_i \right\rangle_{q_{i} (s_i)} = \left\langle s_i \right\rangle_{q_{i} (s_i)}$ because $s_i \in \{0, 1\}$.
 
@@
 
-\[ \langle \log P(\textbf{s} | \theta)\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \langle s_i \rangle_{q_{i} (s_i)} \log \pi_i + (1-\langle s_i\rangle_{q_{i} (s_i)}) \log (1-\pi_i)}\]
+\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \left\langle s_i \right\rangle_{q_{i} (s_i)} \log \pi_i + (1-\left\langle s_i\right\rangle_{q_{i} (s_i)}) \log (1-\pi_i)}\]
 
@@
 
-\[ \langle \log P(\textbf{s} | \theta)\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \lambda_i \log \pi_i + (1-\lambda_i) \log (1-\pi_i)}\]
+\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \lambda_i \log \pi_i + (1-\lambda_i) \log (1-\pi_i)}\]
 
@@
 
-\[\frac{\partial}{\partial q_i}\left( \mathcal{F}(q, \theta) + \lambda^{LG} \int q_i -1)\right) = \langle \log P(\textbf{x}, \textbf{s} | \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} - \log q_i(s_i) - 1 + \lambda^{LG}\]
+\[\frac{\partial}{\partial q_i}\left( \mathcal{F}(q, \theta) + \lambda^{LG} \int q_i -1)\right) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} - \log q_i(s_i) - 1 + \lambda^{LG}\]
 
@@
 
-\[\log q_i(s_i) = \langle \log P(\textbf{x}, \textbf{s} | \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} - 1 + \lambda^{LG}\]
+\[\log q_i(s_i) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} - 1 + \lambda^{LG}\]
 
@@
 
-\[\langle \log P(\textbf{x} |  \textbf{s}, \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} \propto  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} \langle s_k \rangle_{\prod_{j\neq i} q_j(s_j)} \mu_i   + \sum_{k=1}^{K} \sum_{j=1}^{K} \langle s_k s_j \rangle_{\prod_{j\neq i} q_j(s_j)} \right)\]
+\[\left\langle \log P(\textbf{x} |  \textbf{s}, \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} \propto  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} \left\langle s_k \right\rangle_{\prod_{j\neq i} q_j(s_j)} \mu_i   + \sum_{k=1}^{K} \sum_{j=1}^{K} \left\langle s_k s_j \right\rangle_{\prod_{j\neq i} q_j(s_j)} \right)\]
 
@@
 
-\[ \langle \log P(\textbf{s} | \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} = \sum_{k=1}^{K} \langle s_k \rangle_{\prod_{j\neq i} q_j(s_j)} \log \pi_k + (1-\langle s_k\rangle_{\prod_{j\neq i} q_j(s_j)}) \log (1-\pi_k)}\]
+\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} = \sum_{k=1}^{K} \left\langle s_k \right\rangle_{\prod_{j\neq i} q_j(s_j)} \log \pi_k + (1-\left\langle s_k\right\rangle_{\prod_{j\neq i} q_j(s_j)}) \log (1-\pi_k)}\]
 
@@
 
-\[\log q_i(s_i) \propto  \log P(\textbf{x} |  \textbf{s}, \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} + \langle \log P(\textbf{s} | \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} \]
+\[\log q_i(s_i) \propto  \log P(\textbf{x} |  \textbf{s}, \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} + \left\langle \log P(\textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} \]
 
@@
 
-\[\mu = \left( \langle \textbf{s}\textbf{s}^T \rangle_{q(\textbf{s})} \right)^{-1}\langle \textbf{s}\rangle_{q(\textbf{s})} \textbf{x}\]
+\[\mu = \left( \left\langle \textbf{s}\textbf{s}^T \right\rangle_{q(\textbf{s})} \right)^{-1}\left\langle \textbf{s}\right\rangle_{q(\textbf{s})} \textbf{x}\]
 
@@
 
+We begin with the log joint:
+
+\[\log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) = \log P(\textbf{x}|  \textbf{s}, \textbf{A}, \Psi,  \eta)
++ \log P( \textbf{s}| \pi, \eta) + \log P( \pi| \eta) + \log P( \textbf{A}| \eta) + \log P( \Psi| \eta)
+\]
+
+where $\eta$ is a collection of all hyperparameters.
+
+We know:
+
+\[P(\textbf{x}|  \textbf{s}, \textbf{A}, \Psi,  \eta) = \frac{1}{(2\pi)^{\frac{D}{2}} |\Psi|^{\frac{1}{2}}} \exp\left(-\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As})\right)\]
+
+\[P( \textbf{s}| \pi, \eta) = \prod_{k=1}^{K} \pi_k^{s_i} (1-\pi_k)^{1-s_k}\]
+
+\[P( \pi| \eta) = \prod_{k=1}^{K} \frac{\pi_k^{\alpha - 1}(1-\pi_k)^{\beta-1}}{B(\alpha, \beta)}\]
+
+We choose the conjugate priors:
+
+\[P( \textbf{A}| \eta) = \mathcal{N}(\textbf{A}| \mu_{\textbf{A}}, \Sigma_{\textbf{A}}) =\frac{1}{(2\pi)^{\frac{DK}{2}} |\Sigma_{\textbf{A}}|^{\frac{1}{2}}} \exp\left(-\frac{1}{2}(\textbf{A} - \mu_{\textbf{A}})^T \Sigma_{\textbf{A}}^{-1}(\textbf{A} - \mu_{\textbf{A}})\right)\]
+
+\[P( \Psi| \eta) = \prod_{d=1}^D InvGamma(\Psi_{dd}| a_d, b_d) = \prod_{d=1}^D \frac{b_d^{a_d}}{\Gamma(a_d)}\Psi_{dd}^{-a_d-1}\exp(-\frac{b_d}{\Psi_{dd}})\]
+
+a Gaussian prior on $\textbf{A}$ and a product of inverse gamma distributions on $\Psi$ where we assume $\Psi$ is a diagonal matrix.
+
+Combining, we have our expression:
+
+\[
+\begin{array}{l}
+\log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) = -\frac{D}{2} \log (2 \pi) - \frac{1}{2} \sum_{d=1}^{D}\log\Psi_{dd}   -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \\
+\tab \tab \tab \tab \tab + \sum_{k=1}^{K} s_k \log \pi_k + (1-s_k) \log (1-\pi_k)}\\
+\tab \tab \tab \tab \tab + \sum_{i=1}^K (\alpha-1) \log \pi_k + (\beta - 1) \log (1-\pi_k) - \log B(\alpha, \beta)\\
+\tab \tab \tab \tab \tab -\frac{DK}{2} \log (2 \pi) - \frac{1}{2} \log(|\Sigma_{\textbf{A}}|)   -\frac{1}{2}(\textbf{A} - \mu_{\textbf{A}})^T \Sigma_{\textbf{A}}^{-1}(\textbf{A} - \mu_{\textbf{A}})\\
+\tab \tab \tab \tab \tab + \sum_{d=1}^D a_d \log b_d + (-a_d - 1) \log\Psi_{dd} - \frac{b_d}{\Psi_{dd}} - \log \Gamma(a_d)\\
+\end{array}
+\]
+
+%Simplifying:
+%
+%\[
+%\begin{array}{l}
+%\log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) = -\frac{D}{2} \log (2 \pi)  -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \\
+%\tab \tab \tab \tab \tab + \sum_{i=1}^K (s_k+\alpha-1) \log \pi_k + (\beta - s_k) \log (1-\pi_k) - \log B(\alpha, \beta)\\
+%\tab \tab \tab \tab \tab -\frac{DK}{2} \log (2 \pi) - \frac{1}{2} \log(|\Sigma_{\textbf{A}}|)   -\frac{1}{2}\textbf{A}^T \Sigma_{\textbf{A}}^{-1}\textbf{A}\\
+%\tab \tab \tab \tab \tab + \sum_{d=1}^D a_d \log b_d + (-a_d - \frac{3}{2}) \log\Psi_{dd} - \frac{b_d}{\Psi_{dd}} - \log \Gamma(a_d)\\
+%\end{array}
+%\]
+
+For the Variational Bayes expectation step, we minimise $\textbf{KL}[q_{\texbf{s}}(\textbf{s} | \text{everything else}) \| P(\textbf{s} | \text{everything else})]$ by setting:
+
+\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q(\theta)}\]
+
+where $\theta$ denotes the parameters $\pi$, $\textbf{A}$, $\Psi$, $\eta$.
+
+Substituting the relevant terms:
+
+\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) + \sum_{k=1}^{K} s_k \log \pi_k + (1-s_k) \log (1-\pi_k)}\right\rangle_{q(\theta)}\]
+
+Simplifying:
+
+\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle -\frac{1}{2}\left(\textbf{s}^T\textbf{A}^T\Psi^{-1}\textbf{As} - 2\textbf{s}^T\left(\textbf{A}^T\Psi^{-1} \textbf{x}  + 2\log\frac{\pi}{1-\pi}\right)\right)\right\rangle_{q(\theta)}\]
+
+By inspection, we can see:
+
+\[q_{\texbf{s}}(\textbf{s}) \propto \mathcal{N}(s | \mu_\textbf{s}^*, \Sigma_\textbf{s}^*)\]
+
+where
+
+\[\Sigma_\textbf{s}^* = \left\langle \left(\textbf{A}^T\Psi^{-1}\textbf{A}\right)^{-1}\right\rangle_{q(\theta)}\]
+
+and
+
+\[\mu_\textbf{s}^* = \left\langle \left(\textbf{A}^T\Psi^{-1}\textbf{A}\right)^{-1}\left(\textbf{A}^T\Psi^{-1} \textbf{x}  + 2\log\frac{\pi}{1-\pi}\right)\right\rangle_{q(\theta)}\]
+
+the E step updates.
+
+For the Variational Bayes maximisation step, we set:
+
+\[q_{\theta}(\theta) \propto P(\theta) \exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q(\textbf{s})}\]
+
+assuming the factorisation:
+
+\[q_{\theta}(\theta) = q_{\pi}(\pi)q_{\Psi}(\Psi)q_{\textbf{A}}(\textbf{A})\]
+
+we can calculate each factor independently.
+
+For $q_{\pi}(\pi)$:
+
+\[q_{\pi}(\pi) \propto P(\pi)\exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \pi}(\theta)}\]
+
+Substituting the appropriate terms:
+
+\[q_{\pi}(\pi) \propto \left(\prod_{k=1}^{K} \frac{\pi_k^{\alpha - 1}(1-\pi_k)^{\beta-1}}{B(\alpha, \beta)} \right)\exp \left\langle \sum_{i=1}^K s_k \log \pi_k + (1 - s_k) \log (1-\pi_k) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \pi}(\theta)}\]
+
+We see:
+
+\[q_{\pi}(\pi) \propto \prod_{k=1}^{K} \frac{\pi_k^{\alpha  + \left\langle s_k\right\rangle_{q_{s_k}} - 1}(1-\pi_k)^{\beta-\left\langle s_k\right\rangle_{q_{s_k}}}}{B(\alpha, \beta)}\]
+
+\[q_{\pi}(\pi) = \prod_{k=1}^{K}  Beta(\alpha + \left\langle s_k\right\rangle_{q_{s_k}} , \beta+(1-\left\langle s_k\right\rangle_{q_{s_k}}}))\]
+
+%Substituting the appropriate terms:
+%
+%\[q_{\pi}(\pi) \propto \left(\prod_{k=1}^{K} \frac{\pi_k^{\alpha - 1}(1-\pi_k)^{\beta-1}}{B(\alpha, \beta)} \right)\exp \left\langle \sum_{i=1}^K (s_k+\alpha-1) \log \pi_k + (\beta - s_k) \log (1-\pi_k) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \pi}(\theta)}\]
+%
+%We see:
+%
+%\[q_{\pi}(\pi) \propto \prod_{k=1}^{K} \frac{\pi_k^{2\alpha - 2 + \left\langle s_k\right\rangle_{q_{s_k}}}(1-\pi_k)^{2\beta-\left\langle s_k\right\rangle_{q_{s_k}}}}{B(\alpha, \beta)}\]
+%
+%\[q_{\pi}(\pi) = \prod_{k=1}^{K}  Beta(2\alpha -2 + \left\langle s_k\right\rangle_{q_{s_k}} , 2\beta+1-\left\langle s_k\right\rangle_{q_{s_k}}})\]
+
+For $q_{\Psi}(\Psi)$:
+
+\[q_{\Psi}(\Psi) \propto P(\Psi)\exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \Psi}(\theta)}\]
+
+Substituting the appropriate terms:
+
+\[
+q_{\Psi}(\Psi) \propto \left( \prod_{d=1}^D \frac{b_d^{a_d}}{\Gamma(a_d)}\Psi_{dd}^{-a_d-1}\exp(-\frac{b_d}{\Psi_{dd}}) \right)\exp \left\langle  - \frac{1}{2} \sum_{d=1}^{D}\log\Psi_{dd}   -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \Psi(\theta)}}
+\]
+
+We see:
+
+\[
+q_{\Psi}(\Psi) \propto  \prod_{d=1}^D \frac{b_d^{a_d}}{\Gamma(a_d)}\Psi_{dd}^{-(a_d+\frac{1}{2})-1}\exp\left(-\frac{b_d+\frac{1}{2} \left\langle (x_d-\textbf{A}_d\textbf{s})^2 \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}_d}(\textbf{A}_d)}}{\Psi_{dd}}\right) }
+\]
+
+where $\textbf{A}_d \in \mathbb{R}^{1 \times K}$ is the $d^{th}$ row of $\textbf{A}$.
+
+Thus,
+
+\[
+q_{\Psi}(\Psi) = \prod_{d=1}^D  InvGamma\left(\Psi_{dd} \left| a_d+\frac{1}{2}, b_d+\frac{1}{2} \left\langle (x_d-\textbf{A}_d\textbf{s})^2 \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}_d}(\textbf{A}_d)}\right)
+\]
+
+For $q_{\textbf{A}}(\textbf{A})$:
+
+\[q_{\textbf{A}}(\textbf{A})  \propto P(\textbf{A})\exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}}(\theta)}\]
+
+\[q_{\textbf{A}}(\textbf{A}) \propto \exp\left(-\frac{1}{2}(\textbf{A} - \mu_{\textbf{A}})^T \Sigma_{\textbf{A}}^{-1}(\textbf{A} - \mu_{\textbf{A}})\right) \exp \left\langle -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}}(\theta)}\]
+
+\[q_{\textbf{A}}(\textbf{A}) \propto \exp\left(-\frac{1}{2}(\textbf{A}^T \Sigma_{\textbf{A}}^{-1}\textbf{A} -2 \textbf{A}^T \Sigma_{\textbf{A}}^{-1}  \mu_{\textbf{A}})\right) \exp \left\langle -\frac{1}{2}(\textbf{s}^T\textbf{A}^T \Psi^{-1} \textbf{A}\textbf{s} - 2 \textbf{s}^T\textbf{A}^T \Psi^{-1} \textbf{x}) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}}(\theta)}\]
+
 \newpage
@@
 
+We can use automatic relevance determination (ARD) as a hyperparameter method to select relevant features
+
+Place prior on $\sigma^2$ and optimise with respect to the distributions would cause some to diverge and only relevant latent dimensions will remain.
+This gives us a value for $K$, the number of latent factors that haven't diverged.
+
 \newpage

### 2022-12-28 main.py

@@ -107,11 +107,24 @@ if __name__ == "__main__":
     Q3_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q3")
     if not os.path.exists(Q3_OUTPUT_FOLDER):
         os.makedirs(Q3_OUTPUT_FOLDER)
+    number_of_images = 1000
+    x = generate_images(n=number_of_images)
+    binary_latent_factor_model = q3.e_and_f(
+        x=x,
+        k=8,
+        em_iterations=1000,
+        e_maximum_steps=200,
+        e_convergence_criterion=0,
+        save_path=os.path.join(Q3_OUTPUT_FOLDER, "f"),
+    )
 
-    # q3.learn_binary_factors(
-    #     x=generate_images(),
-    #     k=8,
-    #     em_maximum_iterations=5,
-    #     e_maximum_steps=100,
-    #     e_convergence_criterion=0,
-    # )
+    q3.g(
+        x=x[:1, :],
+        binary_latent_factor_model=binary_latent_factor_model,
+        sigmas=[1, 2, 3],
+        k=8,
+        em_iterations=1000,
+        e_maximum_steps=200,
+        e_convergence_criterion=0,
+        save_path=os.path.join(Q3_OUTPUT_FOLDER, "g"),
+    )

### 2022-12-28 src/generate_images.py

@@ -3,11 +3,12 @@ import numpy as np
 from src.constants import DEFAULT_SEED, M1, M2, M3, M4, M5, M6, M7, M8
 
 
-def generate_images(n: int = 400, seed: int = DEFAULT_SEED):
+def generate_images(n: int = 400, seed: int = DEFAULT_SEED, sigma: float = 0.1):
     """
 
     :param n: number of data points
     :param seed: random seed
+    :param sigma: Gaussian noise
     :return:
     """
     d = 16  # dimensionality of the data
@@ -36,4 +37,4 @@ def generate_images(n: int = 400, seed: int = DEFAULT_SEED):
 
     # Generate Data - The Data is stored in Y
 
-    return np.dot(s, mut) + np.random.randn(n, d) * 0.1  # some Gaussian noise is added
+    return np.dot(s, mut) + np.random.randn(n, d) * sigma  # some Gaussian noise is added

### 2022-12-28 src/models/mean_field_learning.py

@@ -4,6 +4,8 @@ import numpy as np
 
 from demo_code.MStep import m_step
 
+from typing import List, Tuple
+
 
 @dataclass
 class MeanFieldApproximation:
@@ -249,7 +251,7 @@ def variational_expectation_step(
     mean_field_approximation: MeanFieldApproximation,
     max_steps: int,
     convergence_criterion: float,
-) -> MeanFieldApproximation:
+) -> Tuple[MeanFieldApproximation, np.ndarray]:
     """Variational E step
 
     :param x: data matrix (number_of_points, number_of_dimensions)
@@ -259,9 +261,9 @@ def variational_expectation_step(
     :param convergence_criterion: early stopping if change in free energy <  convergence_criterion
     :return: mean field approximation
     """
-    previous_free_energy = compute_free_energy(
+    free_energy = [compute_free_energy(
         x, binary_latent_factor_model, mean_field_approximation
-    )
+    )]
     for i in range(max_steps):
         for latent_factor in range(binary_latent_factor_model.k):
             mean_field_approximation.lambda_matrix[
@@ -269,13 +271,12 @@ def variational_expectation_step(
             ] = partial_expectation_step(
                 x, binary_latent_factor_model, mean_field_approximation, latent_factor
             )
-        free_energy = compute_free_energy(
+        free_energy.append(compute_free_energy(
             x, binary_latent_factor_model, mean_field_approximation
-        )
-        if free_energy - previous_free_energy <= convergence_criterion:
+        ))
+        if free_energy[-1] - free_energy[-2] <= convergence_criterion:
             break
-        previous_free_energy = free_energy
-    return mean_field_approximation
+    return mean_field_approximation, free_energy
 
 
 def maximisation_step(
@@ -296,21 +297,34 @@ def maximisation_step(
     )
 
 
+def is_converge(free_energies, new_mean_field_approximation, mean_field_approximation):
+    return (abs(free_energies[-1] - free_energies[-2]) == 0) and np.linalg.norm(
+        mean_field_approximation.lambda_matrix
+        - new_mean_field_approximation.lambda_matrix
+    ) ==0
+
+
 def learn_binary_factors(
     x: np.ndarray,
     k: int,
-    em_maximum_iterations: int,
+    em_iterations: int,
     e_maximum_steps: int,
     e_convergence_criterion: float,
+    mean_field_approximation: MeanFieldApproximation = None,
+    binary_latent_factor_model: BinaryLatentFactorModel = None,
 ):
     n = x.shape[0]
[…]

### 2022-12-28 src/solutions/q3.py

@@ -1,0 +1,79 @@
+import numpy as np
+from src.models.mean_field_learning import learn_binary_factors, BinaryLatentFactorModel, compute_free_energy, init_mean_field_approximation, variational_expectation_step, is_converge
+from src.generate_images import generate_images
+import matplotlib.pyplot as plt
+from typing import List
+
+
+def e_and_f(
+    x: np.ndarray,
+    k: int,
+    em_iterations: int,
+    e_maximum_steps: int,
+    e_convergence_criterion: float,
+    save_path: str,
+):
+    _, binary_latent_factor_model, free_energy = learn_binary_factors(
+        x, k, em_iterations, e_maximum_steps, e_convergence_criterion
+    )
+    fig, ax = plt.subplots(1, k, figsize=(k * 2, 2))
+    for i in range(k):
+        ax[i].imshow(binary_latent_factor_model.mu[:, i].reshape(4, 4))
+        ax[i].set_title(f"Latent Feature mu_{i}")
+    fig.suptitle("Learned Features")
+    plt.tight_layout()
+    plt.savefig(save_path + "-latent-factors", bbox_inches="tight")
+    plt.close()
+
+    plt.title("Free Energy")
+    plt.xlabel("t (EM steps)")
+    plt.ylabel("Free Energy")
+    plt.plot(free_energy)
+    plt.savefig(save_path + "-free-energy", bbox_inches="tight")
+    plt.close()
+    return binary_latent_factor_model
+
+
+def g(
+    x: np.ndarray,
+    binary_latent_factor_model: BinaryLatentFactorModel,
+    sigmas: List[float],
+    k: int,
+    em_iterations: int,
+    e_maximum_steps: int,
+    e_convergence_criterion: float,
+    save_path: str,
+):
+    n = x.shape[0]
+    free_energies = []
+    for sigma in sigmas:
+        binary_latent_factor_model.sigma = sigma
+        mean_field_approximation = init_mean_field_approximation(k, n)
+        free_energy: List[float] = [
+            compute_free_energy(x, binary_latent_factor_model, mean_field_approximation)
+        ]
+        for _ in range(em_iterations):
+            new_mean_field_approximation, new_free_energy = variational_expectation_step(
+                x=x,
+                binary_latent_factor_model=binary_latent_factor_model,
+                mean_field_approximation=mean_field_approximation,
+                max_steps=e_maximum_steps,
+                convergence_criterion=e_convergence_criterion,
+            )
[…]

### 2022-12-29 submission.tex
update belief prop derivation

@@
 
-\[\log p(\textbf{s}, \textbf{x})  =    \sum_{i=1}^{K} \left(\left(\frac{\textbf{x}^T \mu_i}{\sigma^2} +\log\frac{\pi_i}{1-\pi_i} - \frac{\mu_i^T \mu_j}{2\sigma^2} \right) s_i\right) + \sum_{i=1}^{K}\sum_{j=1}^{i-1}\left( \frac{ - \mu_i^T \mu_j}{\sigma^2} s_i s_j \right) + C\]
+\[\log p(\textbf{s}, \textbf{x})  =    \sum_{i=1}^{K} \left(\left(\frac{\textbf{x}^T \mu_i}{\sigma^2} +\log\frac{\pi_i}{1-\pi_i} - \frac{\mu_i^T \mu_i}{2\sigma^2} \right) s_i\right) + \sum_{i=1}^{K}\sum_{j=1}^{i-1}\left( \frac{ - \mu_i^T \mu_j}{\sigma^2} s_i s_j \right) + C\]
 
@@
 
-\[\log f_i(s_i) = \left(\frac{\textbf{x}^T \mu_i}{\sigma^2} +\log\frac{\pi_i}{1-\pi_i} - \frac{\mu_i^T \mu_j}{2\sigma^2} \right) s_i\]
+\[\log f_i(s_i) = \left(\frac{\textbf{x}^T \mu_i}{\sigma^2} +\log\frac{\pi_i}{1-\pi_i} - \frac{\mu_i^T \mu_i}{2\sigma^2} \right) s_i\]
 
@@
 
-as required. Note that $C$ can simply be absorbed into any one of these factors.
+as required.
 
@@
 
-\[b_i = -\left(\frac{\textbf{x}^T \mu_i}{\sigma^2} +\log\frac{\pi_i}{1-\pi_i} - \frac{\mu_i^T \mu_j}{2\sigma^2}\right)\]
+\[b_i = -\left(\frac{\textbf{x}^T \mu_i}{\sigma^2} +\log\frac{\pi_i}{1-\pi_i} - \frac{\mu_i^T \mu_i}{2\sigma^2}\right)\]
 
@@
 
-\[\tilde{g}_{ij}(s_i, s_j) = \left( \theta_i^{s_i} + (1-\theta_i)^{1-s_i}\right)\left( \theta_j^{s_j} + (1-\theta_j)^{1-s_j}\right)\]
+\[\tilde{g}_{ij}(s_i, s_j) = \tilde{g}_{ij, \neg s_j}(s_i)\tilde{g}_{ij, \neg s_i}(s_j)\]
+
+where
+
+\[\tilde{g}_{ij, \neg s_j}(s_i) = (\theta_{ji})^{s_i} + (1-\theta_{ji})^{1-s_i}\]
+
+and
+
+\[\tilde{g}_{ij, \neg s_i}(s_j) = (\theta_{ij})^{s_j} + (1-\theta_{ij})^{1-s_j}\]
 
@@
 
-\[\log \tilde{g}_{ij}(s_i, s_j) \propto \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j\]
-
-To derive the a message passing scheme, we first define:
-
-\[q(\textbf{s}) = \prod_{i'=1}^{K} \left( \tilde{f}_{i'}(s_{i'})
-\prod_{j\in ne(i)}^{K} \mathcal{M}_{j \rightarrow i}
-\right)
-\]
-
-where $ne(i)$ are indices of neighbouring nodes of node $i$ and $\mathcal{M}_{j \rightarrow i}$ is the message from node $j$ to $i$.
-
-Thus, the cavity distributions:
-
-\[q_{\neg \tilde{f}_i(s_i)}(\textbf{s}) =
-\prod_{j\in ne(i)}^{K} \mathcal{M}_{j \rightarrow i}
+\[\log \tilde{g}_{ij}(s_i, s_j) \propto \log \left(\frac{\theta_{ji}}{1-\theta_{ji}} \right) s_i + \log \left(\frac{\theta_{ij}}{1-\theta_{ij}} \right) s_j\]
+
+To derive the a message passing scheme, we first define the incoming message to node $i$ from the singleton factor:
+
+\[\mathcal{M}_{i}(s_i) = \tilde{f}_i(s_i)\]
+
+and the message incoming message to node $i$ from node $j$:
+
+%\[\mathcal{M}_{j \rightarrow i} = \sum_{\substack{s_m \in \{0, 1\} \\ m \in \{1, ..., j-1, j+1,... K\}}} \tilde{f}_j(s_j) \prod_{k\in ne(j), k\neq i}^{K}} g_{kj}(s_k, s_j)\]
+
+%\[\mathcal{M}_{j \rightarrow i}(s_i) = \sum_{s_1 \in \{0, 1\} } \cdots \sum_{s_{i-1} \in \{0, 1\} } \sum_{s_{i+1} \in \{0, 1\} } \cdots \sum_{s_1 \in \{0, 1\} } \tilde{f}_j(s_j) \tilde{g}_{ji}(s_j, s_i) \prod_{k\in ne(j), k\neq i}^{K}} \tilde{g}_{kj}(s_k, s_j)\]
+
+\[\mathcal{M}_{j \rightarrow i}(s_i) = \sum_{s_1 \in \{0, 1\} } \cdots \sum_{s_{i-1} \in \{0, 1\} } \sum_{s_{i+1} \in \{0, 1\} } \cdots \sum_{s_1 \in \{0, 1\} } \tilde{f}_j(s_j) \tilde{g}_{ji}(s_j, s_i) \prod_{k\in ne(j), k\neq i}^{K}} \mathcal{M}_{k \rightarrow j}(s_j)\]
+
+where $ne(j)$ are indices of neighbouring nodes of node $j$.
+
+Because $\tilde{g}_{ji}(s_j, s_i)$ is a product:
+
+\[\mathcal{M}_{j \rightarrow i}(s_i) = \tilde{g}_{ji, \neg s_j}(s_i) \sum_{s_1 \in \{0, 1\} } \cdots \sum_{s_{i-1} \in \{0, 1\} } \sum_{s_{i+1} \in \{0, 1\} } \cdots \sum_{s_1 \in \{0, 1\} } \tilde{f}_j(s_j) \tilde{g}_{ji, \neg s_i}(s_j) \prod_{k\in ne(j), k\neq i}^{K}} \mathcal{M}_{k \rightarrow j}(s_j)\]
+
+Simplifying:
+
+\[\mathcal{M}_{j \rightarrow i}(s_i) = \tilde{g}_{ji, \neg s_j}(s_i)\]
+
+and,
+
+\[\mathcal{M}_{j \rightarrow i}(s_i) \propto \exp\left(\log\left(\frac{\theta_{ji}}{1-\theta_{ji}} \right) s_i\right)\]
+
+%\[\mathcal{M}_{j \rightarrow i}(s_i) = \tilde{g}_{ji, \neg s_j}(s_i)\]
+
+%\[\mathcal{M}_{j \rightarrow i} = \tilde{f}_j(s_j) \prod_{k\in ne(j), k\neq i}^{K}}\tilde{g}_{kj, \neg s_k}(s_j)  \sum_{\substack{s_m \in \{0, 1\} \\ m \in \{1, ..., j-1, j+1,...
+K\}}} \prod_{k\in ne(j), k\neq i}^{K}} \tilde{g}_{kj, \neg s_j}(s_k)\]
+
+%\[\mathcal{M}_{j \rightarrow i}(s_j) = \tilde{f}_j(s_j) \prod_{k\in ne(j), k\neq i}^{K}}\tilde{g}_{kj, \neg s_k}(s_j) \prod_{k\in ne(j), k\neq i}^{K}} \left(\sum_{s_k \in \{0, 1\} } \tilde{g}_{kj, \neg s_j}(s_k) \right)\]
+%
+%\[\mathcal{M}_{j \rightarrow i}(s_j) = \tilde{f}_j(s_j) \prod_{k\in ne(j), k\neq i}^{K}}\tilde{g}_{kj, \neg s_k}(s_j)\]
+
+Thus, the cavity distributions are:
+
+\[q_{\neg \tilde{f}_i(s_i)}(s_i) =
+\prod_{j\in ne(i)}^{K} \mathcal{M}_{j \rightarrow i}(s_i)
 \]
@@
 
-\[q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) = \left( \tilde{f}_{i}(s_{i}) \prod_{k\in ne(i), k\neq j}^{K} \mathcal{M}_{k \rightarrow i}
-\right)\left( \tilde{f}_{j}(s_{j}) \prod_{k\in ne(j), k\neq j}^{K} \mathcal{M}_{k \rightarrow j}
+\[q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) = \left( \mathcal{M}_{i}(s_i) \prod_{k\in ne(i), k\neq j}^{K} \mathcal{M}_{k \rightarrow i}(s_i)
+\right)\left( \mathcal{M}_{j}(s_j) \prod_{k\in ne(j), k\neq i}^{K} \mathcal{M}_{k \rightarrow j}(s_j)
 \right)
@@
 
-\[\tilde{f}_{i}(s_{i}) = \arg \min_{\tilde{f}_{i}(s_{i})} \textbf{KL} \left[ f_{i}(s_{i}) q_{\neg \tilde{f}_i(s_i)}(\textbf{s}) \| \tilde{f}_{i}(s_{i}) q_{\neg \tilde{f}_i(s_i)}(\textbf{s}) \right]\]
+\[\tilde{f}_{i}(s_{i}) = \arg \min_{\tilde{f}_{i}(s_{i})} \textbf{KL} \left[ f_{i}(s_{i}) q_{\neg \tilde{f}_i(s_i)}(s_i) \| \tilde{f}_{i}(s_{i}) q_{\neg \tilde{f}_i(s_i)}(s_i) \right]\]
 
@@
 
-\[\tilde{g}_{ij}(s_i, s_j) = \arg \min_{\tilde{f}_{i}(s_{i})} \textbf{KL} \left[ g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \| \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \right]\]
-
-Defining the natural parameters $\eta_{ij}$ and $\eta_{ji}$ for the messages:
-
-\[\mathcal{M}_{i \rightarrow j} = \exp(\eta_{ij} s_i) = \tilde{f}_{i}(s_{i}) \prod_{k\in ne(i), k\neq j}^{K} \mathcal{M}_{k \rightarrow i} \]
-
-\[\mathcal{M}_{j \rightarrow i}  = \exp(\eta_{ji} s_j) = \tilde{f}_{j}(s_{j}) \prod_{k\in ne(j), k\neq i}^{K} \mathcal{M}_{k \rightarrow j}\]
-
-Note that because $\tilde{g}_{ij}(s_i, s_j)$ is the product of two Bernoulli distributions, updates to this site approximation involves updating to the natural parameters:
-
-\[ \phi_{ij}(\theta) = \begin{bmatrix} \theta_i \\ \theta_j\right] \end{bmatrix}\]
-
-the mean of $s_i$ and $s_j$ respectively.
+\[\tilde{g}_{ij}(s_i, s_j) = \arg \min_{\tilde{g}_{ij}(s_i, s_j)} \textbf{KL} \left[ g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \left\| \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \right]\]
+
+We can define natural parameters $\eta_{i, \neg s_j}$ and $\eta_{j, \neg s_i}$ for $q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)$ such that:
+
+\[\mathcal{M}_{i}(s_i) \prod_{k\in ne(i), k\neq j}^{K} \mathcal{M}_{k \rightarrow i}(s_i) \propto \exp(\eta_{i, \neg s_j} s_i)\]
+
+\[ \mathcal{M}_{j}(s_j) \prod_{k\in ne(j), k\neq j}^{K} \mathcal{M}_{k \rightarrow j}(s_j) \propto \exp(\eta_{j, \neg s_i} s_j)\]
+
+where:
+
+\[\eta_{i, \neg s_j} =  \log \left(\frac{\lambda_i}{1-\lambda_i}\right) + \sum_{k\in ne(j), k\neq j}^{K} \log\left(\frac{\theta_{kj}}{1-\theta_{kj}} \right)\]
+
+and
+
+\[\eta_{j, \neg s_i} =  \log \left(\frac{\lambda_j}{1-\lambda_j}\right) + \sum_{k\in ne(i), k\neq i}^{K} \log\left(\frac{\theta_{ki}}{1-\theta_{ki}} \right)\]
+
+Note that $\tilde{g}_{ij}(s_i, s_j)$ was chosen as the product of two Bernoulli distributions, updates to this site approximation involves updating to the natural parameters $\theta_i$ and $\theta_j$, the means of $s_i$ and $s_j$ respectively.
 
@@
 
-\[\log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j + \eta_{ij} s_i + \eta_{ji} s_j\]
+\[\log \tilde{g}_{ij}(s_i, s_j)q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \propto \log \left(\frac{\theta_{ji}}{1-\theta_{ji}} \right) s_i + \log \left(\frac{\theta_{ij}}{1-\theta_{ij}} \right) s_j + \eta_{i, \neg s_j} s_i + \eta_{j, \neg s_i} s_j\]
 
@@
 
-\[\log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \log \left(\frac{\theta_i}{1-\theta_i} + \eta_{ij} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} + \eta_{ji} \right) s_j \]
+\[\log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \propto \log \left(\frac{\theta_{ji}}{1-\theta_{ji}} + \eta_{i, \neg s_j} \right) s_i + \log \left(\frac{\theta_{ij}}{1-\theta_{ij}} + \eta_{j, \neg s_i} \right) s_j \]
 
@@
 
-\[\mathbb{E}_{s_i}[\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \frac{1}{1+\exp{\left(-\left(\theta_i} + \eta_{ij}\right)}\right)}\]
+\[\mathbb{E}_{s_i}\left[\sum_{s_j \in \{0, 1\}}\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \frac{1}{1+\exp{\left(-\left(\theta_{ji} + \eta_{i, \neg s_j} \right)}\right)}\]
 
@@
 
-\[\mathbb{E}_{s_j}[\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \frac{1}{1+\exp{\left(-\left(\theta_j + \eta_{ji}\right)}\right)}\]
+\[\mathbb{E}_{s_j}\left[\sum_{s_i \in \{0, 1\}}\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \frac{1}{1+\exp{\left(-\left(\theta_{ij} + \eta_{j, \neg s_i} \right)}\right)}\]
 
@@
 
-\[\log g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto W_{ij} s_i s_j
- + \eta_{ij} s_i + \eta_{ji} s_j\]
+\[\log g_{ij}(s_i, s_j) qq_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \propto W_{ij} s_i s_j
+ + \eta_{i, \neg s_j} s_i + \eta_{j, \neg s_i} s_j\]
 
@@
 
-To derive the first moment for $ g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})$ with respect to $s_i$, we first marginalise out $s_j$:
-
-\[ \sum_{s_j \in \{0, 1\}} g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \exp\left( W_{ij} s_i + \eta_{ij} s_i + \eta_{ji}\right) +  \exp\left(\eta_{ij} s_i \right)
+To derive the first moment for $ g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)$ with respect to $s_i$, we first marginalise out $s_j$:
+
+\[ \sum_{s_j \in \{0, 1\}} g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \exp\left( W_{ij} s_i + \eta_{i, \neg s_j} s_i + \eta_{j, \neg s_i}\right) +  \exp\left(\eta_{i, \neg s_j} s_i \right)
 \]
@@
 
-\[\mathbb{E}_{s_i}[g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \frac{\exp\left( W_{ij} + \eta_{ij} + \eta_{ji}\right) +  \exp\left(\eta_{ij} \right)}{[\exp\left( W_{ij} + \eta_{ij} + \eta_{ji}\right) +  \exp\left(\eta_{ij} \right)]+[\exp\left( \eta_{ji}\right) +  1]}\]
+\[\mathbb{E}_{s_i}\left[\sum_{s_j \in \{0, 1\}}g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \frac{\exp\left( W_{ij} + \eta_{i, \neg s_j}  + \eta_{j, \neg s_i}\right) +  \exp\left(\eta_{i, \neg s_j}\right)}{\left[\exp\left( W_{ij} + \eta_{i, \neg s_j}  + \eta_{j, \neg s_i}\right) +  \exp\left(\eta_{i, \neg s_j}\right)\right]+\left[\exp\left( \eta_{j, \neg s_i}\right) +  1\right]}\]
 
@@
 
-\[\mathbb{E}_{s_i}[g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \frac{\left( \exp\left( W_{ij} + \eta_{ji}\right) +  1 \right)\exp\left(\eta_{ij} \right)}{[\left( \exp\left( W_{ij} + \eta_{ji}\right) +  1 \right)\exp\left(\eta_{ij} \right)]+[\exp\left( \eta_{ji}\right) +  1]}\]
-
-similarly:
-
-\[\mathbb{E}_{s_j}[g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \frac{\left( \exp\left( W_{ij} + \eta_{ij}\right) +  1 \right)\exp\left(\eta_{ji} \right)}{[\left( \exp\left( W_{ij} + \eta_{ij}\right) +  1 \right)\exp\left(\eta_{ji} \right)]+[\exp\left( \eta_{ij}\right) +  1]}\]
+\[\mathbb{E}_{s_i}\left[\sum_{s_j \in \{0, 1\}}g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \frac{\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)  }{\left[\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)  \right]+\left[\exp\left( \eta_{j, \neg s_i}\right) +  1\right]}\]
+
+Similarly:
+
+\[\mathbb{E}_{s_j}\left[\sum_{s_i \in \{0, 1\}}g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \frac{\exp\left(\eta_{j, \neg s_i}\right)\left(\exp\left( W_{ij}  + \eta_{i, \neg s_j}\right) + 1\right)  }{\left[\exp\left(\eta_{j, \neg s_i}\right)\left(\exp\left( W_{ij}  + \eta_{i, \neg s_j}\right) + 1\right) \right]+\left[\exp\left( \eta_{i, \neg s_j}\right) +  1\right]}\]
 
@@
 
-\[\mathbb{E}_{s_i}[\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \mathbb{E}_{s_i}[g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})]\]
+\[\mathbb{E}_{s_i}\left[\sum_{s_j \in \{0, 1\}}\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \mathbb{E}_{s_i}\left[\sum_{s_j \in \{0, 1\}}g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right]\]
 
@@
 
-\[\mathbb{E}_{s_j}[\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})] = \mathbb{E}_{s_j}[g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s})]\]
+\[\mathbb{E}_{s_j}\left[\sum_{s_i \in \{0, 1\}}\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \mathbb{E}_{s_j}\left[\sum_{s_i \in \{0, 1\}}g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right]\]
 
@@
 
-\[\frac{1}{1+\exp{\left(-\left(\theta_i} + \eta_{ij}\right)}\right)}} = \frac{\left( \exp\left( W_{ij} + \eta_{ji}\right) +  1 \right)\exp\left(\eta_{ij} \right)}{[\left( \exp\left( W_{ij} + \eta_{ji}\right) +  1 \right)\exp\left(\eta_{ij} \right)]+[\exp\left( \eta_{ji}\right) +  1]}\]
-
-\[\exp\left( \eta_{ji}\right) +  1 =\exp{\left(-\left(\theta_i + \eta_{ij}\right)}\right)}\left( \exp\left( W_{ij} + \eta_{ji}\right) +  1 \right)\exp\left(\eta_{ij} \right)\]
-
-\[\frac{\exp\left( \eta_{ji}\right) +  1}{ \exp\left( W_{ij} + \eta_{ji}\right) +  1 } =\exp{\left(-\theta_i\right)}\]
-
-\[\theta_i = \log \left( \frac{ 1+\exp\left( W_{ij} + \eta_{ji}\right) }{1+\exp\left( \eta_{ji}\right)} \right)\]
+\[\frac{1}{1+\exp{\left(-\left(\theta_{ji} + \eta_{i, \neg s_j} \right)}\right)}} = \frac{\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)  }{\left[\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)  \right]+\left[\exp\left( \eta_{j, \neg s_i}\right) +  1\right]}\]
+
+\[\exp\left( \eta_{j, \neg s_i}\right) +  1=\exp{\left(-\left(\theta_{ji} + \eta_{i, \neg s_j} \right)}\right)}\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)\]
+
+\[\frac{\exp\left( \eta_{j, \neg s_i}\right) +  1}{ \exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1} =\exp{\left(-\theta_{ji}\right)}\]
+
+\[\theta_{ji} = \log \left( \frac{1+\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right)}{1+\exp\left( \eta_{j, \neg s_i}\right)} \right)\]
 
@@
 
-\[\theta_j = \log \left( \frac{ 1+ \exp\left( W_{ij} + \eta_{ij}\right)  }{1+\exp\left( \eta_{ij}\right)} \right)\]
+\[\theta_{ij} = \log \left( \frac{1+\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right)}{1+\exp\left( \eta_{i, \neg s_j}\right)} \right)\]
 

### 2022-12-29 submission.tex
update eta definition to use b
fix loopy bp derivation

@@
 
+we can define $\xi_{ji} = \log \left(\frac{\theta_{ji}}{1-\theta_{ji}} \right)$ and $\xi_{ij} = \log \left(\frac{\theta_{ij}}{1-\theta_{ij}} \right)$:
+
+\[\log \tilde{g}_{ij}(s_i, s_j) \propto \xi_{ji} s_i + \xi_{ij} s_j\]
+
 To derive the a message passing scheme, we first define the incoming message to node $i$ from the singleton factor:
@@
 
-\[\mathcal{M}_{j \rightarrow i}(s_i) \propto \exp\left(\log\left(\frac{\theta_{ji}}{1-\theta_{ji}} \right) s_i\right)\]
+\[\mathcal{M}_{j \rightarrow i}(s_i) \propto \exp\left(\xi_{ji} s_i\right)\]
 
@@
 
+Knowing $\log \left(\frac{\lambda_i}{1-\lambda_i} \right)s_i = b_i s_i$ and $\xi_{kj} = \log\left(\frac{\theta_{kj}}{1-\theta_{kj}} \right)$:
+
+\[\eta_{i, \neg s_j} =  b_i + \sum_{k\in ne(j), k\neq j}^{K} \xi_{kj}\]
+
 and
 
-\[\eta_{j, \neg s_i} =  \log \left(\frac{\lambda_j}{1-\lambda_j}\right) + \sum_{k\in ne(i), k\neq i}^{K} \log\left(\frac{\theta_{ki}}{1-\theta_{ki}} \right)\]
-
-Note that $\tilde{g}_{ij}(s_i, s_j)$ was chosen as the product of two Bernoulli distributions, updates to this site approximation involves updating to the natural parameters $\theta_i$ and $\theta_j$, the means of $s_i$ and $s_j$ respectively.
+\[\eta_{j, \neg s_i} =  b_j + \sum_{k\in ne(i), k\neq i}^{K} \xi_{ki}\]
+
+Note that $\tilde{g}_{ij}(s_i, s_j)$ was chosen as the product of two Bernoulli distributions, updates to this site approximation involves updating the  parameters $\xi_{ij}$ and $\xi_{ji}$, for $s_i$ and $s_j$ respectively.
 
@@
 
-\[\log \tilde{g}_{ij}(s_i, s_j)q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \propto \log \left(\frac{\theta_{ji}}{1-\theta_{ji}} \right) s_i + \log \left(\frac{\theta_{ij}}{1-\theta_{ij}} \right) s_j + \eta_{i, \neg s_j} s_i + \eta_{j, \neg s_i} s_j\]
+\[\log \tilde{g}_{ij}(s_i, s_j)q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \propto \xi_{ji} s_i + \xi_{ij} s_j + \eta_{i, \neg s_j} s_i + \eta_{j, \neg s_i} s_j\]
 
@@
 
-\[\log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \propto \log \left(\frac{\theta_{ji}}{1-\theta_{ji}} + \eta_{i, \neg s_j} \right) s_i + \log \left(\frac{\theta_{ij}}{1-\theta_{ij}} + \eta_{j, \neg s_i} \right) s_j \]
+\[\log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \propto \left(\xi_{ji}+ \eta_{i, \neg s_j} \right) s_i + \left(\xi_{ij} + \eta_{j, \neg s_i} \right) s_j \]
 
@@
 
-\[\mathbb{E}_{s_i}\left[\sum_{s_j \in \{0, 1\}}\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \frac{1}{1+\exp{\left(-\left(\theta_{ji} + \eta_{i, \neg s_j} \right)}\right)}\]
+\[\mathbb{E}_{s_i}\left[\sum_{s_j \in \{0, 1\}}\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \frac{1}{1+\exp{\left(-\left(\xi_{ji} + \eta_{i, \neg s_j} \right)}\right)}\]
 
@@
 
-\[\mathbb{E}_{s_j}\left[\sum_{s_i \in \{0, 1\}}\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \frac{1}{1+\exp{\left(-\left(\theta_{ij} + \eta_{j, \neg s_i} \right)}\right)}\]
+\[\mathbb{E}_{s_j}\left[\sum_{s_i \in \{0, 1\}}\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \frac{1}{1+\exp{\left(-\left(\xi_{ij} + \eta_{j, \neg s_i} \right)}\right)}\]
 
@@
 
-\[\frac{1}{1+\exp{\left(-\left(\theta_{ji} + \eta_{i, \neg s_j} \right)}\right)}} = \frac{\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)  }{\left[\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)  \right]+\left[\exp\left( \eta_{j, \neg s_i}\right) +  1\right]}\]
-
-\[\exp\left( \eta_{j, \neg s_i}\right) +  1=\exp{\left(-\left(\theta_{ji} + \eta_{i, \neg s_j} \right)}\right)}\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)\]
-
-\[\frac{\exp\left( \eta_{j, \neg s_i}\right) +  1}{ \exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1} =\exp{\left(-\theta_{ji}\right)}\]
-
-\[\theta_{ji} = \log \left( \frac{1+\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right)}{1+\exp\left( \eta_{j, \neg s_i}\right)} \right)\]
+\[\frac{1}{1+\exp{\left(-\left(\xi_{ji} + \eta_{i, \neg s_j} \right)}\right)}} = \frac{\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)  }{\left[\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)  \right]+\left[\exp\left( \eta_{j, \neg s_i}\right) +  1\right]}\]
+
+Simplifying:
+
+\[\exp\left( \eta_{j, \neg s_i}\right) +  1=\exp{\left(-\left(\xi_{ji} + \eta_{i, \neg s_j} \right)}\right)}\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)\]
+
+\[\frac{\exp\left( \eta_{j, \neg s_i}\right) +  1}{ \exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1} =\exp{\left(-\xi_{ji}\right)}\]
+
+Our parameter update:
+
+\[\xi_{ji} = \log \left( \frac{1+\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right)}{1+\exp\left( \eta_{j, \neg s_i}\right)} \right)\]
 
@@
 
-\[\theta_{ij} = \log \left( \frac{1+\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right)}{1+\exp\left( \eta_{i, \neg s_j}\right)} \right)\]
+\[\xi_{ij} = \log \left( \frac{1+\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right)}{1+\exp\left( \eta_{i, \neg s_j}\right)} \right)\]
 

### 2022-12-30 submission.tex
add loopy bp and refactor ep models

@@
 \newpage
+The Python code for the binary latent factor model:
+\lstinputlisting[language=Python]{src/models/binary_latent_factor_model.py}
+
+\newpage
 The Python code for mean field learning:
@@
 
-\[\eta_{i, \neg s_j} =  \log \left(\frac{\lambda_i}{1-\lambda_i}\right) + \sum_{k\in ne(j), k\neq j}^{K} \log\left(\frac{\theta_{kj}}{1-\theta_{kj}} \right)\]
-
-Knowing $\log \left(\frac{\lambda_i}{1-\lambda_i} \right)s_i = b_i s_i$ and $\xi_{kj} = \log\left(\frac{\theta_{kj}}{1-\theta_{kj}} \right)$:
-
-\[\eta_{i, \neg s_j} =  b_i + \sum_{k\in ne(j), k\neq j}^{K} \xi_{kj}\]
+\[\eta_{i, \neg s_j} =  \log \left(\frac{\lambda_i}{1-\lambda_i}\right) + \sum_{k\in ne(i), k\neq j}^{K} \log\left(\frac{\theta_{ki}}{1-\theta_{ki}} \right)\]
+
+Knowing $b_i=\log \left(\frac{\lambda_i}{1-\lambda_i}\right)$ and $\xi_{ki}=\log\left(\frac{\theta_{ki}}{1-\theta_{ki}} \right)$:
+
+\[\eta_{i, \neg s_j} =  b_i + \sum_{k\in ne(i), k\neq j}^{K} \xi_{ki}\]
 
@@
 
-\[\eta_{j, \neg s_i} =  b_j + \sum_{k\in ne(i), k\neq i}^{K} \xi_{ki}\]
+\[\eta_{j, \neg s_i} =  b_j + \sum_{k\in ne(j), k\neq i}^{K} \xi_{kj}\]
 
@@
 
-\[\xi_{ij} = \log \left( \frac{1+\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right)}{1+\exp\left( \eta_{i, \neg s_j}\right)} \right)\]
+\[\xi_{ij} = \log \left( \frac{1+\exp\left( W_{ij}  + \eta_{i, \neg s_j}\right)}{1+\exp\left( \eta_{i, \neg s_j}\right)} \right)\]
 
@@
 \newpage
+\section{Question 6}
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.4]{outputs/q6/all-latent-factors}
+\caption{Latent factors learned with EP/Loopy-BP}
+\label{fig:6-latent-factors}
+\end{figure}
+
+\newpage
+The Python code for the Boltzmann machine:
+\lstinputlisting[language=Python]{src/models/boltzmann_machine.py}
+
+\newpage
+The Python code for message passing:
+\lstinputlisting[language=Python]{src/models/message_passing.py}
+
+\newpage
+The rest of the Python code for question 6:
+\lstinputlisting[language=Python]{src/solutions/q6.py}
+
+\newpage
 \section*{Appendix 1: constants.py}

### 2022-12-30 main.py

@@ -10,7 +10,7 @@ from src.generate_images import generate_images
 from src.models.bayesian_linear_regression import LinearRegressionParameters
 from src.models.kernels import CombinedKernel, CombinedKernelParameters
 from src.models.gaussian_process_regression import GaussianProcessParameters
-from src.solutions import q2, q3, q4, q5, q6
+from src.solutions import q2, q3, q4, q6
 from dataclasses import asdict
 
 jax.config.update("jax_enable_x64", True)
@@ -132,3 +132,10 @@ if __name__ == "__main__":
         e_convergence_criterion=e_convergence_criterion,
         save_path=os.path.join(Q3_OUTPUT_FOLDER, "g"),
     )
+
+    # Question 6
+    Q6_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q6")
+    if not os.path.exists(Q6_OUTPUT_FOLDER):
+        os.makedirs(Q6_OUTPUT_FOLDER)
+    em_iterations = 50
+    q6.run(x, k, em_iterations, save_path=os.path.join(Q6_OUTPUT_FOLDER, "all"))

### 2022-12-30 src/models/binary_latent_factor_model.py

@@ -1,0 +1,253 @@
+from __future__ import annotations
+
+import numpy as np
+
+from demo_code.MStep import m_step
+from typing import List
+from abc import ABC, abstractmethod
+
+
+class BinaryLatentFactorModel:
+    """
+    mu: matrix of means (number_of_dimensions, number_of_latent_variables)
+    sigma: gaussian noise parameter
+    pi: vector of priors (1, number_of_latent_variables)
+    """
+
+    def __init__(
+        self,
+        mu: np.ndarray,
+        sigma: float,
+        pi: np.ndarray,
+    ):
+        self.mu = mu
+        self.sigma = sigma
+        self.pi = pi
+
+    def mu_exclude(self, exclude_latent_index: int) -> np.ndarray:
+        #  (number_of_dimensions, number_of_latent_variables-1)
+        return np.concatenate(
+            (self.mu[:, :exclude_latent_index], self.mu[:, exclude_latent_index + 1 :]),
+            axis=1,
+        )
+
+    @property
+    def log_pi(self):
+        return np.log(self.pi)
+
+    @property
+    def log_one_minus_pi(self):
+        return np.log(1 - self.pi)
+
+    @property
+    def variance(self):
+        return self.sigma**2
+
+    @property
+    def precision(self):
+        return 1 / self.variance
+
+    @property
+    def d(self):
+        return self.mu.shape[0]
+
+    @property
+    def k(self):
+        return self.mu.shape[1]
+
+    @staticmethod
+    def calculate_maximisation_parameters(
+        x: np.ndarray,
+        binary_latent_factor_approximation: BinaryLatentFactorApproximation,
+    ):
+
+        expectation_s = binary_latent_factor_approximation.lambda_matrix
+        expectation_ss = (
+            binary_latent_factor_approximation.lambda_matrix.T
+            @ binary_latent_factor_approximation.lambda_matrix
+        )
+        np.fill_diagonal(
+            expectation_ss, binary_latent_factor_approximation.lambda_matrix.sum(axis=0)
+        )
+        return m_step(x, expectation_s, expectation_ss)
+
+    def maximisation_step(
+        self,
+        x: np.ndarray,
+        binary_latent_factor_approximation: BinaryLatentFactorApproximation,
+    ):
+        mu, sigma, pi = self.calculate_maximisation_parameters(
+            x, binary_latent_factor_approximation
+        )
+        self.mu = mu
+        self.sigma = sigma
[…]

### 2022-12-30 src/models/boltzmann_machine.py

@@ -1,0 +1,71 @@
+import numpy as np
+from src.models.binary_latent_factor_model import (
+    BinaryLatentFactorModel,
+    BinaryLatentFactorApproximation,
+)
+
+
+class BoltzmannMachine(BinaryLatentFactorModel):
+    """
+    mu: matrix of means (number_of_dimensions, number_of_latent_variables)
+    sigma: gaussian noise parameter
+    pi: vector of priors (1, number_of_latent_variables)
+    """
+
+    def __init__(
+        self,
+        mu: np.ndarray,
+        sigma: float,
+        pi: np.ndarray,
+    ):
+        super().__init__(mu, sigma, pi)
+
+    @property
+    def w_matrix(self):
+        # (number_of_latent_variables, number_of_latent_variables)
+        return -self.precision * (self.mu.T @ self.mu)
+
+    def w_matrix_index(self, i, j):
+        # (number_of_latent_variables, number_of_latent_variables)
+        return -self.precision * (self.mu[:, i] @ self.mu[:, j])
+
+    def b(self, x):
+        """
+
+        :param x: design matrix (number_of_points, number_of_dimensions)
+        :return:
+        """
+        # (number_of_points, number_of_latent_variables)
+        return -(
+            self.precision * x @ self.mu
+            + self.log_pi_ratio
+            - 0.5 * self.precision * np.multiply(self.mu, self.mu).sum(axis=0)
+        )
+
+    def b_index(self, x, node_index) -> float:
+        # (number_of_points, 1)
+        return -(
+            self.precision * x @ self.mu[:, node_index]
+            + (self.log_pi[0, node_index] - self.log_one_minus_pi[0, node_index])
+            - 0.5 * self.precision * self.mu[:, node_index] @ self.mu[:, node_index]
+        ).reshape(
+            -1,
+        )
+
+    @property
+    def log_pi_ratio(self):
+        return self.log_pi - self.log_one_minus_pi
+
+
+def init_boltzmann_machine(
+    x: np.ndarray,
+    binary_latent_factor_approximation: BinaryLatentFactorApproximation,
+) -> BinaryLatentFactorModel:
+    mu, sigma, pi = BinaryLatentFactorModel.calculate_maximisation_parameters(
+        x, binary_latent_factor_approximation
+    )
+    return BoltzmannMachine(
+        mu=mu,
+        sigma=sigma,
+        pi=pi,
+    )

### 2022-12-30 src/models/mean_field_learning.py

@@ -1,19 +1,30 @@
-from dataclasses import dataclass
-
 import numpy as np
 
-from demo_code.MStep import m_step
+from src.models.binary_latent_factor_model import (
+    BinaryLatentFactorModel,
+    BinaryLatentFactorApproximation,
+)
 
-from typing import List, Tuple
 
-
-@dataclass
-class MeanFieldApproximation:
+class MeanFieldApproximation(BinaryLatentFactorApproximation):
     """
     lambda_matrix: parameters variational approximation (number_of_points, number_of_latent_variables)
     """
 
-    lambda_matrix: np.ndarray
+    _lambda_matrix: np.ndarray
+
+    def __init__(self, lambda_matrix, max_steps, convergence_criterion):
+        self.lambda_matrix = lambda_matrix
+        self.max_steps = max_steps
+        self.convergence_criterion = convergence_criterion
+
+    @property
+    def lambda_matrix(self):
+        return self._lambda_matrix
+
+    @lambda_matrix.setter
+    def lambda_matrix(self, value):
+        self._lambda_matrix = value
 
     def lambda_matrix_exclude(self, exclude_latent_index: int) -> np.ndarray:
         #  (number_of_points, number_of_latent_variables-1)
@@ -25,322 +36,80 @@ class MeanFieldApproximation:
             axis=1,
         )
 
-    @property
-    def log_lambda_matrix(self):
-        return np.log(self.lambda_matrix)
+    def _partial_expectation_step(
+        self,
+        x: np.ndarray,
+        binary_latent_factor_model: BinaryLatentFactorModel,
+        latent_factor: int,
+    ) -> np.ndarray:
+        """Partial Variational E step for factor i for all data points
 
-    @property
-    def log_one_minus_lambda_matrix(self):
-        return np.log(1 - self.lambda_matrix)
+        :param x: data matrix (number_of_points, number_of_dimensions)
+        :param binary_latent_factor_model: a binary_latent_factor_model
+        :param latent_factor: latent factor to compute partial update
+        :return: lambda_vector: new lambda parameters for the latent factor (number_of_points, 1)
+        """
+        lambda_matrix_excluded = self.lambda_matrix_exclude(latent_factor)
+        mu_excluded = binary_latent_factor_model.mu_exclude(latent_factor)
 
-    @property
-    def n(self):
-        return self.lambda_matrix.shape[0]
[…]

### 2022-12-30 src/models/message_passing.py

@@ -1,0 +1,116 @@
+import numpy as np
+from src.models.binary_latent_factor_model import BinaryLatentFactorApproximation
+from src.models.boltzmann_machine import BoltzmannMachine
+
+from typing import List
+
+
+class MessagePassing(BinaryLatentFactorApproximation):
+    """
+    eta_matrix: off diagonal matrix of parameters eta_matrix[n, i, j] corresponds to \tilda{g}_{ij, \neg s_i}(s_j) for
+                data point n
+                (number_of_points, number_of_latent_variables, number_of_latent_variables)
+    """
+
+    def __init__(self, eta_matrix: np.ndarray):
+        self.eta_matrix = eta_matrix
+
+    @property
+    def lambda_matrix(self):
+        return 1 / (1 + np.exp(-self.xi.sum(axis=1)))
+
+    @property
+    def xi(self):
+        return np.log(np.divide(self.eta_matrix, 1 - self.eta_matrix))
+
+    def aggregate_incoming_binary_factor_messages(
+        self, node_index: int, excluded_node_index: int
+    ) -> np.ndarray:
+        # (number_of_points, )
+        #  exclude message from excluded_node_index -> node_index
+        return (
+            np.sum(self.xi[:, :excluded_node_index, node_index], axis=1)
+            + np.sum(self.xi[:, excluded_node_index + 1 :, node_index], axis=1)
+        ).reshape(
+            -1,
+        )
+
+    def set_xi(self, i, j, value):
+        eta_values = 1 / (1 + np.exp(-value))
+        eta_values[eta_values == 0] = 1e-10
+        eta_values[eta_values == 1] = 1 - 1e-10
+        self.eta_matrix[:, i, j] = eta_values
+
+    def variational_expectation_step(
+        self, x, binary_latent_factor_model: BoltzmannMachine
+    ) -> List[float]:
+        free_energy = [self.compute_free_energy(x, binary_latent_factor_model)]
+        for i in range(self.k):
+            for j in range(self.k):
+                self.update_message(
+                    binary_latent_factor_model,
+                    x,
+                    i,
+                    j,
+                )
+            free_energy.append(self.compute_free_energy(x, binary_latent_factor_model))
+        return free_energy
+
+    def update_message(
+        self,
+        boltzmann_machine: BoltzmannMachine,
+        x,
+        start_node: int,
+        end_node: int,
+    ):
[…]

### 2022-12-30 src/solutions/q6.py

@@ -1,0 +1,32 @@
+from src.generate_images import generate_images
+import matplotlib.pyplot as plt
+from src.models.binary_latent_factor_model import learn_binary_factors
+from src.models.boltzmann_machine import init_boltzmann_machine
+from src.models.message_passing import init_message_passing
+
+
+def run(x, k, em_iterations, save_path):
+    n = x.shape[0]
+    message_passing = init_message_passing(k, n)
+    boltzmann_machine = init_boltzmann_machine(x, message_passing)
+    message_passing, boltzmann_machine, free_energy = learn_binary_factors(
+        x=x,
+        em_iterations=em_iterations,
+        binary_latent_factor_model=boltzmann_machine,
+        binary_latent_factor_approximation=message_passing,
+    )
+    fig, ax = plt.subplots(1, k, figsize=(k * 2, 2))
+    for i in range(k):
+        ax[i].imshow(boltzmann_machine.mu[:, i].reshape(4, 4))
+        ax[i].set_title(f"Latent Feature mu_{i}")
+    fig.suptitle("Learned Features (Loopy BP)")
+    plt.tight_layout()
+    plt.savefig(save_path + "-latent-factors", bbox_inches="tight")
+    plt.close()
+
+    plt.title("Free Energy (Loopy BP)")
+    plt.xlabel("t (EM steps)")
+    plt.ylabel("Free Energy")
+    plt.plot(free_energy)
+    plt.savefig(save_path + "-free-energy", bbox_inches="tight")
+    plt.close()

### 2022-12-30 main.py
update images for 3 and 6

@@ -110,8 +110,8 @@ if __name__ == "__main__":
     number_of_images = 1000
     x = generate_images(n=number_of_images)
     k = 8
-    em_iterations = 1000
-    e_maximum_steps = 200
+    em_iterations = 200
+    e_maximum_steps = 100
     e_convergence_criterion = 0
 
     binary_latent_factor_model = q3.e_and_f(
@@ -137,5 +137,4 @@ if __name__ == "__main__":
     Q6_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q6")
     if not os.path.exists(Q6_OUTPUT_FOLDER):
         os.makedirs(Q6_OUTPUT_FOLDER)
-    em_iterations = 50
     q6.run(x, k, em_iterations, save_path=os.path.join(Q6_OUTPUT_FOLDER, "all"))

### 2022-12-30 submission.tex
update images for 3 and 6
update vb question

@@
 
-We choose the conjugate priors:
-
-\[P( \textbf{A}| \eta) = \mathcal{N}(\textbf{A}| \mu_{\textbf{A}}, \Sigma_{\textbf{A}}) =\frac{1}{(2\pi)^{\frac{DK}{2}} |\Sigma_{\textbf{A}}|^{\frac{1}{2}}} \exp\left(-\frac{1}{2}(\textbf{A} - \mu_{\textbf{A}})^T \Sigma_{\textbf{A}}^{-1}(\textbf{A} - \mu_{\textbf{A}})\right)\]
-
-\[P( \Psi| \eta) = \prod_{d=1}^D InvGamma(\Psi_{dd}| a_d, b_d) = \prod_{d=1}^D \frac{b_d^{a_d}}{\Gamma(a_d)}\Psi_{dd}^{-a_d-1}\exp(-\frac{b_d}{\Psi_{dd}})\]
-
-a Gaussian prior on $\textbf{A}$ and a product of inverse gamma distributions on $\Psi$ where we assume $\Psi$ is a diagonal matrix.
+For $\textbf{A}$ we choose a factorised conjugate prior:
+
+\[P( \textbf{A}| \eta) = \prod_{k=1}^{K}P( \textbf{A}_{: k}| \eta)\]
+
+where $\textbf{A}_{: k} \in \mathbb{R}^{D \times 1}$ is the $k^{th}$ column of $\textbf{A}$. For each column we choose:
+
+\[P( \textbf{A}_{: k}| \eta) = \mathcal{N}(\textbf{A}_{: k}| \mu_{\textbf{A}_{: k}}, \Sigma_{\textbf{A}_{: k}}) =\frac{1}{(2\pi)^{\frac{D}{2}} |\Sigma_{\textbf{A}_{: k}}|^{\frac{1}{2}}} \exp\left(-\frac{1}{2}(\textbf{A}_{: k} - \mu_{\textbf{A}_{: k}})^T \Sigma_{\textbf{A}_{: k}}^{-1}(\textbf{A}_d - \mu_{\textbf{A}_d})\right)\]
+
+a Gaussian prior with diagonal covariance $\Sigma_{\textbf{A}_{: k}} = \alpha_k^{2}\textbf{I}$ and mean zero, so we can simplify:
+
+\[P( \textbf{A}_{: k}| \alpha_k) = (2\pi\alpha_k^{2})^{\frac{-D}{2}} \exp\left(-\frac{\textbf{A}_{: k}^T \textbf{A}_{: k}}{2\alpha_k^{2}} \right)\]
+
+For $\Psi}$ we choose a conjugate prior:
+
+\[P( \Psi| \eta) = \prod_{d=1}^D InvGamma(\Psi_{dd}| a, b) = \prod_{d=1}^D \frac{b^{a}}{\Gamma(a)}\Psi_{dd}^{-a-1}\exp(-\frac{b}{\Psi_{dd}})\]
+
+a product of inverse gamma distributions on $\Psi$ where we assume $\Psi$ is a diagonal matrix.
 
@@
 \tab \tab \tab \tab \tab + \sum_{i=1}^K (\alpha-1) \log \pi_k + (\beta - 1) \log (1-\pi_k) - \log B(\alpha, \beta)\\
-\tab \tab \tab \tab \tab -\frac{DK}{2} \log (2 \pi) - \frac{1}{2} \log(|\Sigma_{\textbf{A}}|)   -\frac{1}{2}(\textbf{A} - \mu_{\textbf{A}})^T \Sigma_{\textbf{A}}^{-1}(\textbf{A} - \mu_{\textbf{A}})\\
-\tab \tab \tab \tab \tab + \sum_{d=1}^D a_d \log b_d + (-a_d - 1) \log\Psi_{dd} - \frac{b_d}{\Psi_{dd}} - \log \Gamma(a_d)\\
+\tab \tab \tab \tab \tab + \sum_{i=1}^K -\frac{D}{2} \log (2 \pi\alpha_k^2)  -\frac{\textbf{A}_{: k}^T \textbf{A}_{: k}}{2\alpha_k^{2}}\\
+\tab \tab \tab \tab \tab + \sum_{d=1}^D a \log b + (-a - 1) \log\Psi_{dd} - \frac{b}{\Psi_{dd}} - \log \Gamma(a)\\
 \end{array}
@@
 \[
-q_{\Psi}(\Psi) \propto \left( \prod_{d=1}^D \frac{b_d^{a_d}}{\Gamma(a_d)}\Psi_{dd}^{-a_d-1}\exp(-\frac{b_d}{\Psi_{dd}}) \right)\exp \left\langle  - \frac{1}{2} \sum_{d=1}^{D}\log\Psi_{dd}   -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \Psi(\theta)}}
+q_{\Psi}(\Psi) \propto \left( \prod_{d=1}^D \frac{b^{a}}{\Gamma(a)}\Psi_{dd}^{-a-1}\exp(-\frac{b}{\Psi_{dd}}) \right)\exp \left\langle  - \frac{1}{2} \sum_{d=1}^{D}\log\Psi_{dd}   -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \Psi(\theta)}}
 \]
@@
 \[
-q_{\Psi}(\Psi) \propto  \prod_{d=1}^D \frac{b_d^{a_d}}{\Gamma(a_d)}\Psi_{dd}^{-(a_d+\frac{1}{2})-1}\exp\left(-\frac{b_d+\frac{1}{2} \left\langle (x_d-\textbf{A}_d\textbf{s})^2 \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}_d}(\textbf{A}_d)}}{\Psi_{dd}}\right) }
-\]
-
-where $\textbf{A}_d \in \mathbb{R}^{1 \times K}$ is the $d^{th}$ row of $\textbf{A}$.
+q_{\Psi}(\Psi) \propto  \prod_{d=1}^D \frac{b^{a}}{\Gamma(a)}\Psi_{dd}^{-(a+\frac{1}{2})-1}\exp\left(-\frac{b+\frac{1}{2} \left\langle (x_d-\textbf{A}_{d :}\textbf{s})^2 \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}_{d :}}(\textbf{A}_{d :})}}{\Psi_{dd}}\right) }
+\]
+
+where $\textbf{A}_{d :} \in \mathbb{R}^{1 \times K}$ is the $d^{th}$ row of $\textbf{A}$.
 
@@
 \[
-q_{\Psi}(\Psi) = \prod_{d=1}^D  InvGamma\left(\Psi_{dd} \left| a_d+\frac{1}{2}, b_d+\frac{1}{2} \left\langle (x_d-\textbf{A}_d\textbf{s})^2 \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}_d}(\textbf{A}_d)}\right)
-\]
-
-For $q_{\textbf{A}}(\textbf{A})$:
-
-\[q_{\textbf{A}}(\textbf{A})  \propto P(\textbf{A})\exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}}(\theta)}\]
-
-\[q_{\textbf{A}}(\textbf{A}) \propto \exp\left(-\frac{1}{2}(\textbf{A} - \mu_{\textbf{A}})^T \Sigma_{\textbf{A}}^{-1}(\textbf{A} - \mu_{\textbf{A}})\right) \exp \left\langle -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}}(\theta)}\]
-
-\[q_{\textbf{A}}(\textbf{A}) \propto \exp\left(-\frac{1}{2}(\textbf{A}^T \Sigma_{\textbf{A}}^{-1}\textbf{A} -2 \textbf{A}^T \Sigma_{\textbf{A}}^{-1}  \mu_{\textbf{A}})\right) \exp \left\langle -\frac{1}{2}(\textbf{s}^T\textbf{A}^T \Psi^{-1} \textbf{A}\textbf{s} - 2 \textbf{s}^T\textbf{A}^T \Psi^{-1} \textbf{x}) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}}(\theta)}\]
+q_{\Psi}(\Psi) = \prod_{d=1}^D  InvGamma\left(\Psi_{dd} \left| a+\frac{1}{2}, b+\frac{1}{2} \left\langle (x_d-\textbf{A}_{d :}\textbf{s})^2 \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}_{d :}}(\textbf{A}_{d :})}\right)
+\]
+
+For $q_{\textbf{A}_{: k}}(\textbf{A}_{: k})$:
+
+\[q_{\textbf{A}_{: k}}(\textbf{A}_{: k})  \propto P(\textbf{A}_{: k})\exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}_{: k}}(\theta)}\]
+
+Substituting the appropriate terms:
+
+\[q_{\textbf{A}_{: k}}(\textbf{A}_{: k}) \propto \exp\left(-\frac{\textbf{A}_{: k}^T \textbf{A}_{: k}}{2\alpha_k^{2}} \right) \exp \left\langle -\frac{1}{2}(\textbf{x}-\textbf{A}_{: k} s_k)^T \Psi^{-1} (\textbf{x}-\textbf{A}_{: k} s_k) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}_{: k}}(\theta)}\]
+
+\[q_{\textbf{A}_{: k}}(\textbf{A}_{: k}) \propto \exp\left(-\frac{\textbf{A}_{: k}^T \textbf{A}_{: k}}{2\alpha_k^{2}} \right) \exp \left\langle -\frac{1}{2} \textbf{A}_{: k}^T\frac{1 }{ s_k^2 \Psi_{dd}}\textbf{A}_{: k}  - 2 \frac{s_k}{\Psi_{dd}} \textbf{x}^T\textbf{A}_{: k} \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}_{: k}}(\theta)}\]
+
+We see that
+
+\[q_{\textbf{A}_{: k}}(\textbf{A}_{: k}) = \mathcal{N}\left(\mu_{\textbf{A}_{: k}}, \Sigma_{\textbf{A}_{: k}}\right)}\]
+
+where:
+
+\[ \Sigma_{\textbf{A}_{: k}} = \left(\frac{1}{\alpha_k^2} + \frac{1}{s_k^2\Psi_{dd}}\right)^{-1} \textbf{I}\]
+
+and
+
+\[\mu_{\textbf{A}_{: k}} = \Sigma_{\textbf{A}_{: k}}^{-1} \frac{s_k}{\Psi_{dd}}\textbf{x}\]
+
+By optimising with respect to the the distributions $\Psi$ and $\alpha$ in turn causes some $\alpha_i^2$ to diverge, the number of remaining $\alpha_i^2$ provide our determination for the value of $K$.
+This is automatic relevance determination through factor analysis.
 
@@
 
-\newpage
-
 \subsection*{(c)}
@@
 \newpage
-\section{Question 6}
+\section*{Question 6}
 

### 2022-12-30 submission.tex
update gp question

@@
 \begin{minipage}{.5\textwidth}
+\includegraphics[scale = 0.35]{outputs/q2/c-parameter-samples}
+\caption{Samples for different parameters}
+\label{fig:fig2-c-parameter-samples}
+  \centering
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
   \centering
@@
 \label{fig:fig2-c-parameter-grams}
-\end{minipage}%
-\begin{minipage}{.5\textwidth}
-  \centering
-\includegraphics[scale = 0.35]{outputs/q2/c-parameter-samples}
-\caption{Samples for different parameters}
-\label{fig:fig2-c-parameter-samples}
 \end{minipage}
@@
     The covariance matrix shows how increasing $\theta$ visually reveals the striped periodic component.
-        This is expected because it is the parameter that adjusts the weight of $\exp\left( - \frac{2\sin^2(\pi(s-t)/\tau)}{\sigma^2}\right)$.
-\item [$\sigma$:] As $\sigma$ increases, we see smoother periodic behaviour in the sample function.
+        This is expected because it is the parameter that adjusts the weight of the periodic component.
+\item [$\sigma$:] As $\sigma$ increases, we see reduced periodic behaviour in the sample function.
     The covariance matrix shows how increasing $\sigma$ will increase covariance values in the off-diagonals.
-    This is expected because it adjusts the lengthscale of the periodic portion of the kernel.
-\item [$\phi$:] As $\phi$ increases, we see less smooth behaviour in the sample function.
-    The covariance matrix shows how increasing $\sigma$ will increase covariance values in the off-diagonals.
-    This is expected because it adjusts the lengthscale of the periodic portion of the kernel.
-\item [$\eta$:]
-\item [$\tau$:]
-\item [$\zeta$:]
+    This is expected because it adjusts the lengthscale of the periodic portion of the kernel, which ends up dominating the function.
+\item [$\phi$:] As $\phi$ increases, we see the ratio of the amplitude of the periodicity component of the sample function reduces compared to the baseline.
+    The covariance matrix shows how increasing $\phi$ will start to increase the non-periodic component.
+    This is expected because it adjusts the weight of the non-periodic portion of the kernel, thus the periodic component remains the same (i.e.same amplitude) but the large baseline shifts from increasing $\phi$ ends up dominating the function visually.
+\item [$\eta$:] As $\eta$ increases we see smoother sample functions.
+This is expected because the $\eta$ increases the lengthscale of the non-periodic component, allowing for smoother functions.
+This causes the off-diagonals of the gram matrix to increase, however the periodic component is still maintained because $\eta$ doesn't affect the relative weight of the two components.
+\item [$\tau$:] As $\tau$ increases, the period of the periodic function increases.
+We can see this reflected in the stripes in the gram matrix getting further apart.
+This makes sense because we are adjusting the period in the sinusoid function of the periodic term with $\tau$.
+\item [$\zeta$:] As $\zeta$ increases, the function becomes less smooth.
+    This is because the $\zeta$ parameter adjusts the weight of the $\delta_{s=t}$ parameter.
+    This places stronger emphasis on the independence of each timestep, which can be seen with the reduction of relative magnitude of off-diagonals in the gram matrix.
+    However, this is simply masking the periodic and squared-exponential terms as we can see with the increased magnitude of the functions as $\zeta$ increases.
 \end{itemize}
@@
 \subsection*{(e)}
-
+Suitable values for hyper-parameters can be chosen through a combination of visual inspection and prior knowledge.
+For example, it is a reasonable assumption that the $CO_2$ concentration levels have a strong yearly seasonality behaviour due to the cyclic changes in temperature, humidity, etc.
+Thus we can choose $\tau=1$ to ensure functions with a period of one year to reflect this knowledge.
+It can be difficult to quantitatively choose values for the other parameters as they can relate to the uncertainty exhibited in the data (i.e.the smoothness of the function).
+One approach is to maximise:
+
+\[\log P(\textbf{Y} | \textbf{X}) = -\frac{1}{2} \textbf{Y}^T (\textbf{K} + \sigma^2\textbf{I})^{-1} \textbf{Y} - \frac{1}{2} \log |\textbf{K}+ \sigma^2\textbf{I}| - \frac{n}{2}\log(2\pi)\]
+
+the log-likelihood  of the posterior distribution with respect to the given data where $\textbf{K}$ is the gram matrix for the kernel (equation 2.30 from http://gaussianprocess.org/gpml/chapters/RW2.pdf).
+We can define a loss function as the negative log-likelihood and employ gradient-based algorithms to find optimal parameters.
+
+Comparing the hyperparameters corresponding to before and after training side by side:
+
+\begin{figure}[h]
+\centering
+\begin{minipage}{.5\textwidth}
+  \centering
+\includegraphics[scale=0.5]{outputs/q2/f-untrained-parameters}
+\caption{Untrained hyperparameters}
+\label{fig:f-untrained-parameters}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+\includegraphics[scale=0.5]{outputs/q2/f-trained-parameters}
+\caption{Trained Hyperparmaeters}
+\label{fig:f-trained-parameters}
+\end{minipage}
+\end{figure}
+
+We can analyse some of the changes in these parameters after training to gain some insights.
+We can see that $\tau$ remains the same as we would expect given the yearly seasonality we have prior knowledge of.
+On the other hand, the value for $\zeta$ is significantly reduced signifying that $\delta_{s=t}$ is not a very good kernel for representing the data as datapoints at different timesteps do exhibit correlations.
+
+\newpage
 \subsection*{(f)}
+Extrapolating the $CO_2$ concentration levels:
 
@@
 \includegraphics[scale=0.5]{outputs/q2/f-extrapolation-untrained}
-\caption{Without hyperparameter tuning}
+\caption{Untrained extrapolation}
 \label{fig:f-extrapolation-untrained}
@@
 \includegraphics[scale=0.5]{outputs/q2/f-extrapolation-trained}
-\caption{With hyperparameter tuning}
+\caption{Trained extrapolation}
 \label{fig:f-extrapolation-trained}
@@
 \end{figure}
-%
-%\begin{figure}[h]
-%\centering
-%\includegraphics[scale=0.5]{outputs/q2/f-extrapolation-untrained}
-%\caption{Extrapolation of $CO_2$ concentration levels}
-%\label{fig:f-extrapolation-untrained}
-%\end{figure}
-%
-%\begin{figure}[h]
-%\centering
-%\includegraphics[scale=0.5]{outputs/q2/f-extrapolation-trained}
-%\caption{Extrapolation of $CO_2$ concentration levels}
-%\label{fig:f-extrapolation-trained}
-%\end{figure}
+
+We can see that the extrapolation shows a continued increase in $CO_2$ in the future.
+This follows our expectations given that the levels has been steadily increasing in the past.
+Moreover, the concentration continues to exhibit yearly seasonality (for the trained extrapolation) as we would expect.
+We can see that the conclusions can be quite sensitive to kernel hyperparameters when comparing the values from before and after training.
+Prior to training, the extrapolated prediction is not representative of the given data, with pretty much no seasonal behaviour and very large uncertainty.
+After training, we can see that the prediction is much more reasonable, and qualitatively the uncertainty bounds seem to exhibit the historical variability in the data.
 
 \subsection*{(g)}
+
+This procedure is not fully Bayesian because despite using a posterior estimate of our linear regression terms, we only use a point estimate when making prediction.
+For a fully Bayesian approach, we should also incorporate the uncertainty of the linear regression parameters into our extrapolation/uncertainty bounds.
+For our procedure, we only include the uncertainty of $g(t)$ however it can be observed in the plots that the trend is not perfectly linear so this should be reflected in the uncertainty of our extrapolation.
+Another approach could be to add a linear kernel to our combined kernel function and model $f(t)$ directly with our kernel, removing the linear regression component in our procedure.
+Thus our kernel extrapolation would incorporate the uncertainty of all components of our signal.
 

### 2022-12-30 main.py

@@ -76,14 +76,22 @@ if __name__ == "__main__":
     q2.c(
         kernel=kernel,
         kernel_parameters=kernel_parameters,
-        log_theta_range=jnp.log(jnp.linspace(1e-1, 2, 5)),
+        log_theta_range=jnp.log(jnp.linspace(1e-2, 5, 5)),
         t=t[:50].reshape(-1, 1),
         number_of_samples=3,
         save_path=os.path.join(Q2_OUTPUT_FOLDER, "c"),
     )
 
+    init_kernel_parameters = CombinedKernelParameters(
+        log_theta=jnp.log(1e-1),
+        log_sigma=jnp.log(1),
+        log_phi=jnp.log(5),
+        log_eta=jnp.log(1e-1),
+        log_tau=jnp.log(1),
+        log_zeta=jnp.log(1e-1),
+    )
     gaussian_process_parameters = GaussianProcessParameters(
-        kernel=asdict(kernel_parameters),
+        kernel=asdict(init_kernel_parameters),
         log_sigma=jnp.log(1),
     )
     years_to_predict = 15
@@ -99,42 +107,42 @@ if __name__ == "__main__":
         kernel=kernel,
         gaussian_process_parameters=gaussian_process_parameters,
         learning_rate=1e-2,
-        number_of_iterations=100,
+        number_of_iterations=500,
         save_path=os.path.join(Q2_OUTPUT_FOLDER, "f"),
     )
 
-    # Question 3
-    Q3_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q3")
-    if not os.path.exists(Q3_OUTPUT_FOLDER):
-        os.makedirs(Q3_OUTPUT_FOLDER)
-    number_of_images = 1000
-    x = generate_images(n=number_of_images)
-    k = 8
-    em_iterations = 200
-    e_maximum_steps = 100
-    e_convergence_criterion = 0
-
-    binary_latent_factor_model = q3.e_and_f(
-        x=x,
-        k=k,
-        em_iterations=em_iterations,
-        e_maximum_steps=e_maximum_steps,
-        e_convergence_criterion=e_convergence_criterion,
-        save_path=os.path.join(Q3_OUTPUT_FOLDER, "f"),
-    )
-    q3.g(
-        x=x[:1, :],
[…]

### 2022-12-30 src/models/message_passing.py

@@ -17,7 +17,10 @@ class MessagePassing(BinaryLatentFactorApproximation):
 
     @property
     def lambda_matrix(self):
-        return 1 / (1 + np.exp(-self.xi.sum(axis=1)))
+        lambda_matrix = 1 / (1 + np.exp(-self.xi.sum(axis=1)))
+        lambda_matrix[lambda_matrix == 0] = 1e-10
+        lambda_matrix[lambda_matrix == 1] = 1 - 1e-10
+        return lambda_matrix
 
     @property
     def xi(self):

### 2022-12-30 src/solutions/q2.py

@@ -244,11 +244,38 @@ def f(
     plt.savefig(save_path + "-extrapolation-untrained", bbox_inches="tight")
     plt.close()
 
+    df_parameters = pd.DataFrame(
+        [
+            [
+                x.strip("log_") + " (kernel)",
+                np.exp(gaussian_process_parameters.kernel[x]),
+            ]
+            for x in gaussian_process_parameters.kernel.keys()
+        ]
+        + [["sigma", float(gaussian_process_parameters.sigma)]],
+        columns=["parameter", "value"],
+    )
+    df_parameters = df_parameters.set_index("parameter")
+    dfi.export(df_parameters, save_path + "-untrained-parameters.png")
+
     # Train Gaussian Process Regression (Hyperparameter Tune)
     optimizer = optax.adam(learning_rate)
     gaussian_process_parameters = gaussian_process.train(
         optimizer, number_of_iterations, **asdict(gaussian_process_parameters)
     )
+    df_parameters = pd.DataFrame(
+        [
+            [
+                x.strip("log_") + " (kernel)",
+                np.exp(gaussian_process_parameters.kernel[x]),
+            ]
+            for x in gaussian_process_parameters.kernel.keys()
+        ]
+        + [["sigma", float(gaussian_process_parameters.sigma)]],
+        columns=["parameter", "value"],
+    )
+    df_parameters = df_parameters.set_index("parameter")
+    dfi.export(df_parameters, save_path + "-trained-parameters.png")
 
     # Prediction
     x_test = construct_design_matrix(t_test)

### 2022-12-30 submission.tex
update mean field

@@
 
+\newpage
 \subsection*{(e)}
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.4]{outputs/q3/f-free-energy}
+\caption{Free Energy}
+\label{fig:3f-free-energy}
+\end{figure}
+
 \subsection*{(f)}
 
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.4]{outputs/q3/f-init-latent-factors}
+\caption{Initial Latent Factors}
+\label{fig:3f-init-latent-factors}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.4]{outputs/q3/f-latent-factors}
+\caption{Learned Latent Factors}
+\label{fig:3f-latent-factors}
+\end{figure}
+
 \subsection*{(g)}
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.4]{outputs/q3/g-free-energy-diff-sigma}
+\caption{Free energy convergence vs $\sigma$}
+\label{fig:3g-free-energy-diff-sigma}
+\end{figure}
 
@@
 
-%\[\mathcal{M}_{j \rightarrow i}(s_i) = \tilde{g}_{ji, \neg s_j}(s_i)\]
-
-%\[\mathcal{M}_{j \rightarrow i} = \tilde{f}_j(s_j) \prod_{k\in ne(j), k\neq i}^{K}}\tilde{g}_{kj, \neg s_k}(s_j)  \sum_{\substack{s_m \in \{0, 1\} \\ m \in \{1, ..., j-1, j+1,...
-K\}}} \prod_{k\in ne(j), k\neq i}^{K}} \tilde{g}_{kj, \neg s_j}(s_k)\]
-
-%\[\mathcal{M}_{j \rightarrow i}(s_j) = \tilde{f}_j(s_j) \prod_{k\in ne(j), k\neq i}^{K}}\tilde{g}_{kj, \neg s_k}(s_j) \prod_{k\in ne(j), k\neq i}^{K}} \left(\sum_{s_k \in \{0, 1\} } \tilde{g}_{kj, \neg s_j}(s_k) \right)\]
-%
-%\[\mathcal{M}_{j \rightarrow i}(s_j) = \tilde{f}_j(s_j) \prod_{k\in ne(j), k\neq i}^{K}}\tilde{g}_{kj, \neg s_k}(s_j)\]
-
 Thus, the cavity distributions are:
@@
 
-where:
-
-\[\eta_{i, \neg s_j} =  \log \left(\frac{\lambda_i}{1-\lambda_i}\right) + \sum_{k\in ne(i), k\neq j}^{K} \log\left(\frac{\theta_{ki}}{1-\theta_{ki}} \right)\]
-
-Knowing $b_i=\log \left(\frac{\lambda_i}{1-\lambda_i}\right)$ and $\xi_{ki}=\log\left(\frac{\theta_{ki}}{1-\theta_{ki}} \right)$:
-
-\[\eta_{i, \neg s_j} =  b_i + \sum_{k\in ne(i), k\neq j}^{K} \xi_{ki}\]
-
-and
-
-\[\eta_{j, \neg s_i} =  b_j + \sum_{k\in ne(j), k\neq i}^{K} \xi_{kj}\]
-
 Note that $\tilde{g}_{ij}(s_i, s_j)$ was chosen as the product of two Bernoulli distributions, updates to this site approximation involves updating the  parameters $\xi_{ij}$ and $\xi_{ji}$, for $s_i$ and $s_j$ respectively.
@@
 
-%\[
-%\begin{array}{l}
-% \log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j\\
-%\tab \tab \tab \tab \tab + \sum_{i'=1}^{K}  \log \left(\frac{\lambda_{i'}}{1-\lambda_{i'}} \right) s_{i'} \\
-%\tab \tab \tab \tab \tab +  \sum_{i'=1}^{K}\sum_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1}  \log \left(\frac{\theta_{i'}}{1-\theta_{i'}} \right) s_{i'} + \log \left(\frac{\theta_{j'}}{1-\theta_{j'}} \right) s_{j'}
-%\end{array}
-%\]
-
-%Only including terms with $s_i$ and $s_j$:
-%
-%\[
-%\begin{array}{l}
-% \log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \left((K-1) \log \left(\frac{\theta_i}{1-\theta_i} \right) + \log \left(\frac{\lambda_i}{1-\lambda_i} \right)\right) s_i\\
-%\tab \tab \tab \tab \tab +\left(
-%(K-1) \log \left(\frac{\theta_j}{1-\theta_j} \right) + \log \left(\frac{\lambda_j}{1-\lambda_j} \right)\right) s_j
-%\end{array}
-%\]
-
 Thus, the first moments:
@@
 
-%\[
-%\begin{array}{l}
-% \log g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \frac{- \mu_i^T \mu_j}{\sigma^2} s_i s_j\\
-%\tab \tab \tab \tab \tab + \sum_{i'=1}^{K}  \log \left(\frac{\lambda_{i'}}{1-\lambda_{i'}} \right) s_{i'} \\
-%\tab \tab \tab \tab \tab +  \sum_{i'=1}^{K}\sum_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1}  \log \left(\frac{\theta_{i'}}{1-\theta_{i'}} \right) s_{i'} + \log \left(\frac{\theta_{j'}}{1-\theta_{j'}} \right) s_{j'}
-%\end{array}
-%\]
-
-%Only including terms with $s_i$ and $s_j$:
-%
-%\[
-%\begin{array}{l}
-% \log g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto \frac{- \mu_i^T \mu_j}{\sigma^2} s_i s_j\\
-%\tab \tab \tab \tab \tab + \left((K-2)\log \left(\frac{\theta_i}{1-\theta_i} \right) + \log \left(\frac{\lambda_i}{1-\lambda_i} \right)\right) s_i \\
-%\tab \tab \tab \tab \tab +  \left((K-2)\log \left(\frac{\theta_j}{1-\theta_j} \right) + \log \left(\frac{\lambda_j}{1-\lambda_j} \right)\right) s_j
-%\end{array}
-%\]
-%
-%We can write this as:
-%
-%\[
-%\log g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(\textbf{s}) \propto W_{ij} s_i s_j+ A_{s_i} s_i + B_{s_j} s_j
-%\]
-%
-%where:
-%
-%\[A_{s_i} = (K-2)\log \left(\frac{\theta_i}{1-\theta_i} \right) + \log \left(\frac{\lambda_i}{1-\lambda_i} \right)\]
-%
-%and
-%
-%\[B_{s_j} = (K-2)\log \left(\frac{\theta_j}{1-\theta_j} \right) + \log \left(\frac{\lambda_j}{1-\lambda_j} \right)\]
-
 To derive the first moment for $ g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)$ with respect to $s_i$, we first marginalise out $s_j$:
@@
 
-Our message passing approximations:
-
-\[\exp(\eta_{ij} s_i) = \tilde{f}_{i}(s_{i}) \prod_{k\in ne(i), k\neq j}^{K} \mathcal{M}_{k \rightarrow i} \]
-
-\[\exp(\eta_{ji} s_j) = \tilde{f}_{j}(s_{j}) \prod_{k\in ne(j), k\neq i}^{K} \mathcal{M}_{k \rightarrow j}\]
-
-where each message $\mathcal{M}_{j \rightarrow i}$ has a factored approximation:
-
-\[\mathcal{M}_{k \rightarrow i} = \exp(\eta_{ki} s_k) \]
-
-because each site $\tilde{g}_{jk}(s_j s_k)$ is approximated as a product of two messages $\mathcal{M}_{j \rightarrow k}\mathcal{M}_{k \rightarrow j}$, each a Bernoulli.
-
-Thus, the natural parameters of the messages are updated with:
-
-\[\eta_{ij} = b_i + \sum_{k\in ne(i), k\neq j}^{K} \eta_{ki} \]
+Using factored approximate messages, we see that:
+
+\[\eta_{i, \neg s_j} =  \log \left(\frac{\lambda_i}{1-\lambda_i}\right) + \sum_{k\in ne(i), k\neq j}^{K} \log\left(\frac{\theta_{ki}}{1-\theta_{ki}} \right)\]
+
+Knowing $b_i=\log \left(\frac{\lambda_i}{1-\lambda_i}\right)$ and $\xi_{ki}=\log\left(\frac{\theta_{ki}}{1-\theta_{ki}} \right)$:
+
+\[\eta_{i, \neg s_j} =  b_i + \sum_{k\in ne(i), k\neq j}^{K} \xi_{ki}\]
+
+and
+
+\[\eta_{j, \neg s_i} =  b_j + \sum_{k\in ne(j), k\neq i}^{K} \xi_{kj}\]
 
@@
 \centering
+\includegraphics[scale=0.4]{outputs/q6/all-init-latent-factors}
+\caption{Initial Latent factors learned with EP/Loopy-BP}
+\label{fig:6-init-latent-factors}
+\end{figure}
+
+\begin{figure}[h]
+\centering
 \includegraphics[scale=0.4]{outputs/q6/all-latent-factors}
-\caption{Latent factors learned with EP/Loopy-BP}
+\caption{Learned Latent factors learned with EP/Loopy-BP}
 \label{fig:6-latent-factors}

### 2022-12-30 src/models/mean_field_learning.py

@@ -99,7 +99,11 @@ class MeanFieldApproximation(BinaryLatentFactorApproximation):
                 self.lambda_matrix[:, latent_factor] = self._partial_expectation_step(
                     x, binary_latent_factor_model, latent_factor
                 )
-            free_energy.append(self.compute_free_energy(x, binary_latent_factor_model))
+                free_energy.append(
+                    self.compute_free_energy(x, binary_latent_factor_model)
+                )
+                if free_energy[-1] - free_energy[-2] <= self.convergence_criterion:
+                    break
             if free_energy[-1] - free_energy[-2] <= self.convergence_criterion:
                 break
         return free_energy

### 2022-12-30 src/solutions/q3.py

@@ -27,6 +27,14 @@ def e_and_f(
     binary_latent_factor_model = init_binary_latent_factor_model(
         x, mean_field_approximation
     )
+    fig, ax = plt.subplots(1, k, figsize=(k * 2, 2))
+    for i in range(k):
+        ax[i].imshow(binary_latent_factor_model.mu[:, i].reshape(4, 4))
+        ax[i].set_title(f"Latent Feature mu_{i}")
+    fig.suptitle("Initial Features (Mean Field Learning)")
+    plt.tight_layout()
+    plt.savefig(save_path + "-init-latent-factors", bbox_inches="tight")
+    plt.close()
     _, binary_latent_factor_model, free_energy = learn_binary_factors(
         x,
         em_iterations,
@@ -75,12 +83,19 @@ def g(
             mean_field_approximation.compute_free_energy(x, binary_latent_factor_model)
         ]
         for _ in range(em_iterations):
+            free_energy.pop(-1)
             previous_lambda_matrix = np.copy(mean_field_approximation.lambda_matrix)
             new_free_energy = mean_field_approximation.variational_expectation_step(
                 binary_latent_factor_model=binary_latent_factor_model,
                 x=x,
             )
             free_energy.extend(new_free_energy)
+            if (
+                free_energy[-1] - free_energy[-2]
+                <= mean_field_approximation.convergence_criterion
+            ):
+                free_energy.pop(-1)
+                break
             if is_converge(
                 free_energy,
                 mean_field_approximation.lambda_matrix,
@@ -90,9 +105,9 @@ def g(
         free_energies.append(free_energy)
 
     for i, free_energy in enumerate(free_energies):
+        diffs = np.log(np.diff(free_energy))
         plt.plot(
-            np.arange(len(free_energy) - 1),
-            np.log(np.diff(np.array(free_energy))),
+            diffs,
             label=f"sigma={sigmas[i]}",
         )
     plt.title(f"log(F(t)-F(t-1)")

### 2022-12-30 src/solutions/q6.py

@@ -9,6 +9,14 @@ def run(x, k, em_iterations, save_path):
     n = x.shape[0]
     message_passing = init_message_passing(k, n)
     boltzmann_machine = init_boltzmann_machine(x, message_passing)
+    fig, ax = plt.subplots(1, k, figsize=(k * 2, 2))
+    for i in range(k):
+        ax[i].imshow(boltzmann_machine.mu[:, i].reshape(4, 4))
+        ax[i].set_title(f"Latent Feature mu_{i}")
+    fig.suptitle("Initial Features (Loopy BP)")
+    plt.tight_layout()
+    plt.savefig(save_path + "-init-latent-factors", bbox_inches="tight")
+    plt.close()
     message_passing, boltzmann_machine, free_energy = learn_binary_factors(
         x=x,
         em_iterations=em_iterations,

### 2022-12-31 submission.tex
attempt vb and update report

@@
 
-\newpage
+Examining the generated images, we can see eight features:
+\begin{enumerate}[\hspace{2cm}(1)]
+\item a cross
+\item a border
+\item a two by two square in the middle
+\item a two by two square in the bottom left corner
+\item a diagonal from top left to bottom right
+\item a vertical line in the second column
+\item a vertical line in the fourth column
+\item a a horizontal line in the first row
+\end{enumerate}
+
+Factor analysis assumes a model:
+
+\[\textbf{x} = \textbf{W}\textbf{s} + \epsilon\]
+
+where $\epsilon \sim \mathcal{N}(\mu_{\epsilon}, \Sigma_{\epsilon})$ and $\textbf{s} \sim \mathcal{N}(\mu_{\textbf{s}}, \Sigma_{\textbf{s}})$.
+Factor analysis would be inappropriate for this data because the our latent variables are binary (i.e.
+whether or not a feature is present) and not Gaussians.
+Moreover, the presence of each feature is independent of the presence of another which is not enforced in this model with a covariance matrix that might not be diagonal.
+
+A mixture of Gaussians assumes as model:
+
+\[\textbf{x} = \sum_{k=1}^K\pi_k \mu_k + \epsilon\]
+
+where $\epsilon \sim \mathcal{N}(0, \Sigma_{\epsilon})$.
+This also wouldn't be appropriate because each mixture component (feature) is assumed to have some covariance, whereas our mixtures are defined as binary vectors (a cross, a border, etc) and added together before adding some noise.
+
+The independent component analysis assumes a model:
+
+\[\textbf{x} = \textbf{W}\textbf{s} + \epsilon\]
+
+where $\epsilon \sim \mathcal{N}(0, \sigma^2\textbf{I})$ and $p(\textbf{s}) = \prod_{k=1}^K p(s_k)$.
+This is appropriate for our data because we are linearly combining different features and then adding noise.
+
+Thus, it would be expected that ICA does a good job modelling this data while factor analysis and mixture of Gaussians would not.
+
 \subsection*{(e)}
@@
 \end{figure}
+\newpage
 
 \subsection*{(f)}
+
+The initialised features:
 
@@
 
+The features learned by the algorithm:
+
 \begin{figure}[h]
@@
 
+We can see that it has learned some of previously identified features, such as the vertical line in the second column, the two by two square in the bottom left corner, the border, and the a two by two square in the middle.
+THe other features seem to be some linear combination of two or more features, such as $\mu_4$ which looks like a combination of the cross and two by two square in the middle.
+
+A possible way to improve our algorithm is reinitialising our algorithm a few times to find better potential convergence results (i.e. choose model with best free energy).
+Another way to improve the algorithm could be to increase the $K$, although it may learn some duplicate features, there is also a higher chance of capturing all the features. We can visualise this:
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.25]{outputs/q3/f-larger-k-latent-factors}
+\caption{Increasing Number of Latent Factors}
+\label{fig:3f-larger-k-latent-factors}
+\end{figure}
+
+Here we can identify a few more features such as the vertical line in the fourth column the cross, and some of the diagonal feature in $\mu_7$.
+
+When implementing the algorithm, the mean field parameters were initialised randomly, each independently from a uniform distribution.
+However $\pi$, $\sigma$, and $\mu$ by running the maximisation step using the randomly initialised mean field parameters.
+$K$ was set to eight, after visually identifying eight features in part d.
+
+\newpage
 \subsection*{(g)}
 
-\begin{figure}[h]
+Plotting the convergence of the variational approximation for different $\sigma$'s:
+
+\begin{figure}[h]
+\centering
+\begin{minipage}{.5\textwidth}
+  \centering
+\includegraphics[scale=0.4]{outputs/q3/g-free-energy-sigma}
+\caption{Free energy vs $\sigma$}
+\label{fig:3g-free-energy-diff-sigma}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
   \centering
@@
 \label{fig:3g-free-energy-diff-sigma}
-\end{figure}
+\end{minipage}
+\end{figure}
+
+We can see that when $\sigma$ is smaller, we are able to converge at a higher free energy. However, convergence is only reached after more steps, as seen in the plot of $\log(F(t)-F(t-1))$
 
@@
 The Python code for mean field learning:
-\lstinputlisting[language=Python]{src/models/mean_field_learning.py}
+\lstinputlisting[language=Python]{src/models/mean_field_approximation.py}
 
@@
 
-where $\eta$ is a collection of all hyperparameters.
+where $\textbf{A} \in \mathbb{R}^{D \times K}$ is a concatenation of $\mu_k$ for $k \in \{1, ..., K\}$  and \eta$ is a collection of all hyperparameters.
 
@@
 
-For $\textbf{A}$ we choose a factorised conjugate prior:
-
-\[P( \textbf{A}| \eta) = \prod_{k=1}^{K}P( \textbf{A}_{: k}| \eta)\]
-
-where $\textbf{A}_{: k} \in \mathbb{R}^{D \times 1}$ is the $k^{th}$ column of $\textbf{A}$. For each column we choose:
-
-\[P( \textbf{A}_{: k}| \eta) = \mathcal{N}(\textbf{A}_{: k}| \mu_{\textbf{A}_{: k}}, \Sigma_{\textbf{A}_{: k}}) =\frac{1}{(2\pi)^{\frac{D}{2}} |\Sigma_{\textbf{A}_{: k}}|^{\frac{1}{2}}} \exp\left(-\frac{1}{2}(\textbf{A}_{: k} - \mu_{\textbf{A}_{: k}})^T \Sigma_{\textbf{A}_{: k}}^{-1}(\textbf{A}_d - \mu_{\textbf{A}_d})\right)\]
-
-a Gaussian prior with diagonal covariance $\Sigma_{\textbf{A}_{: k}} = \alpha_k^{2}\textbf{I}$ and mean zero, so we can simplify:
-
-\[P( \textbf{A}_{: k}| \alpha_k) = (2\pi\alpha_k^{2})^{\frac{-D}{2}} \exp\left(-\frac{\textbf{A}_{: k}^T \textbf{A}_{: k}}{2\alpha_k^{2}} \right)\]
+For $\textbf{A}$ we choose a factorised conjugate prior for each column, $\mu_k$:
+
+\[P( \textbf{A}| \eta) = \prod_{k=1}^{K}P( \mu_{k}| \eta)\]
+
+For each column we choose:
+
+\[P( \mu_{k}| \eta) = \mathcal{N}(\mu_{k}| \mu_{\mu_{k}}, \Sigma_{\mu_{k}}) =\frac{1}{(2\pi)^{\frac{D}{2}} |\Sigma_{\mu_{k}}|^{\frac{1}{2}}} \exp\left(-\frac{1}{2}(\mu_{k}} - \mu_{\mu_{k}})^T \Sigma_{\mu_{k}}^{-1}(\mu_{k} - \mu_{\mu_{k}})\right)\]
+
+a Gaussian prior with diagonal covariance $\Sigma_{\mu_{k}} = \alpha_k^{2}\textbf{I}$ and mean zero, so we can simplify:
+
+\[P( \mu_{k}| \alpha_k) = (2\pi\alpha_k^{2})^{\frac{-D}{2}} \exp\left(-\frac{\mu_{k}^T \mu_{k}}{2\alpha_k^{2}} \right)\]
 
@@
 
-\[P( \Psi| \eta) = \prod_{d=1}^D InvGamma(\Psi_{dd}| a, b) = \prod_{d=1}^D \frac{b^{a}}{\Gamma(a)}\Psi_{dd}^{-a-1}\exp(-\frac{b}{\Psi_{dd}})\]
-
-a product of inverse gamma distributions on $\Psi$ where we assume $\Psi$ is a diagonal matrix.
+\[P( \Psi| \eta) = \prod_{d=1}^D InvGamma(\sigma^2| a, b) = \prod_{d=1}^D \frac{b^{a}}{\Gamma(a)}(\sigma^2)^{-a-1}\exp(-\frac{b}{\sigma^2})\]
+
+a product of inverse gamma distributions, assuming $\Psi$ is a diagonal matrix with elements \sigma^2.
 
@@
 \begin{array}{l}
-\log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) = -\frac{D}{2} \log (2 \pi) - \frac{1}{2} \sum_{d=1}^{D}\log\Psi_{dd}   -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \\
-\tab \tab \tab \tab \tab + \sum_{k=1}^{K} s_k \log \pi_k + (1-s_k) \log (1-\pi_k)}\\
-\tab \tab \tab \tab \tab + \sum_{i=1}^K (\alpha-1) \log \pi_k + (\beta - 1) \log (1-\pi_k) - \log B(\alpha, \beta)\\
-\tab \tab \tab \tab \tab + \sum_{i=1}^K -\frac{D}{2} \log (2 \pi\alpha_k^2)  -\frac{\textbf{A}_{: k}^T \textbf{A}_{: k}}{2\alpha_k^{2}}\\
-\tab \tab \tab \tab \tab + \sum_{d=1}^D a \log b + (-a - 1) \log\Psi_{dd} - \frac{b}{\Psi_{dd}} - \log \Gamma(a)\\
+\log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) = \\
+\tab \tab -\frac{-D}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K}  s_k  \mu_k   + \sum_{k=1}^{K} \sum_{k'=1, k'\neq k}^{K}  s_k s_{k'} \mu_k^T\mu_{k'} + \sum_{k=1}^{K}  s_k \mu_k^T\mu_k \right) \\
+\tab \tab + \sum_{k=1}^{K} s_k \log \pi_k + (1-s_k) \log (1-\pi_k)}\\
+\tab \tab + \sum_{i=1}^K (\alpha-1) \log \pi_k + (\beta - 1) \log (1-\pi_k) - \log B(\alpha, \beta)\\
+\tab \tab + \sum_{i=1}^K -\frac{D}{2} \log (2 \pi\alpha_k^2)  -\frac{\mu_{k}^T \mu_{k}}{2\alpha_k^{2}}\\
+\tab \tab + \sum_{d=1}^D a \log b + (-a - 1) \log(\sigma^2) - \frac{b}{\sigma^2} - \log \Gamma(a)\\
 \end{array}
@@
 %\tab \tab \tab \tab \tab -\frac{DK}{2} \log (2 \pi) - \frac{1}{2} \log(|\Sigma_{\textbf{A}}|)   -\frac{1}{2}\textbf{A}^T \Sigma_{\textbf{A}}^{-1}\textbf{A}\\
-%\tab \tab \tab \tab \tab + \sum_{d=1}^D a_d \log b_d + (-a_d - \frac{3}{2}) \log\Psi_{dd} - \frac{b_d}{\Psi_{dd}} - \log \Gamma(a_d)\\
+%\tab \tab \tab \tab \tab + \sum_{d=1}^D a_d \log b_d + (-a_d - \frac{3}{2}) \log\sigma^2 - \frac{b_d}{\sigma^2} - \log \Gamma(a_d)\\
 %\end{array}
@@
 
-\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) + \sum_{k=1}^{K} s_k \log \pi_k + (1-s_k) \log (1-\pi_k)}\right\rangle_{q(\theta)}\]
-
-Simplifying:
-
-\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle -\frac{1}{2}\left(\textbf{s}^T\textbf{A}^T\Psi^{-1}\textbf{As} - 2\textbf{s}^T\left(\textbf{A}^T\Psi^{-1} \textbf{x}  + 2\log\frac{\pi}{1-\pi}\right)\right)\right\rangle_{q(\theta)}\]
-
-By inspection, we can see:
-
-\[q_{\texbf{s}}(\textbf{s}) \propto \mathcal{N}(s | \mu_\textbf{s}^*, \Sigma_\textbf{s}^*)\]
-
-where
-
-\[\Sigma_\textbf{s}^* = \left\langle \left(\textbf{A}^T\Psi^{-1}\textbf{A}\right)^{-1}\right\rangle_{q(\theta)}\]
-
-and
-
-\[\mu_\textbf{s}^* = \left\langle \left(\textbf{A}^T\Psi^{-1}\textbf{A}\right)^{-1}\left(\textbf{A}^T\Psi^{-1} \textbf{x}  + 2\log\frac{\pi}{1-\pi}\right)\right\rangle_{q(\theta)}\]
-
-the E step updates.
+\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle -\frac{1}{2 \sigma^2} \left( - 2 \textbf{x}^T\sum_{k=1}^{K}  s_k  \mu_k   + \sum_{k=1}^{K} \sum_{k'=1, k'\neq k}^{K}  s_k s_{k'} \mu_k^T\mu_{k'} + \sum_{k=1}^{K}  s_k \mu_k^T\mu_k \right) + \sum_{k=1}^{K} s_k \log \frac{\pi_k}{1-\pi_k}\right\rangle_{q(\theta)}\]
+
+%Simplifying:
+%
+%\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle -\frac{1}{2}\left(\textbf{s}^T\textbf{A}^T\Psi^{-1}\textbf{As} - 2\textbf{s}^T\left(\textbf{A}^T\Psi^{-1} \textbf{x}  + 2\log\frac{\pi}{1-\pi}\right)\right)\right\rangle_{q(\theta)}\]
+
+Given our factored approximation $q(\textbf{s}) = \prod_{i=1}^K q_i(s_i)$, we can see that we can derive a similar partial update for $q_i(s_i)$ as in Question 3, by taking the variation derivative of the Lagrangian to enforce the normalisation of $q_i$:
+
+\[\frac{\partial}{\partial q_i}\left( \exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q(\theta)} + \lambda^{LG} \int q_i -1)\right) \propto \exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q(\theta)\prod_{j\neq i} q_j(s_j)} - \log q_i(s_i)\]
+
+Setting this to zero we can solve for $\lambda_i$ where $q_i(s_i) = \lambda_i^{s_i}(1-\lambda_i)^{(1-s_i)}$:
+
+\[ \lambda_i =  \frac{1}{1+\exp\left[ - \left(\frac{  \langle \mu_i\rangle_{q_{\mu_i}}^T}{\langle\sigma^2\rangle_{q(\Psi)}} \left( \textbf{x} -  \frac{\langle \mu_i\rangle_{q_{\mu_i}}}{2} -  \sum_{j=1, j\neq i}^{K}   \lambda_j  \langle \mu_j\rangle_{q_{\mu_j} \right) + \log \frac{\langle \pi_i\rangle_{q_{\pi_i}}}{1-\langle \pi_i\rangle_{q_{\pi_i}}}\right) \right]}\]
+
+we have our partial E step update.
 
@@
 
-\[q_{\pi}(\pi) \propto \left(\prod_{k=1}^{K} \frac{\pi_k^{\alpha - 1}(1-\pi_k)^{\beta-1}}{B(\alpha, \beta)} \right)\exp \left\langle \sum_{i=1}^K s_k \log \pi_k + (1 - s_k) \log (1-\pi_k) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \pi}(\theta)}\]
+\[q_{\pi}(\pi) \propto \left(\prod_{k=1}^{K} \frac{\pi_k^{\alpha - 1}(1-\pi_k)^{\beta-1}}{B(\alpha, \beta)} \right)\exp \left\langle \sum_{k=1}^K s_k \log \pi_k + (1 - s_k) \log (1-\pi_k) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \pi}(\theta)}\]
 
@@
 
-\[q_{\pi}(\pi) \propto \prod_{k=1}^{K} \frac{\pi_k^{\alpha  + \left\langle s_k\right\rangle_{q_{s_k}} - 1}(1-\pi_k)^{\beta-\left\langle s_k\right\rangle_{q_{s_k}}}}{B(\alpha, \beta)}\]
-
-\[q_{\pi}(\pi) = \prod_{k=1}^{K}  Beta(\alpha + \left\langle s_k\right\rangle_{q_{s_k}} , \beta+(1-\left\langle s_k\right\rangle_{q_{s_k}}}))\]
+\[q_{\pi}(\pi) \propto \prod_{k=1}^{K} \frac{\pi_k^{\alpha  + \frac{1}{D}\left(\sum_{k=1}^K \left\langle s_k\right\rangle_{q_{s_k}}\right) - 1}(1-\pi_k)^{\beta-\frac{1}{D}\left(\sum_{k=1}^K \left\langle s_k\right\rangle_{q_{s_k}}\right)}}{B(\alpha, \beta)}\]
+
+\[q_{\pi}(\pi) = \prod_{k=1}^{K}  Beta\left(\alpha + \frac{1}{D}\sum_{k=1}^K \left\langle s_k\right\rangle_{q_{s_k}} , \beta+1- \frac{1}{D}\sum_{k=1}^K \left\langle s_k\right\rangle_{q_{s_k}}\right)\]
 
@@
 \[
-q_{\Psi}(\Psi) \propto \left( \prod_{d=1}^D \frac{b^{a}}{\Gamma(a)}\Psi_{dd}^{-a-1}\exp(-\frac{b}{\Psi_{dd}}) \right)\exp \left\langle  - \frac{1}{2} \sum_{d=1}^{D}\log\Psi_{dd}   -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \Psi(\theta)}}
+q_{\Psi}(\Psi) \propto \left( \prod_{d=1}^D \frac{b^{a}}{\Gamma(a)}(\sigma^2)^{-a-1}\exp\left(-\frac{b}{\sigma^2}\right) \right)\exp \left\langle   -\frac{-D}{2} \log (\sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}-\textbf{A}\textbf{s} \right)^T\left(\textbf{x}-\textbf{A}\textbf{s} \right)  \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \Psi(\theta)}}
 \]
@@
 \[
-q_{\Psi}(\Psi) \propto  \prod_{d=1}^D \frac{b^{a}}{\Gamma(a)}\Psi_{dd}^{-(a+\frac{1}{2})-1}\exp\left(-\frac{b+\frac{1}{2} \left\langle (x_d-\textbf{A}_{d :}\textbf{s})^2 \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}_{d :}}(\textbf{A}_{d :})}}{\Psi_{dd}}\right) }
+q_{\Psi}(\Psi) \propto  \prod_{d=1}^D \frac{b^{a}}{\Gamma(a)}(\sigma^2)^{-(a+\frac{1}{2})-1}\exp\left(-\frac{b+\frac{2}{D} \left\langle \left(\textbf{x}-\textbf{A}\textbf{s} \right)^T\left(\textbf{x}-\textbf{A}\textbf{s} \right) \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}}(\textbf{A})}}{\sigma^2}\right) }
 \]
@@
 \[
-q_{\Psi}(\Psi) = \prod_{d=1}^D  InvGamma\left(\Psi_{dd} \left| a+\frac{1}{2}, b+\frac{1}{2} \left\langle (x_d-\textbf{A}_{d :}\textbf{s})^2 \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}_{d :}}(\textbf{A}_{d :})}\right)
-\]
-
-For $q_{\textbf{A}_{: k}}(\textbf{A}_{: k})$:
-
-\[q_{\textbf{A}_{: k}}(\textbf{A}_{: k})  \propto P(\textbf{A}_{: k})\exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}_{: k}}(\theta)}\]
+q_{\Psi}(\Psi) = \prod_{d=1}^D  InvGamma\left(\sigma^2 \left| a+\frac{1}{2}, b+\frac{2}{D} \left\langle \left(\textbf{x}-\textbf{A}\textbf{s} \right)^T\left(\textbf{x}-\textbf{A}\textbf{s} \right) \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}}(\textbf{A})}\right)
+\]
+
+For $q_{\mu_{k}}(\mu_{k})$:
+
+\[q_{\mu_{k}}(\mu_{k})  \propto P(\mu_{k})\exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \mu_{k}}(\theta)}\]
 
@@
 
-\[q_{\textbf{A}_{: k}}(\textbf{A}_{: k}) \propto \exp\left(-\frac{\textbf{A}_{: k}^T \textbf{A}_{: k}}{2\alpha_k^{2}} \right) \exp \left\langle -\frac{1}{2}(\textbf{x}-\textbf{A}_{: k} s_k)^T \Psi^{-1} (\textbf{x}-\textbf{A}_{: k} s_k) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}_{: k}}(\theta)}\]
-
-\[q_{\textbf{A}_{: k}}(\textbf{A}_{: k}) \propto \exp\left(-\frac{\textbf{A}_{: k}^T \textbf{A}_{: k}}{2\alpha_k^{2}} \right) \exp \left\langle -\frac{1}{2} \textbf{A}_{: k}^T\frac{1 }{ s_k^2 \Psi_{dd}}\textbf{A}_{: k}  - 2 \frac{s_k}{\Psi_{dd}} \textbf{x}^T\textbf{A}_{: k} \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}_{: k}}(\theta)}\]
+\[q_{\mu_{k}}(\mu_{k}) \propto \exp\left(-\frac{\mu_{k}^T \mu_{k}}{2\alpha_k^{2}} \right) \exp \left\langle \frac{-1}{2\sigma^2} \left(- 2 \textbf{x}^T s_k  \mu_k  + s_k \mu_k^T\mu_k  + 2 \sum_{k'=1, k'\neq k}^{K}  s_k s_{k'} \mu_k^T\mu_{k'}}\right) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \mu_{k}}(\theta)}\]
+
+\[q_{\mu_{k}}(\mu_{k}) \propto \exp\left(-\frac{1}{2}\left( \mu_{k}^T\left(\frac{1}{\alpha_k^{2}} + \frac{\langle s_k \rangle_{q_{s_k}}}{\langle\sigma^2\rangle_{q_{\Psi}}}}\right)\mu_{k} + 2 \langle s_k \rangle_{q_{s_k}}\left(-  \textbf{x}^T +  \sum_{k'=1, k'\neq k}^{K} \langle s_{k'} \rangle_{q_{s_{k'}}} \langle \mu_{k'} \rangle_{q_{\mu_{k'}}}^T\right)\mu_k\right)\]
 
@@
 
-\[q_{\textbf{A}_{: k}}(\textbf{A}_{: k}) = \mathcal{N}\left(\mu_{\textbf{A}_{: k}}, \Sigma_{\textbf{A}_{: k}}\right)}\]
+\[q_{\mu_{k}}(\mu_{k}) = \mathcal{N}\left(\mu_{\mu_{k}}, \Sigma_{\mu_{k}}\right)}\]
 
@@
 
-\[ \Sigma_{\textbf{A}_{: k}} = \left(\frac{1}{\alpha_k^2} + \frac{1}{s_k^2\Psi_{dd}}\right)^{-1} \textbf{I}\]
+\[ \Sigma_{\mu_{k}} = \left(\frac{1}{\alpha_k^{2}} + \frac{\langle s_k \rangle_{q_{s_k}}}{\langle\sigma^2\rangle_{q_{\Psi}}}}\right)^{-1} \textbf{I}\]
 
@@
 
-\[\mu_{\textbf{A}_{: k}} = \Sigma_{\textbf{A}_{: k}}^{-1} \frac{s_k}{\Psi_{dd}}\textbf{x}\]
-
-By optimising with respect to the the distributions $\Psi$ and $\alpha$ in turn causes some $\alpha_i^2$ to diverge, the number of remaining $\alpha_i^2$ provide our determination for the value of $K$.
+\[\mu_{\mu_{k}} = \Sigma_{\mu_{k}} \left(\textbf{x}^T - \sum_{k'=1, k'\neq k}^{K} \langle s_{k'} \rangle_{q_{s_{k'}}} \langle \mu_{k'} \rangle_{q_{\mu_{k'}}}^T\right)\langle s_k \rangle_{q_{s_k}}\]
+
+Optimising the parameters $\{a, b\}$, and $\alpha^2$ in turn, will cause some $\alpha_i^2$ to diverge, the number of remaining $\alpha_i^2$ provide our determination for the value of $K$.
 This is automatic relevance determination through factor analysis.
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.4]{outputs/q3/f-latent-factors}
+\caption{Learned Latent Factors with Mean Field Approximation}
+\label{fig:3f-latent-factors}
+\end{figure}
+
+\newpage
+
+\begin{figure}[h]
+\centering
+\begin{minipage}{.5\textwidth}
+  \centering
+\includegraphics[scale=0.4]{outputs/q3/f-free-energy}
+\caption{Mean Field Approximation}
+\label{fig:3g-free-energy-diff-sigma}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+\includegraphics[scale=0.4]{outputs/q6/all-free-energy}
+\caption{Loopy BP}
+\label{fig:3g-free-energy-diff-sigma}
+\end{minipage}
+\end{figure}
+
 \newpage

### 2022-12-31 demo_code/MStep.py

@@ -1,35 +1,35 @@
 import numpy as np
 
 
-def m_step(X, ES, ESS):
+def m_step(x, es, ess):
     """
-    mu, sigma, pie = MStep(X,ES,ESS)
+    mu, sigma, pie = MStep(x,es,ess)
 
     Inputs:
     -----------------
-           X: shape (N, D) data matrix
-          ES: shape (N, K) E_q[s]
-         ESS: shape (K, K) sum over data points of E_q[ss'] (N, K, K)
-                           if E_q[ss'] is provided, the sum over N is done for you.
+           x: shape (n, d) data matrix
+          es: shape (n, k) E_q[s]
+         ess: shape (k, k) sum over data points of E_q[ss'] (n, k, k)
+                           if E_q[ss'] is provided, the sum over n is done for you.
 
     Outputs:
     --------
-          mu: shape (D, K) matrix of means in p(y|{s_i},mu,sigma)
+          mu: shape (d, k) matrix of means in p(y|{s_i},mu,sigma)
        sigma: shape (,)    standard deviation in same
-         pie: shape (1, K) vector of parameters specifying generative distribution for s
+         pie: shape (1, k) vector of parameters specifying generative distribution for s
     """
-    N, D = X.shape
-    if ES.shape[0] != N:
-        raise TypeError('ES must have the same number of rows as X')
-    K = ES.shape[1]
-    if ESS.shape == (N, K, K):
-        ESS = np.sum(ESS, axis=0)
-    if ESS.shape != (K, K):
-        raise TypeError('ESS must be square and have the same number of columns as ES')
+    n, d = x.shape
+    if es.shape[0] != n:
+        raise TypeError('es must have the same number of rows as x')
+    k = es.shape[1]
+    if ess.shape == (n, k, k):
+        ess = np.sum(ess, axis=0)
+    if ess.shape != (k, k):
+        raise TypeError('ess must be square and have the same number of columns as es')
 
-    mu = np.dot(np.dot(np.linalg.inv(ESS), ES.T), X).T
-    sigma = np.sqrt((np.trace(np.dot(X.T, X)) + np.trace(np.dot(np.dot(mu.T, mu), ESS))
-                     - 2 * np.trace(np.dot(np.dot(ES.T, X), mu))) / (N * D))
-    pie = np.mean(ES, axis=0, keepdims=True)
[…]

### 2022-12-31 main.py

@@ -11,6 +11,7 @@ from src.generate_images import generate_images
 from src.models.bayesian_linear_regression import LinearRegressionParameters
 from src.models.gaussian_process_regression import GaussianProcessParameters
 from src.models.kernels import CombinedKernel, CombinedKernelParameters
+from src.models.variational_bayes import Beta, Gaussian, InverseGamma
 from src.solutions import q2, q3, q4, q6
 
 jax.config.update("jax_enable_x64", True)
@@ -85,12 +86,12 @@ if __name__ == "__main__":
     )
 
     init_kernel_parameters = CombinedKernelParameters(
-        log_theta=jnp.log(1e-1),
-        log_sigma=jnp.log(1),
-        log_phi=jnp.log(5),
-        log_eta=jnp.log(1e-1),
+        log_theta=jnp.log(5),
+        log_sigma=jnp.log(5),
+        log_phi=jnp.log(10),
+        log_eta=jnp.log(5),
         log_tau=jnp.log(1),
-        log_zeta=jnp.log(1e-1),
+        log_zeta=jnp.log(2),
     )
     gaussian_process_parameters = GaussianProcessParameters(
         kernel=asdict(init_kernel_parameters),
@@ -109,42 +110,79 @@ if __name__ == "__main__":
         kernel=kernel,
         gaussian_process_parameters=gaussian_process_parameters,
         learning_rate=1e-2,
-        number_of_iterations=500,
+        number_of_iterations=100,
         save_path=os.path.join(Q2_OUTPUT_FOLDER, "f"),
     )
-
-    # Question 3
-    Q3_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q3")
-    if not os.path.exists(Q3_OUTPUT_FOLDER):
-        os.makedirs(Q3_OUTPUT_FOLDER)
-    number_of_images = 1000
-    x = generate_images(n=number_of_images)
-    k = 8
-    em_iterations = 200
-    e_maximum_steps = 100
-    e_convergence_criterion = 0
-
-    binary_latent_factor_model = q3.e_and_f(
-        x=x,
-        k=k,
-        em_iterations=em_iterations,
-        e_maximum_steps=e_maximum_steps,
[…]

### 2022-12-31 src/models/mean_field_approximation.py

@@ -3,8 +3,8 @@ from typing import List
 import numpy as np
 
 from src.models.binary_latent_factor_model import (
+    AbstractBinaryLatentFactorModel,
     BinaryLatentFactorApproximation,
-    BinaryLatentFactorModel,
 )
 
 
@@ -41,7 +41,7 @@ class MeanFieldApproximation(BinaryLatentFactorApproximation):
     def _partial_expectation_step(
         self,
         x: np.ndarray,
-        binary_latent_factor_model: BinaryLatentFactorModel,
+        binary_latent_factor_model: AbstractBinaryLatentFactorModel,
         latent_factor: int,
     ) -> np.ndarray:
         """Partial Variational E step for factor i for all data points
@@ -88,7 +88,7 @@ class MeanFieldApproximation(BinaryLatentFactorApproximation):
         return lambda_vector
 
     def variational_expectation_step(
-        self, x: np.ndarray, binary_latent_factor_model: BinaryLatentFactorModel
+        self, x: np.ndarray, binary_latent_factor_model: AbstractBinaryLatentFactorModel
     ) -> List[float]:
         """Variational E step
 

### 2022-12-31 src/models/variational_bayes.py

@@ -1,0 +1,153 @@
+from abc import ABC, abstractmethod
+from dataclasses import dataclass
+
+import numpy as np
+
+from src.models.binary_latent_factor_model import (
+    AbstractBinaryLatentFactorModel,
+    BinaryLatentFactorApproximation,
+)
+
+
+@dataclass
+class Distribution(ABC):
+    @property
+    @abstractmethod
+    def mean(self):
+        pass
+
+
+@dataclass
+class Beta(Distribution):
+    alpha: np.ndarray
+    beta: np.ndarray
+
+    @property
+    def mean(self) -> np.ndarray:
+        return np.divide(self.alpha, (self.alpha + self.beta))
+
+
+@dataclass
+class InverseGamma(Distribution):
+    a: float
+    b: float
+
+    @property
+    def mean(self) -> float:
+        return self.b / (self.a - 1)
+
+
+@dataclass
+class Gaussian(Distribution):
+    mu: np.ndarray  # (number_of_dimensions,  number_of_latent_variables)
+    variance: np.ndarray  # (number_of_latent_variables, )
+
+    @property
+    def precision(self) -> np.ndarray:
+        return 1 / self.variance
+
+    @property
+    def mean(self) -> np.ndarray:
+        return self.mu
+
+
+class VariationalBayesBinaryLatentFactorModel(AbstractBinaryLatentFactorModel):
+    def __init__(self, mu: Gaussian, variance: InverseGamma, pi: Beta):
+        self._mu = mu
+        self._variance = variance
+        self._pi = pi
+
+    @property
+    def variance(self) -> float:
+        return self._variance.mean
+
+    @property
+    def pi(self) -> np.ndarray:
+        return self._pi.mean
+
+    @property
+    def mu(self) -> np.ndarray:
+        return self._mu.mean
+
+    def _update_pi(
+        self,
+        binary_latent_factor_approximation: BinaryLatentFactorApproximation,
+    ):
+        self._pi.alpha += np.sum(
+            binary_latent_factor_approximation.expectation_s, axis=0
+        ).reshape(1, -1)
+        self._pi.beta += binary_latent_factor_approximation.n - np.sum(
+            binary_latent_factor_approximation.expectation_s, axis=0
+        ).reshape(1, -1)
+
+    def _update_variance(
+        self,
+        x: np.ndarray,  # (number_of_points, number_of_dimensions)
+        binary_latent_factor_approximation: BinaryLatentFactorApproximation,
[…]

### 2022-12-31 src/solutions/q2.py

@@ -232,8 +232,12 @@ def f(
     )
     plt.fill_between(
         t_test + min_year,
-        linear_prediction + mean_prediction - 1 * jnp.diagonal(covariance_prediction),
-        linear_prediction + mean_prediction + 1 * jnp.diagonal(covariance_prediction),
+        linear_prediction
+        + mean_prediction
+        - 1 * jnp.sqrt(jnp.diagonal(covariance_prediction)),
+        linear_prediction
+        + mean_prediction
+        + 1 * jnp.sqrt(jnp.diagonal(covariance_prediction)),
         facecolor=(0.8, 0.8, 0.8),
         label="error bound (one stdev)",
     )
@@ -256,7 +260,7 @@ def f(
         + [["sigma", float(gaussian_process_parameters.sigma)]],
         columns=["parameter", "value"],
     )
-    df_parameters = df_parameters.set_index("parameter")
+    df_parameters = df_parameters.set_index("parameter").sort_values(by=["parameter"])
     dfi.export(df_parameters, save_path + "-untrained-parameters.png")
 
     # Train Gaussian Process Regression (Hyperparameter Tune)
@@ -275,7 +279,7 @@ def f(
         + [["sigma", float(gaussian_process_parameters.sigma)]],
         columns=["parameter", "value"],
     )
-    df_parameters = df_parameters.set_index("parameter")
+    df_parameters = df_parameters.set_index("parameter").sort_values(by=["parameter"])
     dfi.export(df_parameters, save_path + "-trained-parameters.png")
 
     # Prediction
@@ -304,8 +308,12 @@ def f(
     )
     plt.fill_between(
         t_test + min_year,
-        linear_prediction + mean_prediction - 1 * jnp.diagonal(covariance_prediction),
-        linear_prediction + mean_prediction + 1 * jnp.diagonal(covariance_prediction),
+        linear_prediction
+        + mean_prediction
+        - 1 * jnp.sqrt(jnp.diagonal(covariance_prediction)),
+        linear_prediction
+        + mean_prediction
+        + 1 * jnp.sqrt(jnp.diagonal(covariance_prediction)),
         facecolor=(0.8, 0.8, 0.8),
         label="error bound (one stdev)",
     )

### 2022-12-31 src/solutions/q3.py

@@ -4,14 +4,13 @@ import matplotlib.pyplot as plt
 import numpy as np
 
 from src.models.binary_latent_factor_model import (
+    AbstractBinaryLatentFactorModel,
+    BinaryLatentFactorModel,
     init_binary_latent_factor_model,
     is_converge,
     learn_binary_factors,
 )
-from src.models.mean_field_learning import (
-    BinaryLatentFactorModel,
-    init_mean_field_approximation,
-)
+from src.models.mean_field_approximation import init_mean_field_approximation
 
 
 def e_and_f(
@@ -21,7 +20,7 @@ def e_and_f(
     e_maximum_steps: int,
     e_convergence_criterion: float,
     save_path: str,
-) -> BinaryLatentFactorModel:
+) -> AbstractBinaryLatentFactorModel:
     n = x.shape[0]
     mean_field_approximation = init_mean_field_approximation(
         k, n, max_steps=e_maximum_steps, convergence_criterion=e_convergence_criterion
@@ -63,7 +62,7 @@ def e_and_f(
 
 def g(
     x: np.ndarray,
-    binary_latent_factor_model: BinaryLatentFactorModel,
+    binary_latent_factor_model: AbstractBinaryLatentFactorModel,
     sigmas: List[float],
     k: int,
     em_iterations: int,
@@ -105,6 +104,19 @@ def g(
             ):
                 break
         free_energies.append(free_energy)
+
+    for i, free_energy in enumerate(free_energies):
+        plt.plot(
+            free_energy,
+            label=f"sigma={sigmas[i]}",
+        )
+    plt.title(f"F(t)")
+    plt.xlabel("t (Variational E steps)")
+    plt.ylabel("F(t)")
+    plt.tight_layout()
+    plt.legend()
+    plt.savefig(save_path + f"-free-energy-sigma.png", bbox_inches="tight")
+    plt.close()
 
     for i, free_energy in enumerate(free_energies):
         diffs = np.log(np.diff(free_energy))

### 2022-12-31 src/solutions/q4.py

@@ -1,0 +1,88 @@
+import matplotlib.pyplot as plt
+import numpy as np
+
+from src.models.binary_latent_factor_model import (
+    BinaryLatentFactorModel,
+    learn_binary_factors,
+)
+from src.models.mean_field_approximation import init_mean_field_approximation
+from src.models.variational_bayes import (
+    Beta,
+    Gaussian,
+    InverseGamma,
+    VariationalBayesBinaryLatentFactorModel,
+)
+
+
+def b(
+    x: np.ndarray,
+    k: int,
+    em_iterations: int,
+    e_maximum_steps: int,
+    e_convergence_criterion: float,
+    # mu: Gaussian,
+    # variance: InverseGamma,
+    # pi: Beta,
+    save_path: str,
+) -> None:
+    n = x.shape[0]
+    mean_field_approximation = init_mean_field_approximation(
+        k, n, max_steps=e_maximum_steps, convergence_criterion=e_convergence_criterion
+    )
+    (
+        mu_max_step,
+        sigma,
+        pi_max_step,
+    ) = BinaryLatentFactorModel.calculate_maximisation_parameters(
+        x, mean_field_approximation
+    )
+    mu = Gaussian(
+        mu=mu_max_step,
+        variance=mu_max_step.std(axis=0) ** 2,
+    )
+    variance = InverseGamma(
+        a=2,
+        b=sigma**2,
+    )
+    pi = Beta(
+        alpha=np.ones((1, k)),
+        beta=np.divide(1 - pi_max_step, pi_max_step),
+    )
+    binary_latent_factor_model = VariationalBayesBinaryLatentFactorModel(
+        mu=mu,
+        variance=variance,
+        pi=pi,
+    )
+    fig, ax = plt.subplots(1, k, figsize=(k * 2, 2))
+    for i in range(k):
+        ax[i].imshow(binary_latent_factor_model.mu[:, i].reshape(4, 4))
+        ax[i].set_title(f"Latent Feature mu_{i}")
+    fig.suptitle("Initial Features (Variational Bayes)")
+    plt.tight_layout()
+    plt.savefig(save_path + "-init-latent-factors", bbox_inches="tight")
+    plt.close()
+    (
+        mean_field_approximation,
+        binary_latent_factor_model,
+        free_energy,
+    ) = learn_binary_factors(
+        x=x,
+        em_iterations=em_iterations,
[…]

### 2022-12-31 main.py
formatting

@@ -113,76 +113,76 @@ if __name__ == "__main__":
         number_of_iterations=100,
         save_path=os.path.join(Q2_OUTPUT_FOLDER, "f"),
     )
-    #
-    # # Question 3
-    # Q3_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q3")
-    # if not os.path.exists(Q3_OUTPUT_FOLDER):
-    #     os.makedirs(Q3_OUTPUT_FOLDER)
-    # number_of_images = 2000
-    # x = generate_images(n=number_of_images)
-    # k = 8
-    # em_iterations = 200
-    # e_maximum_steps = 100
-    # e_convergence_criterion = 0
-    #
-    # binary_latent_factor_model = q3.e_and_f(
-    #     x=x,
-    #     k=k,
-    #     em_iterations=em_iterations,
-    #     e_maximum_steps=e_maximum_steps,
-    #     e_convergence_criterion=e_convergence_criterion,
-    #     save_path=os.path.join(Q3_OUTPUT_FOLDER, "f"),
-    # )
-    # _ = q3.e_and_f(
-    #     x=x,
-    #     k=int(k*1.5),
-    #     em_iterations=em_iterations,
-    #     e_maximum_steps=e_maximum_steps,
-    #     e_convergence_criterion=e_convergence_criterion,
-    #     save_path=os.path.join(Q3_OUTPUT_FOLDER, "f-larger-k"),
-    # )
-    # q3.g(
-    #     x=x[:1, :],
-    #     binary_latent_factor_model=binary_latent_factor_model,
-    #     sigmas=[1, 2, 3],
-    #     k=k,
-    #     em_iterations=em_iterations,
-    #     e_maximum_steps=e_maximum_steps,
-    #     e_convergence_criterion=e_convergence_criterion,
-    #     save_path=os.path.join(Q3_OUTPUT_FOLDER, "g"),
-    # )
-    #
-    # # Question 4
-    # Q4_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q4")
-    # if not os.path.exists(Q4_OUTPUT_FOLDER):
-    #     os.makedirs(Q4_OUTPUT_FOLDER)
-    # d = x.shape[1]
-    # pi = Beta(
-    #     alpha=np.random.gamma(1, size=(1, k)),
-    #     beta=np.random.gamma(1, size=(1, k)),
-    # )
[…]

### 2023-01-01 submission.tex
formatting
implement variational bayes for ard

@@
 
-We begin with the log joint:
-
-\[\log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) = \log P(\textbf{x}|  \textbf{s}, \textbf{A}, \Psi,  \eta)
-+ \log P( \textbf{s}| \pi, \eta) + \log P( \pi| \eta) + \log P( \textbf{A}| \eta) + \log P( \Psi| \eta)
+We begin by writing the expression for $x_d$:
+
+\[P(x_d | \texbf{s}, \textbf{w}_d, \sigma^2) = \mathcal{N}\left(\textbf{s}^T\textbf{w}_d, \sigma^2\right)\]
+
+where we know from the diagonal covariance of $P(\textbf{x} | \textbf{s}, \mu, \sigma^2)$ that each dimension is independent.
+Moreover, $\textbf{w}_d \in \mathbb{R}^{K\times 1}$, which is the $d^{th}$ row of $\mu \in \mathbb{R}^{D \times K}$
+
+Thus, we can write the posterior:
+
+\[\log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) = \log P( \textbf{s}| \pi) + \sum_{d=1}^{D} \log P(x_d | \texbf{s}, \textbf{w}_d, \sigma^2) + \log P( \textbf{w}_d| \alpha)
 \]
 
-where $\textbf{A} \in \mathbb{R}^{D \times K}$ is a concatenation of $\mu_k$ for $k \in \{1, ..., K\}$  and \eta$ is a collection of all hyperparameters.
-
-We know:
-
-\[P(\textbf{x}|  \textbf{s}, \textbf{A}, \Psi,  \eta) = \frac{1}{(2\pi)^{\frac{D}{2}} |\Psi|^{\frac{1}{2}}} \exp\left(-\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As})\right)\]
-
-\[P( \textbf{s}| \pi, \eta) = \prod_{k=1}^{K} \pi_k^{s_i} (1-\pi_k)^{1-s_k}\]
-
-\[P( \pi| \eta) = \prod_{k=1}^{K} \frac{\pi_k^{\alpha - 1}(1-\pi_k)^{\beta-1}}{B(\alpha, \beta)}\]
-
-For $\textbf{A}$ we choose a factorised conjugate prior for each column, $\mu_k$:
-
-\[P( \textbf{A}| \eta) = \prod_{k=1}^{K}P( \mu_{k}| \eta)\]
-
-For each column we choose:
-
-\[P( \mu_{k}| \eta) = \mathcal{N}(\mu_{k}| \mu_{\mu_{k}}, \Sigma_{\mu_{k}}) =\frac{1}{(2\pi)^{\frac{D}{2}} |\Sigma_{\mu_{k}}|^{\frac{1}{2}}} \exp\left(-\frac{1}{2}(\mu_{k}} - \mu_{\mu_{k}})^T \Sigma_{\mu_{k}}^{-1}(\mu_{k} - \mu_{\mu_{k}})\right)\]
-
-a Gaussian prior with diagonal covariance $\Sigma_{\mu_{k}} = \alpha_k^{2}\textbf{I}$ and mean zero, so we can simplify:
-
-\[P( \mu_{k}| \alpha_k) = (2\pi\alpha_k^{2})^{\frac{-D}{2}} \exp\left(-\frac{\mu_{k}^T \mu_{k}}{2\alpha_k^{2}} \right)\]
-
-For $\Psi}$ we choose a conjugate prior:
-
-\[P( \Psi| \eta) = \prod_{d=1}^D InvGamma(\sigma^2| a, b) = \prod_{d=1}^D \frac{b^{a}}{\Gamma(a)}(\sigma^2)^{-a-1}\exp(-\frac{b}{\sigma^2})\]
-
-a product of inverse gamma distributions, assuming $\Psi$ is a diagonal matrix with elements \sigma^2.
+where we introduce priors on each $\textbf{w}_k$ with $\alpha \in \mathbb{R}^{K \times 1}$.
+
+We choose each prior to be:
+
+\[P( \textbf{w}_d| \alpha) = \mathcal{N}(0, \textbf{A}^{-1})\]
+
+where $\textbf{A} = diag(\alpha)$, the precision matrix.
 
@@
 \begin{array}{l}
-\log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) = \\
-\tab \tab -\frac{-D}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K}  s_k  \mu_k   + \sum_{k=1}^{K} \sum_{k'=1, k'\neq k}^{K}  s_k s_{k'} \mu_k^T\mu_{k'} + \sum_{k=1}^{K}  s_k \mu_k^T\mu_k \right) \\
+\log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) = \\
+\tab \tab + \sum_{d=1}^D-\frac{-1}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(x_d^2 - 2 x_d \textbf{s}^T \textbf{w}_d   + \textbf{w}_d^T \textbf{s}\textbf{s}^T\textbf{w}_d\right) \\
 \tab \tab + \sum_{k=1}^{K} s_k \log \pi_k + (1-s_k) \log (1-\pi_k)}\\
-\tab \tab + \sum_{i=1}^K (\alpha-1) \log \pi_k + (\beta - 1) \log (1-\pi_k) - \log B(\alpha, \beta)\\
-\tab \tab + \sum_{i=1}^K -\frac{D}{2} \log (2 \pi\alpha_k^2)  -\frac{\mu_{k}^T \mu_{k}}{2\alpha_k^{2}}\\
-\tab \tab + \sum_{d=1}^D a \log b + (-a - 1) \log(\sigma^2) - \frac{b}{\sigma^2} - \log \Gamma(a)\\
+\tab \tab + \sum_{d=1}^D -\frac{K}{2} \log (2 \pi) + \frac{1}{2}  \sum_{k=1}^K \left(\log\alpha_k \right)  -\frac{1}{2}\textbf{w}_{d}^T \textbf{A}\textbf{w}_{d}\\
 \end{array}
@@
 
-%Simplifying:
-%
-%\[
-%\begin{array}{l}
-%\log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) = -\frac{D}{2} \log (2 \pi)  -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \\
-%\tab \tab \tab \tab \tab + \sum_{i=1}^K (s_k+\alpha-1) \log \pi_k + (\beta - s_k) \log (1-\pi_k) - \log B(\alpha, \beta)\\
-%\tab \tab \tab \tab \tab -\frac{DK}{2} \log (2 \pi) - \frac{1}{2} \log(|\Sigma_{\textbf{A}}|)   -\frac{1}{2}\textbf{A}^T \Sigma_{\textbf{A}}^{-1}\textbf{A}\\
-%\tab \tab \tab \tab \tab + \sum_{d=1}^D a_d \log b_d + (-a_d - \frac{3}{2}) \log\sigma^2 - \frac{b_d}{\sigma^2} - \log \Gamma(a_d)\\
-%\end{array}
-%\]
-
 For the Variational Bayes expectation step, we minimise $\textbf{KL}[q_{\texbf{s}}(\textbf{s} | \text{everything else}) \| P(\textbf{s} | \text{everything else})]$ by setting:
 
-\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q(\theta)}\]
-
-where $\theta$ denotes the parameters $\pi$, $\textbf{A}$, $\Psi$, $\eta$.
+\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle \log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha)  \right\rangle_{q(\mu)}\]
 
@@
 
-\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle -\frac{1}{2 \sigma^2} \left( - 2 \textbf{x}^T\sum_{k=1}^{K}  s_k  \mu_k   + \sum_{k=1}^{K} \sum_{k'=1, k'\neq k}^{K}  s_k s_{k'} \mu_k^T\mu_{k'} + \sum_{k=1}^{K}  s_k \mu_k^T\mu_k \right) + \sum_{k=1}^{K} s_k \log \frac{\pi_k}{1-\pi_k}\right\rangle_{q(\theta)}\]
+\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle -\frac{1}{2 \sigma^2} \left( - 2 \textbf{x}^T\sum_{k=1}^{K}  s_k  \mu_k   + \sum_{k=1}^{K} \sum_{k'=1, k'\neq k}^{K}  s_k s_{k'} \mu_k^T\mu_{k'} + \sum_{k=1}^{K}  s_k \mu_k^T\mu_k \right) + \sum_{k=1}^{K} s_k \log \frac{\pi_k}{1-\pi_k}\right\rangle_{q(\mu)}\]
 
@@
 
-\[\frac{\partial}{\partial q_i}\left( \exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q(\theta)} + \lambda^{LG} \int q_i -1)\right) \propto \exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q(\theta)\prod_{j\neq i} q_j(s_j)} - \log q_i(s_i)\]
+\[\frac{\partial}{\partial q_i}\left( \exp \left\langle \log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) \right\rangle_{q(\mu)} + \lambda^{LG} \int q_i -1)\right) \propto \exp \left\langle \log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) \right\rangle_{q(\mu)\prod_{j\neq i} q_j(s_j)} - \log q_i(s_i)\]
 
@@
 
-\[ \lambda_i =  \frac{1}{1+\exp\left[ - \left(\frac{  \langle \mu_i\rangle_{q_{\mu_i}}^T}{\langle\sigma^2\rangle_{q(\Psi)}} \left( \textbf{x} -  \frac{\langle \mu_i\rangle_{q_{\mu_i}}}{2} -  \sum_{j=1, j\neq i}^{K}   \lambda_j  \langle \mu_j\rangle_{q_{\mu_j} \right) + \log \frac{\langle \pi_i\rangle_{q_{\pi_i}}}{1-\langle \pi_i\rangle_{q_{\pi_i}}}\right) \right]}\]
+\[ \lambda_i =  \frac{1}{1+\exp\left[ - \left(\frac{  \langle \mu_i\rangle_{q_{\mu_i}}^T}{\sigma^2} \left( \textbf{x} -  \frac{\langle \mu_i\rangle_{q_{\mu_i}}}{2} -  \sum_{j=1, j\neq i}^{K}   \lambda_j  \langle \mu_j\rangle_{q_{\mu_j} \right) + \log \frac{ \pi_i}{1- \pi_i\right) \right]}\]
 
@@
 
-For the Variational Bayes maximisation step, we set:
-
-\[q_{\theta}(\theta) \propto P(\theta) \exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q(\textbf{s})}\]
-
-assuming the factorisation:
-
-\[q_{\theta}(\theta) = q_{\pi}(\pi)q_{\Psi}(\Psi)q_{\textbf{A}}(\textbf{A})\]
-
-we can calculate each factor independently.
-
-For $q_{\pi}(\pi)$:
-
-\[q_{\pi}(\pi) \propto P(\pi)\exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \pi}(\theta)}\]
+For the maximisation step, we perform maximisation steps for the parameters $\sigma$ and $\pi$ in the same way as question 3.
+However, having defined a prior on $\mu$ (through $\textbf{w}$) so we will have to derive our expression for $\langle \mu_k\rangle_{q_{\mu_k}}$ the expectation of the posterior on $\mu_k$.
+This involves deriving the posterior distribution of $\textbf{w}_d$
+
+\[q_{\textbf{w}_d}(\textbf{w}_d})  \propto P(\textbf{w}_d)\exp \left\langle \log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{w}_d}(\textbf{w}_d)}\]
 
@@
 
-\[q_{\pi}(\pi) \propto \left(\prod_{k=1}^{K} \frac{\pi_k^{\alpha - 1}(1-\pi_k)^{\beta-1}}{B(\alpha, \beta)} \right)\exp \left\langle \sum_{k=1}^K s_k \log \pi_k + (1 - s_k) \log (1-\pi_k) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \pi}(\theta)}\]
-
-We see:
-
-\[q_{\pi}(\pi) \propto \prod_{k=1}^{K} \frac{\pi_k^{\alpha  + \frac{1}{D}\left(\sum_{k=1}^K \left\langle s_k\right\rangle_{q_{s_k}}\right) - 1}(1-\pi_k)^{\beta-\frac{1}{D}\left(\sum_{k=1}^K \left\langle s_k\right\rangle_{q_{s_k}}\right)}}{B(\alpha, \beta)}\]
-
-\[q_{\pi}(\pi) = \prod_{k=1}^{K}  Beta\left(\alpha + \frac{1}{D}\sum_{k=1}^K \left\langle s_k\right\rangle_{q_{s_k}} , \beta+1- \frac{1}{D}\sum_{k=1}^K \left\langle s_k\right\rangle_{q_{s_k}}\right)\]
-
-%Substituting the appropriate terms:
+\[q_{\textbf{w}_d}(\textbf{w}_d})\propto \exp\left(-\frac{1}{2}\textbf{w}_{d}^T \textbf{A}\textbf{w}_{d} \right) \exp \left\langle  -\frac{1}{2 \sigma^2} \left(- 2 x_d \textbf{s}^T \textbf{w}_d   + \textbf{w}_d^T \textbf{s}\textbf{s}^T\textbf{w}_d\right)\right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{w}_d}(\textbf{w}_d)}\]
+
+Simplifying:
+
+\[q_{\textbf{w}_d}(\textbf{w}_d})\propto \exp\left(-\frac{1}{2}\left(\textbf{w}_{d}^T \left(\textbf{A}+\frac{\langle  \textbf{s}\textbf{s}^T \rangle_{q_{\textbf{s}}(\textbf{s})}}{\sigma^2} \right)\textbf{w}_{d} - 2 \left(\frac{x_d \langle\textbf{s}^T \rangle_{q_{\textbf{s}}(\textbf{s})}}{\sigma^2}}\right) \textbf{w}_d  \right)\right)\right)\]
+
+We see that the posterior:
+
+\[q_{\textbf{w}_d}(\textbf{w}_d) = \mathcal{N}\left(\mu_{\textbf{w}_d}, \Sigma_{\textbf{w}_d}\right)}\]
+
+where:
+
+\[ \Sigma_{\textbf{w}_d} = \left(\frac{\langle  \textbf{s}\textbf{s}^T \rangle_{q_{\textbf{s}}(\textbf{s})}}{\sigma^2} + \textbf{A}\right)^{-1}\]
+
+and
+
+\[\mu_{\textbf{w}_d} = \Sigma_{\textbf{w}_d}} \left(\frac{x_d \langle\textbf{s}^T \rangle_{q_{\textbf{s}}(\textbf{s})}}{\sigma^2}}\right)\]
+
+Thus, $\langle \mu_k\rangle_{q_{\mu_k}} \in \mathbb{R}^{D \times 1}$ is simply the concatenation of the $k^{th}$ elements of $\mu_{\textbf{w}_d}$ for $d\in \{1,...,D\}$
+
+For ARD, we must also optimise $\alpha$ with a hyper-M step.
+We start by choose $Ga(\alpha_k| a, b)$, a Gamma prior on \alpha_k$, with $a$ and $b$ being hyperparameters.
+Thus, to optimise $\alpha$ we want to maximise the penalised objective:
+
+\[ \alpha =  \arg \max_{\alpha} \left\langle \log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) \right\rangle_{q(\textbf{w})} + \sum_{k=1}^K \log P( \alpha_k| a, b)\]
+
+Substituting the appropriate terms, we have our penalised objective $\mathcal{Q}$:
+
+\[\mathcal{Q} =  \left\langle \sum_{d=1}^D \frac{1}{2}  \sum_{k=1}^K \left(\log\alpha_k \right)  -\frac{1}{2}\textbf{w}_{d}^T \textbf{A}\textbf{w}_{d} \right\rangle_{q(\textbf{w})} + \sum_{k=1}^K (a-1)\log\alpha_k -b\alpha_k\]
+
+Simplifying:
+
+\[\mathcal{Q} =  \frac{D}{2}  \sum_{k=1}^K \left( \log\alpha_k \right)  - \frac{1}{2}\sum_{d=1}^D \left( tr \left[ \textbf{A}\left\langle  \textbf{w}_{d}\textbf{w}_{d}^T \right\rangle_{q(\textbf{w}_d)} \right]\right)+ \sum_{k=1}^K (a-1)\log\alpha_k -b\alpha_k\]
+
+%Knowing $\left\langle  \textbf{w}_{d}\textbf{w}_{d}^T \right\rangle_{q(\textbf{w})} = \mu_{\textbf{w}_d}\mu_{\textbf{w}_d}^T + \Sigma_{\textbf{w}_d}$:
 %
-%\[q_{\pi}(\pi) \propto \left(\prod_{k=1}^{K} \frac{\pi_k^{\alpha - 1}(1-\pi_k)^{\beta-1}}{B(\alpha, \beta)} \right)\exp \left\langle \sum_{i=1}^K (s_k+\alpha-1) \log \pi_k + (\beta - s_k) \log (1-\pi_k) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \pi}(\theta)}\]
-%
-%We see:
-%
-%\[q_{\pi}(\pi) \propto \prod_{k=1}^{K} \frac{\pi_k^{2\alpha - 2 + \left\langle s_k\right\rangle_{q_{s_k}}}(1-\pi_k)^{2\beta-\left\langle s_k\right\rangle_{q_{s_k}}}}{B(\alpha, \beta)}\]
-%
-%\[q_{\pi}(\pi) = \prod_{k=1}^{K}  Beta(2\alpha -2 + \left\langle s_k\right\rangle_{q_{s_k}} , 2\beta+1-\left\langle s_k\right\rangle_{q_{s_k}}})\]
-
-For $q_{\Psi}(\Psi)$:
-
-\[q_{\Psi}(\Psi) \propto P(\Psi)\exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \Psi}(\theta)}\]
-
-Substituting the appropriate terms:
-
-\[
-q_{\Psi}(\Psi) \propto \left( \prod_{d=1}^D \frac{b^{a}}{\Gamma(a)}(\sigma^2)^{-a-1}\exp\left(-\frac{b}{\sigma^2}\right) \right)\exp \left\langle   -\frac{-D}{2} \log (\sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}-\textbf{A}\textbf{s} \right)^T\left(\textbf{x}-\textbf{A}\textbf{s} \right)  \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \Psi(\theta)}}
-\]
-
-We see:
-
-\[
-q_{\Psi}(\Psi) \propto  \prod_{d=1}^D \frac{b^{a}}{\Gamma(a)}(\sigma^2)^{-(a+\frac{1}{2})-1}\exp\left(-\frac{b+\frac{2}{D} \left\langle \left(\textbf{x}-\textbf{A}\textbf{s} \right)^T\left(\textbf{x}-\textbf{A}\textbf{s} \right) \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}}(\textbf{A})}}{\sigma^2}\right) }
-\]
-
-where $\textbf{A}_{d :} \in \mathbb{R}^{1 \times K}$ is the $d^{th}$ row of $\textbf{A}$.
-
-Thus,
-
-\[
-q_{\Psi}(\Psi) = \prod_{d=1}^D  InvGamma\left(\sigma^2 \left| a+\frac{1}{2}, b+\frac{2}{D} \left\langle \left(\textbf{x}-\textbf{A}\textbf{s} \right)^T\left(\textbf{x}-\textbf{A}\textbf{s} \right) \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}}(\textbf{A})}\right)
-\]
-
-For $q_{\mu_{k}}(\mu_{k})$:
-
-\[q_{\mu_{k}}(\mu_{k})  \propto P(\mu_{k})\exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \mu_{k}}(\theta)}\]
-
-Substituting the appropriate terms:
-
-\[q_{\mu_{k}}(\mu_{k}) \propto \exp\left(-\frac{\mu_{k}^T \mu_{k}}{2\alpha_k^{2}} \right) \exp \left\langle \frac{-1}{2\sigma^2} \left(- 2 \textbf{x}^T s_k  \mu_k  + s_k \mu_k^T\mu_k  + 2 \sum_{k'=1, k'\neq k}^{K}  s_k s_{k'} \mu_k^T\mu_{k'}}\right) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \mu_{k}}(\theta)}\]
-
-\[q_{\mu_{k}}(\mu_{k}) \propto \exp\left(-\frac{1}{2}\left( \mu_{k}^T\left(\frac{1}{\alpha_k^{2}} + \frac{\langle s_k \rangle_{q_{s_k}}}{\langle\sigma^2\rangle_{q_{\Psi}}}}\right)\mu_{k} + 2 \langle s_k \rangle_{q_{s_k}}\left(-  \textbf{x}^T +  \sum_{k'=1, k'\neq k}^{K} \langle s_{k'} \rangle_{q_{s_{k'}}} \langle \mu_{k'} \rangle_{q_{\mu_{k'}}}^T\right)\mu_k\right)\]
-
-We see that
-
-\[q_{\mu_{k}}(\mu_{k}) = \mathcal{N}\left(\mu_{\mu_{k}}, \Sigma_{\mu_{k}}\right)}\]
-
-where:
-
-\[ \Sigma_{\mu_{k}} = \left(\frac{1}{\alpha_k^{2}} + \frac{\langle s_k \rangle_{q_{s_k}}}{\langle\sigma^2\rangle_{q_{\Psi}}}}\right)^{-1} \textbf{I}\]
-
-and
-
-\[\mu_{\mu_{k}} = \Sigma_{\mu_{k}} \left(\textbf{x}^T - \sum_{k'=1, k'\neq k}^{K} \langle s_{k'} \rangle_{q_{s_{k'}}} \langle \mu_{k'} \rangle_{q_{\mu_{k'}}}^T\right)\langle s_k \rangle_{q_{s_k}}\]
-
-Optimising the parameters $\{a, b\}$, and $\alpha^2$ in turn, will cause some $\alpha_i^2$ to diverge, the number of remaining $\alpha_i^2$ provide our determination for the value of $K$.
+%\[\mathcal{Q} =  \frac{D}{2}  \sum_{k=1}^K \left( \log\alpha_k \right)  - \frac{1}{2}\sum_{d=1}^D \left( tr \left[ \textbf{A}\left(\mu_{\textbf{w}_d}\mu_{\textbf{w}_d}^T + \Sigma_{\textbf{w}_d}\right)} \right] \right)+ \sum_{k=1}^K (a-1)\log\alpha_k -b\alpha_k\]
+
+Setting $\frac{d\mathcal{Q}}{d\alpha_k} = 0$ we get:
+
+\[ \frac{D}{2\alpha_k} -\frac{1}{2} \sum_{d=1}^D\left\langle (w_{d, k})^2 \right\rangle_{q(\textbf{w}_{d})} + \frac{a-1}{\alpha_k} - b = 0\]
+
+where $w_{d, k}$ is the $k^{th}$ element of $\textbf{w}_d$.
+
+Knowing $\left\langle (w_{d, k})^2 \right\rangle_{q(\textbf{w}_{d})} = (\mu_{\textbf{w}_{d, k}})^2 + \Sigma_{\textbf{w}_{d, (k, k)}}$, we can solve for $\alpha_k$:
+
+\[ \alpha_k = \frac{2a+D-2}{2b+ \sum_{d=1}^D \left((\mu_{\textbf{w}_{d, k}})^2 + \Sigma_{\textbf{w}_{d, (k, k)}}\right) }\]
+
+we have our hyper-M steps for optimising $\alpha$.
+
+%Optimising the parameters $\{a, b\}$, and $\alpha^2$ in turn, will cause some $\alpha_i^2$ to diverge, the number of remaining $\alpha_i^2$ provide our determination for the value of $K$.
 This is automatic relevance determination through factor analysis.
+
+\newpage
+The Python code for variational Bayes:
+\lstinputlisting[language=Python]{src/models/variational_bayes.py}
+
+\newpage
+The rest of the Python code for question 4:
+\lstinputlisting[language=Python]{src/solutions/q4.py}
 
@@
 \lstinputlisting[language=Python]{main.py}
+\newpage
+\section*{Appendix 3: generate\_images.py}
+\lstinputlisting[language=Python]{src/generate_images.py}
+\newpage
+\section*{Appendix 4: MStep.py}
+\lstinputlisting[language=Python]{demo_code/MStep.py}
 \end{document}

### 2023-01-01 main.py

@@ -11,7 +11,7 @@ from src.generate_images import generate_images
 from src.models.bayesian_linear_regression import LinearRegressionParameters
 from src.models.gaussian_process_regression import GaussianProcessParameters
 from src.models.kernels import CombinedKernel, CombinedKernelParameters
-from src.models.variational_bayes import Beta, Gaussian, InverseGamma
+from src.models.variational_bayes import GaussianPrior
 from src.solutions import q2, q3, q4, q6
 
 jax.config.update("jax_enable_x64", True)
@@ -22,97 +22,97 @@ if __name__ == "__main__":
     if not os.path.exists(OUTPUTS_FOLDER):
         os.makedirs(OUTPUTS_FOLDER)
 
-    # Question 2
-    Q2_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q2")
-    if not os.path.exists(Q2_OUTPUT_FOLDER):
-        os.makedirs(Q2_OUTPUT_FOLDER)
-    with open(CO2_FILE_PATH) as file:
-        lines = [line.rstrip().split() for line in file]
-
-    df_co2 = pd.DataFrame(
-        np.array([line for line in lines if line[0] != "#"]).astype(float)
-    )
-    column_names = lines[max([i for i, line in enumerate(lines) if line[0] == "#"])][1:]
-    df_co2.columns = column_names
-    t = df_co2.decimal.values[:] - np.min(df_co2.decimal.values[:])
-    y = df_co2.average.values[:].reshape(1, -1)
-
-    sigma = 1
-    mean = np.array([0, 360]).reshape(-1, 1)
-    covariance = np.array(
-        [
-            [10**2, 0],
-            [0, 100**2],
-        ]
-    )
-    kernel = CombinedKernel()
-    kernel_parameters = CombinedKernelParameters(
-        log_theta=jnp.log(1),
-        log_sigma=jnp.log(1),
-        log_phi=jnp.log(1),
-        log_eta=jnp.log(1),
-        log_tau=jnp.log(1),
-        log_zeta=jnp.log(1e-1),
-    )
-
-    prior_linear_regression_parameters = LinearRegressionParameters(
-        mean=mean,
-        covariance=covariance,
-    )
-    posterior_linear_regression_parameters = q2.a(
-        t,
-        y,
[…]

### 2023-01-01 src/models/variational_bayes.py

@@ -1,153 +1,95 @@
-from abc import ABC, abstractmethod
-from dataclasses import dataclass
-
 import numpy as np
 
 from src.models.binary_latent_factor_model import (
     AbstractBinaryLatentFactorModel,
     BinaryLatentFactorApproximation,
+    BinaryLatentFactorModel,
 )
 
 
-@dataclass
-class Distribution(ABC):
-    @property
-    @abstractmethod
-    def mean(self):
-        pass
+class GaussianPrior:
+    def __init__(self, a, b, d, k):
+        self.a = a
+        self.b = b
+        self.mu = np.zeros((d, k))
+        self.alpha = np.ones((k,))  # np.random.gamma(a, b, size=(k,))
+        self.w_covariance = np.zeros((k, k))
 
+    def mu_k(self, k):  # (number_of_dimensions,  1)
+        return self.mu[:, k : k + 1]
 
-@dataclass
-class Beta(Distribution):
-    alpha: np.ndarray
-    beta: np.ndarray
-
-    @property
-    def mean(self) -> np.ndarray:
-        return np.divide(self.alpha, (self.alpha + self.beta))
-
-
-@dataclass
-class InverseGamma(Distribution):
-    a: float
-    b: float
+    def w_d(self, d):  # (1,  number_of_latent_variables)
+        return self.mu[d : d + 1, :]
 
     @property
-    def mean(self) -> float:
-        return self.b / (self.a - 1)
-
-
-@dataclass
-class Gaussian(Distribution):
-    mu: np.ndarray  # (number_of_dimensions,  number_of_latent_variables)
-    variance: np.ndarray  # (number_of_latent_variables, )
-
-    @property
-    def precision(self) -> np.ndarray:
-        return 1 / self.variance
-
-    @property
-    def mean(self) -> np.ndarray:
-        return self.mu
+    def a_matrix(self) -> np.ndarray:
+        #  precision matrix for w_d
+        return np.diag(self.alpha)
 
 
 class VariationalBayesBinaryLatentFactorModel(AbstractBinaryLatentFactorModel):
-    def __init__(self, mu: Gaussian, variance: InverseGamma, pi: Beta):
-        self._mu = mu
+    def __init__(self, mu: GaussianPrior, variance: float, pi: np.ndarray):
+        self.gaussian_prior = mu
         self._variance = variance
         self._pi = pi
 
     @property
     def variance(self) -> float:
-        return self._variance.mean
+        return self._variance
 
[…]

### 2023-01-01 src/solutions/q4.py

@@ -7,9 +7,7 @@ from src.models.binary_latent_factor_model import (
 )
 from src.models.mean_field_approximation import init_mean_field_approximation
 from src.models.variational_bayes import (
-    Beta,
-    Gaussian,
-    InverseGamma,
+    GaussianPrior,
     VariationalBayesBinaryLatentFactorModel,
 )
 
@@ -20,39 +18,24 @@ def b(
     em_iterations: int,
     e_maximum_steps: int,
     e_convergence_criterion: float,
-    # mu: Gaussian,
-    # variance: InverseGamma,
-    # pi: Beta,
+    mu: GaussianPrior,
     save_path: str,
 ) -> None:
     n = x.shape[0]
     mean_field_approximation = init_mean_field_approximation(
         k, n, max_steps=e_maximum_steps, convergence_criterion=e_convergence_criterion
     )
-    (
-        mu_max_step,
-        sigma,
-        pi_max_step,
-    ) = BinaryLatentFactorModel.calculate_maximisation_parameters(
+    (_, sigma, pi,) = BinaryLatentFactorModel.calculate_maximisation_parameters(
         x, mean_field_approximation
     )
-    mu = Gaussian(
-        mu=mu_max_step,
-        variance=mu_max_step.std(axis=0) ** 2,
+    binary_latent_factor_model: VariationalBayesBinaryLatentFactorModel = (
+        VariationalBayesBinaryLatentFactorModel(
+            mu=mu,
+            variance=sigma**2,
+            pi=pi,
+        )
     )
-    variance = InverseGamma(
-        a=2,
-        b=sigma**2,
-    )
-    pi = Beta(
-        alpha=np.ones((1, k)),
-        beta=np.divide(1 - pi_max_step, pi_max_step),
-    )
-    binary_latent_factor_model = VariationalBayesBinaryLatentFactorModel(
-        mu=mu,
-        variance=variance,
-        pi=pi,
-    )
+
     fig, ax = plt.subplots(1, k, figsize=(k * 2, 2))
     for i in range(k):
         ax[i].imshow(binary_latent_factor_model.mu[:, i].reshape(4, 4))
@@ -61,6 +44,7 @@ def b(
     plt.tight_layout()
     plt.savefig(save_path + "-init-latent-factors", bbox_inches="tight")
     plt.close()
+
     (
         mean_field_approximation,
         binary_latent_factor_model,
@@ -71,13 +55,26 @@ def b(
[…]

### 2023-01-01 main.py
update figures and report

@@ -11,7 +11,6 @@ from src.generate_images import generate_images
 from src.models.bayesian_linear_regression import LinearRegressionParameters
 from src.models.gaussian_process_regression import GaussianProcessParameters
 from src.models.kernels import CombinedKernel, CombinedKernelParameters
-from src.models.variational_bayes import GaussianPrior
 from src.solutions import q2, q3, q4, q6
 
 jax.config.update("jax_enable_x64", True)
@@ -22,97 +21,97 @@ if __name__ == "__main__":
     if not os.path.exists(OUTPUTS_FOLDER):
         os.makedirs(OUTPUTS_FOLDER)
 
-    # # Question 2
-    # Q2_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q2")
-    # if not os.path.exists(Q2_OUTPUT_FOLDER):
-    #     os.makedirs(Q2_OUTPUT_FOLDER)
-    # with open(CO2_FILE_PATH) as file:
-    #     lines = [line.rstrip().split() for line in file]
-    #
-    # df_co2 = pd.DataFrame(
-    #     np.array([line for line in lines if line[0] != "#"]).astype(float)
-    # )
-    # column_names = lines[max([i for i, line in enumerate(lines) if line[0] == "#"])][1:]
-    # df_co2.columns = column_names
-    # t = df_co2.decimal.values[:] - np.min(df_co2.decimal.values[:])
-    # y = df_co2.average.values[:].reshape(1, -1)
-    #
-    # sigma = 1
-    # mean = np.array([0, 360]).reshape(-1, 1)
-    # covariance = np.array(
-    #     [
-    #         [10**2, 0],
-    #         [0, 100**2],
-    #     ]
-    # )
-    # kernel = CombinedKernel()
-    # kernel_parameters = CombinedKernelParameters(
-    #     log_theta=jnp.log(1),
-    #     log_sigma=jnp.log(1),
-    #     log_phi=jnp.log(1),
-    #     log_eta=jnp.log(1),
-    #     log_tau=jnp.log(1),
-    #     log_zeta=jnp.log(1e-1),
-    # )
-    #
-    # prior_linear_regression_parameters = LinearRegressionParameters(
-    #     mean=mean,
-    #     covariance=covariance,
-    # )
-    # posterior_linear_regression_parameters = q2.a(
[…]

### 2023-01-01 src/solutions/q4.py

@@ -1,3 +1,5 @@
+from typing import List, Tuple
+
 import matplotlib.pyplot as plt
 import numpy as np
 
@@ -12,21 +14,27 @@ from src.models.variational_bayes import (
 )
 
 
-def b(
+def _run_automatic_relevance_determination(
     x: np.ndarray,
+    a_parameter: int,
+    b_parameter: int,
     k: int,
     em_iterations: int,
     e_maximum_steps: int,
     e_convergence_criterion: float,
-    mu: GaussianPrior,
-    save_path: str,
-) -> None:
+) -> Tuple[VariationalBayesBinaryLatentFactorModel, List[float]]:
     n = x.shape[0]
     mean_field_approximation = init_mean_field_approximation(
         k, n, max_steps=e_maximum_steps, convergence_criterion=e_convergence_criterion
     )
     (_, sigma, pi,) = BinaryLatentFactorModel.calculate_maximisation_parameters(
         x, mean_field_approximation
+    )
+    mu = GaussianPrior(
+        a=a_parameter,
+        b=b_parameter,
+        k=k,
+        d=x.shape[1],
     )
     binary_latent_factor_model: VariationalBayesBinaryLatentFactorModel = (
         VariationalBayesBinaryLatentFactorModel(
@@ -35,51 +43,93 @@ def b(
             pi=pi,
         )
     )
-
-    fig, ax = plt.subplots(1, k, figsize=(k * 2, 2))
-    for i in range(k):
-        ax[i].imshow(binary_latent_factor_model.mu[:, i].reshape(4, 4))
-        ax[i].set_title(f"Latent Feature mu_{i}")
-    fig.suptitle("Initial Features (Variational Bayes)")
-    plt.tight_layout()
-    plt.savefig(save_path + "-init-latent-factors", bbox_inches="tight")
-    plt.close()
-
-    (
-        mean_field_approximation,
-        binary_latent_factor_model,
-        free_energy,
-    ) = learn_binary_factors(
+    (_, binary_latent_factor_model, free_energy,) = learn_binary_factors(
         x=x,
         em_iterations=em_iterations,
         binary_latent_factor_model=binary_latent_factor_model,
         binary_latent_factor_approximation=mean_field_approximation,
     )
+    return binary_latent_factor_model, free_energy
 
-    sort_indices = np.argsort(binary_latent_factor_model.gaussian_prior.alpha)
 
[…]

### 2023-01-01 src/expectation_maximisation.py
split out em and binary approximation model

@@ -1,0 +1,60 @@
+from __future__ import annotations
+
+from typing import TYPE_CHECKING, List, Tuple
+
+import numpy as np
+
+if TYPE_CHECKING:
+    from src.models.binary_latent_factor_model import AbstractBinaryLatentFactorModel
+    from src.models.binary_latent_factor_model_approximation import (
+        BinaryLatentFactorApproximation,
+    )
+
+
+def is_converge(
+    free_energies: List[float],
+    current_lambda_matrix: np.ndarray,
+    previous_lambda_matrix: np.ndarray,
+) -> bool:
+    return (abs(free_energies[-1] - free_energies[-2]) == 0) and np.linalg.norm(
+        current_lambda_matrix - previous_lambda_matrix
+    ) == 0
+
+
+def learn_binary_factors(
+    x: np.ndarray,
+    em_iterations: int,
+    binary_latent_factor_model: AbstractBinaryLatentFactorModel,
+    binary_latent_factor_approximation: BinaryLatentFactorApproximation,
+) -> Tuple[
+    BinaryLatentFactorApproximation, AbstractBinaryLatentFactorModel, List[float]
+]:
+    free_energies: List[float] = [
+        binary_latent_factor_approximation.compute_free_energy(
+            x, binary_latent_factor_model
+        )
+    ]
+    for _ in range(em_iterations):
+        previous_lambda_matrix = np.copy(
+            binary_latent_factor_approximation.lambda_matrix
+        )
+        binary_latent_factor_approximation.variational_expectation_step(
+            x=x,
+            binary_latent_factor_model=binary_latent_factor_model,
+        )
+        binary_latent_factor_model.maximisation_step(
+            x,
+            binary_latent_factor_approximation,
+        )
+        free_energies.append(
+            binary_latent_factor_approximation.compute_free_energy(
+                x, binary_latent_factor_model
+            )
+        )
+        if is_converge(
+            free_energies,
+            binary_latent_factor_approximation.lambda_matrix,
+            previous_lambda_matrix,
+        ):
+            break
+    return binary_latent_factor_approximation, binary_latent_factor_model, free_energies

### 2023-01-01 src/models/binary_latent_factor_model.py

@@ -1,11 +1,16 @@
 from __future__ import annotations
 
 from abc import ABC, abstractmethod
-from typing import List, Tuple
+from typing import TYPE_CHECKING, List, Tuple
 
 import numpy as np
 
 from demo_code.MStep import m_step
+
+if TYPE_CHECKING:
+    from src.models.binary_latent_factor_model_approximation import (
+        BinaryLatentFactorApproximation,
+    )
 
 
 class AbstractBinaryLatentFactorModel(ABC):
@@ -136,178 +141,3 @@ def init_binary_latent_factor_model(
         x, binary_latent_factor_approximation
     )
     return BinaryLatentFactorModel(mu, sigma, pi)
-
-
-class BinaryLatentFactorApproximation(ABC):
-    @property
-    @abstractmethod
-    def lambda_matrix(self) -> np.ndarray:
-        pass
-
-    @abstractmethod
-    def variational_expectation_step(
-        self,
-        x: np.ndarray,
-        binary_latent_factor_model: AbstractBinaryLatentFactorModel,
-    ) -> List[float]:
-        pass
-
-    @property
-    def expectation_s(self):
-        return self.lambda_matrix
-
-    @property
-    def expectation_ss(self):
-        ess = self.lambda_matrix.T @ self.lambda_matrix
-        np.fill_diagonal(ess, self.lambda_matrix.sum(axis=0))
-        return ess
-
-    @property
-    def log_lambda_matrix(self) -> np.ndarray:
-        return np.log(self.lambda_matrix)
-
-    @property
-    def log_one_minus_lambda_matrix(self) -> np.ndarray:
-        return np.log(1 - self.lambda_matrix)
-
-    @property
-    def n(self) -> int:
-        return self.lambda_matrix.shape[0]
-
-    @property
-    def k(self) -> int:
-        return self.lambda_matrix.shape[1]
-
-    def compute_free_energy(
-        self,
-        x: np.ndarray,
-        binary_latent_factor_model: AbstractBinaryLatentFactorModel,
-    ) -> float:
-        """
-        free energy associated with current EM parameters and data x
-
-        :param x: data matrix (number_of_points, number_of_dimensions)
-        :param binary_latent_factor_model: a binary_latent_factor_model
-        :return: average free energy per data point
-        """
-        expectation_log_p_x_s_given_theta = (
-            self._compute_expectation_log_p_x_s_given_theta(
-                x, binary_latent_factor_model
-            )
-        )
-        approximation_model_entropy = self._compute_approximation_model_entropy()
[…]

### 2023-01-01 src/models/binary_latent_factor_model_approximation.py

@@ -1,0 +1,135 @@
+from __future__ import annotations
+
+from abc import ABC, abstractmethod
+from typing import TYPE_CHECKING, List
+
+if TYPE_CHECKING:
+    from src.models.binary_latent_factor_model import AbstractBinaryLatentFactorModel
+
+import numpy as np
+
+
+class BinaryLatentFactorApproximation(ABC):
+    @property
+    @abstractmethod
+    def lambda_matrix(self) -> np.ndarray:
+        pass
+
+    @abstractmethod
+    def variational_expectation_step(
+        self,
+        x: np.ndarray,
+        binary_latent_factor_model: AbstractBinaryLatentFactorModel,
+    ) -> List[float]:
+        pass
+
+    @property
+    def expectation_s(self):
+        return self.lambda_matrix
+
+    @property
+    def expectation_ss(self):
+        ess = self.lambda_matrix.T @ self.lambda_matrix
+        np.fill_diagonal(ess, self.lambda_matrix.sum(axis=0))
+        return ess
+
+    @property
+    def log_lambda_matrix(self) -> np.ndarray:
+        return np.log(self.lambda_matrix)
+
+    @property
+    def log_one_minus_lambda_matrix(self) -> np.ndarray:
+        return np.log(1 - self.lambda_matrix)
+
+    @property
+    def n(self) -> int:
+        return self.lambda_matrix.shape[0]
+
+    @property
+    def k(self) -> int:
+        return self.lambda_matrix.shape[1]
+
+    def compute_free_energy(
+        self,
+        x: np.ndarray,
+        binary_latent_factor_model: AbstractBinaryLatentFactorModel,
+    ) -> float:
+        """
+        free energy associated with current EM parameters and data x
+
+        :param x: data matrix (number_of_points, number_of_dimensions)
+        :param binary_latent_factor_model: a binary_latent_factor_model
+        :return: average free energy per data point
+        """
+        expectation_log_p_x_s_given_theta = (
+            self._compute_expectation_log_p_x_s_given_theta(
+                x, binary_latent_factor_model
+            )
+        )
+        approximation_model_entropy = self._compute_approximation_model_entropy()
+        return (
+            expectation_log_p_x_s_given_theta + approximation_model_entropy
+        ) / self.n
+
+    def _compute_expectation_log_p_x_s_given_theta(
+        self,
+        x: np.ndarray,
+        binary_latent_factor_model: AbstractBinaryLatentFactorModel,
+    ) -> float:
+        """
[…]

### 2023-01-01 src/models/boltzmann_machine.py

@@ -1,8 +1,8 @@
 import numpy as np
 
-from src.models.binary_latent_factor_model import (
+from src.models.binary_latent_factor_model import BinaryLatentFactorModel
+from src.models.binary_latent_factor_model_approximation import (
     BinaryLatentFactorApproximation,
-    BinaryLatentFactorModel,
 )
 
 

### 2023-01-01 src/models/mean_field_approximation.py

@@ -2,8 +2,8 @@ from typing import List
 
 import numpy as np
 
-from src.models.binary_latent_factor_model import (
-    AbstractBinaryLatentFactorModel,
+from src.models.binary_latent_factor_model import AbstractBinaryLatentFactorModel
+from src.models.binary_latent_factor_model_approximation import (
     BinaryLatentFactorApproximation,
 )
 

### 2023-01-01 src/models/message_passing.py

@@ -2,7 +2,9 @@ from typing import List
 
 import numpy as np
 
-from src.models.binary_latent_factor_model import BinaryLatentFactorApproximation
+from src.models.binary_latent_factor_model_approximation import (
+    BinaryLatentFactorApproximation,
+)
 from src.models.boltzmann_machine import BoltzmannMachine
 
 

### 2023-01-01 src/models/variational_bayes.py

@@ -2,8 +2,10 @@ import numpy as np
 
 from src.models.binary_latent_factor_model import (
     AbstractBinaryLatentFactorModel,
-    BinaryLatentFactorApproximation,
     BinaryLatentFactorModel,
+)
+from src.models.binary_latent_factor_model_approximation import (
+    BinaryLatentFactorApproximation,
 )
 
 

### 2023-01-01 src/solutions/q3.py

@@ -3,12 +3,10 @@ from typing import List
 import matplotlib.pyplot as plt
 import numpy as np
 
+from src.expectation_maximisation import is_converge, learn_binary_factors
 from src.models.binary_latent_factor_model import (
     AbstractBinaryLatentFactorModel,
-    BinaryLatentFactorModel,
     init_binary_latent_factor_model,
-    is_converge,
-    learn_binary_factors,
 )
 from src.models.mean_field_approximation import init_mean_field_approximation
 

### 2023-01-01 src/solutions/q4.py

@@ -3,10 +3,8 @@ from typing import List, Tuple
 import matplotlib.pyplot as plt
 import numpy as np
 
-from src.models.binary_latent_factor_model import (
-    BinaryLatentFactorModel,
-    learn_binary_factors,
-)
+from src.expectation_maximisation import learn_binary_factors
+from src.models.binary_latent_factor_model import BinaryLatentFactorModel
 from src.models.mean_field_approximation import init_mean_field_approximation
 from src.models.variational_bayes import (
     GaussianPrior,

### 2023-01-01 src/solutions/q6.py

@@ -1,7 +1,7 @@
 import matplotlib.pyplot as plt
 import numpy as np
 
-from src.models.binary_latent_factor_model import learn_binary_factors
+from src.expectation_maximisation import learn_binary_factors
 from src.models.boltzmann_machine import init_boltzmann_machine
 from src.models.message_passing import init_message_passing
 

### 2023-01-01 submission.tex
update figures and report
split out em and binary approximation model
add new python files to report

@@
 \newpage
+The Python code for an abstract binary latent factor model approximation:
+\lstinputlisting[language=Python]{src/models/binary_latent_factor_model_approximation.py}
+
+\newpage
 The Python code for mean field learning:
 \lstinputlisting[language=Python]{src/models/mean_field_approximation.py}
+
+\newpage
+The Python code for expectation maximisation:
+\lstinputlisting[language=Python]{src/expectation_maximisation.py}
 
@@
 
-%Optimising the parameters $\{a, b\}$, and $\alpha^2$ in turn, will cause some $\alpha_i^2$ to diverge, the number of remaining $\alpha_i^2$ provide our determination for the value of $K$.
-This is automatic relevance determination through factor analysis.
-
-\newpage
-The Python code for variational Bayes:
+\newpage
+\begin{figure}[h]
+\centering
+\begin{minipage}{.6\textwidth}
+  \centering
+\includegraphics[scale=0.15]{outputs/q4/b-latent-factors}
+\caption{Learned Latent Factors}
+\label{fig:}
+\end{minipage}%
+\begin{minipage}{.2\textwidth}
+  \centering
+\includegraphics[scale=0.11]{outputs/q4/b-alpha-trained}
+\caption{Alpha values}
+\label{fig:}
+\end{minipage}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.8]{outputs/q4/b-free-energy}
+\caption{Free Energy for different values of k}
+\label{fig:}
+\end{figure}
+
+\newpage
+The Python code for Variational Bayes:
 \lstinputlisting[language=Python]{src/models/variational_bayes.py}

### 2023-01-01 main.py
update gamma prior parameters for vb

@@ -159,7 +159,7 @@ if __name__ == "__main__":
     q4.b(
         x=x,
         a_parameter=1,
-        b_parameter=1,
+        b_parameter=0,
         ks=ks,
         em_iterations=em_iterations,
         e_maximum_steps=e_maximum_steps,

### 2023-01-01 src/solutions/q4.py

@@ -96,28 +96,18 @@ def b(
     plt.close()
 
     fig, ax = plt.subplots(len(ks), 1, figsize=(12, 6 + 3 * len(ks)))
-    max_alpha = (
-        np.max(
-            [
-                np.max(binary_latent_factor_models[i].gaussian_prior.alpha)
-                for i in range(len(ks))
-            ]
-        )
-        * 1.1
-    )
     for i, k in enumerate(ks):
         sort_indices = np.argsort(binary_latent_factor_models[i].gaussian_prior.alpha)
         ax[i].set_title(f"{k=}")
         ax[i].set_xlabel("Latent Factor")
-        ax[i].set_ylabel("Alpha")
+        ax[i].set_ylabel("Inverse Alpha")
         ax[i].bar(
             [str(x + 1) for x in sort_indices]
             + [" " * (j + 1) for j in range(np.max(ks) - k)],
-            list(binary_latent_factor_models[i].gaussian_prior.alpha[sort_indices])
+            list(1 / binary_latent_factor_models[i].gaussian_prior.alpha[sort_indices])
             + [0] * (np.max(ks) - k),
         )
-        ax[i].set_ylim([0, max_alpha])
-    fig.suptitle("Alpha values (after optimisation)")
+    fig.suptitle("Inverse Alpha values (after optimisation)")
     plt.tight_layout()
     plt.savefig(save_path + "-alpha-trained", bbox_inches="tight")
     plt.close()

### 2023-01-02 main.py
update plots

@@ -120,8 +120,8 @@ if __name__ == "__main__":
     number_of_images = 2000
     x = generate_images(n=number_of_images)
     k = 8
-    em_iterations = 200
-    e_maximum_steps = 100
+    em_iterations = 100
+    e_maximum_steps = 50
     e_convergence_criterion = 0
 
     binary_latent_factor_model = q3.e_and_f(
@@ -150,23 +150,44 @@ if __name__ == "__main__":
         e_convergence_criterion=e_convergence_criterion,
         save_path=os.path.join(Q3_OUTPUT_FOLDER, "g"),
     )
-
     # Question 4
     Q4_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q4")
     if not os.path.exists(Q4_OUTPUT_FOLDER):
         os.makedirs(Q4_OUTPUT_FOLDER)
-    ks = np.arange(int(k / 2), int(2 * k) + 1)
+    max_k = 21
     q4.b(
         x=x,
         a_parameter=1,
         b_parameter=0,
-        ks=ks,
+        ks=np.arange(4, 22),
+        max_k=max_k,
         em_iterations=em_iterations,
         e_maximum_steps=e_maximum_steps,
         e_convergence_criterion=e_convergence_criterion,
         save_path=os.path.join(Q4_OUTPUT_FOLDER, "b"),
     )
-
+    q4.b(
+        x=x,
+        a_parameter=1,
+        b_parameter=0,
+        ks=np.arange(4, 13),
+        max_k=max_k,
+        em_iterations=em_iterations,
+        e_maximum_steps=e_maximum_steps,
+        e_convergence_criterion=e_convergence_criterion,
+        save_path=os.path.join(Q4_OUTPUT_FOLDER, "b-1"),
+    )
+    q4.b(
+        x=x,
+        a_parameter=1,
+        b_parameter=0,
+        ks=np.arange(13, 22),
+        max_k=max_k,
+        em_iterations=em_iterations,
+        e_maximum_steps=e_maximum_steps,
+        e_convergence_criterion=e_convergence_criterion,
+        save_path=os.path.join(Q4_OUTPUT_FOLDER, "b-2"),
+    )
     # Question 6
     Q6_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q6")
     if not os.path.exists(Q6_OUTPUT_FOLDER):

### 2023-01-02 src/solutions/q4.py

@@ -1,7 +1,9 @@
+import os
 from typing import List, Tuple
 
 import matplotlib.pyplot as plt
 import numpy as np
+from matplotlib.offsetbox import AnnotationBbox, OffsetImage
 
 from src.expectation_maximisation import learn_binary_factors
 from src.models.binary_latent_factor_model import BinaryLatentFactorModel
@@ -10,6 +12,24 @@ from src.models.variational_bayes import (
     GaussianPrior,
     VariationalBayesBinaryLatentFactorModel,
 )
+
+
+def offset_image(coord, path, ax):
+    img = plt.imread(path)
+    im = OffsetImage(img, zoom=0.72)
+    im.image.axes = ax
+
+    ab = AnnotationBbox(
+        im,
+        (coord, 0),
+        xybox=(0.0, -19.0),
+        frameon=False,
+        xycoords="data",
+        boxcoords="offset points",
+        pad=0,
+    )
+
+    ax.add_artist(ab)
 
 
 def _run_automatic_relevance_determination(
@@ -55,6 +75,7 @@ def b(
     a_parameter: int,
     b_parameter: int,
     ks: List[int],
+    max_k: int,
     em_iterations: int,
     e_maximum_steps: int,
     e_convergence_criterion: float,
@@ -95,21 +116,35 @@ def b(
     plt.savefig(save_path + "-latent-factors", bbox_inches="tight")
     plt.close()
 
-    fig, ax = plt.subplots(len(ks), 1, figsize=(12, 6 + 3 * len(ks)))
     for i, k in enumerate(ks):
         sort_indices = np.argsort(binary_latent_factor_models[i].gaussian_prior.alpha)
+        for j, idx in enumerate(sort_indices):
+            fig = plt.figure(figsize=(0.3, 0.3))
+            ax = plt.Axes(fig, [0.0, 0.0, 1.0, 1.0])
+            ax.set_axis_off()
+            fig.add_axes(ax)
+            ax.imshow(binary_latent_factor_models[i].mu[:, idx].reshape(4, 4))
+            fig.savefig(save_path + f"-latent-factor-{i}-{j}", bbox_inches="tight")
+            plt.close()
+
+    fig, ax = plt.subplots(len(ks), 1, figsize=(10, 2 * len(ks)))
+    plt.subplots_adjust(hspace=1)
+    for i, k in enumerate(ks):
+        sort_indices = np.argsort(binary_latent_factor_models[i].gaussian_prior.alpha)
+        y = list(
[…]

### 2023-01-02 submission.tex
update gamma prior parameters for vb
update plots
refactor folder structure for models

@@
 The Python code for the binary latent factor model:
-\lstinputlisting[language=Python]{src/models/binary_latent_factor_model.py}
-
-\newpage
-The Python code for an abstract binary latent factor model approximation:
-\lstinputlisting[language=Python]{src/models/binary_latent_factor_model_approximation.py}
+\lstinputlisting[language=Python]{src/models/binary_latent_factor_models/binary_latent_factor_model.py}
 
@@
 The Python code for mean field learning:
-\lstinputlisting[language=Python]{src/models/mean_field_approximation.py}
+\lstinputlisting[language=Python]{src/models/binary_latent_factor_approximations/mean_field_approximation.py}
 
@@
 \centering
-\begin{minipage}{.6\textwidth}
-  \centering
-\includegraphics[scale=0.15]{outputs/q4/b-latent-factors}
-\caption{Learned Latent Factors}
+\includegraphics[scale=0.47]{outputs/q4/b-1-latent-factors-comparison}
+\caption{Learned Latent Factors vs Inverse Alpha}
 \label{fig:}
-\end{minipage}%
-\begin{minipage}{.2\textwidth}
-  \centering
-\includegraphics[scale=0.11]{outputs/q4/b-alpha-trained}
-\caption{Alpha values}
+\end{figure}
+
+\newpage
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.47]{outputs/q4/b-2-latent-factors-comparison}
+\caption{Learned Latent Factors vs Inverse Alpha}
 \label{fig:}
-\end{minipage}
 \end{figure}
@@
 The Python code for Variational Bayes:
-\lstinputlisting[language=Python]{src/models/variational_bayes.py}
+\lstinputlisting[language=Python]{src/models/binary_latent_factor_models/variational_bayes.py}
 
@@
 The Python code for the Boltzmann machine:
-\lstinputlisting[language=Python]{src/models/boltzmann_machine.py}
+\lstinputlisting[language=Python]{src/models/binary_latent_factor_models/boltzmann_machine.py}
 
@@
 The Python code for message passing:
-\lstinputlisting[language=Python]{src/models/message_passing.py}
+\lstinputlisting[language=Python]{src/models/binary_latent_factor_approximations/message_passing.py}
 
@@
 \newpage
-\section*{Appendix 1: constants.py}
+\section*{Appendix 1: abstract\_binary\_latent\_factor\_model.py}
+\lstinputlisting[language=Python]{src/models/binary_latent_factor_models/abstract_binary_latent_factor_model.py}
+
+\newpage
+\section*{Appendix 2: abstract\_binary\_latent\_factor\_approximation.py}
+\lstinputlisting[language=Python]{src/models/binary_latent_factor_approximations/abstract_binary_latent_factor_approximation.py}
+
+\newpage
+\section*{Appendix 3: main.py}
+\lstinputlisting[language=Python]{main.py}
+
+\newpage
+\section*{Appendix 4: constants.py}
 \lstinputlisting[language=Python]{src/constants.py}
-\newpage
-\section*{Appendix 2: main.py}
-\lstinputlisting[language=Python]{main.py}
-\newpage
-\section*{Appendix 3: generate\_images.py}
+
+\newpage
+\section*{Appendix 5: generate\_images.py}
 \lstinputlisting[language=Python]{src/generate_images.py}
 \newpage
-\section*{Appendix 4: MStep.py}
+\section*{Appendix 6: MStep.py}
 \lstinputlisting[language=Python]{demo_code/MStep.py}

### 2023-01-02 src/expectation_maximisation.py

@@ -5,9 +5,11 @@ from typing import TYPE_CHECKING, List, Tuple
 import numpy as np
 
 if TYPE_CHECKING:
-    from src.models.binary_latent_factor_model import AbstractBinaryLatentFactorModel
-    from src.models.binary_latent_factor_model_approximation import (
-        BinaryLatentFactorApproximation,
+    from src.models.binary_latent_factor_approximations.abstract_binary_latent_factor_approximation import (
+        AbstractBinaryLatentFactorApproximation,
+    )
+    from src.models.binary_latent_factor_models.binary_latent_factor_model import (
+        AbstractBinaryLatentFactorModel,
     )
 
 
@@ -25,9 +27,11 @@ def learn_binary_factors(
     x: np.ndarray,
     em_iterations: int,
     binary_latent_factor_model: AbstractBinaryLatentFactorModel,
-    binary_latent_factor_approximation: BinaryLatentFactorApproximation,
+    binary_latent_factor_approximation: AbstractBinaryLatentFactorApproximation,
 ) -> Tuple[
-    BinaryLatentFactorApproximation, AbstractBinaryLatentFactorModel, List[float]
+    AbstractBinaryLatentFactorApproximation,
+    AbstractBinaryLatentFactorModel,
+    List[float],
 ]:
     free_energies: List[float] = [
         binary_latent_factor_approximation.compute_free_energy(

### 2023-01-02 src/models/binary_latent_factor_approximations/abstract_binary_latent_factor_approximation.py

@@ -4,12 +4,14 @@ from abc import ABC, abstractmethod
 from typing import TYPE_CHECKING, List
 
 if TYPE_CHECKING:
-    from src.models.binary_latent_factor_model import AbstractBinaryLatentFactorModel
+    from src.models.binary_latent_factor_models.binary_latent_factor_model import (
+        AbstractBinaryLatentFactorModel,
+    )
 
 import numpy as np
 
 
-class BinaryLatentFactorApproximation(ABC):
+class AbstractBinaryLatentFactorApproximation(ABC):
     @property
     @abstractmethod
     def lambda_matrix(self) -> np.ndarray:

### 2023-01-02 src/models/binary_latent_factor_approximations/mean_field_approximation.py

@@ -2,13 +2,15 @@ from typing import List
 
 import numpy as np
 
-from src.models.binary_latent_factor_model import AbstractBinaryLatentFactorModel
-from src.models.binary_latent_factor_model_approximation import (
-    BinaryLatentFactorApproximation,
+from src.models.binary_latent_factor_approximations.abstract_binary_latent_factor_approximation import (
+    AbstractBinaryLatentFactorApproximation,
+)
+from src.models.binary_latent_factor_models.binary_latent_factor_model import (
+    AbstractBinaryLatentFactorModel,
 )
 
 
-class MeanFieldApproximation(BinaryLatentFactorApproximation):
+class MeanFieldApproximation(AbstractBinaryLatentFactorApproximation):
     """
     lambda_matrix: parameters variational approximation (number_of_points, number_of_latent_variables)
     """

### 2023-01-02 src/models/binary_latent_factor_approximations/message_passing.py

@@ -2,13 +2,13 @@ from typing import List
 
 import numpy as np
 
-from src.models.binary_latent_factor_model_approximation import (
-    BinaryLatentFactorApproximation,
+from src.models.binary_latent_factor_approximations.abstract_binary_latent_factor_approximation import (
+    AbstractBinaryLatentFactorApproximation,
 )
-from src.models.boltzmann_machine import BoltzmannMachine
+from src.models.binary_latent_factor_models.boltzmann_machine import BoltzmannMachine
 
 
-class MessagePassing(BinaryLatentFactorApproximation):
+class MessagePassing(AbstractBinaryLatentFactorApproximation):
     """
     eta_matrix:  of parameters eta_matrix[n, i, j]
                 off diagonals corresponds to \tilda{g}_{ij, \neg s_i}(s_j) for data point n

### 2023-01-02 src/models/binary_latent_factor_models/abstract_binary_latent_factor_model.py

@@ -1,0 +1,62 @@
+from __future__ import annotations
+
+from abc import ABC, abstractmethod
+from typing import TYPE_CHECKING
+
+import numpy as np
+
+if TYPE_CHECKING:
+    from src.models.binary_latent_factor_approximations.abstract_binary_latent_factor_approximation import (
+        AbstractBinaryLatentFactorApproximation,
+    )
+
+
+class AbstractBinaryLatentFactorModel(ABC):
+    @property
+    @abstractmethod
+    def mu(self) -> np.ndarray:
+        pass
+
+    @property
+    @abstractmethod
+    def variance(self) -> float:
+        pass
+
+    @property
+    @abstractmethod
+    def pi(self) -> np.ndarray:
+        pass
+
+    @abstractmethod
+    def maximisation_step(
+        self,
+        x: np.ndarray,
+        binary_latent_factor_approximation: AbstractBinaryLatentFactorApproximation,
+    ) -> None:
+        pass
+
+    def mu_exclude(self, exclude_latent_index: int) -> np.ndarray:
+        return np.concatenate(  # (number_of_dimensions, number_of_latent_variables-1)
+            (self.mu[:, :exclude_latent_index], self.mu[:, exclude_latent_index + 1 :]),
+            axis=1,
+        )
+
+    @property
+    def log_pi(self) -> np.ndarray:
+        return np.log(self.pi)
+
+    @property
+    def log_one_minus_pi(self) -> np.ndarray:
+        return np.log(1 - self.pi)
+
+    @property
+    def precision(self) -> float:
+        return 1 / self.variance
+
+    @property
+    def d(self) -> int:
+        return self.mu.shape[0]
+
+    @property
+    def k(self) -> int:
+        return self.mu.shape[1]

### 2023-01-02 src/models/binary_latent_factor_models/boltzmann_machine.py

@@ -1,8 +1,10 @@
 import numpy as np
 
-from src.models.binary_latent_factor_model import BinaryLatentFactorModel
-from src.models.binary_latent_factor_model_approximation import (
-    BinaryLatentFactorApproximation,
+from src.models.binary_latent_factor_approximations.abstract_binary_latent_factor_approximation import (
+    AbstractBinaryLatentFactorApproximation,
+)
+from src.models.binary_latent_factor_models.binary_latent_factor_model import (
+    BinaryLatentFactorModel,
 )
 
 
@@ -59,7 +61,7 @@ class BoltzmannMachine(BinaryLatentFactorModel):
 
 def init_boltzmann_machine(
     x: np.ndarray,
-    binary_latent_factor_approximation: BinaryLatentFactorApproximation,
+    binary_latent_factor_approximation: AbstractBinaryLatentFactorApproximation,
 ) -> BinaryLatentFactorModel:
     mu, sigma, pi = BinaryLatentFactorModel.calculate_maximisation_parameters(
         x, binary_latent_factor_approximation

### 2023-01-02 src/models/binary_latent_factor_models/variational_bayes.py

@@ -1,11 +1,11 @@
 import numpy as np
 
-from src.models.binary_latent_factor_model import (
+from src.models.binary_latent_factor_approximations.abstract_binary_latent_factor_approximation import (
+    AbstractBinaryLatentFactorApproximation,
+)
+from src.models.binary_latent_factor_models.binary_latent_factor_model import (
     AbstractBinaryLatentFactorModel,
     BinaryLatentFactorModel,
-)
-from src.models.binary_latent_factor_model_approximation import (
-    BinaryLatentFactorApproximation,
 )
 
 
@@ -48,7 +48,8 @@ class VariationalBayesBinaryLatentFactorModel(AbstractBinaryLatentFactorModel):
         return self.gaussian_prior.mu
 
     def _update_w_d_covariance(
-        self, binary_latent_factor_approximation: BinaryLatentFactorApproximation
+        self,
+        binary_latent_factor_approximation: AbstractBinaryLatentFactorApproximation,
     ):
         #  expectation_s (number_of_points, number_of_latent_variables)
         #  expectation_ss (number_of_latent_variables, number_of_latent_variables)
@@ -60,7 +61,7 @@ class VariationalBayesBinaryLatentFactorModel(AbstractBinaryLatentFactorModel):
     def _update_w_d_mean(
         self,
         x: np.ndarray,  # (number_of_points, number_of_dimensions)
-        binary_latent_factor_approximation: BinaryLatentFactorApproximation,
+        binary_latent_factor_approximation: AbstractBinaryLatentFactorApproximation,
         d: int,
     ):
         # (number_of_latent_variables x 1)
@@ -84,7 +85,7 @@ class VariationalBayesBinaryLatentFactorModel(AbstractBinaryLatentFactorModel):
     def maximisation_step(
         self,
         x: np.ndarray,
-        binary_latent_factor_approximation: BinaryLatentFactorApproximation,
+        binary_latent_factor_approximation: AbstractBinaryLatentFactorApproximation,
     ) -> None:
         _, sigma, pi = BinaryLatentFactorModel.calculate_maximisation_parameters(
             x, binary_latent_factor_approximation

### 2023-01-02 src/solutions/q3.py

@@ -4,11 +4,13 @@ import matplotlib.pyplot as plt
 import numpy as np
 
 from src.expectation_maximisation import is_converge, learn_binary_factors
-from src.models.binary_latent_factor_model import (
+from src.models.binary_latent_factor_approximations.mean_field_approximation import (
+    init_mean_field_approximation,
+)
+from src.models.binary_latent_factor_models.binary_latent_factor_model import (
     AbstractBinaryLatentFactorModel,
     init_binary_latent_factor_model,
 )
-from src.models.mean_field_approximation import init_mean_field_approximation
 
 
 def e_and_f(

### 2023-01-02 src/solutions/q4.py

@@ -6,9 +6,13 @@ import numpy as np
 from matplotlib.offsetbox import AnnotationBbox, OffsetImage
 
 from src.expectation_maximisation import learn_binary_factors
-from src.models.binary_latent_factor_model import BinaryLatentFactorModel
-from src.models.mean_field_approximation import init_mean_field_approximation
-from src.models.variational_bayes import (
+from src.models.binary_latent_factor_approximations.mean_field_approximation import (
+    init_mean_field_approximation,
+)
+from src.models.binary_latent_factor_models.binary_latent_factor_model import (
+    BinaryLatentFactorModel,
+)
+from src.models.binary_latent_factor_models.variational_bayes import (
     GaussianPrior,
     VariationalBayesBinaryLatentFactorModel,
 )

### 2023-01-02 src/solutions/q6.py

@@ -2,8 +2,12 @@ import matplotlib.pyplot as plt
 import numpy as np
 
 from src.expectation_maximisation import learn_binary_factors
-from src.models.boltzmann_machine import init_boltzmann_machine
-from src.models.message_passing import init_message_passing
+from src.models.binary_latent_factor_approximations.message_passing import (
+    init_message_passing,
+)
+from src.models.binary_latent_factor_models.boltzmann_machine import (
+    init_boltzmann_machine,
+)
 
 
 def run(x: np.ndarray, k: int, em_iterations: int, save_path: str) -> None:

### 2023-01-02 submission.tex
update report with factor analysis and q3 convergence

@@
 
+\newpage
+
 \subsection*{(b)} The moralised graph:
@@
 
-\newpage
-
 An effective triangulation:
-
-%\begin{center}
-%\begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}]
-%    \node[main] (A) {$A$};
-%    \node[main] (B) [below left of=A] {$B$};
-%    \node[main] (C) [below right of=A] {$C$};
-%    \node[main] (E) [below right of=C] {$E$};
-%    \node[main] (D) [below left of=C] {$D$};
-%    \node[main] (H) [below of=E] {$H$};
-%    \node[main] (G) [right of=E] {$G$};
-%    \node[main] (I) [below left of=H] {$I$};
-%    \node[main] (F) [above left of=I] {$F$};
-%
-%    \draw (A) -- (B);
-%    \draw (A) -- (C);
-%    \draw (B) -- (F);
-%    \draw (B) -- (D);
-%    \draw (C) -- (D);
-%    \draw (C) -- (E);
-%    \draw (E) -- (G);
-%    \draw (G) -- (H);
-%    \draw (D) -- (F);
-%    \draw (E) -- (H);
-%    \draw (F) -- (H);
-%    \draw (F) -- (I);
-%    \draw (H) -- (I);
-%
-%    \draw[dashed] (I) -- (E);
-%    \draw[dashed] (F) -- (E);
-%    \draw[dashed] (C) -- (F);
-%    \draw[dashed] (B) -- (D);
-%    \draw[dashed] (B) -- (C);
-%
-%\end{tikzpicture}
-%\end{center}
 
@@
 where the dashed lines are edges added to triangulate the moralised graph.
-
 \newpage
@@
 %\begin{center}
-%\begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}, square/.style={regular polygon,regular polygon sides=4}]
+%\begin{tikzpicture}[node distance={25mm}, main/.style = {draw, circle}, square/.style={regular polygon,regular polygon sides=4}]
 %    \node[main] (ABC) {$ABC$};
-%    \node[square] (BC) [square,draw] [below of=ABC] {$BC$};
-%    \node[main] (BCDF) [below of=BC] {$BCDF$};
-%    \node[square] (CF) [square,draw] [below of=BCDF] {$CF$};
-%    \node[main] (CEF) [below of=CF] {$CEF$};
-%    \node[square] (EF) [square,draw] [below of=CEF] {$EF$};
-%    \node[main] (EFHI) [below of=EF] {$EFHI$};
-%    \node[square] (EH) [square,draw] [below of=EFHI] {$EH$};
-%    \node[main] (EGH) [below of=EH] {$EGH$};
-%    \draw (ABC) -- (BC);
-%    \draw (BC) -- (BCDF);
-%    \draw (BCDF) -- (CF);
-%    \draw (CF) -- (CEF);
-%    \draw (CEF) -- (EF);
-%    \draw (EF) -- (EFHI);
-%    \draw (EFHI) -- (EH);
-%    \draw (EH) -- (EGH);
+%    \node[main] (BCDF) [below of=ABC] {$BCDF$};
+%    \node[main] (CFH) [below of=BCDF] {$CFH$};
+%    \node[main] (FHI) [below left of=CFH] {$FHI$};
+%    \node[main] (CEH) [below right of=CFH] {$CEH$};
+%    \node[main] (EGH) [below of=CEH] {$EGH$};
+%    \draw (ABC) -- (BCDF);
+%    \draw (BCDF) -- (CFH);
+%    \draw (CFH) -- (FHI);
+%    \draw (CFH)  -- (CEH);
+%    \draw (EGH)  -- (CEH);
 %\end{tikzpicture}
 %\end{center}
-
-\begin{center}
-\begin{tikzpicture}[node distance={25mm}, main/.style = {draw, circle}, square/.style={regular polygon,regular polygon sides=4}]
-    \node[main] (ABC) {$ABC$};
-    \node[main] (BCDF) [below of=ABC] {$BCDF$};
-    \node[main] (CFH) [below of=BCDF] {$CFH$};
-    \node[main] (FHI) [below left of=CFH] {$FHI$};
-    \node[main] (CEH) [below right of=CFH] {$CEH$};
-    \node[main] (EGH) [below of=CEH] {$EGH$};
-    \draw (ABC) -- (BCDF);
-    \draw (BCDF) -- (CFH);
-    \draw (CFH) -- (FHI);
-    \draw (CFH)  -- (CEH);
-    \draw (EGH)  -- (CEH);
-\end{tikzpicture}
-\end{center}
-
-where the circular nodes are cliques.
-
-\newpage
-
-The junction tree redrawn as a factor graph:
 
@@
 
+The junction tree redrawn as a factor graph:
+
+\begin{center}
+\begin{tikzpicture}[node distance={10mm}, main/.style = {draw, circle}, square/.style={fill=black, regular polygon,regular polygon sides=4}]
+    \node[main] (A) {$A$};
+    \node[square] (ABC) [square,draw] [below of=A];
+    \node[main] (B) [below left of=ABC] {$B$};
+    \node[main] (C) [below right of=ABC] {$C$};
+    \node[square] (BCDF) [square,draw] [below left of=C];
+    \node[square] (CEH) [right of=C];
+    \node[main] (E) [above right of=CEH] {$E$};
+    \node[square] (GEH) [below right of=E];
+    \node[main] (G) [right of=GEH] {$G$};
+    \node[main] (D) [below left of=BCDF] {$D$};
+    \node[square] (CFH) [below right of=C];
+    \node[main] (H) [right of=CFH] {$H$};
+    \node[square] (FHI) [below left of = H];
+    \node[main] (I) [below right of=FHI] {$I$};
+    \node[main] (F) [below right of=BCDF] {$F$};
+
+    \draw (ABC) -- (A);
+    \draw (ABC) -- (C);
+    \draw (ABC) -- (B);
+    \draw (BCDF) -- (B);
+    \draw (BCDF) -- (C);
+    \draw (BCDF) -- (D);
+    \draw (BCDF) -- (F);
+    \draw (CEH) -- (C);
+    \draw (CEH) -- (E);
+    \draw (CEH) -- (H);
+    \draw (GEH) -- (G);
+    \draw (GEH) -- (E);
+    \draw (GEH) -- (H);
+    \draw (CFH) -- (C);
+    \draw (CFH) -- (F);
+    \draw (CFH) -- (H);
+    \draw (FHI) -- (F);
+    \draw (FHI) -- (H);
+    \draw (FHI) -- (I);
+\end{tikzpicture}
+\end{center}
 \newpage
@@
 
-\newpage
+Using our factor analysis model, we can describe the biochemical pathway as:
+
+\[\delta[\textbf{x}] = \Lambda \textbf{z} + \epsilon\]
+
+where $\delta[\textbf{x}]$ are the concentration perturbations, $\epsilon \sim \mathcal{N} (0, \Psi)$, and the latent factors $z \sim \mathcal{N} (0, I)$.
+
+From the graph structure, we know that:
+
+\[\Lambda = \begin{bmatrix}
+                 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0  \\
+                 \Lambda_{BA} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0  \\
+                 \Lambda_{CA} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0  \\
+                 0 & 0 & \Lambda_{DC} & 0 & 0 & 0 & 0 & 0 & 0  \\
+                 0 & 0 & \Lambda_{EC} & 0 & 0 & 0 & 0 & 0 & 0  \\
+                 0 & \Lambda_{FB} & 0 & \Lambda_{FD} & 0 & 0 & 0 & 0 & 0  \\
+                 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0  \\
+                 0 & 0 & 0 & 0 & \Lambda_{HE} & 0 & \Lambda_{HG} & 0 & 0  \\
+                 0 & 0 & 0 & 0 & 0 & \Lambda_{IF} & 0 & \Lambda_{IH} & 0 \\
+         \end{bmatrix} \text{ and }
+    \textbf{z} = \begin{bmatrix}
+                 z_A  \\
+                 z_B  \\
+                 z_C  \\
+                 z_D  \\
+                 z_E  \\
+                 z_F  \\
+                 z_G  \\
+                 z_H  \\
+                 z_I \\
+         \end{bmatrix}
+\]
+
+Having observations for $\delta[B]$, $\delta[D]$, $\delta[E]$ and $\delta[G]$:
+
+\[
+\begin{bmatrix}
+                 \delta[B]  \\
+                 \delta[D]  \\
+                 \delta[E]  \\
+                 \delta[G]  \\
+         \end{bmatrix} = \begin{bmatrix}
+                 \Lambda_{BA} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0  \\
+                 0 & 0 & \Lambda_{DC} & 0 & 0 & 0 & 0 & 0 & 0  \\
+                 0 & 0 & \Lambda_{EC} & 0 & 0 & 0 & 0 & 0 & 0  \\
+                 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0  \\
+         \end{bmatrix} \begin{bmatrix}
+                 z_A  \\
+                 z_B  \\
+                 z_C  \\
+                 z_D  \\
+                 z_E  \\
+                 z_F  \\
+                 z_G  \\
+                 z_H  \\
+                 z_I \\
+         \end{bmatrix} +  \begin{bmatrix}
+                 \epsilon_B  \\
+                 \epsilon_D  \\
+                 \epsilon_E  \\
+                 \epsilon_G  \\
+         \end{bmatrix}
+\]
+
+We can see that these simplify to the equations:
+
+\begin{gather*}
+    \delta[B] = \Lambda_{BA}z_A + \epsilon_B\\
+    \delta[D] = \Lambda_{DC}z_C  + \epsilon_D\\
+    \delta[E] = \Lambda_{EC}z_C  + \epsilon_E\\
+    \delta[G] = \epsilon_G\\
+\end{gather*}
+
+Thus, we see that the only latent variables present are $z_A$ and $z_C$, so would expect to recover the factors of $A$ and $C$, the two parent nodes of the observations.
 
@@
 
-Plotting the convergence of the variational approximation for different $\sigma$'s:
+Plotting the free energy at each partial expectation step of the variational approximation for different $\sigma$'s:
 
@@
 
-We can see that when $\sigma$ is smaller, we are able to converge at a higher free energy. However, convergence is only reached after more steps, as seen in the plot of $\log(F(t)-F(t-1))$
+We know that our free energy is a upper bounded on the log likelihood:
+
+\[\log P(\mathcal{X} | \theta) \geq \mathcal{F}(q, \theta)\]
+
+In the variational expectation step, $\log P(\mathcal{X} | \theta)$ is fixed and we adjust our approximation $q$ to reach this upper bound.
+We know that $\sigma$ quantifies the noise of $\textbf{x}$, thus a higher $\sigma$ means a wider spread in our distribution $\log P(\mathcal{X} | \theta)$, meaning we are reducing our upper bound for $\mathcal{F}(q, \theta)$.
+As such, we can see in the plot for free energy above that when $\sigma$ is increased, our free energy converges to a lower value, due to being bounded above by a lower log-likelihood.
+Moreover, by reducing the upper bound, we see in the plot of $\log(F(t)-F(t-1))$ that our free energy is able to converge faster.
+Because we have reduced the upper bound by increasing $\sigma$, our free energy can reach this upper bound faster.
 

### 2023-01-02 src/solutions/q4.py

@@ -131,7 +131,7 @@ def b(
             fig.savefig(save_path + f"-latent-factor-{i}-{j}", bbox_inches="tight")
             plt.close()
 
-    fig, ax = plt.subplots(len(ks), 1, figsize=(10, 2 * len(ks)))
+    fig, ax = plt.subplots(len(ks), 1, figsize=(12, 2 * len(ks)))
     plt.subplots_adjust(hspace=1)
     for i, k in enumerate(ks):
         sort_indices = np.argsort(binary_latent_factor_models[i].gaussian_prior.alpha)
@@ -151,6 +151,9 @@ def b(
     fig.savefig(save_path + f"-latent-factors-comparison", bbox_inches="tight")
     plt.close()
 
+    fig = plt.figure()
+    fig.set_figwidth(10)
+    fig.set_figheight(10)
     shades = np.flip(np.linspace(0, 0.7, len(ks)))
     for i, k in enumerate(ks):
         plt.plot(free_energies[i], label=f"{k=}", color=np.ones(3) * shades[i])

### 2023-01-02 submission.tex
add vb response

@@
 \section*{Question 4}
+\subsection*{(a)}
 
@@
 \newpage
+\subsection*{(b)}
+
+Running variational Bayes for different values of $k$, we can visualise the learned features $\mu_k$ and corresponding $\alpha_k^{-1}$:
+
 \begin{figure}[h]
@@
 \end{figure}
-
-\newpage
-\begin{figure}[h]
-\centering
-\includegraphics[scale=0.8]{outputs/q4/b-free-energy}
+As we expect, when running the algorithm for higher $K$ values, many of the features have $\alpha_k \rightarrow \infty$, depicted as $\alpha^{-1}_k$ for visual convenience.
+Moreover, visualising the learned features, we can see the clearest features often have the highest $\alpha^{-1}_k$ while the features deemed irrelevant are often noisy or duplicates.
+
+\newpage
+
+Comparing the free energy plots of models trained on different $K$ values:
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.4]{outputs/q4/b-free-energy}
 \caption{Free Energy for different values of k}
@@
 \end{figure}
+
+We can see that initially, for $k=4$ to $k=8$, increasing $k$ significantly increases the convergence value of the free energy.
+However, beyond $k=8$ there is a no clear trend of $k$ versus the free energy convergence value.
+We can see that this corresponds to the visualisation of $\alpha^{-1}$ where beyond $k=11$, the effective number of features remains more or less the same.
+We know that there are only eight latent features, thus models with $k>8$ should be learning duplicate or irrelevant features.
+As such, we wouldn't expect a model to be able to increase it's free energy significantly when provided with additional degrees of freedom by increasing the value of $k$ beyond eight.
+We see that for models with $k>>8$, there are typically ten or eleven features that might be deemed relevant (depending on how you threshold) and this is likely from slight overfitting, noise in the data, or duplicate features.
+Thus, the relationship between the free energy and the effective number of latent features for each model is as we would expect with ARD.
 
@@
 
-We can use automatic relevance determination (ARD) as a hyperparameter method to select relevant features
-
-Place prior on $\sigma^2$ and optimise with respect to the distributions would cause some to diverge and only relevant latent dimensions will remain.
-This gives us a value for $K$, the number of latent factors that haven't diverged.
+Similar to question 3, we can use automatic relevance determination (ARD) as a hyperparameter method to select relevant features by placing a prior on $\mu_i$.
+With a hyper-M step, certain features will have diverging precision, indicating that they are not relevant to the model output.
+Thus, the number of remaining features will be our selection for $K$.
 
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[scale=0.4]{outputs/q6/all-init-latent-factors}
-\caption{Initial Latent factors learned with EP/Loopy-BP}
-\label{fig:6-init-latent-factors}
-\end{figure}
+Implementing the EP/loopy-BP algorithm, we can compare the learned latent factors with those of the variational mean-field algorithm:
 
@@
 
-\newpage
+We can see that the mean field algorithm seems to learn better latent features.
+In particular there's are fewer duplicates, unlike loopy BP that has a few duplicates of the two by two square in the middle.
+Moreover, the learned features have less noise.
+For example $\mu_0$ for the mean field algorithm looks almost like a binary image.
+We can understand the reason for this by comparing the free energies of the two algorithms:
 
@@
 
+We can observe that the free energy of the mean field algorithm converges while our loopy belief propagation is unable to converge to a free energy.
+Because loopy BP does not have convergence guarantees, this is one of the limitations of this approach.
+
 \newpage
@@
 The Python code for message passing:
-\lstinputlisting[language=Python]{src/models/binary_latent_factor_approximations/message_passing.py}
+\lstinputlisting[language=Python]{src/models/binary_latent_factor_approximations/message_passing_approximation.py}
 

### 2023-01-02 main.py

@@ -96,7 +96,7 @@ if __name__ == "__main__":
         kernel=asdict(init_kernel_parameters),
         log_sigma=jnp.log(1),
     )
-    years_to_predict = 15
+    years_to_predict = 14
     t_new = t[-1] + np.linspace(0, years_to_predict, years_to_predict * 12)
     t_test = np.concatenate((t, t_new))
     q2.f(

### 2023-01-02 src/models/binary_latent_factor_models/binary_latent_factor_model.py

@@ -1,18 +1,14 @@
-from __future__ import annotations
-
 from typing import TYPE_CHECKING, Tuple
 
 import numpy as np
 
 from demo_code.MStep import m_step
+from src.models.binary_latent_factor_approximations.abstract_binary_latent_factor_approximation import (
+    AbstractBinaryLatentFactorApproximation,
+)
 from src.models.binary_latent_factor_models.abstract_binary_latent_factor_model import (
     AbstractBinaryLatentFactorModel,
 )
-
-if TYPE_CHECKING:
-    from src.models.binary_latent_factor_approximations.abstract_binary_latent_factor_approximation import (
-        AbstractBinaryLatentFactorApproximation,
-    )
 
 
 class BinaryLatentFactorModel(AbstractBinaryLatentFactorModel):

### 2023-01-02 src/solutions/q6.py

@@ -2,7 +2,7 @@ import matplotlib.pyplot as plt
 import numpy as np
 
 from src.expectation_maximisation import learn_binary_factors
-from src.models.binary_latent_factor_approximations.message_passing import (
+from src.models.binary_latent_factor_approximations.message_passing_approximation import (
     init_message_passing,
 )
 from src.models.binary_latent_factor_models.boltzmann_machine import (

### 2023-01-02 src/expectation_maximisation.py
add comments

@@ -1,16 +1,15 @@
 from __future__ import annotations
 
-from typing import TYPE_CHECKING, List, Tuple
+from typing import List, Tuple
 
 import numpy as np
 
-if TYPE_CHECKING:
-    from src.models.binary_latent_factor_approximations.abstract_binary_latent_factor_approximation import (
-        AbstractBinaryLatentFactorApproximation,
-    )
-    from src.models.binary_latent_factor_models.binary_latent_factor_model import (
-        AbstractBinaryLatentFactorModel,
-    )
+from src.models.binary_latent_factor_approximations.abstract_binary_latent_factor_approximation import (
+    AbstractBinaryLatentFactorApproximation,
+)
+from src.models.binary_latent_factor_models.binary_latent_factor_model import (
+    AbstractBinaryLatentFactorModel,
+)
 
 
 def is_converge(
@@ -18,6 +17,14 @@ def is_converge(
     current_lambda_matrix: np.ndarray,
     previous_lambda_matrix: np.ndarray,
 ) -> bool:
+    """
+    Check for convergence of free energy and lambda matrix
+
+    :param free_energies: list of free energies
+    :param current_lambda_matrix: current lambda matrix
+    :param previous_lambda_matrix: previous lambda matrix
+    :return: boolean indicating convergence
+    """
     return (abs(free_energies[-1] - free_energies[-2]) == 0) and np.linalg.norm(
         current_lambda_matrix - previous_lambda_matrix
     ) == 0
@@ -33,6 +40,16 @@ def learn_binary_factors(
     AbstractBinaryLatentFactorModel,
     List[float],
 ]:
+    """
+    Expectation maximisation algorithm to learn binary factors.
+
+    :param x: data matrix (number_of_points, number_of_dimensions)
+    :param em_iterations: number of iterations to run EM
+    :param binary_latent_factor_model: a binary_latent_factor_model
+    :param binary_latent_factor_approximation: a binary_latent_factor_approximation
+    :return: a Tuple containing the updated binary_latent_factor_model, updated binary_latent_factor_approximation,
+             and free energies during each step of EM
+    """
     free_energies: List[float] = [
         binary_latent_factor_approximation.compute_free_energy(
             x, binary_latent_factor_model
@@ -42,14 +59,19 @@ def learn_binary_factors(
         previous_lambda_matrix = np.copy(
             binary_latent_factor_approximation.lambda_matrix
         )
+
+        # E step
         binary_latent_factor_approximation.variational_expectation_step(
             x=x,
[…]

### 2023-01-02 src/generate_images.py

@@ -5,11 +5,12 @@ from src.constants import DEFAULT_SEED, M1, M2, M3, M4, M5, M6, M7, M8
 
 def generate_images(n: int = 400, seed: int = DEFAULT_SEED, sigma: float = 0.1):
     """
+    Image generation, adapted from provided demo code
 
     :param n: number of data points
     :param seed: random seed
     :param sigma: Gaussian noise
-    :return:
+    :return: images as a data matrix (number_of_points, number_of_dimensions)
     """
     d = 16  # dimensionality of the data
     np.random.seed(seed)

### 2023-01-02 src/models/bayesian_linear_regression.py

@@ -5,14 +5,24 @@ import numpy as np
 
 @dataclass
 class LinearRegressionParameters:
-    mean: np.ndarray
-    covariance: np.ndarray
+    """
+    Parameters for linear regression
+    """
+
+    mean: np.ndarray  # weight vector (1, number of features)
+    covariance: np.ndarray  # covariance matrix on mean (number of features, number of features)
 
     @property
     def precision(self) -> np.ndarray:
         return np.linalg.inv(self.covariance)
 
     def predict(self, x: np.ndarray) -> np.ndarray:
+        """
+        Linear regression prediction.
+
+        :param x: design matrix (number of features, number of data points)
+        :return: predicted response matrix (1, number of data points)
+        """
         return self.mean.T @ x
 
 

### 2023-01-02 src/models/binary_latent_factor_approximations/mean_field_approximation.py

@@ -11,19 +11,18 @@ from src.models.binary_latent_factor_models.binary_latent_factor_model import (
 
 
 class MeanFieldApproximation(AbstractBinaryLatentFactorApproximation):
-    """
-    lambda_matrix: parameters variational approximation (number_of_points, number_of_latent_variables)
-    """
-
-    _lambda_matrix: np.ndarray
-
-    def __init__(self, lambda_matrix, max_steps, convergence_criterion):
-        self.lambda_matrix = lambda_matrix
+    def __init__(
+        self, lambda_matrix: np.ndarray, max_steps: int, convergence_criterion: float
+    ):
+        self._lambda_matrix = lambda_matrix
         self.max_steps = max_steps
         self.convergence_criterion = convergence_criterion
 
     @property
     def lambda_matrix(self) -> np.ndarray:
+        """
+        lambda_matrix: parameters variational approximation (number_of_points, number_of_latent_variables)
+        """
         return self._lambda_matrix
 
     @lambda_matrix.setter

### 2023-01-02 src/models/binary_latent_factor_approximations/message_passing_approximation.py

@@ -8,7 +8,7 @@ from src.models.binary_latent_factor_approximations.abstract_binary_latent_facto
 from src.models.binary_latent_factor_models.boltzmann_machine import BoltzmannMachine
 
 
-class MessagePassing(AbstractBinaryLatentFactorApproximation):
+class MessagePassingApproximation(AbstractBinaryLatentFactorApproximation):
     """
     eta_matrix:  of parameters eta_matrix[n, i, j]
                 off diagonals corresponds to \tilda{g}_{ij, \neg s_i}(s_j) for data point n
@@ -21,6 +21,10 @@ class MessagePassing(AbstractBinaryLatentFactorApproximation):
 
     @property
     def lambda_matrix(self) -> np.ndarray:
+        """
+        Aggregate messages and compute parameter for Bernoulli distribution
+        :return:
+        """
         lambda_matrix = 1 / (1 + np.exp(-self.xi.sum(axis=1)))
         lambda_matrix[lambda_matrix == 0] = 1e-10
         lambda_matrix[lambda_matrix == 1] = 1 - 1e-10
@@ -52,8 +56,15 @@ class MessagePassing(AbstractBinaryLatentFactorApproximation):
     def variational_expectation_step(
         self, x: np.ndarray, binary_latent_factor_model: BoltzmannMachine
     ) -> List[float]:
+        """
+        Iteratively update singleton and binary factors
+        :param x: data matrix (number_of_points, number_of_dimensions)
+        :param binary_latent_factor_model: a binary_latent_factor_model
+        :return: free energies after each update
+        """
         free_energy = [self.compute_free_energy(x, binary_latent_factor_model)]
         for i in range(self.k):
+            # singleton factor update
             xi_new_ii = self.calculate_singleton_message_update(
                 boltzmann_machine=binary_latent_factor_model,
                 x=x,
@@ -63,6 +74,7 @@ class MessagePassing(AbstractBinaryLatentFactorApproximation):
             free_energy.append(self.compute_free_energy(x, binary_latent_factor_model))
 
             for j in range(i):
+                # binary factor update
                 xi_new_ij = self.calculate_binary_message_update(
                     boltzmann_machine=binary_latent_factor_model,
                     x=x,
@@ -89,6 +101,15 @@ class MessagePassing(AbstractBinaryLatentFactorApproximation):
         i: int,
         j: int,
     ) -> float:
+        """
+        Calculate new parameters for a binary factored message.
+
+        :param x: data matrix (number_of_points, number_of_dimensions)
[…]

### 2023-01-02 src/models/binary_latent_factor_models/abstract_binary_latent_factor_model.py

@@ -15,16 +15,25 @@ class AbstractBinaryLatentFactorModel(ABC):
     @property
     @abstractmethod
     def mu(self) -> np.ndarray:
+        """
+        matrix of means (number_of_dimensions, number_of_latent_variables)
+        """
         pass
 
     @property
     @abstractmethod
     def variance(self) -> float:
+        """
+        gaussian noise parameter
+        """
         pass
 
     @property
     @abstractmethod
     def pi(self) -> np.ndarray:
+        """
+        (1, number_of_latent_variables)
+        """
         pass
 
     @abstractmethod

### 2023-01-02 src/models/binary_latent_factor_models/boltzmann_machine.py

@@ -9,35 +9,47 @@ from src.models.binary_latent_factor_models.binary_latent_factor_model import (
 
 
 class BoltzmannMachine(BinaryLatentFactorModel):
-    """
-    mu: matrix of means (number_of_dimensions, number_of_latent_variables)
-    sigma: gaussian noise parameter
-    pi: vector of priors (1, number_of_latent_variables)
-    """
-
     def __init__(
         self,
         mu: np.ndarray,
         sigma: float,
         pi: np.ndarray,
     ):
+        """
+        Binary latent factor model as a Boltzmann Machine
+
+        :param mu: matrix of means (number_of_dimensions, number_of_latent_variables)
+        :param sigma: gaussian noise parameter
+        :param pi: vector of priors (1, number_of_latent_variables)
+        """
         super().__init__(mu, sigma, pi)
 
     @property
     def w_matrix(self) -> np.ndarray:
-        # (number_of_latent_variables, number_of_latent_variables)
+        """
+        Weight matrix of the Boltzmann machine
+
+        :return: matrix of weights (number_of_latent_variables, number_of_latent_variables)
+        """
         return -self.precision * (self.mu.T @ self.mu)
 
     def w_matrix_index(self, i, j) -> float:
+        """
+        Weight matrix at a specific index
+
+        :param i: row index
+        :param j: column index
+        :return: weight value
+        """
         return -self.precision * (self.mu[:, i] @ self.mu[:, j])
 
     def b(self, x) -> np.ndarray:
         """
+        b term in the Boltzmann machine for all data points
 
         :param x: design matrix (number_of_points, number_of_dimensions)
-        :return:
+        :return: matrix of shape (number_of_points, number_of_latent_variables)
         """
-        # (number_of_points, number_of_latent_variables)
         return -(
             self.precision * x @ self.mu
             + self.log_pi_ratio
@@ -45,7 +57,13 @@ class BoltzmannMachine(BinaryLatentFactorModel):
         )
 
     def b_index(self, x, node_index) -> float:
-        # (number_of_points, 1)
+        """
+        b term for a specific node in the Boltzmann machine for all data points
+
+        :param x: design matrix (number_of_points, number_of_dimensions)
+        :param node_index: node index
+        :return: vector of shape (number_of_points, 1)
+        """
         return -(
             self.precision * x @ self.mu[:, node_index]
[…]

### 2023-01-02 src/models/binary_latent_factor_models/variational_bayes.py

@@ -11,26 +11,56 @@ from src.models.binary_latent_factor_models.binary_latent_factor_model import (
 
 class GaussianPrior:
     def __init__(self, a, b, d, k):
+        """
+        Gaussian prior on mu matrix
+
+        :param a: alpha parameter of Gamma Prior
+        :param b: beta parameter of Gamma Prior
+        :param d: number of dimensions
+        :param k: number of latent factors
+        """
         self.a = a
         self.b = b
         self.mu = np.zeros((d, k))
-        self.alpha = np.ones((k,))  # np.random.gamma(a, b, size=(k,))
+        self.alpha = np.ones((k,))
         self.w_covariance = np.zeros((k, k))
 
-    def mu_k(self, k):  # (number_of_dimensions,  1)
+    def mu_k(self, k):
+        """
+        Column vector of mu matrix, the latent feature vector
+
+        :param k: latent factor index
+        :return: column vector (number_of_dimensions,  1)
+        """
         return self.mu[:, k : k + 1]
 
-    def w_d(self, d):  # (1,  number_of_latent_variables)
+    def w_d(self, d):
+        """
+        Row vector of mu matrix, the weight vector for a particular dimension (pixel) of the data
+
+        :param d: data dimension index
+        :return: row vector (1,  number_of_latent_variables)
+        """
         return self.mu[d : d + 1, :]
 
     @property
     def a_matrix(self) -> np.ndarray:
-        #  precision matrix for w_d
+        """
+        Precision matrix for a weight vector w_d
+        :return: matrix of shape (number_of_latent_variables, number_of_latent_variables)
+        """
         return np.diag(self.alpha)
 
 
 class VariationalBayesBinaryLatentFactorModel(AbstractBinaryLatentFactorModel):
     def __init__(self, mu: GaussianPrior, variance: float, pi: np.ndarray):
+        """
+        Variational Bayes implementation with prior on mu
+
+        :param mu: Gaussian prior on latent features
+        :param variance: Gaussian noise parameter
+        :param pi: vector of priors (1, number_of_latent_variables)
+        """
         self.gaussian_prior = mu
         self._variance = variance
         self._pi = pi
@@ -51,8 +81,7 @@ class VariationalBayesBinaryLatentFactorModel(AbstractBinaryLatentFactorModel):
         self,
         binary_latent_factor_approximation: AbstractBinaryLatentFactorApproximation,
     ):
[…]

### 2023-01-02 src/solutions/q4.py

@@ -18,24 +18,6 @@ from src.models.binary_latent_factor_models.variational_bayes import (
 )
 
 
-def offset_image(coord, path, ax):
-    img = plt.imread(path)
-    im = OffsetImage(img, zoom=0.72)
-    im.image.axes = ax
-
-    ab = AnnotationBbox(
-        im,
-        (coord, 0),
-        xybox=(0.0, -19.0),
-        frameon=False,
-        xycoords="data",
-        boxcoords="offset points",
-        pad=0,
-    )
-
-    ax.add_artist(ab)
-
-
 def _run_automatic_relevance_determination(
     x: np.ndarray,
     a_parameter: int,
@@ -45,6 +27,18 @@ def _run_automatic_relevance_determination(
     e_maximum_steps: int,
     e_convergence_criterion: float,
 ) -> Tuple[VariationalBayesBinaryLatentFactorModel, List[float]]:
+    """
+    Run automatic relevance determination with variational Bayes.
+
+    :param x: data matrix (number_of_points, number_of_dimensions)
+    :param a_parameter: alpha parameter for gamma prior
+    :param b_parameter: beta parameter for gamma prior
+    :param k: number of latent variables
+    :param em_iterations: number of iterations to run EM
+    :param e_maximum_steps: maximum number of iterations of partial expectation steps
+    :param e_convergence_criterion: minimum required change in free energy for each partial expectation step
+    :return: a Tuple containing the optimised VB model and a list of free energies during each EM step
+    """
     n = x.shape[0]
     mean_field_approximation = init_mean_field_approximation(
         k, n, max_steps=e_maximum_steps, convergence_criterion=e_convergence_criterion
@@ -65,13 +59,37 @@ def _run_automatic_relevance_determination(
             pi=pi,
         )
     )
-    (_, binary_latent_factor_model, free_energy,) = learn_binary_factors(
+    _, binary_latent_factor_model, free_energy = learn_binary_factors(
         x=x,
         em_iterations=em_iterations,
         binary_latent_factor_model=binary_latent_factor_model,
         binary_latent_factor_approximation=mean_field_approximation,
     )
     return binary_latent_factor_model, free_energy
+
+
+def _offset_image(coord: int, path: str, ax: plt.axis):
+    """
+    Add image to matplotlib axis.
+
[…]

### 2023-01-02 src/solutions/q6.py

@@ -14,6 +14,8 @@ def run(x: np.ndarray, k: int, em_iterations: int, save_path: str) -> None:
     n = x.shape[0]
     message_passing = init_message_passing(k, n)
     boltzmann_machine = init_boltzmann_machine(x, message_passing)
+
+    # pre-training features plot
     fig, ax = plt.subplots(1, k, figsize=(k * 2, 2))
     for i in range(k):
         ax[i].imshow(boltzmann_machine.mu[:, i].reshape(4, 4))
@@ -22,12 +24,16 @@ def run(x: np.ndarray, k: int, em_iterations: int, save_path: str) -> None:
     plt.tight_layout()
     plt.savefig(save_path + "-init-latent-factors", bbox_inches="tight")
     plt.close()
+
+    # EM
     message_passing, boltzmann_machine, free_energy = learn_binary_factors(
         x=x,
         em_iterations=em_iterations,
         binary_latent_factor_model=boltzmann_machine,
         binary_latent_factor_approximation=message_passing,
     )
+
+    # post training features plot
     fig, ax = plt.subplots(1, k, figsize=(k * 2, 2))
     for i in range(k):
         ax[i].imshow(boltzmann_machine.mu[:, i].reshape(4, 4))
@@ -37,6 +43,7 @@ def run(x: np.ndarray, k: int, em_iterations: int, save_path: str) -> None:
     plt.savefig(save_path + "-latent-factors", bbox_inches="tight")
     plt.close()
 
+    # free energy plot
     plt.title("Free Energy (Loopy BP)")
     plt.xlabel("t (EM steps)")
     plt.ylabel("Free Energy")

### 2023-01-02 src/automatic_relevance_determination.py
split out ard code

@@ -1,0 +1,65 @@
+from typing import List, Tuple
+
+import numpy as np
+
+from src.expectation_maximisation import learn_binary_factors
+from src.models.binary_latent_factor_approximations.mean_field_approximation import (
+    init_mean_field_approximation,
+)
+from src.models.binary_latent_factor_models.binary_latent_factor_model import (
+    BinaryLatentFactorModel,
+)
+from src.models.binary_latent_factor_models.variational_bayes import (
+    GaussianPrior,
+    VariationalBayesBinaryLatentFactorModel,
+)
+
+
+def run_automatic_relevance_determination(
+    x: np.ndarray,
+    a_parameter: int,
+    b_parameter: int,
+    k: int,
+    em_iterations: int,
+    e_maximum_steps: int,
+    e_convergence_criterion: float,
+) -> Tuple[VariationalBayesBinaryLatentFactorModel, List[float]]:
+    """
+    Run automatic relevance determination with variational Bayes.
+
+    :param x: data matrix (number_of_points, number_of_dimensions)
+    :param a_parameter: alpha parameter for gamma prior
+    :param b_parameter: beta parameter for gamma prior
+    :param k: number of latent variables
+    :param em_iterations: number of iterations to run EM
+    :param e_maximum_steps: maximum number of iterations of partial expectation steps
+    :param e_convergence_criterion: minimum required change in free energy for each partial expectation step
+    :return: a Tuple containing the optimised VB model and a list of free energies during each EM step
+    """
+    n = x.shape[0]
+    mean_field_approximation = init_mean_field_approximation(
+        k, n, max_steps=e_maximum_steps, convergence_criterion=e_convergence_criterion
+    )
+    (_, sigma, pi,) = BinaryLatentFactorModel.calculate_maximisation_parameters(
+        x, mean_field_approximation
+    )
+    mu = GaussianPrior(
+        a=a_parameter,
+        b=b_parameter,
+        k=k,
+        d=x.shape[1],
+    )
+    binary_latent_factor_model: VariationalBayesBinaryLatentFactorModel = (
+        VariationalBayesBinaryLatentFactorModel(
+            mu=mu,
+            variance=sigma**2,
+            pi=pi,
+        )
+    )
+    _, binary_latent_factor_model, free_energy = learn_binary_factors(
+        x=x,
+        em_iterations=em_iterations,
+        binary_latent_factor_model=binary_latent_factor_model,
[…]

### 2023-01-02 src/solutions/q4.py

@@ -1,71 +1,11 @@
 import os
-from typing import List, Tuple
+from typing import List
 
 import matplotlib.pyplot as plt
 import numpy as np
 from matplotlib.offsetbox import AnnotationBbox, OffsetImage
 
-from src.expectation_maximisation import learn_binary_factors
-from src.models.binary_latent_factor_approximations.mean_field_approximation import (
-    init_mean_field_approximation,
-)
-from src.models.binary_latent_factor_models.binary_latent_factor_model import (
-    BinaryLatentFactorModel,
-)
-from src.models.binary_latent_factor_models.variational_bayes import (
-    GaussianPrior,
-    VariationalBayesBinaryLatentFactorModel,
-)
-
-
-def _run_automatic_relevance_determination(
-    x: np.ndarray,
-    a_parameter: int,
-    b_parameter: int,
-    k: int,
-    em_iterations: int,
-    e_maximum_steps: int,
-    e_convergence_criterion: float,
-) -> Tuple[VariationalBayesBinaryLatentFactorModel, List[float]]:
-    """
-    Run automatic relevance determination with variational Bayes.
-
-    :param x: data matrix (number_of_points, number_of_dimensions)
-    :param a_parameter: alpha parameter for gamma prior
-    :param b_parameter: beta parameter for gamma prior
-    :param k: number of latent variables
-    :param em_iterations: number of iterations to run EM
-    :param e_maximum_steps: maximum number of iterations of partial expectation steps
-    :param e_convergence_criterion: minimum required change in free energy for each partial expectation step
-    :return: a Tuple containing the optimised VB model and a list of free energies during each EM step
-    """
-    n = x.shape[0]
-    mean_field_approximation = init_mean_field_approximation(
-        k, n, max_steps=e_maximum_steps, convergence_criterion=e_convergence_criterion
-    )
-    (_, sigma, pi,) = BinaryLatentFactorModel.calculate_maximisation_parameters(
-        x, mean_field_approximation
-    )
-    mu = GaussianPrior(
-        a=a_parameter,
-        b=b_parameter,
-        k=k,
-        d=x.shape[1],
-    )
-    binary_latent_factor_model: VariationalBayesBinaryLatentFactorModel = (
-        VariationalBayesBinaryLatentFactorModel(
-            mu=mu,
-            variance=sigma**2,
-            pi=pi,
-        )
-    )
-    _, binary_latent_factor_model, free_energy = learn_binary_factors(
-        x=x,
-        em_iterations=em_iterations,
[…]

### 2023-01-03 submission.tex
add comments
split out ard code
edit report and rename variables for readibility

@@
 \maketitle
-\section*{Question 1}
-
-\subsection*{(a)} The directed acyclic graph:
+\section*{Question 1: A biochemical pathway}
+
+\subsection*{(a)} We are given that:
+
+\begin{itemize}
+    \item The concentrations of both species B and C depend on that of A.
+    \item Molecule C seems to redirect the process that produces enzyme D to produce enzyme E instead.
+    \item D catalyses the production of F from B, while E catalyses the production of H from G.
+    \item F and H then combine to form the end product I.
+\end{itemize}
+
+The corresponding directed acyclic graph:
 
@@
 The resulting junction tree:
-
-%\begin{center}
-%\begin{tikzpicture}[node distance={25mm}, main/.style = {draw, circle}, square/.style={regular polygon,regular polygon sides=4}]
-%    \node[main] (ABC) {$ABC$};
-%    \node[main] (BCDF) [below of=ABC] {$BCDF$};
-%    \node[main] (CFH) [below of=BCDF] {$CFH$};
-%    \node[main] (FHI) [below left of=CFH] {$FHI$};
-%    \node[main] (CEH) [below right of=CFH] {$CEH$};
-%    \node[main] (EGH) [below of=CEH] {$EGH$};
-%    \draw (ABC) -- (BCDF);
-%    \draw (BCDF) -- (CFH);
-%    \draw (CFH) -- (FHI);
-%    \draw (CFH)  -- (CEH);
-%    \draw (EGH)  -- (CEH);
-%\end{tikzpicture}
-%\end{center}
 
@@
 The set $\{A, D, E, F, H\}$ is a non-unique smallest set of molecules such that if the concentrations of the species within the set are known, the concentrations of the others $\{B, C, G, I\}$ would all be independent (conditioned on the measured ones).
+\newpage
 
@@
 
-where $\delta[\textbf{x}]$ are the concentration perturbations, $\epsilon \sim \mathcal{N} (0, \Psi)$, and the latent factors $z \sim \mathcal{N} (0, I)$.
+where $\delta[\textbf{x}]$ are the concentration perturbations, $\epsilon \sim \mathcal{N} (0, \Psi)$ are the reaction specific Gaussian noise variables, and $z \sim \mathcal{N} (0, I)$ are the latent factor, the parent concentration perturbations.
 
@@
 
-We can see that these simplify to the equations:
-
-\begin{gather*}
-    \delta[B] = \Lambda_{BA}z_A + \epsilon_B\\
-    \delta[D] = \Lambda_{DC}z_C  + \epsilon_D\\
-    \delta[E] = \Lambda_{EC}z_C  + \epsilon_E\\
-    \delta[G] = \epsilon_G\\
-\end{gather*}
-
-Thus, we see that the only latent variables present are $z_A$ and $z_C$, so would expect to recover the factors of $A$ and $C$, the two parent nodes of the observations.
+Knowing \textbf{z} are the parent concentration perturbations, we can see that these simplify to a factor analysis set up:
+
+\[
+    \begin{bmatrix}
+                 \delta[B]  \\
+                 \delta[D]  \\
+                 \delta[E]  \\
+         \end{bmatrix} = \begin{bmatrix}
+                 \Lambda_{BA} & 0   \\
+                 0 & \Lambda_{DC} \\
+                 0 & \Lambda_{EC} \\
+         \end{bmatrix} \begin{bmatrix}
+                 \delta[A]  \\
+                 \delta[C]  \\
+         \end{bmatrix} +  \begin{bmatrix}
+                 \epsilon_B  \\
+                 \epsilon_D  \\
+                 \epsilon_E  \\
+         \end{bmatrix}
+\]
+
+Thus, we would expect to recover the factors of $A$ and $C$, $\delta[A]$ and $\delta[C]$, the two parent nodes of the observations.
+\newpage
 
@@
 
-\newpage
-\section*{Question 2}
+From factor analysis, we would recover $\delta[A]$ and $\delta[C]$.
+In theory, if these concentration perturbations had unknown parent factors, we could use the results to recover the concentration perturbations of those parent features.
+However, in our DAG this is not hte case, so we would not be able to recover any other species in the cascade because  they are all downstream from the known concentration perturbations and those recovered from factor analysis.
+
+\newpage
+\section*{Question 2: Bayesian linear and Gaussian process regression}
 
@@
 Our distribution for $\textbf{w}$:
-\[P(\textbf{w}) = \mathcal{N} \left(
+\[P(\textbf{w}) = \mathcal{N}(\mu_{\textbf{w}}, \Sigma_{\textbf{w}}) = \mathcal{N} \left(
 \begin{bmatrix}
@@
 \right)
-  = \mathcal{N}(\mu_{\textbf{w}}, \Sigma_{\textbf{w}})
-\]
-
-Moreover, for our data $\mathcal{D} = \{\textbf{X}, \textbf{Y}\}$:
-\[P(\mathcal{D} | \textbf{w}) = \mathcal{N} \left( \textbf{Y} - \textbf{w}^T \textbf{X}, \sigma^2 \textbf{I}
+
+\]
+
+For our data $\mathcal{D} = \{\textbf{X}, \textbf{y}\}$:
+\[P(\mathcal{D} | \textbf{w}) = \prod_{t=1}^{T}\mathcal{N} \left( y_t - \textbf{w}^T \textbf{x}_t, \sigma^2
 \right)
@@
 
-where $\textbf{X} =  \begin{bmatrix}
-                 t_1 & t_2 \cdots t_N \\
-                 1 & 1 \cdots 1
-         \end{bmatrix} \in  \mathbb{R}^{2 \times N}$ and $\textbf{Y} \in \mathbb{R}^{1 \times N}$.
+where $\textbf{x}_t =  \begin{bmatrix}
+                 t_1 \\
+                 1
+         \end{bmatrix} \in  \mathbb{R}^{2 \times 1}$ and $y_t \in \mathbb{R}^{1 \times 1}$ our response $f_{obs}(t_1)$.
 
@@
 \[P(\textbf{w} | \mathcal{D}) \propto
-  \exp \left(\frac{-1}{2 \sigma^2} \left( \textbf{Y} - \textbf{w}^T \textbf{X}\right) \left( \textbf{Y} - \textbf{w}^T \textbf{X}\right)^T \right)
+  \exp \left(\frac{-1}{2 \sigma^2} \left( \textbf{y} - \textbf{w}^T \textbf{X}\right) \left( \textbf{y} - \textbf{w}^T \textbf{X}\right)^T \right)
 \exp \left(\frac{-1}{2} \left( \textbf{w} -\mu_{\textbf{w}}\right)^T \Sigma_{\textbf{w}}^{-1} \left( \textbf{w} -\mu_{\textbf{w}}\right)\right)
@@
 
-expanding:
+where $\textbf{X} =  \begin{bmatrix}
+                 t_1 & t_2 \cdots t_T \\
+                 1 & 1 \cdots 1
+         \end{bmatrix} \in  \mathbb{R}^{2 \times T}$ and $\textbf{y} \in \mathbb{R}^{1 \times T}$.
+Expanding:
+
 \[\log P(\textbf{w} | \mathcal{D}) \propto
-  \frac{-1}{2} \left( \frac{\textbf{Y} \textbf{Y}^T}{\sigma^2} - 2\textbf{w}^T \frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \textbf{w}^T \frac{\textbf{X} \textbf{X}^T}{\sigma^2} \textbf{w} +  \textbf{w}^T \Sigma_{\textbf{w}}^{-1}  \textbf{w} - 2\textbf{w}^T \Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} + \mu_{\textbf{w}}^T\Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} \right)
+  \frac{-1}{2} \left( \frac{\textbf{y} \textbf{y}^T}{\sigma^2} - 2\textbf{w}^T \frac{\textbf{X}\textbf{y}^T}{\sigma^2} + \textbf{w}^T \frac{\textbf{X} \textbf{X}^T}{\sigma^2} \textbf{w} +  \textbf{w}^T \Sigma_{\textbf{w}}^{-1}  \textbf{w} - 2\textbf{w}^T \Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} + \mu_{\textbf{w}}^T\Sigma_{\textbf{w}}^{-1}\mu_{\textbf{w}} \right)
 \]
@@
 \[\log P(\textbf{w} | \mathcal{D}) \propto
-  \frac{-1}{2} \left( \textbf{w}^T \left(\frac{\textbf{X} \textbf{X}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \right)  \textbf{w} - 2\textbf{w}^T \left(\frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right)  \right)
-\]
-
-Knowing that the posterior $P(\textbf{w} | \mathcal{D})$ will be Gaussian with mean $\bar{\mu}_w$ and covariance $\bar{\Sigma}_w$, we can see that expanding the exponent component would have the form:
+  \frac{-1}{2} \left( \textbf{w}^T \left(\frac{\textbf{X} \textbf{X}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \right)  \textbf{w} - 2\textbf{w}^T \left(\frac{\textbf{X}\textbf{y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right)  \right)
+\]
+
+Knowing that the posterior $P(\textbf{w} | \mathcal{D})$ will be Gaussian with mean $\bar{\mu}_w$ and covariance $\bar{\Sigma}_w$, we can see that expanding the exponent component of a Gaussian would have the form:
+
 \[\left( \textbf{w} - \bar{\mu}_w \right)^T \bar{\Sigma}_w^{-1} \left( \textbf{w} - \bar{\mu}_w \right) = \textbf{w}^T \bar{\Sigma}_w^{-1} \textbf{w} -2 \textbf{w}^T \bar{\Sigma}_w^{-1} \bar{\mu}_w + \bar{\mu}_w^T \bar{\Sigma}_w^{-1} \bar{\mu}_w\]
@@
 and the posterior mean:
-\[\bar{\mu}_w = \bar{\Sigma}_w \left(\frac{\textbf{X}\textbf{Y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right) \]
+\[\bar{\mu}_w = \bar{\Sigma}_w \left(\frac{\textbf{X}\textbf{y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right) \]
 
@@
 
-Plotting the residuals:
+Plotting the residuals $g_{obs}(t) = f_{obs}(t) - (a_{MAP}t + b_{MAP})$:
 
@@
 The density estimation shows that a mean of zero is a reasonable prior belief however the data does not seem to exhibit unit variance.
-Also we know it's not iid because timeseries.
+Also, observing $g_{obs}(t)$ we see that there is periodic structure in the timeseries meaning that the data is not independently and identically distributed as $e(t) \sim \mathcal{N}(0, 1)$ would suggest.
+Otherwise, we would expect $g(t)$ to look like random noise.
 
@@
 
-We can make qualitative observations this kernel by visualising the covariance (gram) matrix:
+We can make qualitative observations of this kernel by visualising the covariance (gram) matrix:
 
@@
 We can observe a striped pattern which indicate higher covariance at regular intervals.
-This can be attributed to the sinusoidal term in the kernel and encourages sinusoidal functions.
+This can be attributed to the sinusoidal term in the kernel and encourages periodic structure in functions.
 Additionally, we can see that covariance values also decay as they are further away from the diagonal.
 This can be attributed to the exponential term in the kernel, encouraging points closer in time to be more correlated and vice versa.
-From our $CO_2$ data, we would want a class of functions which exhibit both of these behaviours as the data looks sinusoidal (seasonal with respect to each year) and correlations locally.
+From visualising $g(t)$, we would want to model this with a class of functions which exhibit both of these behaviours as $g(t)$ looks periodic (seasonal with respect to each year) and locally correlated (smooth).
 
@@
 \begin{itemize}
-\item [$\theta$:] As $\theta$ increases, we see more pronounced periodic behavior in the sample function.
+\item [$\theta$:] As $\theta$ increases, we see more pronounced periodic behavior in the sample function by increasing its amplitude.
     The covariance matrix shows how increasing $\theta$ visually reveals the striped periodic component.
-        This is expected because it is the parameter that adjusts the weight of the periodic component.
-\item [$\sigma$:] As $\sigma$ increases, we see reduced periodic behaviour in the sample function.
-    The covariance matrix shows how increasing $\sigma$ will increase covariance values in the off-diagonals.
-    This is expected because it adjusts the lengthscale of the periodic portion of the kernel, which ends up dominating the function.
+        This is expected because it is the parameter that adjusts the amplitude of the periodic component.
+\item [$\sigma$:] As $\sigma$ increases, we see smoother periodic behaviour in the sample function.
+    The covariance matrix shows how increasing $\sigma$ will increase covariance values around each periodic stripe of the covariance matrix.
+    This is expected because it adjusts the lengthscale of the periodic portion of the kernel.
 \item [$\phi$:] As $\phi$ increases, we see the ratio of the amplitude of the periodicity component of the sample function reduces compared to the baseline.
-    The covariance matrix shows how increasing $\phi$ will start to increase the non-periodic component.
-    This is expected because it adjusts the weight of the non-periodic portion of the kernel, thus the periodic component remains the same (i.e.same amplitude) but the large baseline shifts from increasing $\phi$ ends up dominating the function visually.
+    The covariance matrix shows how increasing $\phi$ will start to increase the non-periodic component masking the periodic stripes.
+    This is expected because it adjusts the weight of the squared exponential portion of the kernel, thus the periodic component remains the same (i.e.same amplitude) but the large baseline shifts from increasing $\phi$ ends up dominating the function visually.
 \item [$\eta$:] As $\eta$ increases we see smoother sample functions.
 This is expected because the $\eta$ increases the lengthscale of the non-periodic component, allowing for smoother functions.
-This causes the off-diagonals of the gram matrix to increase, however the periodic component is still maintained because $\eta$ doesn't affect the relative weight of the two components.
+This causes the off-diagonals of the gram matrix to increase, however the periodic component is still maintained because $\eta$ doesn't affect the relative weighting between the two components.
 \item [$\tau$:] As $\tau$ increases, the period of the periodic function increases.
-We can see this reflected in the stripes in the gram matrix getting further apart.
-This makes sense because we are adjusting the period in the sinusoid function of the periodic term with $\tau$.
+We can see this reflected in the stripes in the gram matrix getting further apart and the period of the samples.
+This makes sense because we are adjusting the period parameter in the sinusoid function of the periodic term with $\tau$.
 \item [$\zeta$:] As $\zeta$ increases, the function becomes less smooth.
-    This is because the $\zeta$ parameter adjusts the weight of the $\delta_{s=t}$ parameter.
+    This is because the $\zeta$ parameter adjusts the weight of the $\delta_{s=t}$ kernel.
     This places stronger emphasis on the independence of each timestep, which can be seen with the reduction of relative magnitude of off-diagonals in the gram matrix.
-    However, this is simply masking the periodic and squared-exponential terms as we can see with the increased magnitude of the functions as $\zeta$ increases.
+    This masks the periodic and squared-exponential components as we can see with the increased magnitude of the functions as $\zeta$ increases.
 \end{itemize}
-\newpage
 \subsection*{(e)}
@@
 For example, it is a reasonable assumption that the $CO_2$ concentration levels have a strong yearly seasonality behaviour due to the cyclic changes in temperature, humidity, etc.
-Thus we can choose $\tau=1$ to ensure functions with a period of one year to reflect this knowledge.
+This behaviour cannot be modelled by our Bayesian linear regression and is reflected in our residuals $g(t)$.
+Thus we can choose $\tau=1$ to ensure functions with a period of one year to reflect this prior knowledge.
 It can be difficult to quantitatively choose values for the other parameters as they can relate to the uncertainty exhibited in the data (i.e.the smoothness of the function).
@@
 
-\[\log P(\textbf{Y} | \textbf{X}) = -\frac{1}{2} \textbf{Y}^T (\textbf{K} + \sigma^2\textbf{I})^{-1} \textbf{Y} - \frac{1}{2} \log |\textbf{K}+ \sigma^2\textbf{I}| - \frac{n}{2}\log(2\pi)\]
+\[\log P(\textbf{y} | \textbf{X}) = -\frac{1}{2} \textbf{y}^T (\textbf{K} + \sigma^2\textbf{I})^{-1} \textbf{y} - \frac{1}{2} \log |\textbf{K}+ \sigma^2\textbf{I}| - \frac{n}{2}\log(2\pi)\]
 
 the log-likelihood  of the posterior distribution with respect to the given data where $\textbf{K}$ is the gram matrix for the kernel (equation 2.30 from http://gaussianprocess.org/gpml/chapters/RW2.pdf).
-We can define a loss function as the negative log-likelihood and employ gradient-based algorithms to find optimal parameters.
+We define a loss function as the negative log-likelihood and employ gradient-based algorithms to find optimal values for these parameters.
 
@@
 
-We can analyse some of the changes in these parameters after training to gain some insights.
-We can see that $\tau$ remains the same as we would expect given the yearly seasonality we have prior knowledge of.
+We can analyse some of the changes in these parameters from optimisation, to gain some insights.
+We can see that $\tau$ remains the same as we would expect, given the yearly seasonality we knew apriori.
 On the other hand, the value for $\zeta$ is significantly reduced signifying that $\delta_{s=t}$ is not a very good kernel for representing the data as datapoints at different timesteps do exhibit correlations.
@@
 \subsection*{(f)}
-Extrapolating the $CO_2$ concentration levels:
+Extrapolating the $CO_2$ concentration levels using both the untrained and trained hyperparameters:
 
@@
 We can see that the extrapolation shows a continued increase in $CO_2$ in the future.
-This follows our expectations given that the levels has been steadily increasing in the past.
-Moreover, the concentration continues to exhibit yearly seasonality (for the trained extrapolation) as we would expect.
-We can see that the conclusions can be quite sensitive to kernel hyperparameters when comparing the values from before and after training.
-Prior to training, the extrapolated prediction is not representative of the given data, with pretty much no seasonal behaviour and very large uncertainty.
+This follows our expectations given that the levels has been steadily increasing in the past and our Bayesian linear regression model assumption.
+Moreover, the concentration continues to exhibit yearly seasonality (for the trained extrapolation) as we would expect through the choice of our kernel for modelling the residuals $g(t)$.
+We can see that the conclusions can be quite sensitive to kernel hyperparameters when comparing the extrapolations from before and after training.
+Prior to training, the extrapolated prediction is not representative of the given data with very large uncertainty.
 After training, we can see that the prediction is much more reasonable, and qualitatively the uncertainty bounds seem to exhibit the historical variability in the data.
@@
 
-This procedure is not fully Bayesian because despite using a posterior estimate of our linear regression terms, we only use a point estimate when making prediction.
+This procedure is not fully Bayesian because despite using a posterior estimate of our linear regression terms, we only use a MAP point estimate when making prediction.
 For a fully Bayesian approach, we should also incorporate the uncertainty of the linear regression parameters into our extrapolation/uncertainty bounds.
@@
 \newpage
-\section*{Question 3}
+\section*{Question 3: Mean-field learning}
 
@@
 
-The free energy is can be calculated as:
+The free energy is can be expressed as:
 \[\mathcal{F}(q, \theta) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{q(\textbf{s})} + H[Q(\textbf{s})]\]
@@
 
-Moreover, our mean field approximation:
-\[q(\textbf{s}) = \prod_{i=1}^{K} q_{i} (s_i) \]
-
-where $q_{i} (s_i) = \lambda_i^{s_i} (1-\lambda_i)^{(1-s_i)}$.
-
-To compute the first term:
-
-\[ P(\textbf{x} | \textbf{s}, \theta)  = \mathcal{N}\left( \sum_{i=1}^{K} s_i \mu_i, \sigma^2 \textbf{I} \right)}\]
-
-substituting the appropriate terms:
-
-\[ P(\textbf{x} | \textbf{s}, \theta)  = 2\pi^{-\frac{d}{2}} |\sigma^2 \textbf{I}|^{-\frac{1}{2}} \exp \left( -\frac{1}{2} \left(\textbf{x} - \sum_{i=1}^{K} s_i \mu_i\right)^T \frac{1}{\sigma^2} \textbf{I}  \left(\textbf{x} - \sum_{i=1}^{K} s_i \mu_i\right) \right) \]
+Moreover, our mean field approximation is:
+\[q(\textbf{s}) = \prod_{k=1}^{K} q_{k} (s_k) \]
+
+where $q_{k} (s_k) = (\lambda_k)^{s_k} (1-\lambda_k)^{(1-s_k)}$.
+
+Given that:
+
+\[ P(\textbf{x} | \textbf{s}, \theta)  = \mathcal{N}\left( \sum_{k=1}^{K} s_k \mu_k, \sigma^2 \textbf{I} \right)}\]
+
+we can substitute the appropriate terms:
+
+\[ P(\textbf{x} | \textbf{s}, \theta)  = 2\pi^{-\frac{d}{2}} |\sigma^2 \textbf{I}|^{-\frac{1}{2}} \exp \left( -\frac{1}{2} \left(\textbf{x} - \sum_{k=1}^{K} s_k \mu_k\right)^T \frac{1}{\sigma^2} \textbf{I}  \left(\textbf{x} - \sum_{k=1}^{K} s_k \mu_k\right) \right) \]
 
@@
 
-\[ \log P(\textbf{x} | \textbf{s}, \theta)  = -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} s_i \mu_i   + \sum_{i=1}^{K} \sum_{i=1}^{K} s_i s_j \mu_i^T\mu_j \right) \]
+\[ \log P(\textbf{x} | \textbf{s}, \theta)  = -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} s_k \mu_k   + \sum_{k=1}^{K} \sum_{k=1}^{K} s_k s_{k'} \mu_k^T\mu_{k'} \right) \]
 
@@
 \left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} \\
-= -\frac{d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} \left\langle s_i \right\rangle_{q_{i} (s_i)} \mu_i   + \sum_{i=1}^{K} \sum_{j=1}^{K} \left\langle s_i s_j \right\rangle_{q_{i} (s_i) q_{j} (s_j)} \mu_i^T\mu_j \right)\]
+= -\frac{d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} \left\langle s_k \right\rangle_{q_{k} (s_k)} \mu_k   + \sum_{k=1}^{K} \sum_{k'=1}^{K} \left\langle s_k s_{k'} \right\rangle_{q_{k} (s_k) q_{k\} (s_{k'})} \mu_k^T\mu_{k'} \right)\]
 
@@
 \left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} \\
-= -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right)\]
-
-where $\left\langle s_i s_i \right\rangle_{q_{i} (s_i)} = \left\langle s_i \right\rangle_{q_{i} (s_i)}$ because $s_i \in \{0, 1\}$.
-
-To compute the second term:
-\[ P(\textbf{s} | \theta) = \prod_{i=1}^{K}\pi_i^{s_i} (1-\pi_i)^{(1-s_i)}\]
+= -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K}  \lambda_k  \mu_k   + \sum_{k=1}^{K} \sum_{k'=1, k'\neq k}^{K}  \lambda_k \lambda_{k'} \mu_k^T\mu_{k'} + \sum_{k=1}^{K}  \lambda_k \mu_k^T\mu_k \right)\]
+
+where $\left\langle s_k s_k \right\rangle_{q_{k} (s_k)} = \left\langle s_k \right\rangle_{q_{k} (s_k)}$ because $s_k \in \{0, 1\}$.
+
+Given that:
+
+\[ P(\textbf{s} | \theta) = \prod_{k=1}^{K}\pi_k^{s_k} (1-\pi_k)^{(1-s_k)}\]
 
@@
 
-\[ \log P(\textbf{s} | \theta) = \sum_{i=1}^{K}s_i \log \pi_i + (1-s_i) \log (1-\pi_i)}\]
+\[ \log P(\textbf{s} | \theta) = \sum_{k=1}^{K}s_k \log \pi_k + (1-s_k) \log (1-\pi_k)}\]
 
@@
 
-\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \left\langle s_i \right\rangle_{q_{i} (s_i)} \log \pi_i + (1-\left\langle s_i\right\rangle_{q_{i} (s_i)}) \log (1-\pi_i)}\]
-
-Evaluating the expectations:
-
-\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \lambda_i \log \pi_i + (1-\lambda_i) \log (1-\pi_i)}\]
+\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})}= \sum_{k=1}^{K} \left\langle s_k \right\rangle_{q_{k} (s_k)} \log \pi_k + (1-\left\langle s_k\right\rangle_{q_{k} (s_k)}) \log (1-\pi_k)}\]
+
+Evaluating the expectations with respect to $q(\textbf{s})$:
+
+\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})}= \sum_{k=1}^{K} \lambda_k \log \pi_k + (1-\lambda_k) \log (1-\pi_k)}\]
 
@@
 
-\[H\left[ q(\textbf{s})\right] = \sum_{i=1}^K H\left[ q_{i} (s_i)\right] \]
+\[H\left[ q(\textbf{s})\right] = \sum_{k=1}^K H\left[ q_{k} (s_k)\right] \]
 
@@
 
-\[H\left[ q(\textbf{s})\right] = - \sum_{i=1}^K \sum_{s_i \in \{0, 1\}} q_{i} (s_i) \log q_{i} (s_i) \]
+\[H\left[ q(\textbf{s})\right] = - \sum_{k=1}^K \sum_{s_k \in \{0, 1\}} q_{k} (s_k) \log q_{k} (s_k) \]
 
@@
 
-\[H\left[ q(\textbf{s})\right] = - \sum_{i=1}^K \lambda_i \log \lambda_i + (1-\lambda_i) \log (1-\lambda_i) \]
+\[H\left[ q(\textbf{s})\right] = - \sum_{k=1}^K \lambda_k \log \lambda_k + (1-\lambda_k) \log (1-\lambda_k) \]
 
@@
 \tab \mathcal{F}(q, \theta) = \\
-\tab \tab \frac{-d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right) \\
-\tab \tab + \sum_{i=1}^{K} \lambda_i \log \pi_i + (1-\lambda_i) \log (1-\pi_i)}\\
-\tab \tab - \sum_{i=1}^K \lambda_i \log \lambda_i + (1-\lambda_i) \log (1-\lambda_i)
+\tab \tab \frac{-d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K}  \lambda_k  \mu_k   + \sum_{k=1}^{K} \sum_{k'=1, k'\neq k}^{K}  \lambda_k \lambda_{k'} \mu_k^T\mu_{k'} + \sum_{k=1}^{K}  \lambda_k \mu_k^T\mu_k \right) \\
+\tab \tab + \sum_{k=1}^{K} \lambda_k \log \pi_k + (1-\lambda_k) \log (1-\pi_k)}\\
+\tab \tab - \sum_{k=1}^K \lambda_k \log \lambda_k + (1-\lambda_k) \log (1-\lambda_k)
 \end{array}
@@
 
-To derive the partial update for $q_i(s_i)$ we take the variational derivative of the Lagrangian, enforcing the normalisation of $q_i$:
-
-\[\frac{\partial}{\partial q_i}\left( \mathcal{F}(q, \theta) + \lambda^{LG} \int q_i -1)\right) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} - \log q_i(s_i) - 1 + \lambda^{LG}\]
+To derive the partial update for $q_k(s_k)$ we take the variational derivative of the Lagrangian, enforcing the normalisation of $q_k$:
+
+\[\frac{\partial}{\partial q_k}\left( \mathcal{F}(q, \theta) + \lambda^{LG} \int q_k -1)\right) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{k'\neq k} q_{k'}(s_{k'})} - \log q_k(s_k) - 1 + \lambda^{LG}\]
 
@@
 
-Setting this to zero we can solve for the $\lambda_i$ that maximises the free energy:
-
-\[\log q_i(s_i) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} - 1 + \lambda^{LG}\]
+Setting this to zero we can solve for the $\lambda_k$ that maximises the free energy:
+
+\[\log q_k(s_k) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{k'\neq k} q_{k'}(s_{k'})} - 1 + \lambda^{LG}\]
 
@@
 
-\[\left\langle \log P(\textbf{x} |  \textbf{s}, \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} \propto  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} \left\langle s_k \right\rangle_{\prod_{j\neq i} q_j(s_j)} \mu_i   + \sum_{k=1}^{K} \sum_{j=1}^{K} \left\langle s_k s_j \right\rangle_{\prod_{j\neq i} q_j(s_j)} \right)\]
+\[\left\langle \log P(\textbf{x} |  \textbf{s}, \theta)\right\rangle_{\prod_{k'\neq k} q_{k'}(s_{k'})} \propto  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} \left\langle s_k \right\rangle_{\prod_{k'\neq k} q_{k'}(s_{k'})} \mu_k   + \sum_{k=1}^{K} \sum_{k'=1}^{K} \left\langle s_k s_{k'} \right\rangle_{\prod_{k'\neq k} q_{k'}(s_{k'})} \right)\]
 
@@
 
-\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} = \sum_{k=1}^{K} \left\langle s_k \right\rangle_{\prod_{j\neq i} q_j(s_j)} \log \pi_k + (1-\left\langle s_k\right\rangle_{\prod_{j\neq i} q_j(s_j)}) \log (1-\pi_k)}\]
+\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{\prod_{k'\neq k} q_{k'}(s_{k'})} = \sum_{k=1}^{K} \left\langle s_k \right\rangle_{\prod_{k'\neq k} q_{k'}(s_{k'})} \log \pi_k + (1-\left\langle s_k\right\rangle_{\prod_{k'\neq k} q_{k'}(s_{k'})}) \log (1-\pi_k)}\]
 
@@
 
-\[\log q_i(s_i) \propto  \log P(\textbf{x} |  \textbf{s}, \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} + \left\langle \log P(\textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} \]
+\[\log q_k(s_k) \propto  \log P(\textbf{x} |  \textbf{s}, \theta)\right\rangle_{\prod_{k'\neq k} q_{k'}(s_{k'})} + \left\langle \log P(\textbf{s} | \theta)\right\rangle_{\prod_{k'\neq k} q_{k'}(s_{k'})} \]
 
@@
 
-\[\log q_i(s_i) \propto  -\frac{1}{2 \sigma^2} \left( - 2 s_i \textbf{x}^T \mu_i   + s_i s_i  \mu_i^T\mu_i +  2 \sum_{j=1, j\neq i}^{K}  s_i \lambda_j  \mu_i^T\mu_j  \right) + s_i \log \pi_i + (1- s_i ) \log (1-\pi_i)}
-\]
-
-Knowing $\log q_i(s_i) = s_i \log \lambda_i + (1-s_i) \log (1-\lambda_i)$:
-
-\[\log q_i(s_i) \propto s_i \log \frac{\lambda_i}{1-\lambda_i}\]
+\[\log q_k(s_k) \propto  -\frac{1}{2 \sigma^2} \left( - 2 s_k \textbf{x}^T \mu_k   + s_k s_k  \mu_k^T\mu_k +  2 \sum_{k'=1, k'\neq k}^{K}  s_k \lambda_{k'}  \mu_k^T\mu_{k'}  \right) + s_k \log \pi_k + (1- s_k ) \log (1-\pi_k)}
+\]
+
+Knowing $\log q_k(s_k) = s_k \log \lambda_k + (1-s_k) \log (1-\lambda_k)$:
+
+\[\log q_k(s_k) \propto s_k \log \frac{\lambda_k}{1-\lambda_k}\]
 
@@
 
-\[s_i \log \frac{\lambda_i}{1-\lambda_i} \propto  -\frac{1}{2 \sigma^2} \left( - 2 s_i \textbf{x}^T \mu_i   + s_i s_i  \mu_i^T\mu_i +  2 \sum_{j=1, j\neq i}^{K}  s_i \lambda_j  \mu_i^T\mu_j  \right) + s_i \log \frac{\pi_i}{1-\pi_i}}\]
-
-Also, because $s_i \in \{0, 1\}}$ we know that $s_i^2 = s_i$:
-
-\[s_i \log \frac{\lambda_i}{1-\lambda_i} \propto  -\frac{1}{2 \sigma^2} \left( - 2 s_i \textbf{x}^T \mu_i   +  s_i  \mu_i^T\mu_i +  2 \sum_{j=1, j\neq i}^{K}  s_i \lambda_j  \mu_i^T\mu_j  \right) + s_i \log \frac{\pi_i}{1-\pi_i}}\]
-
-Because we have only kept terms with $s_i$, this is an equality:
-
-\[s_i \log \frac{\lambda_i}{1-\lambda_i} =  \frac{  s_i \mu_i^T}{2 \sigma^2} \left( 2\textbf{x} - \mu_i -  2 \sum_{j=1, j\neq i}^{K}   \lambda_j  \mu_j \right) + s_i \log \frac{\pi_i}{1-\pi_i}}\]
-
-Solving for $\lambda_i$:
-
-\[ \lambda_i =  \frac{1}{1+\exp\left[ - \left(\frac{  \mu_i^T}{\sigma^2} \left( \textbf{x} -  \frac{\mu_i}{2} -  \sum_{j=1, j\neq i}^{K}   \lambda_j  \mu_j \right) + \log \frac{\pi_i}{1-\pi_i}}\right) \right]}\]
-
-we have our partial update.
+\[s_k \log \frac{\lambda_k}{1-\lambda_k} \propto  -\frac{1}{2 \sigma^2} \left( - 2 s_k \textbf{x}^T \mu_k   + s_k s_k  \mu_k^T\mu_k +  2 \sum_{k'=1, k'\neq k}^{K}  s_k \lambda_{k'}  \mu_k^T\mu_{k'}  \right) + s_k \log \frac{\pi_k}{1-\pi_k}}\]
+
+Also, because $s_k \in \{0, 1\}}$ we know that $s_k^2 = s_k$:
+
+\[s_k \log \frac{\lambda_k}{1-\lambda_k} \propto  -\frac{1}{2 \sigma^2} \left( - 2 s_k \textbf{x}^T \mu_k   +  s_k  \mu_k^T\mu_k +  2 \sum_{k'=1, k'\neq k}^{K}  s_k \lambda_{k'}  \mu_k^T\mu_{k'}  \right) + s_k \log \frac{\pi_k}{1-\pi_k}}\]
+
+Because we have only kept terms with $s_k$, this is an equality:
+
+\[s_k \log \frac{\lambda_k}{1-\lambda_k} =  \frac{  s_k \mu_k^T}{2 \sigma^2} \left( 2\textbf{x} - \mu_k -  2 \sum_{k'=1, k'\neq k}^{K}   \lambda_{k'}  \mu_{k'} \right) + s_k \log \frac{\pi_k}{1-\pi_k}}\]
+
+Solving for $\lambda_k$:
+
+\[ \lambda_k =  \frac{1}{1+\exp\left[ - \left(\frac{  \mu_k^T}{\sigma^2} \left( \textbf{x} -  \frac{\mu_k}{2} -  \sum_{k'=1, k'\neq k}^{K}   \lambda_{k'}  \mu_{k'} \right) + \log \frac{\pi_k}{1-\pi_k}}\right) \right]}\]
+
+we have our partial update for $q_k(s_k)$
 
@@
 
-\[\hat{\beta} = (\textbf{X}\textbf{X}^T) \textbf{X} \textbf{Y}\]
-
-for the linear regression problem $\textbf{Y} = \textbf{X}^T \beta$ where $\beta$ corresponds to the mean parameters $\mu$, the design matrix $\textbf{X}$ corresponds to the input $\textbf{s}$ and the response $Y$ corresponds to the image pixels denoted $\textbf{x}$.
-This makes sense because our resulting images $\textbf{x}$ are modeled as linear combinations of features $\mu$, weighted by $\textbf{s}$.
+\[\hat{\beta} = (\textbf{X}\textbf{X}^T)^{-1} \textbf{X} \textbf{Y}\]
+
+for the linear regression problem $\textbf{Y} = \textbf{X}^T \beta$ where $\beta$ corresponds to our mean parameters $\mu$, the design matrix $\textbf{X}$ corresponds to our input $\textbf{s}$ and the response $Y$ corresponds to the image pixels we denoted as $\textbf{x}$.
+This makes sense because our resulting images $\textbf{x}$ are modeled as linear combinations of features $\mu$, weighted by $\textbf{s}$, $\textbf{x} = \mu \textbf{s}$ where $\textbf{x} \in \mathbb{R}^{D \times 1}$, $\mu \in \mathbb{R}^{D \times K}$, and $\textbf{s} \in \mathbb{R}^{K \times 1}$.
 
@@
 \item The dot product $\text{ESS}^{-1} \text{ES}^T$ where $\text{ESS}^{-1} \in \mathbb{R}^{K \times K}$ and $\text{ES} \in \mathbb{R}^{N \times K}$ is  $\mathcal{O}(K^2N)$
-\item The dot product $(\text{ESS}^{-1} \text{ES}^T)\textbf{x}$ where $(\text{ESS}^{-1} \text{ES}^T)\ \in \mathbb{R}^{K \times N}$ and $\textbf{x} \in \mathbb{R}^{N \times D}$ is  $\mathcal{O}(KND )$
+\item The dot product $(\text{ESS}^{-1} \text{ES}^T)\textbf{x}$ where $(\text{ESS}^{-1} \text{ES}^T)\ \in \mathbb{R}^{K \times N}$ and $\textbf{x} \in \mathbb{R}^{N \times D}$ is  $\mathcal{O}(KND)$
 \end{itemize}
@@
 \item The dot product $(\mu^{T} \mu)\text{ESS}$ where $\mu^{T} \mu \in \mathbb{R}^{K \times K}$ and $\text{ESS} \in \mathbb{R}^{K \times K}$ is $\mathcal{O}(K^3)$
+\item The dot product  $\text{ESS}^T\textbf{x}$ where $\text{ES} \in \mathbb{R}^{N \times K}$ and $\textbf{x} \in \mathbb{R}^{N \times D}$ is  $\mathcal{O}(KND)$
+\item The dot product $(\text{ESS}^T\textbf{x}) \mu$ where $\text{ESS}^T\textbf{x} \in \mathbb{R}^{K\times D}$ and $\mu \in \mathbb{R}^{D \times K}$ is $\mathcal{O}(K^2 D)$
 \end{itemize}
@@
 \item a vertical line in the fourth column
-\item a a horizontal line in the first row
+\item a horizontal line in the first row
 \end{enumerate}
@@
 
-where $\epsilon \sim \mathcal{N}(\mu_{\epsilon}, \Sigma_{\epsilon})$ and $\textbf{s} \sim \mathcal{N}(\mu_{\textbf{s}}, \Sigma_{\textbf{s}})$.
+where $\epsilon \sim \mathcal{N}(0, \Psi)$ and $\textbf{s} \sim \mathcal{N}(0, \textbf{I})$.
 Factor analysis would be inappropriate for this data because the our latent variables are binary (i.e.
 whether or not a feature is present) and not Gaussians.
-Moreover, the presence of each feature is independent of the presence of another which is not enforced in this model with a covariance matrix that might not be diagonal.
 
@@
 
-\[\textbf{x} = \sum_{k=1}^K\pi_k \mu_k + \epsilon\]
+\[\textbf{x} = \sum_{k=1}^K\pi_k s_k + \epsilon \text{ and } \sum_{k=1}^K s_k = 1\]
 
 where $\epsilon \sim \mathcal{N}(0, \Sigma_{\epsilon})$.
-This also wouldn't be appropriate because each mixture component (feature) is assumed to have some covariance, whereas our mixtures are defined as binary vectors (a cross, a border, etc) and added together before adding some noise.
-
-The independent component analysis assumes a model:
+This also wouldn't be appropriate because it assumes a dirichlet distribution on the vector $\textbf{s}$.
+Our data is constructed by having each $s_k$ independently sampled from a Bernoulli distribution for that feature.
+
+Independent component analysis assumes a model:
 
@@
 
-where $\epsilon \sim \mathcal{N}(0, \sigma^2\textbf{I})$ and $p(\textbf{s}) = \prod_{k=1}^K p(s_k)$.
-This is appropriate for our data because we are linearly combining different features and then adding noise.
-
-Thus, it would be expected that ICA does a good job modelling this data while factor analysis and mixture of Gaussians would not.
+where $\epsilon \sim \mathcal{N}(0, \sigma^2\textbf{I})$ and $P(\textbf{s}) = \prod_{k=1}^K P(s_k)$.
+This is appropriate for our data because $P(s_k)$ is our Bernoulli distribution and we are linearly combining different features with $\textbf{w} in \mathbb{R}^{D \times K}$and then adding noise $\epsilon$.
+
+Thus, it would be expected that ICA does a good job modelling this data while factor analysis and mixture of Gaussians models would not.
 
@@
 
-We can plot the free energy to make sure it increases each iteration:
+We can plot the free energy at each EM step to make sure it increases each iteration:
 
@@
 
-We can see that it has learned some of previously identified features, such as the vertical line in the second column, the two by two square in the bottom left corner, the border, and the a two by two square in the middle.
-THe other features seem to be some linear combination of two or more features, such as $\mu_4$ which looks like a combination of the cross and two by two square in the middle.
-
-A possible way to improve our algorithm is reinitialising our algorithm a few times to find better potential convergence results (i.e. choose model with best free energy).
-Another way to improve the algorithm could be to increase the $K$, although it may learn some duplicate features, there is also a higher chance of capturing all the features. We can visualise this:
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=0.25]{outputs/q3/f-larger-k-latent-factors}
-\caption{Increasing Number of Latent Factors}
-\label{fig:3f-larger-k-latent-factors}
-\end{figure}
-
-Here we can identify a few more features such as the vertical line in the fourth column the cross, and some of the diagonal feature in $\mu_7$.
-
-When implementing the algorithm, the mean field parameters were initialised randomly, each independently from a uniform distribution.
-However $\pi$, $\sigma$, and $\mu$ by running the maximisation step using the randomly initialised mean field parameters.
+We can see that it has learned some of the previously identified features, such as the vertical line in the second column, the two by two square in the bottom left corner, the border, the horizontal line in the first row, and the a two by two square in the middle.
+The other features seem to be some linear combination of two or more features, such as $\mu_4$ which looks like a combination of the cross and two by two square in the middle.
+
+A possible way to improve our algorithm is reinitialising our algorithm a few times to find better potential convergence results (i.e.
+choose model with highest free energy convergence).
+We can also increase the number of data samples, with the hopes of learning better features.
+Finally, we can perform Variational Bayes by setting priors on $\pi$, $\sigma^2$, and $\mu$ for better estimation of these parameters.
+
+When implementing the algorithm, the mean field parameters were initialised randomly, each independently from a uniform distribution on $\[ 0, 1\]$.
+$\pi$, $\sigma$, and $\mu$ were initialised by running a maximisation step using the randomly initialised mean field parameters.
 $K$ was set to eight, after visually identifying eight features in part d.
+Moreover, for computational stability, when calculating $\log(\lambda_i)$ and $\log(1-\lambda_i)$, we manually shifted $\lambda_i$ by $1e^{-10}$ to prevent $\lambda_i=0$ or $\lambda_i = 1$.
 
@@
 
-In the variational expectation step, $\log P(\mathcal{X} | \theta)$ is fixed and we adjust our approximation $q$ to reach this upper bound.
+In the variational expectation step, $\log P(\mathcal{X} | \theta)$ is fixed by the parameters $\pi$, $\sigma$, and $\mu$ and we adjust our approximation $q$ with parameter $\lambda$ to try to reach this upper bound by increasing $\mathcal{F}(q, \theta)$.
 We know that $\sigma$ quantifies the noise of $\textbf{x}$, thus a higher $\sigma$ means a wider spread in our distribution $\log P(\mathcal{X} | \theta)$, meaning we are reducing our upper bound for $\mathcal{F}(q, \theta)$.
@@
 Moreover, by reducing the upper bound, we see in the plot of $\log(F(t)-F(t-1))$ that our free energy is able to converge faster.
-Because we have reduced the upper bound by increasing $\sigma$, our free energy can reach this upper bound faster.
+Because we have reduced the upper bound by increasing $\sigma$, our free energy reaches the upper bound faster.
 
@@
 \newpage
-\section*{Question 4}
+\section*{Question 4: Variational Bayes for binary factors}
 \subsection*{(a)}
 
-We begin by writing the expression for $x_d$:
+To derive a Variational Bayesian hyperparameter optimisation algorithm to automatically determine $K$, the number of hidden binary variables in this model, we begin by writing the expression for $x_d$:
 
@@
 
-Thus, we can write the posterior:
+Thus, we can express the posterior as:
 
@@
 \begin{array}{l}
-\log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) = \\
-\tab \tab + \sum_{d=1}^D-\frac{-1}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(x_d^2 - 2 x_d \textbf{s}^T \textbf{w}_d   + \textbf{w}_d^T \textbf{s}\textbf{s}^T\textbf{w}_d\right) \\
-\tab \tab + \sum_{k=1}^{K} s_k \log \pi_k + (1-s_k) \log (1-\pi_k)}\\
-\tab \tab + \sum_{d=1}^D -\frac{K}{2} \log (2 \pi) + \frac{1}{2}  \sum_{k=1}^K \left(\log\alpha_k \right)  -\frac{1}{2}\textbf{w}_{d}^T \textbf{A}\textbf{w}_{d}\\
+\log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) = \sum_{d=1}^D\frac{-1}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(x_d^2 - 2 x_d \textbf{s}^T \textbf{w}_d   + \textbf{w}_d^T \textbf{s}\textbf{s}^T\textbf{w}_d\right) \\
+\tab \tab \tab \tab + \sum_{k=1}^{K} s_k \log \pi_k + (1-s_k) \log (1-\pi_k)}\\
+\tab \tab \tab \tab + \sum_{d=1}^D \left( \frac{-K}{2} \log (2 \pi) + \frac{1}{2}  \sum_{k=1}^K \left(\log\alpha_k \right)  -\frac{1}{2}\textbf{w}_{d}^T \textbf{A}\textbf{w}_{d} \right)\\
 \end{array}
@@
 
-Given our factored approximation $q(\textbf{s}) = \prod_{i=1}^K q_i(s_i)$, we can see that we can derive a similar partial update for $q_i(s_i)$ as in Question 3, by taking the variation derivative of the Lagrangian to enforce the normalisation of $q_i$:
-
-\[\frac{\partial}{\partial q_i}\left( \exp \left\langle \log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) \right\rangle_{q(\mu)} + \lambda^{LG} \int q_i -1)\right) \propto \exp \left\langle \log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) \right\rangle_{q(\mu)\prod_{j\neq i} q_j(s_j)} - \log q_i(s_i)\]
-
-Setting this to zero we can solve for $\lambda_i$ where $q_i(s_i) = \lambda_i^{s_i}(1-\lambda_i)^{(1-s_i)}$:
-
-\[ \lambda_i =  \frac{1}{1+\exp\left[ - \left(\frac{  \langle \mu_i\rangle_{q_{\mu_i}}^T}{\sigma^2} \left( \textbf{x} -  \frac{\langle \mu_i\rangle_{q_{\mu_i}}}{2} -  \sum_{j=1, j\neq i}^{K}   \lambda_j  \langle \mu_j\rangle_{q_{\mu_j} \right) + \log \frac{ \pi_i}{1- \pi_i\right) \right]}\]
+Given our factored approximation $q(\textbf{s}) = \prod_{k=1}^K q_k(s_k)$, we can see that we can derive a similar partial update for $q_k(q_k)$ as in Question 3, by taking the variation derivative of the Lagrangian to enforce the normalisation of $q_k$:
+
+\[
+\begin{array}{l}
+\frac{\partial}{\partial q_k}\left( \exp \left\langle \log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) \right\rangle_{q(\mu)} + \lambda^{LG} \int q_k -1)\right) \propto \exp \left\langle \log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) \right\rangle_{q(\mu)\prod_{k'\neq k} q_{k'}(s_{k})} - \log q_k(s_k)\\
+\end{array}
+\]
+
+Setting this to zero we can solve for $\lambda_i$ where $q_k(s_k) = \lambda_k^{s_k}(1-\lambda_k)^{(1-s_k)}$:
+
+\[ \lambda_k =  \frac{1}{1+\exp\left[ - \left(\frac{  \langle \mu_k\rangle_{q_{\mu_k}}^T}{\sigma^2} \left( \textbf{x} -  \frac{\langle \mu_k\rangle_{q_{\mu_k}}}{2} -  \sum_{k'=1, k'\neq k}^{K}   \lambda_{k'}  \langle \mu_k\rangle_{q_{\mu_k} \right) + \log \frac{ \pi_k}{1- \pi_k\right) \right]}\]
 
@@
 
-\[q_{\textbf{w}_d}(\textbf{w}_d})  \propto P(\textbf{w}_d)\exp \left\langle \log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{w}_d}(\textbf{w}_d)}\]
+\[q_{\textbf{w}_d}(\textbf{w}_d})  \propto P(\textbf{w}_d)\exp \left\langle \log P(\textbf{x}, \textbf{s}, \mu| \pi , \sigma^2,  \alpha) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{w}_d}}\]
 
@@
 
-\[q_{\textbf{w}_d}(\textbf{w}_d})\propto \exp\left(-\frac{1}{2}\textbf{w}_{d}^T \textbf{A}\textbf{w}_{d} \right) \exp \left\langle  -\frac{1}{2 \sigma^2} \left(- 2 x_d \textbf{s}^T \textbf{w}_d   + \textbf{w}_d^T \textbf{s}\textbf{s}^T\textbf{w}_d\right)\right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{w}_d}(\textbf{w}_d)}\]
+\[q_{\textbf{w}_d}(\textbf{w}_d})\propto \exp\left(-\frac{1}{2}\textbf{w}_{d}^T \textbf{A}\textbf{w}_{d} \right) \exp \left\langle  -\frac{1}{2 \sigma^2} \left(- 2 x_d \textbf{s}^T \textbf{w}_d   + \textbf{w}_d^T \textbf{s}\textbf{s}^T\textbf{w}_d\right)\right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{w}_d}}\]
 
@@
 
-Thus, $\langle \mu_k\rangle_{q_{\mu_k}} \in \mathbb{R}^{D \times 1}$ is simply the concatenation of the $k^{th}$ elements of $\mu_{\textbf{w}_d}$ for $d\in \{1,...,D\}$
+Thus, $\langle \mu_k\rangle_{q_{\mu_k}} \in \mathbb{R}^{D \times 1}$ is the concatenation of the $k^{th}$ elements of $\mu_{\textbf{w}_d} \in \mathbb{R}^{K \times 1}$ for $d\in \{1,...,D\}$
 
 For ARD, we must also optimise $\alpha$ with a hyper-M step.
-We start by choose $Ga(\alpha_k| a, b)$, a Gamma prior on \alpha_k$, with $a$ and $b$ being hyperparameters.
+We start by choosing $Ga(\alpha_k| a, b)$, a Gamma prior on \alpha_k$, with $a$ and $b$ being hyperparameters.
 Thus, to optimise $\alpha$ we want to maximise the penalised objective:
@@
 
-%Knowing $\left\langle  \textbf{w}_{d}\textbf{w}_{d}^T \right\rangle_{q(\textbf{w})} = \mu_{\textbf{w}_d}\mu_{\textbf{w}_d}^T + \Sigma_{\textbf{w}_d}$:
-%
-%\[\mathcal{Q} =  \frac{D}{2}  \sum_{k=1}^K \left( \log\alpha_k \right)  - \frac{1}{2}\sum_{d=1}^D \left( tr \left[ \textbf{A}\left(\mu_{\textbf{w}_d}\mu_{\textbf{w}_d}^T + \Sigma_{\textbf{w}_d}\right)} \right] \right)+ \sum_{k=1}^K (a-1)\log\alpha_k -b\alpha_k\]
-
 Setting $\frac{d\mathcal{Q}}{d\alpha_k} = 0$ we get:
@@
 
-Knowing $\left\langle (w_{d, k})^2 \right\rangle_{q(\textbf{w}_{d})} = (\mu_{\textbf{w}_{d, k}})^2 + \Sigma_{\textbf{w}_{d, (k, k)}}$, we can solve for $\alpha_k$:
+Knowing $\left\langle (w_{d, k})^2 \right\rangle_{q(\textbf{w}_{d})} = (\mu_{\textbf{w}_{d, k}})^2 + \Sigma_{\textbf{w}_{d, (k, k)}}$, the $k^{th}$ element of $\mu_{\textbf{w}_d}$ and element $(k, k)$ of $\Sigma_{\textbf{w}_d$ respectively, we can solve for $\alpha_k$:
 
@@
 \end{figure}
-As we expect, when running the algorithm for higher $K$ values, many of the features have $\alpha_k \rightarrow \infty$, depicted as $\alpha^{-1}_k$ for visual convenience.
+As we expect, when running the algorithm for higher $K$ values, many of the features have $\alpha_k \rightarrow \infty$, depicted as $\alpha^{-1}_k \rightarrow 0$ for visual convenience.
 Moreover, visualising the learned features, we can see the clearest features often have the highest $\alpha^{-1}_k$ while the features deemed irrelevant are often noisy or duplicates.
@@
 
-We can see that initially, for $k=4$ to $k=8$, increasing $k$ significantly increases the convergence value of the free energy.
-However, beyond $k=8$ there is a no clear trend of $k$ versus the free energy convergence value.
-We can see that this corresponds to the visualisation of $\alpha^{-1}$ where beyond $k=11$, the effective number of features remains more or less the same.
-We know that there are only eight latent features, thus models with $k>8$ should be learning duplicate or irrelevant features.
+We can see that initially for $K=4$ to $K=8$, increasing $K$ significantly increases the convergence value of the free energy.
+However, beyond $K=8$ there is a no clear trend of $K$ versus the free energy convergence value.
+We can see that this corresponds to the visualisation of $\alpha^{-1}$ where beyond $K=11$, the effective number of features remains more or less the same.
+We know that there are only eight latent features, thus models with $K>8$ should be learning duplicate or irrelevant features.
 As such, we wouldn't expect a model to be able to increase it's free energy significantly when provided with additional degrees of freedom by increasing the value of $k$ beyond eight.
-We see that for models with $k>>8$, there are typically ten or eleven features that might be deemed relevant (depending on how you threshold) and this is likely from slight overfitting, noise in the data, or duplicate features.
+We see that for models with $K>>8$, there are typically ten or eleven features that might be deemed relevant (depending on how you threshold) and this is likely from slight overfitting, noise in the data, or duplicate features.
 Thus, the relationship between the free energy and the effective number of latent features for each model is as we would expect with ARD.
@@
 \newpage
+The Python code for Automatic Relevance Determination:
+\lstinputlisting[language=Python]{src/automatic_relevance_determination.py}
+
+\newpage
 The rest of the Python code for question 4:
@@
 \newpage
-\section*{Question 5}
+\section*{Question 5: EP for the binary factor model}
 
@@
 
-The log-joint probability for a single observation-source pair:
+To derive an EP algorithm to infer marginals on the source variables in the binary latent factor model from Question 3, we first write the log-joint probability for a single observation-source pair:
 
@@
 
-\[\tilde{f}_i(s_i) = \lambda_i^{s_i} + (1-\lambda_i)^{1-s_i}\]
+\[\tilde{f}_i(s_i) = (\theta_{ii})^{s_i} + (1-\theta_{ii})^{1-s_i}\]
 
@@
 
-\[\log \tilde{f}_i(s_i) \propto \log \left(\frac{\lambda_i}{1-\lambda_i} \right)s_i\]
+\[\log \tilde{f}_i(s_i) \propto \log \left(\frac{\theta_{ii}}{1-\theta_{ii}} \right)s_i\]
+
+we can define $\eta_{ii} = \log \left(\frac{\theta_{ii}}{1-\theta_{ii}} \right)$:
+
+\[\log \tilde{f}_i(s_i) \propto \eta_{ii}s_i\]
 
@@
 
-we can define $\xi_{ji} = \log \left(\frac{\theta_{ji}}{1-\theta_{ji}} \right)$ and $\xi_{ij} = \log \left(\frac{\theta_{ij}}{1-\theta_{ij}} \right)$:
-
-\[\log \tilde{g}_{ij}(s_i, s_j) \propto \xi_{ji} s_i + \xi_{ij} s_j\]
-
-To derive the a message passing scheme, we first define the incoming message to node $i$ from the singleton factor:
-
-\[\mathcal{M}_{i}(s_i) = \tilde{f}_i(s_i)\]
-
-and the message incoming message to node $i$ from node $j$:
+we can define $\eta_{ji} = \log \left(\frac{\theta_{ji}}{1-\theta_{ji}} \right)$ and $\eta_{ij} = \log \left(\frac{\theta_{ij}}{1-\theta_{ij}} \right)$:
+
+\[\log \tilde{g}_{ij}(s_i, s_j) \propto \eta_{ji} s_i + \eta_{ij} s_j\]
+
+To derive a message passing scheme, we define the incoming message to node $i$ from the singleton factor:
+
+\[\mathcal{M}_{i \rightarrow i}(s_i) = \tilde{f}_i(s_i)\]
+
+and the incoming message to node $i$ from node $j$:
 
@@
 
-\[\mathcal{M}_{j \rightarrow i}(s_i) = \sum_{s_1 \in \{0, 1\} } \cdots \sum_{s_{i-1} \in \{0, 1\} } \sum_{s_{i+1} \in \{0, 1\} } \cdots \sum_{s_1 \in \{0, 1\} } \tilde{f}_j(s_j) \tilde{g}_{ji}(s_j, s_i) \prod_{k\in ne(j), k\neq i}^{K}} \mathcal{M}_{k \rightarrow j}(s_j)\]
+\[\mathcal{M}_{j \rightarrow i}(s_i) = \sum_{s_1 \in \{0, 1\} } \cdots \sum_{s_{i-1} \in \{0, 1\} } \sum_{s_{i+1} \in \{0, 1\} } \cdots \sum_{s_1 \in \{0, 1\} }  \tilde{g}_{ji}(s_j, s_i) \mathcal{M}_{j \rightarrow j}(s_j)\prod_{k\in ne(j), k\neq i}^{K}} \mathcal{M}_{k \rightarrow j}(s_j)\]
 
 where $ne(j)$ are indices of neighbouring nodes of node $j$.
+This is the product of the incoming messages to node $j$ from its neighbours, the singleton message from node $j$, and $\tilde{g}_{ji}(s_j, s_i)$ with all nodes except $s_i$ marginalised out.
 
@@
 
-Simplifying:
+We are left with:
 
@@
 
-\[\mathcal{M}_{j \rightarrow i}(s_i) \propto \exp\left(\xi_{ji} s_i\right)\]
-
-Thus, the cavity distributions are:
+\[\mathcal{M}_{j \rightarrow i}(s_i) \propto \exp\left(\eta_{ji} s_i\right)\]
+
+The cavity distributions are:
 
@@
 
-\[q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) = \left( \mathcal{M}_{i}(s_i) \prod_{k\in ne(i), k\neq j}^{K} \mathcal{M}_{k \rightarrow i}(s_i)
+\[q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) = \left( \mathcal{M}_{i \rightarrow i}(s_i) \prod_{k\in ne(i), k\neq j}^{K} \mathcal{M}_{k \rightarrow i}(s_i)
 \right)\left( \mathcal{M}_{j}(s_j) \prod_{k\in ne(j), k\neq i}^{K} \mathcal{M}_{k \rightarrow j}(s_j)
@@
 We know that the factor $\log f_i(s_i)$ is a Bernoulli of the form $b_i s_i$.
-Because our approximation for this site is also Bernoulli, we can simply solve for $\lambda_i$ in $\log \tilde{f}_{i}(s_{i})$:
+Because our approximation for this site is also Bernoulli, we can simply solve for $\theta_{ii}$ in $\log \tilde{f}_{i}(s_{i})$:
 
@@
 
-\[\log \left(\frac{\lambda_i}{1-\lambda_i} \right)s_i = b_i s_i\]
-
-\[\lambda_i = \frac{1}{1+\exp(-b_i)}\]
+\[\log \left(\frac{\theta_{ii}}{1-\theta_{ii}} \right)s_i = b_i s_i\]
+
+\[\theta_{ii} = \frac{1}{1+\exp(-b_i)}\]
 
@@
 
-\[\mathcal{M}_{i}(s_i) \prod_{k\in ne(i), k\neq j}^{K} \mathcal{M}_{k \rightarrow i}(s_i) \propto \exp(\eta_{i, \neg s_j} s_i)\]
+\[\mathcal{M}_{i \rightarrow i}(s_i) \prod_{k\in ne(i), k\neq j}^{K} \mathcal{M}_{k \rightarrow i}(s_i) \propto \exp(\eta_{i, \neg s_j} s_i)\]
 
@@
 
-Note that $\tilde{g}_{ij}(s_i, s_j)$ was chosen as the product of two Bernoulli distributions, updates to this site approximation involves updating the  parameters $\xi_{ij}$ and $\xi_{ji}$, for $s_i$ and $s_j$ respectively.
+Note that $\tilde{g}_{ij}(s_i, s_j)$ was chosen as the product of two Bernoulli distributions, updates to this site approximation involves updating the  parameters $\eta_{ij}$ and $\eta_{ji}$, for $s_i$ and $s_j$ respectively.
 
 We can write:
-%\[\log \tilde{g}_{ij}q_{\neg \tilde{g}_{ij}} \propto \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j + \sum_{i=1}^{K}  \log \left(\frac{\lambda_i}{1-\lambda_i} \right) +  \sum_{i'=1}^{K}\sum_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1}  \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j\]
-
-\[\log \tilde{g}_{ij}(s_i, s_j)q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \propto \xi_{ji} s_i + \xi_{ij} s_j + \eta_{i, \neg s_j} s_i + \eta_{j, \neg s_i} s_j\]
+%\[\log \tilde{g}_{ij}q_{\neg \tilde{g}_{ij}} \propto \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j + \sum_{i=1}^{K}  \log \left(\frac{\theta_{ii}}{1-\theta_{ii}} \right) +  \sum_{i'=1}^{K}\sum_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1}  \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j\]
+
+\[\log \left(\tilde{g}_{ij}(s_i, s_j)q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right) \propto \eta_{ji} s_i + \eta_{ij} s_j + \eta_{i, \neg s_j} s_i + \eta_{j, \neg s_i} s_j\]
 
@@
 
-\[\log \tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \propto \left(\xi_{ji}+ \eta_{i, \neg s_j} \right) s_i + \left(\xi_{ij} + \eta_{j, \neg s_i} \right) s_j \]
+\[\log \left(\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \right)\propto \left(\eta_{ji}+ \eta_{i, \neg s_j} \right) s_i + \left(\eta_{ij} + \eta_{j, \neg s_i} \right) s_j \]
 
@@
 
-\[\mathbb{E}_{s_i}\left[\sum_{s_j \in \{0, 1\}}\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \frac{1}{1+\exp{\left(-\left(\xi_{ji} + \eta_{i, \neg s_j} \right)}\right)}\]
+\[\mathbb{E}_{s_i}\left[\sum_{s_j \in \{0, 1\}}\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \frac{1}{1+\exp{\left(-\left(\eta_{ji} + \eta_{i, \neg s_j} \right)}\right)}\]
 
@@
 
-\[\mathbb{E}_{s_j}\left[\sum_{s_i \in \{0, 1\}}\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \frac{1}{1+\exp{\left(-\left(\xi_{ij} + \eta_{j, \neg s_i} \right)}\right)}\]
+\[\mathbb{E}_{s_j}\left[\sum_{s_i \in \{0, 1\}}\tilde{g}_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j)\right] = \frac{1}{1+\exp{\left(-\left(\eta_{ij} + \eta_{j, \neg s_i} \right)}\right)}\]
 
@@
 
-\[\log g_{ij}(s_i, s_j) qq_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \propto W_{ij} s_i s_j
+\[\log \left(g_{ij}(s_i, s_j) q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) \right)\propto W_{ij} s_i s_j
  + \eta_{i, \neg s_j} s_i + \eta_{j, \neg s_i} s_j\]
@@
 
-\[\frac{1}{1+\exp{\left(-\left(\xi_{ji} + \eta_{i, \neg s_j} \right)}\right)}} = \frac{\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)  }{\left[\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)  \right]+\left[\exp\left( \eta_{j, \neg s_i}\right) +  1\right]}\]
+\[\frac{1}{1+\exp{\left(-\left(\eta_{ji} + \eta_{i, \neg s_j} \right)}\right)}} = \frac{\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)  }{\left[\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)  \right]+\left[\exp\left( \eta_{j, \neg s_i}\right) +  1\right]}\]
 
@@
 
-\[\exp\left( \eta_{j, \neg s_i}\right) +  1=\exp{\left(-\left(\xi_{ji} + \eta_{i, \neg s_j} \right)}\right)}\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)\]
-
-\[\frac{\exp\left( \eta_{j, \neg s_i}\right) +  1}{ \exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1} =\exp{\left(-\xi_{ji}\right)}\]
+\[\exp\left( \eta_{j, \neg s_i}\right) +  1=\exp{\left(-\left(\eta_{ji} + \eta_{i, \neg s_j} \right)}\right)}\exp\left(\eta_{i, \neg s_j}\right)\left(\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1\right)\]
+
+\[\frac{\exp\left( \eta_{j, \neg s_i}\right) +  1}{ \exp\left( W_{ij}  + \eta_{j, \neg s_i}\right) + 1} =\exp{\left(-\eta_{ji}\right)}\]
 
@@
 
-\[\xi_{ji} = \log \left( \frac{1+\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right)}{1+\exp\left( \eta_{j, \neg s_i}\right)} \right)\]
+\[\eta_{ji} = \log \left( \frac{1+\exp\left( W_{ij}  + \eta_{j, \neg s_i}\right)}{1+\exp\left( \eta_{j, \neg s_i}\right)} \right)\]
 
@@
 
-\[\xi_{ij} = \log \left( \frac{1+\exp\left( W_{ij}  + \eta_{i, \neg s_j}\right)}{1+\exp\left( \eta_{i, \neg s_j}\right)} \right)\]
+\[\eta_{ij} = \log \left( \frac{1+\exp\left( W_{ij}  + \eta_{i, \neg s_j}\right)}{1+\exp\left( \eta_{i, \neg s_j}\right)} \right)\]
 
@@
 
-\[\eta_{i, \neg s_j} =  \log \left(\frac{\lambda_i}{1-\lambda_i}\right) + \sum_{k\in ne(i), k\neq j}^{K} \log\left(\frac{\theta_{ki}}{1-\theta_{ki}} \right)\]
-
-Knowing $b_i=\log \left(\frac{\lambda_i}{1-\lambda_i}\right)$ and $\xi_{ki}=\log\left(\frac{\theta_{ki}}{1-\theta_{ki}} \right)$:
-
-\[\eta_{i, \neg s_j} =  b_i + \sum_{k\in ne(i), k\neq j}^{K} \xi_{ki}\]
+\[\eta_{i, \neg s_j} =  \log \left(\frac{\theta_{ii}}{1-\theta_{ii}}\right) + \sum_{k\in ne(i), k\neq j}^{K} \log\left(\frac{\theta_{ki}}{1-\theta_{ki}} \right)\]
+
+Knowing $\eta_{ii}=\log \left(\frac{\theta_{ii}}{1-\theta_{ii}}\right)$ and $\eta_{ki}=\log\left(\frac{\theta_{ki}}{1-\theta_{ki}} \right)$:
+
+\[\eta_{i, \neg s_j} =  \eta_{ii} + \sum_{k\in ne(i), k\neq j}^{K} \eta_{ki}\]
 
@@
 
-\[\eta_{j, \neg s_i} =  b_j + \sum_{k\in ne(j), k\neq i}^{K} \xi_{kj}\]
-
-The summation of the natural parameters of the singleton factor for node $i$ with the natural parameters of messages from all the neighbouring nodes.
+\[\eta_{j, \neg s_i} =  \eta_{jj} + \sum_{k\in ne(j), k\neq i}^{K} \eta_{kj}\]
+
+The summation of the natural parameters of the singleton factor for node $i$ with the natural parameters of messages from all the neighbouring nodes that aren't $j$.
 
@@
 
+Moreover, our factored approximations:
+
+\[q(s_i) \propto \sum_{j=1}^{K} \eta_{ij} s_i\]
+
+and so:
+
+\[\lambda_i = \frac{1}{1+\exp(-\sum_{j=1}^{K} \eta_{ij})}\]
+
+our parameter for $q(s_i)$.
+
 \subsection*{(d)}
@@
 \newpage
-\section*{Question 6}
+\section*{Question 6: EP/Loopy BP Implementation}
 
@@
 We can see that the mean field algorithm seems to learn better latent features.
-In particular there's are fewer duplicates, unlike loopy BP that has a few duplicates of the two by two square in the middle.
+In particular there are fewer duplicates, unlike loopy BP that has a few duplicates like $\mu_3$ and $\mu_4$.
 Moreover, the learned features have less noise.
@@
 We can observe that the free energy of the mean field algorithm converges while our loopy belief propagation is unable to converge to a free energy.
-Because loopy BP does not have convergence guarantees, this is one of the limitations of this approach.
+Because loopy BP does not have convergence guarantees, one of the limitations of this approach, we can see that in this case loopy BP is unable to identify many latent factors.
 

### 2023-01-03 src/models/binary_latent_factor_approximations/message_passing_approximation.py

@@ -10,14 +10,14 @@ from src.models.binary_latent_factor_models.boltzmann_machine import BoltzmannMa
 
 class MessagePassingApproximation(AbstractBinaryLatentFactorApproximation):
     """
-    eta_matrix:  of parameters eta_matrix[n, i, j]
+    bernoulli_parameter_matrix (theta): matrix of parameters bernoulli_parameter_matrix[n, i, j]
                 off diagonals corresponds to \tilda{g}_{ij, \neg s_i}(s_j) for data point n
                 diagonals correspond to \tilda{f}_{i}(s_i)
                 (number_of_points, number_of_latent_variables, number_of_latent_variables)
     """
 
-    def __init__(self, eta_matrix: np.ndarray):
-        self.eta_matrix = eta_matrix
+    def __init__(self, bernoulli_parameter_matrix: np.ndarray):
+        self.bernoulli_parameter_matrix = bernoulli_parameter_matrix
 
     @property
     def lambda_matrix(self) -> np.ndarray:
@@ -25,14 +25,25 @@ class MessagePassingApproximation(AbstractBinaryLatentFactorApproximation):
         Aggregate messages and compute parameter for Bernoulli distribution
         :return:
         """
-        lambda_matrix = 1 / (1 + np.exp(-self.xi.sum(axis=1)))
+        lambda_matrix = 1 / (1 + np.exp(-self.natural_parameter_matrix.sum(axis=1)))
         lambda_matrix[lambda_matrix == 0] = 1e-10
         lambda_matrix[lambda_matrix == 1] = 1 - 1e-10
         return lambda_matrix
 
     @property
-    def xi(self) -> np.ndarray:
-        return np.log(np.divide(self.eta_matrix, 1 - self.eta_matrix))
+    def natural_parameter_matrix(self) -> np.ndarray:
+        """
+        The matrix containing natural parameters (eta) of each factor
+                off diagonals corresponds to \tilda{g}_{ij, \neg s_i}(s_j) for data point n
+                diagonals correspond to \tilda{f}_{i}(s_i)
+                (number_of_points, number_of_latent_variables, number_of_latent_variables)
+        :return:
+        """
+        return np.log(
+            np.divide(
+                self.bernoulli_parameter_matrix, 1 - self.bernoulli_parameter_matrix
+            )
+        )
 
     def aggregate_incoming_binary_factor_messages(
         self, node_index: int, excluded_node_index: int
[…]

### 2023-01-04 submission.tex
add ard for loop bp and edit report

@@
 \usetikzlibrary{arrows}
+\usepackage{titlesec}
+
+\titleformat{\section}
+  {\normalfont\fontsize{14}{15}\bfseries}{\thesection}{1em}{}
+\titleformat{\subsection}
+  {\normalfont\fontsize{12}{15}\bfseries}{\thesection}{1em}{}
 
@@
 %\author{Jian Shu (James) Wu \\ }
-\date{Jan 4, 2023}
+\date{Jan 6, 2023}
 
@@
 The set $\{A, D, E, F, H\}$ is a non-unique smallest set of molecules such that if the concentrations of the species within the set are known, the concentrations of the others $\{B, C, G, I\}$ would all be independent (conditioned on the measured ones).
-\newpage
 
@@
 Thus, we would expect to recover the factors of $A$ and $C$, $\delta[A]$ and $\delta[C]$, the two parent nodes of the observations.
-\newpage
 
@@
 However, in our DAG this is not hte case, so we would not be able to recover any other species in the cascade because  they are all downstream from the known concentration perturbations and those recovered from factor analysis.
+Similarly, the downstream weights in the DAG from the known measurements would also be unidentifiable.
+Using factor analysis, we would only be able to identify the upstream weights from the measurements, up to an unknown scale factor.
+These are the weights $\Lambda_{BA}$, $\Lambda_{DC}$, and $\Lambda_{EC}$.
+Moreover, knowing $\delta[C]$ from factor analysis, we would also be able to determine $\Lambda_{CA}$, up to an unknown scale factor.
 
@@
 
-\[k(s, t) = \theta^2 \left( \exp\left( - \frac{2\sin^2(\pi(s-t)/\tau)}{\sigma^2}\right) + \phi^2 \exp\left( - \frac{(s-t)^2}{2\eta^2}\right)\right) + \zeta^2 \delta_{s=t}\]
+\[k(s, t) = \theta^2 \exp\left( - \frac{2\sin^2(\pi(s-t)/\tau)}{\sigma^2}\right) + \phi^2 \exp\left( - \frac{(s-t)^2}{2\eta^2}\right) + \zeta^2 \delta_{s=t}\]
 
@@
 \item The dot product $(\mu^{T} \mu)\text{ESS}$ where $\mu^{T} \mu \in \mathbb{R}^{K \times K}$ and $\text{ESS} \in \mathbb{R}^{K \times K}$ is $\mathcal{O}(K^3)$
-\item The dot product  $\text{ESS}^T\textbf{x}$ where $\text{ES} \in \mathbb{R}^{N \times K}$ and $\textbf{x} \in \mathbb{R}^{N \times D}$ is  $\mathcal{O}(KND)$
-\item The dot product $(\text{ESS}^T\textbf{x}) \mu$ where $\text{ESS}^T\textbf{x} \in \mathbb{R}^{K\times D}$ and $\mu \in \mathbb{R}^{D \times K}$ is $\mathcal{O}(K^2 D)$
+\item The dot product  $\text{ES}^T\textbf{x}$ where $\text{ES} \in \mathbb{R}^{N \times K}$ and $\textbf{x} \in \mathbb{R}^{N \times D}$ is  $\mathcal{O}(KND)$
+\item The dot product $(\text{ES}^T\textbf{x}) \mu$ where $\text{ES}^T\textbf{x} \in \mathbb{R}^{K\times D}$ and $\mu \in \mathbb{R}^{D \times K}$ is $\mathcal{O}(K^2 D)$
 \end{itemize}
@@
 
-\[\textbf{x} = \sum_{k=1}^K\pi_k s_k + \epsilon \text{ and } \sum_{k=1}^K s_k = 1\]
+\[\textbf{x} = \sum_{k=1}^K s_k \mu_k + \epsilon \text{ and } \sum_{k=1}^K s_k = 1\]
 
@@
 This also wouldn't be appropriate because it assumes a dirichlet distribution on the vector $\textbf{s}$.
-Our data is constructed by having each $s_k$ independently sampled from a Bernoulli distribution for that feature.
+Our data is constructed by having each $s_k$ independently sampled from a Bernoulli distribution indicating the presence of feature $\mu_k$.
 
@@
 where $\epsilon \sim \mathcal{N}(0, \sigma^2\textbf{I})$ and $P(\textbf{s}) = \prod_{k=1}^K P(s_k)$.
-This is appropriate for our data because $P(s_k)$ is our Bernoulli distribution and we are linearly combining different features with $\textbf{w} in \mathbb{R}^{D \times K}$and then adding noise $\epsilon$.
+This is appropriate for our data because $P(s_k)$ is our Bernoulli distribution and we are linearly combining different features with $\textbf{W} \in \mathbb{R}^{D \times K}$and then adding noise $\epsilon$.
 
@@
 
+\newpage
 \subsection*{(d)}
 
-Similar to question 3, we can use automatic relevance determination (ARD) as a hyperparameter method to select relevant features by placing a prior on $\mu_i$.
-With a hyper-M step, certain features will have diverging precision, indicating that they are not relevant to the model output.
-Thus, the number of remaining features will be our selection for $K$.
+We can use automatic relevance determination (ARD) as a hyperparameter method to select relevant features by placing a prior on $\mu_i$ in the same way as Question 4.
+With a hyper-M step, certain features will have diverging precision, leaving us with the relevant features, from which we get our $K$.
+We can implement this by running the message passing algorithm (loopy BP) which was implemented for Question 6 for the expectation step and the VB algorithm for the maximisation step (along with its hyper-M Step):
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.4]{outputs/q5/d-latent-factors-comparison}
+\caption{Learned Latent Factors vs Inverse Alpha}
+\label{fig:}
+\end{figure}
+
+We can see that ARD has more difficulty finding relevant features when using loop BP.
+Looking at the features themselves, we can see that they are not very clear compared to the mean field approximation or Variational Bayes in Question 4.
+This is probably why we do not consistently have the same number of $\alpha_k$ values that don't diverge as compared with Question 3.
+However, we can see for $K<8$ that all of the features are deemed relevant as we would expect.
+
+\newpage
+Comparing the free energy plots of models trained on different $K$ values:
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.4]{outputs/q5/d-free-energy}
+\caption{Free Energy for different values of k}
+\label{fig:}
+\end{figure}
+
+We can see the difficulties of this approach due to the inability for loopy BP to converge for this particular problem.
+This likely contributes to the difficulty for ARD to find relevant features when the message passing algorithm is unable to converge on generating meaningful features.
+To improve this, we could also put priors on the other parameters $\sigma$ and $\pi$ in the hopes that this will help stabilise the algorithm.
+Moreover, because loopy BP is unable to converge for this problem, it is quite computationally intensive as EM will always run the maximum number of iterations without improvement in the results.
+Knowing that it won't converge, employing early stopping or fewer iterations can also be helpful.
+
+\newpage
+The Python code for question 5:
+\lstinputlisting[language=Python]{src/solutions/q5.py}
 
@@
 \lstinputlisting[language=Python]{src/models/binary_latent_factor_models/abstract_binary_latent_factor_model.py}
-
 \newpage
@@
 \lstinputlisting[language=Python]{src/models/binary_latent_factor_approximations/abstract_binary_latent_factor_approximation.py}
-
 \newpage
@@
 \lstinputlisting[language=Python]{main.py}
-
 \newpage
@@
 \lstinputlisting[language=Python]{src/constants.py}
-
 \newpage
@@
 \newpage
-\section*{Appendix 6: MStep.py}
-\lstinputlisting[language=Python]{demo_code/MStep.py}
+\section*{Appendix 6: m\_step.py}
+\lstinputlisting[language=Python]{demo_code/m_step.py}
 \end{document}

### 2023-01-04 demo_code/m_step.py

@@ -1,9 +1,11 @@
 import numpy as np
 
+from typing import Tuple
 
-def m_step(x, es, ess):
+
+def m_step(x: np.ndarray, es: np.ndarray, ess: np.ndarray) -> Tuple[np.ndarray, float, np.ndarray]:
     """
-    mu, sigma, pie = MStep(x,es,ess)
+    mu, sigma, pie = m_step(x,es,ess)
 
     Inputs:
     -----------------

### 2023-01-04 main.py

@@ -11,7 +11,7 @@ from src.generate_images import generate_images
 from src.models.bayesian_linear_regression import LinearRegressionParameters
 from src.models.gaussian_process_regression import GaussianProcessParameters
 from src.models.kernels import CombinedKernel, CombinedKernelParameters
-from src.solutions import q2, q3, q4, q6
+from src.solutions import q2, q3, q4, q5, q6
 
 jax.config.update("jax_enable_x64", True)
 
@@ -148,18 +148,7 @@ if __name__ == "__main__":
     if not os.path.exists(Q4_OUTPUT_FOLDER):
         os.makedirs(Q4_OUTPUT_FOLDER)
     max_k = 21
-    q4.b(
-        x=x,
-        a_parameter=1,
-        b_parameter=0,
-        ks=np.arange(4, 22),
-        max_k=max_k,
-        em_iterations=em_iterations,
-        e_maximum_steps=e_maximum_steps,
-        e_convergence_criterion=e_convergence_criterion,
-        save_path=os.path.join(Q4_OUTPUT_FOLDER, "b"),
-    )
-    q4.b(
+    free_energies_1 = q4.b(
         x=x,
         a_parameter=1,
         b_parameter=0,
@@ -170,7 +159,7 @@ if __name__ == "__main__":
         e_convergence_criterion=e_convergence_criterion,
         save_path=os.path.join(Q4_OUTPUT_FOLDER, "b-1"),
     )
-    q4.b(
+    free_energies_2 = q4.b(
         x=x,
         a_parameter=1,
         b_parameter=0,
@@ -180,6 +169,33 @@ if __name__ == "__main__":
         e_maximum_steps=e_maximum_steps,
         e_convergence_criterion=e_convergence_criterion,
         save_path=os.path.join(Q4_OUTPUT_FOLDER, "b-2"),
+    )
+    q4.free_energy_plot(
+        ks=np.arange(4, 22),
+        free_energies=free_energies_1 + free_energies_2,
+        model_name="Variational Bayes",
+        save_path=os.path.join(Q4_OUTPUT_FOLDER, "b"),
+    )
+
+    # Question 5
+    Q5_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q5")
+    if not os.path.exists(Q5_OUTPUT_FOLDER):
+        os.makedirs(Q5_OUTPUT_FOLDER)
+    max_k = 20
+    free_energies = q5.d(
[…]

### 2023-01-04 src/automatic_relevance_determination.py

@@ -3,45 +3,39 @@ from typing import List, Tuple
 import numpy as np
 
 from src.expectation_maximisation import learn_binary_factors
-from src.models.binary_latent_factor_approximations.mean_field_approximation import (
-    init_mean_field_approximation,
+from src.models.binary_latent_factor_approximations.abstract_binary_latent_factor_approximation import (
+    AbstractBinaryLatentFactorApproximation,
 )
 from src.models.binary_latent_factor_models.binary_latent_factor_model import (
     BinaryLatentFactorModel,
 )
 from src.models.binary_latent_factor_models.variational_bayes import (
     GaussianPrior,
-    VariationalBayesBinaryLatentFactorModel,
+    VariationalBayes,
 )
 
 
 def run_automatic_relevance_determination(
     x: np.ndarray,
-    a_parameter: int,
-    b_parameter: int,
+    binary_latent_factor_approximation: AbstractBinaryLatentFactorApproximation,
+    a_parameter: float,
+    b_parameter: float,
     k: int,
     em_iterations: int,
-    e_maximum_steps: int,
-    e_convergence_criterion: float,
-) -> Tuple[VariationalBayesBinaryLatentFactorModel, List[float]]:
+) -> Tuple[VariationalBayes, List[float]]:
     """
     Run automatic relevance determination with variational Bayes.
 
     :param x: data matrix (number_of_points, number_of_dimensions)
+    :param binary_latent_factor_approximation: a binary_latent_factor_approximation
     :param a_parameter: alpha parameter for gamma prior
     :param b_parameter: beta parameter for gamma prior
     :param k: number of latent variables
     :param em_iterations: number of iterations to run EM
-    :param e_maximum_steps: maximum number of iterations of partial expectation steps
-    :param e_convergence_criterion: minimum required change in free energy for each partial expectation step
     :return: a Tuple containing the optimised VB model and a list of free energies during each EM step
     """
-    n = x.shape[0]
-    mean_field_approximation = init_mean_field_approximation(
-        k, n, max_steps=e_maximum_steps, convergence_criterion=e_convergence_criterion
-    )
     (_, sigma, pi,) = BinaryLatentFactorModel.calculate_maximisation_parameters(
-        x, mean_field_approximation
+        x, binary_latent_factor_approximation
     )
     mu = GaussianPrior(
         a=a_parameter,
@@ -49,17 +43,15 @@ def run_automatic_relevance_determination(
         k=k,
[…]

### 2023-01-04 src/generate_images.py

@@ -3,7 +3,9 @@ import numpy as np
 from src.constants import DEFAULT_SEED, M1, M2, M3, M4, M5, M6, M7, M8
 
 
-def generate_images(n: int = 400, seed: int = DEFAULT_SEED, sigma: float = 0.1):
+def generate_images(
+    n: int = 400, seed: int = DEFAULT_SEED, sigma: float = 0.1
+) -> np.ndarray:
     """
     Image generation, adapted from provided demo code
 

### 2023-01-04 src/models/binary_latent_factor_models/binary_latent_factor_model.py

@@ -1,8 +1,8 @@
-from typing import TYPE_CHECKING, Tuple
+from typing import Tuple
 
 import numpy as np
 
-from demo_code.MStep import m_step
+from demo_code.m_step import m_step
 from src.models.binary_latent_factor_approximations.abstract_binary_latent_factor_approximation import (
     AbstractBinaryLatentFactorApproximation,
 )
@@ -18,15 +18,9 @@ class BinaryLatentFactorModel(AbstractBinaryLatentFactorModel):
         sigma: float,
         pi: np.ndarray,
     ):
-        """
-
-        :param mu: matrix of means (number_of_dimensions, number_of_latent_variables)
-        :param sigma: Gaussian noise parameter
-        :param pi: vector of priors (1, number_of_latent_variables)
-        """
-        self._mu = mu
+        self._mu = mu  # (number_of_dimensions, number_of_latent_variables)
         self._sigma = sigma
-        self._pi = pi
+        self._pi = pi  # (1, number_of_latent_variables)
 
     @property
     def mu(self):
@@ -85,8 +79,7 @@ def init_binary_latent_factor_model(
     binary_latent_factor_approximation: AbstractBinaryLatentFactorApproximation,
 ) -> BinaryLatentFactorModel:
     """
-    Initialise binary latent factor model by running a maximisation step with the parameters of the
-    binary latent factor approximation
+    Initialise by running a maximisation step with the parameters of the binary latent factor approximation
 
     :param x: data matrix (number_of_points, number_of_dimensions)
     :param binary_latent_factor_approximation: a binary_latent_factor_approximation

### 2023-01-04 src/models/binary_latent_factor_models/boltzmann_machine.py

@@ -16,11 +16,7 @@ class BoltzmannMachine(BinaryLatentFactorModel):
         pi: np.ndarray,
     ):
         """
-        Binary latent factor model as a Boltzmann Machine
-
-        :param mu: matrix of means (number_of_dimensions, number_of_latent_variables)
-        :param sigma: gaussian noise parameter
-        :param pi: vector of priors (1, number_of_latent_variables)
+        Binary latent factor model with Boltzmann Machine terms
         """
         super().__init__(mu, sigma, pi)
 
@@ -82,8 +78,7 @@ def init_boltzmann_machine(
     binary_latent_factor_approximation: AbstractBinaryLatentFactorApproximation,
 ) -> BinaryLatentFactorModel:
     """
-    Initialise a boltzmann machine by running a maximisation step with the parameters of the
-    binary latent factor approximation
+    Initialise by running a maximisation step with the parameters of the binary latent factor approximation
 
     :param x: data matrix (number_of_points, number_of_dimensions)
     :param binary_latent_factor_approximation: a binary_latent_factor_approximation
@@ -92,8 +87,4 @@ def init_boltzmann_machine(
     mu, sigma, pi = BinaryLatentFactorModel.calculate_maximisation_parameters(
         x, binary_latent_factor_approximation
     )
-    return BoltzmannMachine(
-        mu=mu,
-        sigma=sigma,
-        pi=pi,
-    )
+    return BoltzmannMachine(mu=mu, sigma=sigma, pi=pi)

### 2023-01-04 src/models/binary_latent_factor_models/variational_bayes.py

@@ -4,20 +4,20 @@ from src.models.binary_latent_factor_approximations.abstract_binary_latent_facto
     AbstractBinaryLatentFactorApproximation,
 )
 from src.models.binary_latent_factor_models.binary_latent_factor_model import (
-    AbstractBinaryLatentFactorModel,
     BinaryLatentFactorModel,
 )
+from src.models.binary_latent_factor_models.boltzmann_machine import BoltzmannMachine
 
 
 class GaussianPrior:
-    def __init__(self, a, b, d, k):
+    def __init__(self, a: float, b: float, d: int, k: int):
         """
         Gaussian prior on mu matrix
 
         :param a: alpha parameter of Gamma Prior
         :param b: beta parameter of Gamma Prior
         :param d: number of dimensions
-        :param k: number of latent factors
+        :param k: number of latent variables
         """
         self.a = a
         self.b = b
@@ -25,7 +25,7 @@ class GaussianPrior:
         self.alpha = np.ones((k,))
         self.w_covariance = np.zeros((k, k))
 
-    def mu_k(self, k):
+    def mu_k(self, k: int) -> np.ndarray:
         """
         Column vector of mu matrix, the latent feature vector
 
@@ -34,7 +34,7 @@ class GaussianPrior:
         """
         return self.mu[:, k : k + 1]
 
-    def w_d(self, d):
+    def w_d(self, d: int) -> np.ndarray:
         """
         Row vector of mu matrix, the weight vector for a particular dimension (pixel) of the data
 
@@ -52,15 +52,17 @@ class GaussianPrior:
         return np.diag(self.alpha)
 
 
-class VariationalBayesBinaryLatentFactorModel(AbstractBinaryLatentFactorModel):
+class VariationalBayes(BoltzmannMachine):
     def __init__(self, mu: GaussianPrior, variance: float, pi: np.ndarray):
         """
-        Variational Bayes implementation with prior on mu
+        Variational Bayes implementation with prior on mu.
+        Note that we are inheriting from BoltzmannMachine for Question 5d only.
 
         :param mu: Gaussian prior on latent features
         :param variance: Gaussian noise parameter
         :param pi: vector of priors (1, number_of_latent_variables)
         """
+        super().__init__(mu=mu.mu, sigma=np.sqrt(variance), pi=pi)
         self.gaussian_prior = mu
         self._variance = variance
         self._pi = pi
[…]

### 2023-01-04 src/solutions/q4.py

@@ -6,9 +6,13 @@ import numpy as np
 from matplotlib.offsetbox import AnnotationBbox, OffsetImage
 
 from src.automatic_relevance_determination import run_automatic_relevance_determination
+from src.models.binary_latent_factor_approximations.mean_field_approximation import (
+    init_mean_field_approximation,
+)
+from src.models.binary_latent_factor_models.variational_bayes import VariationalBayes
 
 
-def _offset_image(coord: int, path: str, ax: plt.axis):
+def offset_image(coord: int, path: str, ax: plt.axis):
     """
     Add image to matplotlib axis.
 
@@ -32,6 +36,60 @@ def _offset_image(coord: int, path: str, ax: plt.axis):
     ax.add_artist(ab)
 
 
+def plot_factors(
+    variational_bayes_binary_latent_factor_models: List[VariationalBayes],
+    ks: List[int],
+    max_k: int,
+    save_path: str,
+):
+    # store each feature as an image for later use
+    for i, k in enumerate(ks):
+        sort_indices = np.argsort(
+            variational_bayes_binary_latent_factor_models[i].gaussian_prior.alpha
+        )
+        for j, idx in enumerate(sort_indices):
+            fig = plt.figure(figsize=(0.3, 0.3))
+            ax = plt.Axes(fig, [0.0, 0.0, 1.0, 1.0])
+            ax.set_axis_off()
+            fig.add_axes(ax)
+            ax.imshow(
+                variational_bayes_binary_latent_factor_models[i]
+                .mu[:, idx]
+                .reshape(4, 4)
+            )
+            fig.savefig(save_path + f"-latent-factor-{i}-{j}", bbox_inches="tight")
+            plt.close()
+
+    # bar plot of alphas
+    fig, ax = plt.subplots(len(ks), 1, figsize=(12, 2 * len(ks)))
+    plt.subplots_adjust(hspace=1)
+    for i, k in enumerate(ks):
+        sort_indices = np.argsort(
+            variational_bayes_binary_latent_factor_models[i].gaussian_prior.alpha
+        )
+        y = list(
+            1
+            / variational_bayes_binary_latent_factor_models[i].gaussian_prior.alpha[
+                sort_indices
+            ]
+        ) + [0] * (max_k - k)
+        ax[i].set_title(f"{k=}")
+        ax[i].bar(range(max_k), y)
+        ax[i].set_xticks([])
+        ax[i].set_ylabel("Inverse Alpha")
+    # add feature image ticks
[…]

### 2023-01-04 src/solutions/q5.py

@@ -1,0 +1,45 @@
+from typing import List
+
+import numpy as np
+from tqdm import tqdm
+
+from src.automatic_relevance_determination import run_automatic_relevance_determination
+from src.models.binary_latent_factor_approximations.message_passing_approximation import (
+    init_message_passing,
+)
+from src.models.binary_latent_factor_models.variational_bayes import VariationalBayes
+from src.solutions.q4 import plot_factors
+
+
+def d(
+    x: np.ndarray,
+    a_parameter: int,
+    b_parameter: int,
+    ks: List[int],
+    max_k: int,
+    em_iterations: int,
+    save_path: str,
+) -> List[List[float]]:
+
+    variational_bayes_models: List[VariationalBayes] = []
+    free_energies = []
+    for i, k in tqdm(enumerate(ks)):
+        n = x.shape[0]
+        message_passing_approximation = init_message_passing(k, n)
+        (variational_bayes_model, free_energy) = run_automatic_relevance_determination(
+            x=x,
+            binary_latent_factor_approximation=message_passing_approximation,
+            a_parameter=a_parameter,
+            b_parameter=b_parameter,
+            k=k,
+            em_iterations=em_iterations,
+        )
+        variational_bayes_models.append(variational_bayes_model)
+        free_energies.append(free_energy)
+    plot_factors(
+        variational_bayes_models,
+        ks,
+        max_k,
+        save_path,
+    )
+    return free_energies

### 2023-01-04 submission.tex
add plots for loopy bp ard

@@
 
-where $\delta[\textbf{x}]$ are the concentration perturbations, $\epsilon \sim \mathcal{N} (0, \Psi)$ are the reaction specific Gaussian noise variables, and $z \sim \mathcal{N} (0, I)$ are the latent factor, the parent concentration perturbations.
+where $\delta[\textbf{x}]$ are the concentration perturbations, $\epsilon \sim \mathcal{N} (0, \Psi)$ are the reaction specific Gaussian noise variables, \Lambda is the factor loading matrix, and $z \sim \mathcal{N} (0, I)$ are the latent factors, the parent concentration perturbations.
 
@@
 In theory, if these concentration perturbations had unknown parent factors, we could use the results to recover the concentration perturbations of those parent features.
-However, in our DAG this is not hte case, so we would not be able to recover any other species in the cascade because  they are all downstream from the known concentration perturbations and those recovered from factor analysis.
+However, in our DAG this is not the case, so we would not be able to recover any other species in the cascade because  they are all downstream from the known concentration perturbations and those recovered from factor analysis.
 Similarly, the downstream weights in the DAG from the known measurements would also be unidentifiable.
@@
 
-collecting $\textbf{w}$ terms:
+collecting $\textbf{w}$:
 \[\log P(\textbf{w} | \mathcal{D}) \propto
@@
 
-Knowing that the posterior $P(\textbf{w} | \mathcal{D})$ will be Gaussian with mean $\bar{\mu}_w$ and covariance $\bar{\Sigma}_w$, we can see that expanding the exponent component of a Gaussian would have the form:
-
-\[\left( \textbf{w} - \bar{\mu}_w \right)^T \bar{\Sigma}_w^{-1} \left( \textbf{w} - \bar{\mu}_w \right) = \textbf{w}^T \bar{\Sigma}_w^{-1} \textbf{w} -2 \textbf{w}^T \bar{\Sigma}_w^{-1} \bar{\mu}_w + \bar{\mu}_w^T \bar{\Sigma}_w^{-1} \bar{\mu}_w\]
+Knowing that the posterior $P(\textbf{w} | \mathcal{D})$ will be Gaussian with mean $\bar{\mu}_{\textbf{w}}$ and covariance $\bar{\Sigma}_{\textbf{w}}$, we can see that expanding the exponent component of a Gaussian would have the form:
+
+\[\left( \textbf{w} - \bar{\mu}_{\textbf{w}} \right)^T \bar{\Sigma}_{\textbf{w}}^{-1} \left( \textbf{w} - \bar{\mu}_{\textbf{w}} \right) = \textbf{w}^T \bar{\Sigma}_{\textbf{w}}^{-1} \textbf{w} -2 \textbf{w}^T \bar{\Sigma}_{\textbf{w}}^{-1} \bar{\mu}_{\textbf{w}} + \bar{\mu}_{\textbf{w}}^T \bar{\Sigma}_{\textbf{w}}^{-1} \bar{\mu}_{\textbf{w}}\]
 
 Thus we can identify the posterior covariance:
-\[\bar{\Sigma}_w = \left(\frac{\textbf{X} \textbf{X}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \right)^{-1}\]
+\[\bar{\Sigma}_{\textbf{w}} = \left(\frac{\textbf{X} \textbf{X}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \right)^{-1}\]
 
 and the posterior mean:
-\[\bar{\mu}_w = \bar{\Sigma}_w \left(\frac{\textbf{X}\textbf{y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right) \]
+\[\bar{\mu}_{\textbf{w}} = \bar{\Sigma}_{\textbf{w}} \left(\frac{\textbf{X}\textbf{y}^T}{\sigma^2} + \Sigma_{\textbf{w}}^{-1} \mu_{\textbf{w}}\right) \]
 
@@
 Also, observing $g_{obs}(t)$ we see that there is periodic structure in the timeseries meaning that the data is not independently and identically distributed as $e(t) \sim \mathcal{N}(0, 1)$ would suggest.
-Otherwise, we would expect $g(t)$ to look like random noise.
+If this were the case, we would expect $g(t)$ to look like random noise.
 
@@
     The covariance matrix shows how increasing $\phi$ will start to increase the non-periodic component masking the periodic stripes.
-    This is expected because it adjusts the weight of the squared exponential portion of the kernel, thus the periodic component remains the same (i.e.same amplitude) but the large baseline shifts from increasing $\phi$ ends up dominating the function visually.
+    This is expected because it adjusts the weight of the squared exponential portion of the kernel.
+    The periodic component remains the same (i.e.same amplitude) but the large baseline shifts from increasing $\phi$ ends up dominating the function visually.
 \item [$\eta$:] As $\eta$ increases we see smoother sample functions.
-This is expected because the $\eta$ increases the lengthscale of the non-periodic component, allowing for smoother functions.
+This is expected because the $\eta$ increases the lengthscale of the squared exponential component, allowing for smoother functions.
 This causes the off-diagonals of the gram matrix to increase, however the periodic component is still maintained because $\eta$ doesn't affect the relative weighting between the two components.
@@
 We can see this reflected in the stripes in the gram matrix getting further apart and the period of the samples.
-This makes sense because we are adjusting the period parameter in the sinusoid function of the periodic term with $\tau$.
+This makes sense because we are adjusting the period parameter in the sinusoid function of the periodic term.
 \item [$\zeta$:] As $\zeta$ increases, the function becomes less smooth.
@@
 Thus we can choose $\tau=1$ to ensure functions with a period of one year to reflect this prior knowledge.
-It can be difficult to quantitatively choose values for the other parameters as they can relate to the uncertainty exhibited in the data (i.e.the smoothness of the function).
+It can be difficult to quantitatively choose values for the other parameters as they can relate to the uncertainty exhibited in the data (i.e. the smoothness of the function).
 One approach is to maximise:
@@
 
-We can analyse some of the changes in these parameters from optimisation, to gain some insights.
+We can analyse some of the changes in these parameters from optimisation to gain some insights.
 We can see that $\tau$ remains the same as we would expect, given the yearly seasonality we knew apriori.
-On the other hand, the value for $\zeta$ is significantly reduced signifying that $\delta_{s=t}$ is not a very good kernel for representing the data as datapoints at different timesteps do exhibit correlations.
+On the other hand, the value for $\zeta$ is significantly reduced signifying that $\delta_{s=t}$ is not a very good kernel for representing the data as datapoints at different timesteps do exhibit correlations and are not independently and identically distributed.
 
@@
 We can see that the conclusions can be quite sensitive to kernel hyperparameters when comparing the extrapolations from before and after training.
-Prior to training, the extrapolated prediction is not representative of the given data with very large uncertainty.
+Prior to training, the extrapolated prediction is not representative of the given data, with very large uncertainties.
 After training, we can see that the prediction is much more reasonable, and qualitatively the uncertainty bounds seem to exhibit the historical variability in the data.
@@
 
-%Thus,
-%\[\mathcal{F}(Q, \theta) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{i=1}^{K} q_{i} (s_i)} + H\left[ \prod_{i=1}^{K} q_{i} (s_i)\right] \]
-
 Knowing,
@@
 
-\[ P(\textbf{x} | \textbf{s}, \theta)  = 2\pi^{-\frac{d}{2}} |\sigma^2 \textbf{I}|^{-\frac{1}{2}} \exp \left( -\frac{1}{2} \left(\textbf{x} - \sum_{k=1}^{K} s_k \mu_k\right)^T \frac{1}{\sigma^2} \textbf{I}  \left(\textbf{x} - \sum_{k=1}^{K} s_k \mu_k\right) \right) \]
+\[ P(\textbf{x} | \textbf{s}, \theta)  = 2\pi^{-\frac{D}{2}} |\sigma^2 \textbf{I}|^{-\frac{1}{2}} \exp \left( -\frac{1}{2} \left(\textbf{x} - \sum_{k=1}^{K} s_k \mu_k\right)^T \frac{1}{\sigma^2} \textbf{I}  \left(\textbf{x} - \sum_{k=1}^{K} s_k \mu_k\right) \right) \]
 
@@
 
-\[ \log P(\textbf{x} | \textbf{s}, \theta)  = -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} s_k \mu_k   + \sum_{k=1}^{K} \sum_{k=1}^{K} s_k s_{k'} \mu_k^T\mu_{k'} \right) \]
+\[ \log P(\textbf{x} | \textbf{s}, \theta)  = -\frac{D}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} s_k \mu_k   + \sum_{k=1}^{K} \sum_{k=1}^{K} s_k s_{k'} \mu_k^T\mu_{k'} \right) \]
 
@@
 \left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} \\
-= -\frac{d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} \left\langle s_k \right\rangle_{q_{k} (s_k)} \mu_k   + \sum_{k=1}^{K} \sum_{k'=1}^{K} \left\langle s_k s_{k'} \right\rangle_{q_{k} (s_k) q_{k\} (s_{k'})} \mu_k^T\mu_{k'} \right)\]
+= -\frac{D}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} \left\langle s_k \right\rangle_{q_{k} (s_k)} \mu_k   + \sum_{k=1}^{K} \sum_{k'=1}^{K} \left\langle s_k s_{k'} \right\rangle_{q_{k} (s_k) q_{k'} (s_k')} \mu_k^T\mu_{k'} \right)\]
 
@@
 \left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} \\
-= -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K}  \lambda_k  \mu_k   + \sum_{k=1}^{K} \sum_{k'=1, k'\neq k}^{K}  \lambda_k \lambda_{k'} \mu_k^T\mu_{k'} + \sum_{k=1}^{K}  \lambda_k \mu_k^T\mu_k \right)\]
+= -\frac{D}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K}  \lambda_k  \mu_k   + \sum_{k=1}^{K} \sum_{k'=1, k'\neq k}^{K}  \lambda_k \lambda_{k'} \mu_k^T\mu_{k'} + \sum_{k=1}^{K}  \lambda_k \mu_k^T\mu_k \right)\]
 
@@
 \tab \mathcal{F}(q, \theta) = \\
-\tab \tab \frac{-d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K}  \lambda_k  \mu_k   + \sum_{k=1}^{K} \sum_{k'=1, k'\neq k}^{K}  \lambda_k \lambda_{k'} \mu_k^T\mu_{k'} + \sum_{k=1}^{K}  \lambda_k \mu_k^T\mu_k \right) \\
+\tab \tab \frac{-D}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K}  \lambda_k  \mu_k   + \sum_{k=1}^{K} \sum_{k'=1, k'\neq k}^{K}  \lambda_k \lambda_{k'} \mu_k^T\mu_{k'} + \sum_{k=1}^{K}  \lambda_k \mu_k^T\mu_k \right) \\
 \tab \tab + \sum_{k=1}^{K} \lambda_k \log \pi_k + (1-\lambda_k) \log (1-\pi_k)}\\
@@
 where $\epsilon \sim \mathcal{N}(0, \sigma^2\textbf{I})$ and $P(\textbf{s}) = \prod_{k=1}^K P(s_k)$.
-This is appropriate for our data because $P(s_k)$ is our Bernoulli distribution and we are linearly combining different features with $\textbf{W} \in \mathbb{R}^{D \times K}$and then adding noise $\epsilon$.
+This is appropriate for our data because $P(s_k)$ are our independent Bernoulli distributions and we are linearly combining different features with $\textbf{W} \in \mathbb{R}^{D \times K}$and then adding noise $\epsilon$.
 
@@
 
-We can see that it has learned some of the previously identified features, such as the vertical line in the second column, the two by two square in the bottom left corner, the border, the horizontal line in the first row, and the a two by two square in the middle.
+We can see that this approach has learned some of the previously identified features, such as the vertical line in the second column, the two by two square in the bottom left corner, the border, the horizontal line in the first row, and the a two by two square in the middle.
 The other features seem to be some linear combination of two or more features, such as $\mu_4$ which looks like a combination of the cross and two by two square in the middle.
@@
 A possible way to improve our algorithm is reinitialising our algorithm a few times to find better potential convergence results (i.e.
-choose model with highest free energy convergence).
+choose the model with the highest free energy convergence).
 We can also increase the number of data samples, with the hopes of learning better features.
@@
 
-When implementing the algorithm, the mean field parameters were initialised randomly, each independently from a uniform distribution on $\[ 0, 1\]$.
-$\pi$, $\sigma$, and $\mu$ were initialised by running a maximisation step using the randomly initialised mean field parameters.
+When implementing the algorithm, the mean field parameters were initialised randomly, each independently from a uniform distribution on $[ 0, 1]$.
+$\pi$, $\sigma$, and $\mu$ were initialised by running a maximisation step using these randomly initialised mean field parameters.
 $K$ was set to eight, after visually identifying eight features in part d.
@@
 
-\[P(x_d | \texbf{s}, \textbf{w}_d, \sigma^2) = \mathcal{N}\left(\textbf{s}^T\textbf{w}_d, \sigma^2\right)\]
+\[P(x_d | \textbf{s}, \textbf{w}_d, \sigma^2) = \mathcal{N}\left(\textbf{s}^T\textbf{w}_d, \sigma^2\right)\]
 
@@
 
-where we introduce priors on each $\textbf{w}_k$ with $\alpha \in \mathbb{R}^{K \times 1}$.
+where we introduce priors on each $\textbf{w}_d$ with $\alpha \in \mathbb{R}^{K \times 1}$.
 
@@
 
-%Simplifying:
-%
-%\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle -\frac{1}{2}\left(\textbf{s}^T\textbf{A}^T\Psi^{-1}\textbf{As} - 2\textbf{s}^T\left(\textbf{A}^T\Psi^{-1} \textbf{x}  + 2\log\frac{\pi}{1-\pi}\right)\right)\right\rangle_{q(\theta)}\]
-
-Given our factored approximation $q(\textbf{s}) = \prod_{k=1}^K q_k(s_k)$, we can see that we can derive a similar partial update for $q_k(q_k)$ as in Question 3, by taking the variation derivative of the Lagrangian to enforce the normalisation of $q_k$:
+Given our factored approximation $q(\textbf{s}) = \prod_{k=1}^K q_k(s_k)$, we can see that we can derive a similar partial update for $q_k(s_k)$ as in Question 3, by taking the variation derivative of the Lagrangian to enforce the normalisation of $q_k$:
 
@@
 For the maximisation step, we perform maximisation steps for the parameters $\sigma$ and $\pi$ in the same way as question 3.
-However, having defined a prior on $\mu$ (through $\textbf{w}$) so we will have to derive our expression for $\langle \mu_k\rangle_{q_{\mu_k}}$ the expectation of the posterior on $\mu_k$.
+However, having defined a prior on $\mu$ (through $\textbf{w}$) we will have to derive our expression for $\langle \mu_k\rangle_{q_{\mu_k}}$ the expectation of the posterior on $\mu_k$.
 This involves deriving the posterior distribution of $\textbf{w}_d$
@@
 We can see that initially for $K=4$ to $K=8$, increasing $K$ significantly increases the convergence value of the free energy.
-However, beyond $K=8$ there is a no clear trend of $K$ versus the free energy convergence value.
-We can see that this corresponds to the visualisation of $\alpha^{-1}$ where beyond $K=11$, the effective number of features remains more or less the same.
+However, beyond $K=8$, the trend of $K$ versus the free energy convergence value is not as clear.
+We can see that this corresponds to the visualisation of $\alpha^{-1}$ where beyond $K=11$, the number of relevant features remains more or less the same.
 We know that there are only eight latent features, thus models with $K>8$ should be learning duplicate or irrelevant features.
-As such, we wouldn't expect a model to be able to increase it's free energy significantly when provided with additional degrees of freedom by increasing the value of $k$ beyond eight.
+As such, we wouldn't expect a model to be able to increase it's free energy significantly when provided with additional degrees of freedom by increasing the value of $K$ beyond eight.
 We see that for models with $K>>8$, there are typically ten or eleven features that might be deemed relevant (depending on how you threshold) and this is likely from slight overfitting, noise in the data, or duplicate features.
@@
 
-The Boltzmann Machine can be defined:
+The Boltzmann Machine can be defined as:
 
@@
 
-For $g_{ij}(s_i, s_j)$, we will choose a product of Bernoulli's approximation:
+For $g_{ij}(s_i, s_j)$, we can choose a product of Bernoulli distributions for our approximation:
 
@@
 
-and,
+and
 
@@
 \[q_{\neg \tilde{g}_{ij}(s_i, s_j)}(s_i, s_j) = \left( \mathcal{M}_{i \rightarrow i}(s_i) \prod_{k\in ne(i), k\neq j}^{K} \mathcal{M}_{k \rightarrow i}(s_i)
-\right)\left( \mathcal{M}_{j}(s_j) \prod_{k\in ne(j), k\neq i}^{K} \mathcal{M}_{k \rightarrow j}(s_j)
+\right)\left( \mathcal{M}_{j \rightarrow j}(s_j) \prod_{k\in ne(j), k\neq i}^{K} \mathcal{M}_{k \rightarrow j}(s_j)
 \right)
@@
 
-\[ \mathcal{M}_{j}(s_j) \prod_{k\in ne(j), k\neq j}^{K} \mathcal{M}_{k \rightarrow j}(s_j) \propto \exp(\eta_{j, \neg s_i} s_j)\]
-
-Note that $\tilde{g}_{ij}(s_i, s_j)$ was chosen as the product of two Bernoulli distributions, updates to this site approximation involves updating the  parameters $\eta_{ij}$ and $\eta_{ji}$, for $s_i$ and $s_j$ respectively.
+\[ \mathcal{M}_{j \rightarrow j}(s_j) \prod_{k\in ne(j), k\neq j}^{K} \mathcal{M}_{k \rightarrow j}(s_j) \propto \exp(\eta_{j, \neg s_i} s_j)\]
+
+Note that $\tilde{g}_{ij}(s_i, s_j)$ was chosen as the product of two Bernoulli distributions, so updates to this site approximation involves updating the  parameters $\eta_{ji}$ and $\eta_{ij}$, for $s_i$ and $s_j$ respectively.
 
@@
 
-We can use automatic relevance determination (ARD) as a hyperparameter method to select relevant features by placing a prior on $\mu_i$ in the same way as Question 4.
+We can use automatic relevance determination (ARD) as a hyperparameter method to select relevant features by placing a prior on $\mu_k$ in the same way as Question 4.
 With a hyper-M step, certain features will have diverging precision, leaving us with the relevant features, from which we get our $K$.
-We can implement this by running the message passing algorithm (loopy BP) which was implemented for Question 6 for the expectation step and the VB algorithm for the maximisation step (along with its hyper-M Step):
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale=0.4]{outputs/q5/d-latent-factors-comparison}
+We can implement this by running the message passing algorithm (loopy BP) which was implemented for Question 6 for the expectation step and the VB algorithm for the maximisation step (along with its hyper-M Step) which was implemented for Question 4:
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.47]{outputs/q5/d-1-latent-factors-comparison}
 \caption{Learned Latent Factors vs Inverse Alpha}
@@
 
+\newpage
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.47]{outputs/q5/d-2-latent-factors-comparison}
+\caption{Learned Latent Factors vs Inverse Alpha}
+\label{fig:}
+\end{figure}
+
 We can see that ARD has more difficulty finding relevant features when using loop BP.
-Looking at the features themselves, we can see that they are not very clear compared to the mean field approximation or Variational Bayes in Question 4.
-This is probably why we do not consistently have the same number of $\alpha_k$ values that don't diverge as compared with Question 3.
-However, we can see for $K<8$ that all of the features are deemed relevant as we would expect.
+However, looking at the features themselves, we can see that the features are not very clear compared those from the mean field approximation or Variational Bayes in Question 4.
+This is probably why we do not consistently have the same number of $\alpha_k$ values that don't diverge as compared with Question 4.
+Moreover we usually have fewer than eight features that don't converge, likely because loopy BP seems to be unable to identify too many clear features that we would want to deem relevant in the first place.
+However, we can see for $K<8$ that more of the features are deemed relevant as we would expect.
 
@@
 This likely contributes to the difficulty for ARD to find relevant features when the message passing algorithm is unable to converge on generating meaningful features.
-To improve this, we could also put priors on the other parameters $\sigma$ and $\pi$ in the hopes that this will help stabilise the algorithm.
+To improve this, we could also put priors on the other parameters $\sigma^2$ and $\pi$ in the hopes that this will help stabilise the algorithm.
 Moreover, because loopy BP is unable to converge for this problem, it is quite computationally intensive as EM will always run the maximum number of iterations without improvement in the results.
 Knowing that it won't converge, employing early stopping or fewer iterations can also be helpful.
+Moreover, Boltzmann machines do not scale well with respect to the number of nodes, $K$, introducing computational difficulties.
+This was experienced when generating the above results, where for larger $K$ values, loopy BP E steps took much longer to perform.
+To tackle this, a restricted Boltzmann Machine could be implemented to reduce the number of connections in the graph, alleviating these computational difficulties.
 
@@
 We can see that the mean field algorithm seems to learn better latent features.
-In particular there are fewer duplicates, unlike loopy BP that has a few duplicates like $\mu_3$ and $\mu_4$.
-Moreover, the learned features have less noise.
+In particular there are fewer duplicates, unlike loopy BP that has a few duplicates.
+Moreover, the learned features of the mean field algorithm have less noise than those from loopy BP.
 For example $\mu_0$ for the mean field algorithm looks almost like a binary image.
@@
 \lstinputlisting[language=Python]{src/constants.py}
-\newpage
 \section*{Appendix 5: generate\_images.py}

### 2023-01-04 main.py

@@ -181,19 +181,28 @@ if __name__ == "__main__":
     Q5_OUTPUT_FOLDER = os.path.join(OUTPUTS_FOLDER, "q5")
     if not os.path.exists(Q5_OUTPUT_FOLDER):
         os.makedirs(Q5_OUTPUT_FOLDER)
-    max_k = 20
-    free_energies = q5.d(
+    max_k = 21
+    free_energies_1 = q5.d(
         x=x,
         a_parameter=1,
         b_parameter=0,
-        ks=np.arange(4, 22, 2),
+        ks=np.arange(4, 13),
         max_k=max_k,
-        em_iterations=100,
-        save_path=os.path.join(Q5_OUTPUT_FOLDER, "d"),
+        em_iterations=em_iterations,
+        save_path=os.path.join(Q5_OUTPUT_FOLDER, "d-1"),
+    )
+    free_energies_2 = q5.d(
+        x=x,
+        a_parameter=1,
+        b_parameter=0,
+        ks=np.arange(13, 22),
+        max_k=max_k,
+        em_iterations=em_iterations,
+        save_path=os.path.join(Q5_OUTPUT_FOLDER, "d-2"),
     )
     q4.free_energy_plot(
-        ks=np.arange(4, 22, 2),
-        free_energies=free_energies,
+        ks=np.arange(4, 22),
+        free_energies=free_energies_1 + free_energies_2,
         model_name="Loopy BP E Step and Variational Bayes M Step",
         save_path=os.path.join(Q5_OUTPUT_FOLDER, "d"),
     )

### 2023-01-04 submission.tex
update report

@@
 We can write:
-%\[\log \tilde{g}_{ij}q_{\neg \tilde{g}_{ij}} \propto \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j + \sum_{i=1}^{K}  \log \left(\frac{\theta_{ii}}{1-\theta_{ii}} \right) +  \sum_{i'=1}^{K}\sum_{\substack{j'=1 \\ i', j' \neq i, j}}^{i'-1}  \log \left(\frac{\theta_i}{1-\theta_i} \right) s_i + \log \left(\frac{\theta_j}{1-\theta_j} \right) s_j\]
 
@@
 However, looking at the features themselves, we can see that the features are not very clear compared those from the mean field approximation or Variational Bayes in Question 4.
-This is probably why we do not consistently have the same number of $\alpha_k$ values that don't diverge as compared with Question 4.
-Moreover we usually have fewer than eight features that don't converge, likely because loopy BP seems to be unable to identify too many clear features that we would want to deem relevant in the first place.
-However, we can see for $K<8$ that more of the features are deemed relevant as we would expect.
+This is probably why we do not consistently have the same number of $\alpha_k$ values that don't diverge as compared with Question 4, loopy BP seems to be unable to identify too many clear features that we would want to clearly deem relevant.
 
@@
 We can see that the mean field algorithm seems to learn better latent features.
-In particular there are fewer duplicates, unlike loopy BP that has a few duplicates.
-Moreover, the learned features of the mean field algorithm have less noise than those from loopy BP.
-For example $\mu_0$ for the mean field algorithm looks almost like a binary image.
+In particular loopy BP that has a few duplicates and some of the learned features are quite noisy.
+For example $\mu_0$ for the mean field algorithm looks almost like a binary image whereas $\mu_3$ and $\mu_5$ from loopy BP are almost completely noise.
 We can understand the reason for this by comparing the free energies of the two algorithms:

### 2023-01-06 submission.tex
fix typo

@@
 
-where $\delta[\textbf{x}]$ are the concentration perturbations, $\epsilon \sim \mathcal{N} (0, \Psi)$ are the reaction specific Gaussian noise variables, \Lambda is the factor loading matrix, and $z \sim \mathcal{N} (0, I)$ are the latent factors, the parent concentration perturbations.
+where $\delta[\textbf{x}]$ are the concentration perturbations, $\epsilon \sim \mathcal{N} (0, \Psi)$ are the reaction specific Gaussian noise variables, $\Lambda$ is the factor loading matrix, and $z \sim \mathcal{N} (0, I)$ are the latent factors, the parent concentration perturbations.
 
